### 2024-03-02 manuscript/notes.qmd
Initial commit

---
title: "Testing Simple Regression"
author: "FEMA Project testing"

# reader-mode: true
---

# Simple Regression {.unnumbered}

## Overview

<p>

This example assesses whether growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1) differs significantly based on handedness, using height ("anthroheightcalc") as a representative metric of growth. The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 height measurements for each participant; 2) a simple regression analysis is used to test whether sex (boy, girl) predicts the average difference value in participants height from baseline to the 1-Year follow-up. Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between difference scores and sex. The ensuing analysis and interpretations are detailed in the subsequent sections.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# rm(list = ls())

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page

```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_ranked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

  #str(df)

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

:::
:::

## Results

### 2024-03-02 examples.qmd
img uploads

---
#title: "Examples: A Brief Overview"
format:
  html:
    css: styles.css
    theme: cosmo
    #page-layout: full
    #grid:
      #sidebar-width: 250px
      #body-width: 2000px
      #margin-width: 0px
    #gutter-width: 1.5rem

---

## <i class="bi bi-book"></i> Examples

::: {.grid}
::: {.g-col-8}
::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.0; font-family: Georgia, serif;"}
[**Each Example**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} uses data from the ABCD Study Dataset and provides a practical solution to a specific longitudinal method or analytic problem.

:::
:::

::: {.g-col-4}
```{=html}
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css">

<div class="section_our_solution">
  <div class="row">
    <div class="col-lg-12 col-md-12 col-sm-12">
      <div class="our_solution_category">
        <div class="solution_cards_box">
          <div class="solution_card">
            <div class="hover_color_bubble"></div>
            <div class="so_top_icon">
              <svg id="Layer_1" enable-background="new 0 0 512 512" height="50" viewBox="0 0 512 512" width="40" xmlns="http://www.w3.org/2000/svg">
                <g>
                  <g>
                    <g>
                      <g>
                        <path d="m47.478 452.317 295.441 19.76c5.511.369 10.277-3.8 10.645-9.31l28.393-424.517c.369-5.511-3.8-10.276-9.31-10.645l-295.441-19.76c-5.511-.369-10.276 3.8-10.645 9.31l-28.394 424.517c-.368 5.511 3.8 10.277 9.311 10.645z" fill="#fae19e" />
                      </g>
                      <g>
                        <g>
                          <g>
                            <g>
                              <g>
                                <path d="m17.5 504.177h226.14l79.96-79.605v-355.86c0-5.523-4.477-10-10-10h-296.1c-5.523 0-10 4.477-10 10v425.466c0 5.522 4.477 9.999 10 9.999z" fill="#fff9e9" />
                              </g>
                              <path d="m313.601 58.712h-40c5.523 0 10 4.477 10 10v355.861l-.258 40.078 40.258-40.078v-355.861c0-5.523-4.477-10-10-10z" fill="#fff4d6" />
                            </g>
                          </g>
                        </g>
                        <path d="m243.64 504.177v-70.253c0-5.523 4.477-10 10-10h69.96z" fill="#ffeec2" />
                      </g>
                    </g>
                    <g>
                      <path d="m468.636 248.58-33.372.165v-50.826c0-9.183 7.463-16.662 16.673-16.708h.007c9.217-.046 16.693 7.371 16.693 16.562v50.807z" fill="#fed23a" />
                      <path d="m451.96 504.177c-10.362-10.277-16.196-24.263-16.208-38.857l-.062-73.973c0-.644.524-1.169 1.171-1.173l30.038-.149c.647-.003 1.171.517 1.171 1.161l.062 74.079c.012 14.531-5.749 28.472-16.015 38.756z" fill="#54b1ff" />
                      <path d="m451.959 469.333h-.01c-14.434.072-26.14-11.542-26.14-25.935v-213.527c0-6.778 5.477-12.283 12.255-12.316l27.626-.137c6.826-.034 12.378 5.49 12.378 12.316v213.436c0 14.38-11.687 26.091-26.109 26.163z" fill="#fdf385" />
                      <path d="m465.69 217.417-23.769.118c6.037.79 10.708 5.94 10.708 12.198v213.437c0 9.823-5.455 18.397-13.507 22.87 3.79 2.115 8.164 3.317 12.826 3.293h.01c14.422-.072 26.109-11.783 26.109-26.163v-213.436c.001-6.826-5.551-12.351-12.377-12.317z" fill="#faee6e" />
                      <path d="m491.274 247.925-71.615.355c-7.305.036-13.226 5.968-13.226 13.248 0 7.281 5.921 13.153 13.226 13.117l58.389-.29v77.489c0 7.281 5.921 13.153 13.226 13.117 7.305-.036 13.226-5.968 13.226-13.248v-90.672c0-7.28-5.922-13.152-13.226-13.116z" fill="#54b1ff" />
                      <g>
                        <path d="m491.274 247.925-38.441.188-.167 26.311 25.381-.067v77.489c0 7.281 5.921 13.153 13.226 13.117 7.305-.036 13.226-5.968 13.226-13.248v-90.672c.001-7.282-5.921-13.154-13.225-13.118z" fill="#3da7ff" />
                      </g>
                    </g>
                  </g>
                  <g fill="#060606">
                    <path d="m373.147 20.122-295.44-19.761c-9.631-.638-17.984 6.665-18.629 16.293l-2.311 34.557h-39.267c-9.649 0-17.5 7.851-17.5 17.5v425.466c0 9.649 7.851 17.5 17.5 17.5h226.141c1.96 0 3.902-.801 5.292-2.185l34.138-33.987c.347.074.701.133 1.065.157l58.282 3.898c9.302.614 18.005-6.952 18.629-16.293l28.393-424.515c.639-9.528-6.766-17.993-16.293-18.63zm-122.006 465.902v-52.1c0-1.378 1.122-2.5 2.5-2.5h51.9zm94.939-23.757c-.244 1.51-1.131 2.286-2.66 2.327l-46.28-3.096 31.752-31.611c1.414-1.407 2.209-3.32 2.209-5.315v-355.86c0-9.649-7.851-17.5-17.5-17.5h-77.993c-9.697 0-9.697 15 0 15h77.993c1.379 0 2.5 1.122 2.5 2.5v347.712h-62.46c-9.649 0-17.5 7.851-17.5 17.5v62.753h-218.641c-1.378 0-2.5-1.122-2.5-2.5v-425.465c0-1.378 1.122-2.5 2.5-2.5h178.168c9.697 0 9.697-15 0-15h-123.868l2.244-33.556c.244-1.511 1.131-2.286 2.661-2.327l295.44 19.76c1.511.244 2.287 1.131 2.328 2.661z" />
                    <path d="m267.827 237.047h-204.553c-4.142 0-7.5 3.358-7.5 7.5s3.358 7.5 7.5 7.5h204.553c4.143 0 7.5-3.358 7.5-7.5s-3.357-7.5-7.5-7.5z" />
                    <path d="m267.827 289.332h-204.553c-4.142 0-7.5 3.358-7.5 7.5s3.358 7.5 7.5 7.5h204.553c4.143 0 7.5-3.358 7.5-7.5s-3.357-7.5-7.5-7.5z" />
                    <path d="m55.774 192.262c0 4.142 3.358 7.5 7.5 7.5h204.553c4.143 0 7.5-3.358 7.5-7.5s-3.357-7.5-7.5-7.5h-204.553c-4.142 0-7.5 3.358-7.5 7.5z" />
                    <path d="m91.807 139.977c0 4.142 3.358 7.5 7.5 7.5h132.487c4.142 0 7.5-3.358 7.5-7.5s-3.358-7.5-7.5-7.5h-132.487c-4.142 0-7.5 3.358-7.5 7.5z" />
                    <path d="m194.755 438.787c-13.489.036-26.978.065-40.467.086-4.534.007-9.067.013-13.6.016-8.215.006-13.75-1.643-15.59-10.679-1.556-7.64-12.364-6.613-14.464 0-5.19 16.337-13.774 9.936-18.582-1.053-4.797-10.963-6.027-23.233-8.122-34.9-1.54-8.573-14.506-6.17-14.732 1.994-.298 10.751-1.302 21.331-4.031 31.758-2.815 10.758-7.034 21.097-11.222 31.376-3.651 8.961 10.867 12.816 14.464 3.988 3.711-9.108 7.427-18.266 10.193-27.714 5.14 12.36 15.774 26.34 30.927 18.101 2.819-1.533 5.452-3.712 7.763-6.253 7.88 9.106 19.609 8.388 30.584 8.375 15.627-.02 31.254-.054 46.881-.095 9.649-.025 9.667-15.025-.002-15z" />
                    <path d="m505.932 246.439c-3.897-3.878-9.255-5.867-14.695-6.014l-5.668.028v-10.719c0-6.529-3.878-13.427-9.433-16.862v-15.098c0-31.069-48.372-30.934-48.372.146v15.1c-5.659 3.498-9.455 9.741-9.455 16.852v10.982c-24.966 1.7-25.037 39.745.028 41.232.16 33.575.152 66.6-.028 100.737-.049 9.414 14.949 9.966 15 .079.18-34.166.188-67.22.029-100.823l37.211-.185s-.048 110.848-.048 160.784c0 24.338-37.219 24.5-37.219-.253l.013-13.677c.585-9.68-14.387-10.583-14.973-.904v12.834c0 11 3.402 20.316 9.988 26.869.586 15.693 7.198 30.878 18.369 41.956 3.205 3.18 7.642 2.208 10.744-.182 11.365-11.385 17.769-26.394 18.169-42.414 4.951-4.931 9.908-9.896 9.908-26.896l.006-68.351c12.97 3.689 26.494-6.348 26.494-19.946v-90.672c0-5.523-2.155-10.709-6.068-14.603zm-72.623-5.727v-10.841c0-2.219 1.523-4.08 3.573-4.633l30.025-.149c.84.208 1.615.605 2.243 1.231.915.911 1.419 2.123 1.419 3.414v10.794zm18.671-52c4.604 0 9.155 4.514 9.155 9.062v12.166l-18.372.091v-12.111c.001-5.053 4.133-9.183 9.217-9.208zm-.011 303.901c-3.487-4.942-6.009-10.531-7.417-16.406 2.322.503 4.674.765 7.027.765 2.627 0 5.253-.326 7.839-.957-1.374 5.964-3.892 11.587-7.449 16.598zm45.031-140.899c0 7.101-11.452 7.66-11.452.131 0 0 .013-70.974.021-77.48.005-4.196-3.483-7.509-7.558-7.509l-58.389.29c-7.242 0-7.073-11.331.074-11.366l71.615-.355c3.463.295 5.359 2.168 5.688 5.617v90.672z" />
                  </g>
                </g>
              </svg>
            </div>
            <div class="solu_description">
              <p>
                For an entry-level example of each longitudinal model
              </p>
              <button type="button" class="read_more_btn">Go Here</button>
            </div>
          </div>
          </div>
          </div>
        </div>
      </div>
    </div>
```
:::
:::

```{=html}
<section class="light">
	<div class="container py-2">
		<article class="postcard light blue">
			<a class="postcard__img_link" href="#">
				<img class="postcard__img" src="https://picsum.photos/1000/1000" alt="Image Title" />
			</a>
			<div class="postcard__text t-dark">
				<h1 class="postcard__title blue"><a href="#">Traditional Linear Models</a></h1>
				<div class="postcard__bar"></div>
				<ul class="postcard__tagbox">
					<li class="tag__item play blue">
						<a href="./1a_Examples_DifferenceScores_PairedTtests.qmd"><i class="fas fa-play
						mr-2"></i>Difference Scores (Paired T-test)</a>
						</li>
					<li class="tag__item play blue">
						<a href="./1b_Examples_DifferenceScores_SimpleRegression.qmd"><i class="fas fa-play
						mr-2"></i>Difference Scores (Simple Regression)</a>
						</li>
					<li class="tag__item play blue">
						<a href="./2_Examples_ResidualizedChangeScores.qmd"><i class="fas fa-play
						mr-2"></i>Residualized Change Scores</a>
						</li>
					<li class="tag__item play blue">
						<a href="./3a_Examples_LinearMixedModels.qmd"><i class="fas fa-play mr-2"></i>Linear Mixed
						Models (Random Intercept)</a>
					</li>
					<li class="tag__item play blue">
						<a href="./3b_Examples_LinearMixedModels.qmd"><i class="fas fa-play mr-2"></i>Linear Mixed
						Models (Random Slope)</a>
					</li>
				</ul>
			</div>
		</article>
		<article class="postcard light red">
			<a class="postcard__img_link" href="#">
				<img class="postcard__img" src="https://picsum.photos/501/500" alt="Image Title" />	
			</a>
			<div class="postcard__text t-dark">
				<h1 class="postcard__title red"><a href="#">Traditional Nonlinear Models</a></h1>
				<div class="postcard__bar"></div>
				<div class="postcard__preview-txt">
				<ul class="postcard__tagbox">
				  <li class="tag__item play blue">
						<a href="./_4_Examples_SignedRankTest.qmd"><i class="fas fa-play mr-2"></i>Signed-Rank
						Test</a>
						</li>
					<li class="tag__item play blue">
						<a href="./_5_Examples_MarginalModels.qmd"><i class="fas fa-play mr-2"></i>Marginal Models</a>
						</li>
						<li class="tag__item play blue">
						<a href="./_6_Examples_GeneralizedEstimatingEquations.qmd"><i class="fas fa-play
						mr-2"></i>Generalized
						Estimating Equations</a>
						</li>
						<li class="tag__item play blue">
						<a href="./_7_Examples_GeneralizedLinearMixedEffectsModels.qmd"><i class="fas fa-play
						mr-2"></i>Generalized Linear
						Mixed Effects Models</a>
						</li>
				</ul>
			</div>
		</article>
		<article class="postcard light red">
			<a class="postcard__img_link" href="#">
				<img class="postcard__img" src="https://picsum.photos/601/600" alt="Image Title" />	
			</a>
			<div class="postcard__text t-dark">
				<h1 class="postcard__title red"><a href="#">Structural Equation Models</a></h1>
				<div class="postcard__bar"></div>
				<div class="postcard__preview-txt">
				<ul class="postcard__tagbox">
				  <li class="tag__item play blue">
						<a href="./_9_Examples_LatentChangeScoresModels.qmd"><i class="fas fa-play
						mr-2"></i>Latent Change Score Models</a>
						</li>
					<li class="tag__item play blue">
						<a href="./_10_Examples_LatentGrowthCurveModels.qmd"><i class="fas fa-play
						mr-2"></i>Latent Growth Curve Models</a>
						</li>
					<li class="tag__item play blue">
						<a href="./_12_Examples_LatentTransitionAnalysis.qmd"><i class="fas fa-play
						mr-2"></i>Latent Transition Analysis</a>
						</li>
				</ul>
			</div>
		</article>
		<article class="postcard light red">
			<a class="postcard__img_link" href="#">
				<img class="postcard__img" src="https://picsum.photos/500/653" alt="Image Title" />	
			</a>
			<div class="postcard__text t-dark">
				<h1 class="postcard__title red"><a href="#">Advanced SEM</a></h1>
				<div class="postcard__bar"></div>
				<div class="postcard__preview-txt">
				<ul class="postcard__tagbox">
				  <li class="tag__item play blue">
						<a href="./_15_Examples_RandomInterceptCrosslaggedPanelModels.qmd"><i class="fas fa-play
						mr-2"></i>Random-Intercept Crosslagged Panel Model</a>
						</li>
						<li class="tag__item play blue">
						<a href="./_16_Examples_LatentCurveModelsStructuredResiduals.qmd"><i class="fas fa-play
						mr-2"></i>Latent Curve Models with Structured Residuals</a>
						</li>
				</ul>
			</div>
		</article>
		<article class="postcard light red">
			<a class="postcard__img_link" href="#">
				<img class="postcard__img" src="https://picsum.photos/801/800" alt="Image Title" />	
			</a>
			<div class="postcard__text t-dark">
				<h1 class="postcard__title red"><a href="#">Additional Models</a></h1>
				<div class="postcard__bar"></div>
				<div class="postcard__preview-txt">
				<ul class="postcard__tagbox">
				  <li class="tag__item"><i class="fas fa-tag mr-2"></i></li>
					<li class="tag__item"><i class="fas fa-clock mr-2"></i></li>
				</ul>
			</div>
		</article>

	</div>
</section>
```

```{=html}
<div class="d-flex flex-column justify-content-center w-100 h-100">

	<div class="d-flex flex-column justify-content-center align-items-center">
		<h1 class="fw-light text-white m-0">Pure CSS Gradient Background Animation</h1>
		<div class="btn-group my-5">
			<a href="https://codepen-api-export-production.s3.us-west-2.amazonaws.com/zip/PEN/pyBNzX/1578778289271/pure-css-gradient-background-animation.zip" class="btn btn-outline-light" aria-current="page"><i class="fas fa-file-download me-2"></i> SOURCE CODE</a>
			<a href="https://codepen.io/P1N2O/full/pyBNzX" class="btn btn-outline-light">FULL SCREEN <i class="fas fa-expand ms-2"></i></a>
		</div>
		<a href="https://manuel.pinto.dev" class="text-decoration-none">
			<h5 class="fw-light text-white m-0">— Pen by Manuel Pinto —</h5>
		</a>
	</div>
</div>
</div>
```

### 2024-03-02 index.qmd
img uploads

---
#title: "Longitudinal Analysis"

format:
  html:
    css: styles.css
    #theme: cosmo
    page-layout: custom
    grid:
      sidebar-width: 200px
      body-width: 2000px
      margin-width: 0px
      gutter-width: 0rem
---

<link href="output.css" rel="stylesheet">

```{=html}
<div class="bg-gray-900">
  <main>
    <!-- Hero section -->
    <div class="relative isolate overflow-hidden">
      <svg class="absolute inset-0 -z-10 h-full w-full stroke-white/10 [mask-image:radial-gradient(100%_100%_at_top_right,white,transparent)]" aria-hidden="true">
        <defs>
          <pattern id="983e3e4c-de6d-4c3f-8d64-b9761d1534cc" width="200" height="200" x="50%" y="-1" patternUnits="userSpaceOnUse">
            <path d="M.5 200V.5H200" fill="none" />
          </pattern>
        </defs>
        <svg x="50%" y="-1" class="overflow-visible fill-gray-800/20">
          <path d="M-200 0h201v201h-201Z M600 0h201v201h-201Z M-400 600h201v201h-201Z M200 800h201v201h-201Z" stroke-width="0" />
        </svg>
        <rect width="100%" height="100%" stroke-width="0" fill="url(#983e3e4c-de6d-4c3f-8d64-b9761d1534cc)" />
      </svg>
      <div class="absolute left-[calc(50%-4rem)] top-10 -z-10 transform-gpu blur-3xl sm:left-[calc(50%-18rem)] lg:left-48 lg:top-[calc(50%-30rem)] xl:left-[calc(50%-24rem)]" aria-hidden="true">
        <div class="aspect-[1108/632] w-[69.25rem] bg-gradient-to-r from-[#80caff] to-[#4f46e5] opacity-20" style="clip-path: polygon(73.6% 51.7%, 91.7% 11.8%, 100% 46.4%, 97.4% 82.2%, 92.5% 84.9%, 75.7% 64%, 55.3% 47.5%, 46.5% 49.4%, 45% 62.9%, 50.3% 87.2%, 21.3% 64.1%, 0.1% 100%, 5.4% 51.1%, 21.4% 63.9%, 58.9% 0.2%, 73.6% 51.7%)"></div>
      </div>
      <div class="mx-auto max-w-7xl px-6 pb-24 pt-10 sm:pb-40 lg:flex lg:px-8 lg:pt-40">
        <div class="mx-auto max-w-2xl flex-shrink-0 lg:mx-0 lg:max-w-xl lg:pt-8">
          <img class="h-11" src="https://tailwindui.com/img/logos/mark.svg?color=indigo&shade=500" alt="Your Company">
          <div class="mt-24 sm:mt-32 lg:mt-16">
            <a href="#" class="inline-flex space-x-6">
              <span class="rounded-full bg-indigo-500/10 px-3 py-1 text-sm font-semibold leading-6 text-indigo-400 ring-1 ring-inset ring-indigo-500/20">Latest updates</span>
              <span class="inline-flex items-center space-x-2 text-sm font-medium leading-6 text-gray-300">
                <span>Just shipped v1.0</span>
                <svg class="h-5 w-5 text-gray-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
                  <path fill-rule="evenodd" d="M7.21 14.77a.75.75 0 01.02-1.06L11.168 10 7.23 6.29a.75.75 0 111.04-1.08l4.5 4.25a.75.75 0 010 1.08l-4.5 4.25a.75.75 0 01-1.06-.02z" clip-rule="evenodd" />
                </svg>
              </span>
            </a>
          </div>
          <h1 class="mt-10 text-4xl font-bold tracking-tight text-center  text-white sm:text-6xl">Longitudinal Analysis Project</h1>
          <p class="mt-6 text-lg leading-8 text-gray-300">An open-source and collaborative effort guided by the Biostatistics Working Group of the <b> <a href="https://abcdstudy.org/">Adolescent Brain Cognitive Development (ABCD) Study&reg</a></b></p>
          <div class="mt-10 flex items-center gap-x-6">
            <a href="#" class="rounded-md bg-indigo-500 px-3.5 py-2.5 text-sm font-semibold text-white shadow-sm hover:bg-indigo-400 focus-visible:outline focus-visible:outline-2 focus-visible:outline-offset-2 focus-visible:outline-indigo-400">Get started</a>
            <a href="#" class="text-sm font-semibold leading-6 text-white">Live demo <span aria-hidden="true">→</span></a>
          </div>
        </div>
        <div class="mx-auto mt-16 flex max-w-2xl sm:mt-24 lg:ml-10 lg:mr-0 lg:mt-0 lg:max-w-none lg:flex-none xl:ml-32">
          <div class="max-w-3xl flex-none sm:max-w-5xl lg:max-w-none">
            <img src="https://tailwindui.com/img/component-images/dark-project-app-screenshot.png" alt="App screenshot" width="2432" height="1442" class="w-[76rem] rounded-md bg-white/5 shadow-2xl ring-1 ring-white/10">
          </div>
        </div>
      </div>
    </div>

    <!-- Logo cloud -->
    <div class="mx-auto mt-8 max-w-7xl px-6 sm:mt-16 lg:px-8">
      <h2 class="text-center text-lg font-semibold leading-8 text-white">The world’s most innovative companies use our app</h2>
      <div class="mx-auto mt-10 grid max-w-lg grid-cols-4 items-center gap-x-8 gap-y-10 sm:max-w-xl sm:grid-cols-6 sm:gap-x-10 lg:mx-0 lg:max-w-none lg:grid-cols-5">
        <img class="col-span-2 max-h-12 w-full object-contain lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/transistor-logo-white.svg" alt="Transistor" width="158" height="48">
        <img class="col-span-2 max-h-12 w-full object-contain lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/reform-logo-white.svg" alt="Reform" width="158" height="48">
        <img class="col-span-2 max-h-12 w-full object-contain lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/tuple-logo-white.svg" alt="Tuple" width="158" height="48">
        <img class="col-span-2 max-h-12 w-full object-contain sm:col-start-2 lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/savvycal-logo-white.svg" alt="SavvyCal" width="158" height="48">
        <img class="col-span-2 col-start-2 max-h-12 w-full object-contain sm:col-start-auto lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/statamic-logo-white.svg" alt="Statamic" width="158" height="48">
      </div>
    </div>

    <!-- Feature section -->
    <div class="mx-auto mt-32 max-w-7xl px-6 sm:mt-56 lg:px-8">
      <div class="mx-auto max-w-2xl text-center">
        <h2 class="text-base font-semibold leading-7 text-indigo-400">Deploy faster</h2>
        <p class="mt-2 text-3xl font-bold tracking-tight text-white sm:text-4xl">Everything you need to deploy your app</p>
        <p class="mt-6 text-lg leading-8 text-gray-300">Lorem ipsum dolor sit amet consect adipisicing elit. Possimus magnam voluptatum cupiditate veritatis in accusamus quisquam.</p>
      </div>
      <div class="mx-auto mt-16 max-w-2xl sm:mt-20 lg:mt-24 lg:max-w-none">
        <dl class="grid max-w-xl grid-cols-1 gap-x-8 gap-y-16 lg:max-w-none lg:grid-cols-3">
          <div class="flex flex-col">
            <dt class="text-base font-semibold leading-7 text-white">
              <div class="mb-6 flex h-10 w-10 items-center justify-center rounded-lg bg-indigo-500">
                <svg class="h-6 w-6 text-white" fill="none" viewBox="0 0 24 24" stroke-width="1.5" stroke="currentColor" aria-hidden="true">
                  <path stroke-linecap="round" stroke-linejoin="round" d="M3.75 13.5l10.5-11.25L12 10.5h8.25L9.75 21.75 12 13.5H3.75z" />
                </svg>
              </div>
              Server monitoring
            </dt>
            <dd class="mt-1 flex flex-auto flex-col text-base leading-7 text-gray-300">
              <p class="flex-auto">Non quo aperiam repellendus quas est est. Eos aut dolore aut ut sit nesciunt. Ex tempora quia. Sit nobis consequatur dolores incidunt.</p>
              <p class="mt-6">
                <a href="#" class="text-sm font-semibold leading-6 text-indigo-400">Learn more <span aria-hidden="true">→</span></a>
              </p>
            </dd>
          </div>
          <div class="flex flex-col">
            <dt class="text-base font-semibold leading-7 text-white">
              <div class="mb-6 flex h-10 w-10 items-center justify-center rounded-lg bg-indigo-500">
                <svg class="h-6 w-6 text-white" fill="none" viewBox="0 0 24 24" stroke-width="1.5" stroke="currentColor" aria-hidden="true">
                  <path stroke-linecap="round" stroke-linejoin="round" d="M15 19.128a9.38 9.38 0 002.625.372 9.337 9.337 0 004.121-.952 4.125 4.125 0 00-7.533-2.493M15 19.128v-.003c0-1.113-.285-2.16-.786-3.07M15 19.128v.106A12.318 12.318 0 018.624 21c-2.331 0-4.512-.645-6.374-1.766l-.001-.109a6.375 6.375 0 0111.964-3.07M12 6.375a3.375 3.375 0 11-6.75 0 3.375 3.375 0 016.75 0zm8.25 2.25a2.625 2.625 0 11-5.25 0 2.625 2.625 0 015.25 0z" />
                </svg>
              </div>
              Collaborate
            </dt>
            <dd class="mt-1 flex flex-auto flex-col text-base leading-7 text-gray-300">
              <p class="flex-auto">Vero eum voluptatem aliquid nostrum voluptatem. Vitae esse natus. Earum nihil deserunt eos quasi cupiditate. A inventore et molestiae natus.</p>
              <p class="mt-6">
                <a href="#" class="text-sm font-semibold leading-6 text-indigo-400">Learn more <span aria-hidden="true">→</span></a>
              </p>
            </dd>
          </div>
          <div class="flex flex-col">
            <dt class="text-base font-semibold leading-7 text-white">
              <div class="mb-6 flex h-10 w-10 items-center justify-center rounded-lg bg-indigo-500">
                <svg class="h-6 w-6 text-white" fill="none" viewBox="0 0 24 24" stroke-width="1.5" stroke="currentColor" aria-hidden="true">
                  <path stroke-linecap="round" stroke-linejoin="round" d="M6.75 3v2.25M17.25 3v2.25M3 18.75V7.5a2.25 2.25 0 012.25-2.25h13.5A2.25 2.25 0 0121 7.5v11.25m-18 0A2.25 2.25 0 005.25 21h13.5A2.25 2.25 0 0021 18.75m-18 0v-7.5A2.25 2.25 0 015.25 9h13.5A2.25 2.25 0 0121 11.25v7.5m-9-6h.008v.008H12v-.008zM12 15h.008v.008H12V15zm0 2.25h.008v.008H12v-.008zM9.75 15h.008v.008H9.75V15zm0 2.25h.008v.008H9.75v-.008zM7.5 15h.008v.008H7.5V15zm0 2.25h.008v.008H7.5v-.008zm6.75-4.5h.008v.008h-.008v-.008zm0 2.25h.008v.008h-.008V15zm0 2.25h.008v.008h-.008v-.008zm2.25-4.5h.008v.008H16.5v-.008zm0 2.25h.008v.008H16.5V15z" />
                </svg>
              </div>
              Task scheduling
            </dt>
            <dd class="mt-1 flex flex-auto flex-col text-base leading-7 text-gray-300">
              <p class="flex-auto">Et quod quaerat dolorem quaerat architecto aliquam accusantium. Ex adipisci et doloremque autem quia quam. Quis eos molestiae at iure impedit.</p>
              <p class="mt-6">
                <a href="#" class="text-sm font-semibold leading-6 text-indigo-400">Learn more <span aria-hidden="true">→</span></a>
              </p>
            </dd>
          </div>
        </dl>
      </div>
    </div>

    <!-- Feature section -->
    <div class="mt-32 sm:mt-56">
      <div class="mx-auto max-w-7xl px-6 lg:px-8">
        <div class="mx-auto max-w-2xl sm:text-center">
          <h2 class="text-base font-semibold leading-7 text-indigo-400">Everything you need</h2>
          <p class="mt-2 text-3xl font-bold tracking-tight text-white sm:text-4xl">No server? No problem.</p>
          <p class="mt-6 text-lg leading-8 text-gray-300">Lorem ipsum, dolor sit amet consectetur adipisicing elit. Maiores impedit perferendis suscipit eaque, iste dolor cupiditate blanditiis.</p>
        </div>
      </div>
      <div class="relative overflow-hidden pt-16">
        <div class="mx-auto max-w-7xl px-6 lg:px-8">
          <img src="https://tailwindui.com/img/component-images/dark-project-app-screenshot.png" alt="App screenshot" class="mb-[-12%] rounded-xl shadow-2xl ring-1 ring-white/10" width="2432" height="1442">
          <div class="relative" aria-hidden="true">
            <div class="absolute -inset-x-20 bottom-0 bg-gradient-to-t from-gray-900 pt-[7%]"></div>
          </div>
        </div>
      </div>
      <div class="mx-auto mt-16 max-w-7xl px-6 sm:mt-20 md:mt-24 lg:px-8">
        <dl class="mx-auto grid max-w-2xl grid-cols-1 gap-x-6 gap-y-10 text-base leading-7 text-gray-300 sm:grid-cols-2 lg:mx-0 lg:max-w-none lg:grid-cols-3 lg:gap-x-8 lg:gap-y-16">
          <div class="relative pl-9">
            <dt class="inline font-semibold text-white">
              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
                <path fill-rule="evenodd" d="M5.5 17a4.5 4.5 0 01-1.44-8.765 4.5 4.5 0 018.302-3.046 3.5 3.5 0 014.504 4.272A4 4 0 0115 17H5.5zm3.75-2.75a.75.75 0 001.5 0V9.66l1.95 2.1a.75.75 0 101.1-1.02l-3.25-3.5a.75.75 0 00-1.1 0l-3.25 3.5a.75.75 0 101.1 1.02l1.95-2.1v4.59z" clip-rule="evenodd" />
              </svg>
              Push to deploy.
            </dt>
            <dd class="inline">Lorem ipsum, dolor sit amet consectetur adipisicing elit aute id magna.</dd>
          </div>
          <div class="relative pl-9">
            <dt class="inline font-semibold text-white">
              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
                <path fill-rule="evenodd" d="M10 1a4.5 4.5 0 00-4.5 4.5V9H5a2 2 0 00-2 2v6a2 2 0 002 2h10a2 2 0 002-2v-6a2 2 0 00-2-2h-.5V5.5A4.5 4.5 0 0010 1zm3 8V5.5a3 3 0 10-6 0V9h6z" clip-rule="evenodd" />
              </svg>
              SSL certificates.
            </dt>
            <dd class="inline">Anim aute id magna aliqua ad ad non deserunt sunt. Qui irure qui lorem cupidatat commodo.</dd>
          </div>
          <div class="relative pl-9">
            <dt class="inline font-semibold text-white">
              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
                <path fill-rule="evenodd" d="M15.312 11.424a5.5 5.5 0 01-9.201 2.466l-.312-.311h2.433a.75.75 0 000-1.5H3.989a.75.75 0 00-.75.75v4.242a.75.75 0 001.5 0v-2.43l.31.31a7 7 0 0011.712-3.138.75.75 0 00-1.449-.39zm1.23-3.723a.75.75 0 00.219-.53V2.929a.75.75 0 00-1.5 0V5.36l-.31-.31A7 7 0 003.239 8.188a.75.75 0 101.448.389A5.5 5.5 0 0113.89 6.11l.311.31h-2.432a.75.75 0 000 1.5h4.243a.75.75 0 00.53-.219z" clip-rule="evenodd" />
              </svg>
              Simple queues.
            </dt>
            <dd class="inline">Ac tincidunt sapien vehicula erat auctor pellentesque rhoncus.</dd>
          </div>
          <div class="relative pl-9">
            <dt class="inline font-semibold text-white">
              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
                <path fill-rule="evenodd" d="M10 2.5c-1.31 0-2.526.386-3.546 1.051a.75.75 0 01-.82-1.256A8 8 0 0118 9a22.47 22.47 0 01-1.228 7.351.75.75 0 11-1.417-.49A20.97 20.97 0 0016.5 9 6.5 6.5 0 0010 2.5zM4.333 4.416a.75.75 0 01.218 1.038A6.466 6.466 0 003.5 9a7.966 7.966 0 01-1.293 4.362.75.75 0 01-1.257-.819A6.466 6.466 0 002 9c0-1.61.476-3.11 1.295-4.365a.75.75 0 011.038-.219zM10 6.12a3 3 0 00-3.001 3.041 11.455 11.455 0 01-2.697 7.24.75.75 0 01-1.148-.965A9.957 9.957 0 005.5 9c0-.028.002-.055.004-.082a4.5 4.5 0 018.996.084V9.15l-.005.297a.75.75 0 11-1.5-.034c.003-.11.004-.219.005-.328a3 3 0 00-3-2.965zm0 2.13a.75.75 0 01.75.75c0 3.51-1.187 6.745-3.181 9.323a.75.75 0 11-1.186-.918A13.687 13.687 0 009.25 9a.75.75 0 01.75-.75zm3.529 3.698a.75.75 0 01.584.885 18.883 18.883 0 01-2.257 5.84.75.75 0 11-1.29-.764 17.386 17.386 0 002.078-5.377.75.75 0 01.885-.584z" clip-rule="evenodd" />
              </svg>
              Advanced security.
            </dt>
            <dd class="inline">Lorem ipsum, dolor sit amet consectetur adipisicing elit aute id magna.</dd>
          </div>
          <div class="relative pl-9">
            <dt class="inline font-semibold text-white">
              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
                <path fill-rule="evenodd" d="M7.84 1.804A1 1 0 018.82 1h2.36a1 1 0 01.98.804l.331 1.652a6.993 6.993 0 011.929 1.115l1.598-.54a1 1 0 011.186.447l1.18 2.044a1 1 0 01-.205 1.251l-1.267 1.113a7.047 7.047 0 010 2.228l1.267 1.113a1 1 0 01.206 1.25l-1.18 2.045a1 1 0 01-1.187.447l-1.598-.54a6.993 6.993 0 01-1.929 1.115l-.33 1.652a1 1 0 01-.98.804H8.82a1 1 0 01-.98-.804l-.331-1.652a6.993 6.993 0 01-1.929-1.115l-1.598.54a1 1 0 01-1.186-.447l-1.18-2.044a1 1 0 01.205-1.251l1.267-1.114a7.05 7.05 0 010-2.227L1.821 7.773a1 1 0 01-.206-1.25l1.18-2.045a1 1 0 011.187-.447l1.598.54A6.993 6.993 0 017.51 3.456l.33-1.652zM10 13a3 3 0 100-6 3 3 0 000 6z" clip-rule="evenodd" />
              </svg>
              Powerful API.
            </dt>
            <dd class="inline">Anim aute id magna aliqua ad ad non deserunt sunt. Qui irure qui lorem cupidatat commodo.</dd>
          </div>
          <div class="relative pl-9">
            <dt class="inline font-semibold text-white">
              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
                <path d="M4.632 3.533A2 2 0 016.577 2h6.846a2 2 0 011.945 1.533l1.976 8.234A3.489 3.489 0 0016 11.5H4c-.476 0-.93.095-1.344.267l1.976-8.234z" />
                <path fill-rule="evenodd" d="M4 13a2 2 0 100 4h12a2 2 0 100-4H4zm11.24 2a.75.75 0 01.75-.75H16a.75.75 0 01.75.75v.01a.75.75 0 01-.75.75h-.01a.75.75 0 01-.75-.75V15zm-2.25-.75a.75.75 0 00-.75.75v.01c0 .414.336.75.75.75H13a.75.75 0 00.75-.75V15a.75.75 0 00-.75-.75h-.01z" clip-rule="evenodd" />
              </svg>
              Database backups.
            </dt>
            <dd class="inline">Ac tincidunt sapien vehicula erat auctor pellentesque rhoncus.</dd>
          </div>
        </dl>
      </div>
    </div>

    <!-- Stats -->
    <div class="mx-auto mt-32 max-w-7xl px-6 sm:mt-56 lg:px-8">
      <div class="mx-auto max-w-2xl lg:mx-0 lg:max-w-xl">
        <h2 class="text-base font-semibold leading-8 text-indigo-400">Our track record</h2>
        <p class="mt-2 text-3xl font-bold tracking-tight text-white sm:text-4xl">Trusted by thousands of developers&nbsp;worldwide</p>
        <p class="mt-6 text-lg leading-8 text-gray-300">Lorem ipsum, dolor sit amet consectetur adipisicing elit. Maiores impedit perferendis suscipit eaque, iste dolor cupiditate blanditiis ratione.</p>
      </div>
      <dl class="mx-auto mt-16 grid max-w-2xl grid-cols-1 gap-x-8 gap-y-10 text-white sm:mt-20 sm:grid-cols-2 sm:gap-y-16 lg:mx-0 lg:max-w-none lg:grid-cols-4">
        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
          <dt class="text-sm leading-6">Developers on the platform</dt>
          <dd class="order-first text-3xl font-semibold tracking-tight">8,000+</dd>
        </div>
        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
          <dt class="text-sm leading-6">Daily requests</dt>
          <dd class="order-first text-3xl font-semibold tracking-tight">900m+</dd>
        </div>
        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
          <dt class="text-sm leading-6">Uptime guarantee</dt>
          <dd class="order-first text-3xl font-semibold tracking-tight">99.9%</dd>
        </div>
        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
          <dt class="text-sm leading-6">Projects deployed</dt>
          <dd class="order-first text-3xl font-semibold tracking-tight">12m</dd>
        </div>
      </dl>
    </div>

    <!-- CTA section -->
    <div class="relative isolate mt-32 px-6 py-32 sm:mt-56 sm:py-40 lg:px-8">
      <svg class="absolute inset-0 -z-10 h-full w-full stroke-white/10 [mask-image:radial-gradient(100%_100%_at_top_right,white,transparent)]" aria-hidden="true">
        <defs>
          <pattern id="1d4240dd-898f-445f-932d-e2872fd12de3" width="200" height="200" x="50%" y="0" patternUnits="userSpaceOnUse">
            <path d="M.5 200V.5H200" fill="none" />
          </pattern>
        </defs>
        <svg x="50%" y="0" class="overflow-visible fill-gray-800/20">
          <path d="M-200 0h201v201h-201Z M600 0h201v201h-201Z M-400 600h201v201h-201Z M200 800h201v201h-201Z" stroke-width="0" />
        </svg>
        <rect width="100%" height="100%" stroke-width="0" fill="url(#1d4240dd-898f-445f-932d-e2872fd12de3)" />
      </svg>
      <div class="absolute inset-x-0 top-10 -z-10 flex transform-gpu justify-center overflow-hidden blur-3xl" aria-hidden="true">
        <div class="aspect-[1108/632] w-[69.25rem] flex-none bg-gradient-to-r from-[#80caff] to-[#4f46e5] opacity-20" style="clip-path: polygon(73.6% 51.7%, 91.7% 11.8%, 100% 46.4%, 97.4% 82.2%, 92.5% 84.9%, 75.7% 64%, 55.3% 47.5%, 46.5% 49.4%, 45% 62.9%, 50.3% 87.2%, 21.3% 64.1%, 0.1% 100%, 5.4% 51.1%, 21.4% 63.9%, 58.9% 0.2%, 73.6% 51.7%)"></div>
      </div>
      <div class="mx-auto max-w-2xl text-center">
        <h2 class="text-3xl font-bold tracking-tight text-white sm:text-4xl">Boost your productivity.<br>Start using our app today.</h2>
        <p class="mx-auto mt-6 max-w-xl text-lg leading-8 text-gray-300">Incididunt sint fugiat pariatur cupidatat consectetur sit cillum anim id veniam aliqua proident excepteur commodo do ea.</p>
        <div class="mt-10 flex items-center justify-center gap-x-6">
          <a href="#" class="rounded-md bg-white px-3.5 py-2.5 text-sm font-semibold text-gray-900 shadow-sm hover:bg-gray-100 focus-visible:outline focus-visible:outline-2 focus-visible:outline-offset-2 focus-visible:outline-white">Get started</a>
          <a href="#" class="text-sm font-semibold leading-6 text-white">Learn more <span aria-hidden="true">→</span></a>
        </div>
      </div>
    </div>
  </main>

  <!-- Footer -->
  <footer aria-labelledby="footer-heading" class="relative">
    <h2 id="footer-heading" class="sr-only">Footer</h2>
    <div class="mx-auto max-w-7xl px-6 pb-8 pt-4 lg:px-8">
      <div class="border-t border-white/10 pt-8 md:flex md:items-center md:justify-between">
        <div class="flex space-x-6 md:order-2">
          <a href="#" class="text-gray-500 hover:text-gray-400">
            <span class="sr-only">Facebook</span>
            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
              <path fill-rule="evenodd" d="M22 12c0-5.523-4.477-10-10-10S2 6.477 2 12c0 4.991 3.657 9.128 8.438 9.878v-6.987h-2.54V12h2.54V9.797c0-2.506 1.492-3.89 3.777-3.89 1.094 0 2.238.195 2.238.195v2.46h-1.26c-1.243 0-1.63.771-1.63 1.562V12h2.773l-.443 2.89h-2.33v6.988C18.343 21.128 22 16.991 22 12z" clip-rule="evenodd" />
            </svg>
          </a>
          <a href="#" class="text-gray-500 hover:text-gray-400">
            <span class="sr-only">Instagram</span>
            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
              <path fill-rule="evenodd" d="M12.315 2c2.43 0 2.784.013 3.808.06 1.064.049 1.791.218 2.427.465a4.902 4.902 0 011.772 1.153 4.902 4.902 0 011.153 1.772c.247.636.416 1.363.465 2.427.048 1.067.06 1.407.06 4.123v.08c0 2.643-.012 2.987-.06 4.043-.049 1.064-.218 1.791-.465 2.427a4.902 4.902 0 01-1.153 1.772 4.902 4.902 0 01-1.772 1.153c-.636.247-1.363.416-2.427.465-1.067.048-1.407.06-4.123.06h-.08c-2.643 0-2.987-.012-4.043-.06-1.064-.049-1.791-.218-2.427-.465a4.902 4.902 0 01-1.772-1.153 4.902 4.902 0 01-1.153-1.772c-.247-.636-.416-1.363-.465-2.427-.047-1.024-.06-1.379-.06-3.808v-.63c0-2.43.013-2.784.06-3.808.049-1.064.218-1.791.465-2.427a4.902 4.902 0 011.153-1.772A4.902 4.902 0 015.45 2.525c.636-.247 1.363-.416 2.427-.465C8.901 2.013 9.256 2 11.685 2h.63zm-.081 1.802h-.468c-2.456 0-2.784.011-3.807.058-.975.045-1.504.207-1.857.344-.467.182-.8.398-1.15.748-.35.35-.566.683-.748 1.15-.137.353-.3.882-.344 1.857-.047 1.023-.058 1.351-.058 3.807v.468c0 2.456.011 2.784.058 3.807.045.975.207 1.504.344 1.857.182.466.399.8.748 1.15.35.35.683.566 1.15.748.353.137.882.3 1.857.344 1.054.048 1.37.058 4.041.058h.08c2.597 0 2.917-.01 3.96-.058.976-.045 1.505-.207 1.858-.344.466-.182.8-.398 1.15-.748.35-.35.566-.683.748-1.15.137-.353.3-.882.344-1.857.048-1.055.058-1.37.058-4.041v-.08c0-2.597-.01-2.917-.058-3.96-.045-.976-.207-1.505-.344-1.858a3.097 3.097 0 00-.748-1.15 3.098 3.098 0 00-1.15-.748c-.353-.137-.882-.3-1.857-.344-1.023-.047-1.351-.058-3.807-.058zM12 6.865a5.135 5.135 0 110 10.27 5.135 5.135 0 010-10.27zm0 1.802a3.333 3.333 0 100 6.666 3.333 3.333 0 000-6.666zm5.338-3.205a1.2 1.2 0 110 2.4 1.2 1.2 0 010-2.4z" clip-rule="evenodd" />
            </svg>
          </a>
          <a href="#" class="text-gray-500 hover:text-gray-400">
            <span class="sr-only">Twitter</span>
            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
              <path d="M8.29 20.251c7.547 0 11.675-6.253 11.675-11.675 0-.178 0-.355-.012-.53A8.348 8.348 0 0022 5.92a8.19 8.19 0 01-2.357.646 4.118 4.118 0 001.804-2.27 8.224 8.224 0 01-2.605.996 4.107 4.107 0 00-6.993 3.743 11.65 11.65 0 01-8.457-4.287 4.106 4.106 0 001.27 5.477A4.072 4.072 0 012.8 9.713v.052a4.105 4.105 0 003.292 4.022 4.095 4.095 0 01-1.853.07 4.108 4.108 0 003.834 2.85A8.233 8.233 0 012 18.407a11.616 11.616 0 006.29 1.84" />
            </svg>
          </a>
          <a href="#" class="text-gray-500 hover:text-gray-400">
            <span class="sr-only">GitHub</span>
            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
              <path fill-rule="evenodd" d="M12 2C6.477 2 2 6.484 2 12.017c0 4.425 2.865 8.18 6.839 9.504.5.092.682-.217.682-.483 0-.237-.008-.868-.013-1.703-2.782.605-3.369-1.343-3.369-1.343-.454-1.158-1.11-1.466-1.11-1.466-.908-.62.069-.608.069-.608 1.003.07 1.531 1.032 1.531 1.032.892 1.53 2.341 1.088 2.91.832.092-.647.35-1.088.636-1.338-2.22-.253-4.555-1.113-4.555-4.951 0-1.093.39-1.988 1.029-2.688-.103-.253-.446-1.272.098-2.65 0 0 .84-.27 2.75 1.026A9.564 9.564 0 0112 6.844c.85.004 1.705.115 2.504.337 1.909-1.296 2.747-1.027 2.747-1.027.546 1.379.202 2.398.1 2.651.64.7 1.028 1.595 1.028 2.688 0 3.848-2.339 4.695-4.566 4.943.359.309.678.92.678 1.855 0 1.338-.012 2.419-.012 2.747 0 .268.18.58.688.482A10.019 10.019 0 0022 12.017C22 6.484 17.522 2 12 2z" clip-rule="evenodd" />
            </svg>
          </a>
          <a href="#" class="text-gray-500 hover:text-gray-400">
            <span class="sr-only">YouTube</span>
            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
              <path fill-rule="evenodd" d="M19.812 5.418c.861.23 1.538.907 1.768 1.768C21.998 8.746 22 12 22 12s0 3.255-.418 4.814a2.504 2.504 0 0 1-1.768 1.768c-1.56.419-7.814.419-7.814.419s-6.255 0-7.814-.419a2.505 2.505 0 0 1-1.768-1.768C2 15.255 2 12 2 12s0-3.255.417-4.814a2.507 2.507 0 0 1 1.768-1.768C5.744 5 11.998 5 11.998 5s6.255 0 7.814.418ZM15.194 12 10 15V9l5.194 3Z" clip-rule="evenodd" />
            </svg>
          </a>
        </div>
        <p class="mt-8 text-xs leading-5 text-gray-400 md:order-1 md:mt-0">&copy; 2020 Your Company, Inc. All rights reserved.</p>
      </div>
    </div>
  </footer>
</div>

```

### 2024-03-02 manuscript/other_manuscript_draft.qmd
img uploads

---
title: |
  Longitudinal Analysis Manuscript: Working Draft
subtitle: Longitudinal Analysis using data from the ABCD© Study
short-title: Longitudinal Analysis
code-repo: "Access the code, data, and analysis at <http://swhawes.github.io/test/>"
number-sections: true
number-depth: 3
correspondence-prefix: "Correspondence concerning this article should be addressed to"
author:
  - name: Samuel Hawes
    email: shawes@fiu.edu
    title: Associate Research Professor
    affiliations:
      #- id: ds9
        name: Florida International University
        department: Center for Children & Families
        address: 11200 SW 8th St, AHC-IV
        city: Miami
        region: FL
        country: USA
        postal-code: 33199
    attributes:
      corresponding: true
  - name: Andrew K. Littlefield
    email: andrew.littlefield@ttu.edu
    affiliations:
      #- id: ds9
        name: Texas Tech University
        department: Psychological Sciences
        city: Lubbock
        region: TX
        country: USA
  - name: Daniel A. Lopez
    email: Daniel_Lopez2@URMC.Rochester.edu
    affiliations:
      #- id: terok
        name: University of Rochester Medical Center
        department: Department of Public Health
        city: Rochester
        region: NY
        country: USA
      #- ref: ds9
  - name: Kenneth J. Sher
    email: SherK@missouri.edu
    affiliations:
      #- id: ds9
        name: University of Missouri
        department: Psychological Sciences
        city: Columbia
        region: MO
        country: USA
  - name: Erin L. Thompson
    email: erthomps@fiu.edu
    affiliations:
      #- id: terok
        name: Florida International University
        department: Center for Children and Families
        city: Miami
        region: FL
        country: USA
  - name: Raul Gonzalez
    email: gonzara@fiu.edu
    affiliations:
      #- id: ds9
        name: Florida International University
        department: Center for Children and Families
        city: Miami
        region: FL
        country: USA
  - name: Additional Co-authors
    email: xxxxx@fiu.edu
    affiliations:
      #- id: terok
        name: xxxxxx University
        department: xxxxxx Dept
        city: xxxx
        region: xxxxx
        country: xxxxx
  - name: Wesley K. Thompson
    email: wes.stat@gmail.com
    affiliations:
      #- id: terok
        name: Laureate Institute for Brain Research
        department: Center for Population Neuroscience and Genetics
        city: Tulsa
        region: OK
        country: USA

format:
  docx:
    toc: false
  elsevier-pdf:
    journal:
      cite-style: number
      formatting: preprint
      model: 3p
      layout: twocolumn
      #cite-style: number
    include-in-header: 
      text: |
        \newpageafter{author}
  html:
    page-layout: article
    toc: true
    toc-depth: 1
    toc-expand: 2
    toc-title: Table of Contents
    code-tools:
      source: true
      toggle: false
      caption: none
    grid:
      sidebar-width: 350px
      margin-width: 20px

#fontfamily: libertinus
#fontsize: 20pt
#colorlinks: true
#format-links: [html, pdf]
#execute:
  #echo: false
#bibliography: "references.bib"
#biblio-style: unsrt

abstract: |
  The Adolescent Brain Cognitive Development^SM (ABCD) Study provides a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of youths, aged 9-10 at baseline and assessed annually for 10 years. Given the size and complexity of the ABCD Study©, researchers analyzing its data will encounter a myriad of methodological and analytical considerations. This review provides an examination of key concepts and techniques related to longitudinal analyses of the ABCD Study data.  We discuss the importance of longitudinal data, focusing on the types of inferences that are possible when one assesses individuals across multiple time points, including: 1) characterization of the factors associated with variation in developmental trajectories; 2) assessment of how level and timing of exposures may impact subsequent development; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships. We emphasize the importance of selecting appropriate statistical models to address these research questions, e.g., accounting for correlation in repeated measurements, employing linear or non-linear models as indicated by the data, and using link functions that adequately model outcome distributions. By presenting the advantages and potential challenges of longitudinal analyses in the ABCD Study, this review seeks to equip researchers with foundational knowledge and tools to make informed decisions as they navigate and effectively analyze and interpret the multi-dimensional longitudinal data currently available.

keywords:
- ABCD Study
- longitudinal analyses
- development
---

# Introduction

\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (Volkow et al. 2018). At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes. Central to achieving these goals is the commitment of the ABCD Study® and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods, by espousing practices that increase access, integrity, and reproducibility of scientific research. In this sense, the ABCD Study is a collaboration with the larger research community.

The size and scope of the ABCD Study data allow the research community to perform a large variety of developmental analyses of both substantive and methodological interest,  presenting a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes unfold across critical periods of development. In this paper, we describe models and methods for longitudinal analysis of ABCD Study data that can address these fundamental scientific aims, including: 1) characterization of the genetic and environmental factors associated with variation in developmental trajectories; 2) assessment of how the level and timing of exposures may impact subsequent neurodevelopment; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships. We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.0 data with accompanying R scripts. Worked examples are available in Quarto files, accessible at **XXXX**.

## The ABCD Study Data

The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged between 9-11 years at baseline, as well as their parents/guardians. The study sample was recruited from household populations in defined catchment areas for each of the 21 (originally 22) study sites across the United States. Information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at https://abcdstudy.org/. The ABCD Study design is described in more detail in @garavan2018 and @dick2021.

The ABCD Study is currently collecting longitudinal data on a rich variety of outcomes that will enable the construction of complex statistical models potentially incorporating factors from many domains. Each new wave of data collection provides another building block for characterizing developmental trajectories and implementing longitudinal analyses that allow researchers to characterize normative development, to identify variables that presage deviations from normative development, and to assess a range of variables associated with biopsychosocial outcomes of interest. These data include: 1) a neurocognitive battery [@luciana2018a; @thompson2019]; 2) mental and physical health assessments [@barch2018]; 3) measures of culture and environment [@gonzalez2021; @zucker2018]; 4) substance use [@lisdahl2021]; 5) gender identity and sexual health [@potter2022]; 6) biospecimens [@uban2018]; 7) structural and functional brain imaging [@casey2018; @hagler2019]; 8) geolocation-based environmental exposure data [@fan2021]; 9) wearables, and mobile technology [@bagot2018]; and 10) whole genome genotyping [@loughnan2020]. Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual telephone interviews between in-person visits.

Data are publicly released roughly annually, currently through the NIMH Data Archive (NDA). By necessity, the study’s earliest data releases consisted primarily of one or two visits per participant. However, the most recent public release as of the writing of this paper (Release 5.0) contains data collected across five annual visits, including three brain imaging assessments (baseline, year 2 follow-up, and year 4 follow-up visits) for at least a subset of the cohort. Hence, starting with Release 5.0, it is feasible for researchers to begin focusing on the estimation of neurodevelopmental and other trajectories.

##

::: {.callout-note collapse="true" icon="false"}
## Organization and Aims

• Part I. Longitudinal Research

-   Identify fundamental concepts

• Part II. Longitudinal Data

-   Highlight key challenges

• Part III. Longitudinal Analysis

-   Methods & Analysis

• Part IV. Supplemental materials

-   Linked open-source resources
:::

# Developmental Research

\label{sec:headings}

## Basic Concepts and Considerations

There are several important concepts to consider when conducting longitudinal analyses in a developmental context. These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represents the same or different phenomena.

Moreover, in the case of developmentally-focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child’s life course but also brings with it greater methodological opportunities to make use of statistical models that distinguish within- from between-person effects and that loosen constraints that need to be imposed on the furtherance of critical scientific questions.

For example, collecting two or more within-person observations on the same construct at different times enables estimation of individual rates of change (slopes); more observations allow for more precise estimates of individual slopes, as well as characterization of non-linear development. Rate of change or other trajectory characteristics may be more informative about individuals than the simple snapshots of level differences that cross-sectional data are limited to informing. Cross-sectional age-related differences across individuals are poor substitutes for longitudinal trajectory estimates except under highly restrictive assumptions, e.g., parallel trajectories and lack of age cohort and experience effects [@thompson2011]. Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.

### Vulnerable periods.

Adolescent development progresses normatively from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@arseneault2002; @bechtold2016; @hasan2020; @semple2005]; or similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development, corresponding to early adolescence in humans [@spear2016; @crews2000; @ji2018]. In another example, longitudinal data from the National Consortium on Alcohol in Adolescence (NCANDA) show that binge drinking is associated more strongly with decrements in gray matter volume early in adolescence compared to later [@infante2022]. These examples highlight the importance of considering the role of vulnerable periods – e.g., temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.

### Developmental disturbances.

Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times, environmental exposures will tend to suppress stability and disrupt the orderly stochastic process of normative development [e.g., @schulenberg2019]. This situation reflects a developmental disturbance in that the normal course of development is “altered” for a time by some time-limited process. In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period. That is, once the environmental presses are removed (or the individual is removed from the environment), patterns of individual differences (and autoregressive effects) recover to levels similar to those prior to entering the environment. For example, in @infante2022, recent binge drinking appears to be most predictive of gray matter volume trajectories, as opposed to more distal prior binge drinking or cumulative number of binge drinks, suggesting the potential for recovery of gray matter trajectories to prior levels of growth if binge drinking levels subside.

### Developmental snares and cascade effects.

Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior [@moffitt2015]. Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified [e.g., @masten2005; @rogosch2010]. Although conceptually distinct from vulnerable periods, both types of developmental considerations highlight the importance of viewing behavior in the context of development and attempting to determine how various developmental pathways unfold. Longitudinal data are crucial in this context to assess individual levels of development prior to and following onset of experiences or other environmental factors; e.g., the ABCD Study collected data starting at ages 9-10 and hence before the onset of substance use for the vast majority of participants.

**2.1.4 Intermediate processes and feedback loops**
**Something here on the concept of mediation and cross-lagged relationships?**

# Longitudinal Data

\label{sec:headings}

## Interpretation / Issues / Pitfalls & Assumption

The hallmark characteristic of longitudinal data analysis is the administration of repeated measurements of the same constructs on assessment targets (e.g., individuals, families) across time. The primary rationale for collecting longitudinal data is to assess within-person change over time, allowing researchers to estimate individual developmental trajectories and the genetic and/or environmental factors that may impact these trajectories. Administering repeated measurements more frequently or over longer time spans enables researchers to ask more nuanced questions and to make stronger inferences.

### Two Time Points versus Three or More.

Although the clear leap from cross-sectional to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As @rogosa1982 noted over forty years ago, “Two waves of data are better than one, but maybe not much better” (p. 744).

These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for trajectory estimation, model identification and accurate parameter inferences. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (given that linear change is the only estimable form for two assessments waves; [see @duncan2009]). Research designs that include three (but preferably more) time points allow for non-linear trajectory estimation and increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of developmental research.

To illustrate, developmental theories are useful for understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” [@littlefield2021]. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability (e.g., the Latent Curve Model with Structured Residuals; @curran2014) require at least four assessments to parameterize fully and, more generally, increasingly accurate and nuanced parameter estimates are obtained as more assessment occasions are used [@duncan2009].

### Types of stability and change

If one were to try to sum up what developmental trajectories in a living organism are exactly, one could plausibly argue they are the patterns of stability and change in its phenotypes as the organism traverses the life course. Symbolically, developmental trajectories can be expressed as fi(t), a possibly multivariate function of time t, specific to the ith individual and typically taking values in the real numbers for continuous phenotypes and the integers for discrete phenotypes. Ideally, t is a biologically meaningful temporal index (e.g., calendar age) as opposed to an exogenous progression of events (e.g., study visit number). Properties of interest might include rate of change over time, degree of smoothness (e.g., continuously differentiable), shape (e.g., polynomial or asymptotic behavior), how and how much fi(t) differs across individuals, and what factors predict either within-individual variation (at different times) or between-individual variation (either overall or at specific times).

There are a few different ways to think about patterns of stability and change (see Figure 1). Consider measuring school disengagement at the start of middle school and the end of middle school . A common first step may be to compare sixth graders’ average disengagement values and eighth graders’ disengagement values. This comparison of the average scores for the same group of individuals at multiple time points is referred to as “mean-level”, as it provides information about change over time (or lack thereof) for an outcome of interest aggregated across members of a group. In contrast, “between-individual” stability could be assessed, e.g., by calculating the Spearman correlation between the values obtained at different time points (e.g., ‘disengagement in sixth grade’ with ‘disengagement in eighth grade). This analysis focuses on the degree to which individuals retain their relative placement in a group across time. Consider someone who reported the lowest frequencies of disengagement in 6th grade and may report significantly higher disengagement over middle school (i.e., exhibit high levels of change), but report the lowest frequencies of disengagement in eighth grade. That is, the individual is manifesting rank-order stability, even in the context of high mean-level change.

Both types of stability and change are important. Mean-level change in certain traits might help to explain why, in general, populations of individuals tend to be particularly vulnerable to the effects of environmental factors in specific age ranges; rank-order stability might help to quantify the extent to which certain characteristics of the individual are more or less trait-like compared to others. For example, in some areas of development, considerable mean-level change occurs over time (e.g., changes in Big 5 personality traits; @bleidorn2022), but exhibit relatively high rank-order stability, at least over shorter measurement intervals [@bleidorn2022; @roberts2000; @roberts2006]. Despite the useful information afforded by examining mean-level and rank-order stability and change, these approaches are limited in that they provide little information about the overall patterns of within-individual change (i.e., trajectories) and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest [@curran2011]. For example, questions related to the impact of early-onset substance use on brain development focus on changes within a given individual (i.e., intraindividual differences). The longitudinal nature of the ABCD Study, which will provide researchers with over ten time points for some constructs (e.g., substance use) across a ten-year period, allows for the study of within-person processes.

![Types of Stability and Change](./figures/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}

### Use of appropriate longitudinal models
There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2011; @hoffman2015; @littlefield2021].

First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, [e.g., @curran2011]). Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015; latent curve models with structured residuals [LCM-SR], @curran2014). It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].

Second, many statistical models assume certain characteristics about the data to which they are being applied. Common assumptions of parametric statistical models (e.g., linear mixed-effects models) include normality and equality of variances. These assumptions must be carefully considered before finalizing analytical approaches, so that valid inferences can be made from the data, as violation of a model’s assumptions can substantively invalidate the interpretation of results. For example, longitudinal data can exhibit heterogeneous variability (i.e., the variance of the response changes over the duration of the study) that may need to be accounted for within a model. Another pertinent modeling assumption is whether trajectories are linear or non-linear. With 2-3 assessments per individual, only a linear model of within-person change is usually feasible; with more time points, higher level polynomials or models with more flexible trajectory shapes (e.g., curve smoothing via cubic splines) becomes possible [REFs]. Note, however, baseline age in the ABCD Study ranges over two full years; for some outcomes it may be desirable to include a (possibly nonlinear) effect of baseline age along with a linear effect of within-person change in age with only 2-3 assessment times [@thompson2013]. As the study progresses and more times points are assessed, nonparametric curve estimation at the mean [GAMMs] and at the individual level may also become useful [@ramsay2002].

### Continuous and Discrete Outcomes
Implementing valid and efficient statistical models requires an understanding of the type of data being analyzed. For example, repeated assessments within the ABCD Study can be based on continuous or discrete measures. Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., caregiver-reported items measuring emotional and behavioral concerns via the Child Behavior Checklist including the categories of “Not True”, “Somewhat True”, and “Very True”), and count variables (e.g., number of cigarettes smoked per day). In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016]. For example, the Mplus manual [@muthen2017] includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, and d) linear growth models for a count outcome assuming a zero-inflated Poisson model. Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data [@ren2022]. These models can account for issues that can occur when working with discrete outcomes, including overdispersion [when the variance is higher than would be expected based on a given distribution; see @lenz2016]. Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous variables [e.g., @little2013]. However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes [e.g., @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes].

### Issues in attributing longitudinal change to development
One can observe systematic changes over time in a variable of interest and assume this change is attributable to development. However, various pitfalls with longitudinal data can complicate or even invalidate this conclusion. For example, if data missingness or participant dropout are related to the values of the outcome, changing sample composition as the study progresses can bias mean trajectory estimates (we describe this in more detail in Section 3.1.7 below). Another prerequisite for valid developmental interpretations of longitudinal data is to establish whether a construct is measured consistently over time, i.e., longitudinal measurement invariance [@liu2017; @van2015; @willoughby2012]. Establishing longitudinal measurement invariance provides researchers with greater confidence that any change over time identified for a construct is attributable to individual change rather than a measurement artifact. For example, one study using data from the ABCD Study [@brislin2023) found differential item functioning in two items from a brief delinquency measure, revealing significant bias in an arrest item across Black and White youth. More specifically, Black youth were more likely to report being arrested compared to White youth with similar levels of delinquency. Prevalence rates of delinquent behavior would have been severely biased if measurement invariance had not been tested.

Observed patterns of growth and decline often differ between cross-sectional vs. longitudinal effects [@salthouse2014] where subjects gain increasing experience with the assessment with each successive measurement occasion. Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically. In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before). Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; @french2010] and this phenomenon has been well documented in research using structured diagnostic interviews [@robins1985]. While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018].

Some longitudinal studies, e.g., accelerated longitudinal designs [ALDs; @thompson2011] are especially well suited for discovering these effects and modeling them. While ABCD is not an ALD, the variability in age (and grade in school) at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an ALD (e.g., substance use; prosocial behavior; family conflict; screen time). It is possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, wherein number of prior assessments and age are mostly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

### Covariance Structures

A central issue for repeated measurements on an individual is how to account for the correlated nature of the data. Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess the same individuals across time. That is, given that residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results. Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic [@durbin1950], more commonly, independence is assumed to be violated in study designs with repeated assessments. Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.

Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (involving the covariance among pairs of repeated outcomes on an individual), coupled with a model for the mean response and its dependence on covariates (e.g., sex at birth). There exists a range of methods to model covariance structures, each with its own set of tradeoffs between model fit and parsimony and which may be more or less appropriate for each specific application [e.g., see @kincaid2005].

There are various approaches to adjust models for the lack of independence of residuals across time, be it for data with repeated assessments or other situations with so-called nested data (e.g., children nested within schools; siblings nested within families). The most common approach is to use random effects. Essentially, random effects allow for variance estimates around fixed effects. A classic example (from @bryk1992; @singer1998) involves math achievement measured among students nested within schools. In a basic, intercept-only model with no covariates, there would be one fixed effect (the grand mean, or intercept, of math achievement) and two random effects (one representing variation in the intercept, or the variation between schools, and another for the within-school residual, or how much variance in math achievement is left over after taking into account nesting in school). From this framework, each student’s score would be a function of the fixed effect (the overall mean), the variation in scores that exists between schools, and the variation that exists among students within the same school. Assumptions about the variance and covariance components of this model dictate the form of the variance/covariance structure. For example, if we assume the random effects are independent (i.e., uncorrelated), the implied structure would be compound symmetry, where it is assumed the covariance of any two students in a single class is represented by a random effect (the variance between school means) and the covariance of any two students in different classes is zero. The assumptions of this relatively simple covariance structure can be relaxed, resulting in different covariance structures with additional parameters (see @singer1998).

In a similar fashion, individual growth models could be fit to a series of repeated assessments, where assessments are nested within individuals. For example, an unconditional linear growth model would involve two fixed effects – one for the intercept (an estimate of the average score when time is coded zero) and one for the linear slope (an estimate of the change in scores for each unit increase in time). Random effects could include a random effect for intercept (or an estimate of the variation in scores when time is coded as zero), a random effect for the linear slope (or an estimate of the variation in linear change across time), and a random effect for the within-person residual (or an estimate of the left over variance, or residual, of a given assessment when accounting for the intercept and linear slope). Assumptions regarding the covariation among the random effects also indicate different covariance structures. For example, it is typical to assume the random intercept and random slope components covary (e.g., if time is coded such that the first time point = zero, the covariance estimate suggests that an individual’s score at baseline relates to the amount of change exhibited across time). Further, particularly in structural equation model forms of this model, it is typically assumed the random effect for the within-person residual varies across assessments (@curran2003).

One alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design. As the name implies, the structure assumes a subsequent measurement occasion (e.g., assessment at Wave 2) is regressed onto (that is, is predicted by) a prior measurement occasion (e.g., assessment at Wave 1). The most common type of autoregressive design is the AR(1), where assessments at time T + 1 are regressed on assessments at Time T. Identical to compound symmetry, this model assumes the variances are homogenous across time. Diverting from compound symmetry, this model assumes the correlations between repeated assessments decline exponentially across measurement occasions rather than remaining constant. That is, we can think of the underlying process as a stochastic one that wears itself out over time. For example, per the AR(1) structure, if the correlation between Time 1 and Time 2 data is thought to be .5, then the correlation between Time 1 and Time 3 data would be assumed to be .5 × .5 = .25, and the correlation between Time 1 and Time 4 data would be assumed to be .5 × .5 × .5 = .125. As with compound symmetry, the basic AR(1) model is parsimonious in that it only requires two parameters (the variance of the assessments and the autoregressive coefficient). Notably, the assumption of constant autoregressive relations between assessments is often relaxed in commonly employed designs that use autoregressive modeling (e.g., cross-lagged panel models [CLPM]). These designs still typically assume an AR(1) process (e.g., it is sufficient to regress the Time 3 assessment onto the Time 2 assessment and is not necessary to also regress the Time 3 assessment onto the Time 1 assessment, which would result in an AR(2) process). However, the magnitude of these relations is often allowed to differ across different AR(1) pairs of assessment (e.g., the relation between Time 1 and Time 2 can be different from the relation between Time 2 and Time 3). These more commonly employed models also often relax the assumption of equal variances of the repeated assessments. Although the AR(1) structure may involve a more realistic set of assumptions compared to compound symmetry, in that the AR(1) model allows for diminishing correlations across time, the basic AR(1) model, as well as autoregressive models more generally, can also suffer from several limitations in contexts that are common in prospective designs. In particular, recent work demonstrates that if a construct being assessed prospectively across time is trait-like in nature, then a simple AR(1) process fail to adequately account for this trait-like structure, with the downstream consequence that estimates derived from models based on AR structures (such as the CLPM) can be misleading and fail to adequately demarcate between- vs. within-person sources of variance [@hamaker2015]. Finally, it is also worth noting that in many longitudinal contexts, the time intervals between assessments are not equidistant and researchers need to consider carefully how to appropriately model time in their model and what this model implies.

### Missing Data/Attrition

Attrition from a longitudinal panel study such as ABCD is inevitable and represents a potential threat to the validity of longitudinal analyses and cross-sectional analyses conducted at later time points, especially since attrition can only be expected to grow over time[@littlefield2022]. The ABCD Retention Workgroup (RW) employs a data-driven approach to examine, track, and intervene in these issues and while preliminary findings show participant race and parent education level to be associated with late and missing visits, although to date, attrition in ABCD has been minimal [@ewing2022]. Ideally, one tries to minimize attrition through good retention practices from the outset via strategies designed to maintain engagement in the project [@cotter2005; @hill2016; @watson2018]. However, even the best-executed studies need to anticipate growing attrition over the length of the study and implement analytic strategies designed to provide the most valid inferences. Perhaps the most key concern when dealing with data that is missing due to attrition is determining the degree of bias in retained variables that is a consequence of attrition. Such bias can attenuate generalizability, particularly if the pattern of missingness is not random (e.g., certain subsets of the population are more likely to drop out/not attend a visit). Assuming that the data are not missing completely at random, attention to the nature of the missingness and employing techniques designed to mitigate attrition-related biases need to be considered in all longitudinal analyses.

Three types of missingness are considered in the literature [see @little1989], namely: a) missing completely at random (MCAR), b) missing at random (MAR), and c) missing not at random (MNAR). Data that are MCAR are a simple random sample of all data in a given dataset. MAR implies missing data are a random sample (i.e., does not hinge on some unmeasured variables) within strata of the measured covariates in a dataset (e.g., biological sex). Data that are MNAR are missing as a function of unobserved variables and may bias assocations even after conditioning on the observed covariates. @graham2009 provides an excellent and easy-to-digest overview of further details involving missing data considerations.

Modern approaches for handling missing data, such as full information maximum likelihood, propensity weighting, auxiliary variables and multiple imputation avoid the biases of older approaches [see @enders2010; @graham2009]. @graham2009 noted several “myths” regarding missing data. For example, Graham notes many assume the data must be minimally MAR to permit estimating procedures (such as maximum likelihood or multiple imputation) compared to other, more traditional approaches (e.g., using only complete case data). Violations of MAR impact both traditional and more modern data estimation procedures, though as noted by Graham, violations of MAR tend to have a greater effect on older methods. Graham thus suggests that imputing missing data is a better approach compared to listwise deletion in most circumstances, regardless of the model of missingness [i.e., MCAR, MAR, MNAR; see @graham2009].

### Quantifying effect sizes longitudinally

Given that longitudinal data involve multiple sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data. An effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014]. Common effect size metrics include the correlation r between two variables and the standardized difference between two means, Cohen’s *d* [@cohen1988]. An extensive discussion of effect sizes and their relevance for ABCD is given in @dick2021. Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered (e.g., @morris2002]. @wang2019 note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic [see @wang2019, for more details]. Thus, obtaining effect size metrics based on standardized estimates that are relatively simple in cross-sectional data (such as r) becomes more complex in the context of prospective longitudinal data. @feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from traditional analysis [see @feingold2009, for more details]. Given this issue, there have been various proposals for adjusting effect size measures in repeated assessments. @feingold2019 reviews the approach for effect size metrics for analyses based on growth modeling, including when considering linear and non-linear (i.e., quadratic) growth factors. @morris2002 review various equations for effect size calculations relevant to combining estimates in meta-analysis with repeated measures and independent-groups designs. Other approaches to quantifying effect sizes longitudinally may be based on standardized estimates from models that more optimally disentangle between- and within-person sources of variance. As an example, within a random-intercept cross-lagged panel model (RI-CLPM) framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.

### Longitudinal Data Structures

An ideal longitudinal study integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model [@collins2006]. To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure **X**). An understanding of these data structures is helpful, as they can warrant different types of longitudinal data analysis (LDA). Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and Bauer [2019; @bauer2019] provide a nice on-ramp in first distinguishing between the use of “time-to-event” and “repeated measures” data. Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change [@bauer2019]. Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest. This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models) and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events. While much has been written about “time-to-event” data (**see xxx; xxx; xxxx**), including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study (@brislin2023), our emphasis will be given to the modeling of “repeated measures” data.

[![Longitudinal Data Analysis Structural Diagram](./figures/LDAStructure2.png)](https://centerstat.org/resources/)

When discussing longitudinal analysis, we are most often talking about data collected on the same unit (e.g., individuals) across multiple measurement occasions. However, repeated-measures analysis is not a monolith, and it will serve us well to distinguish between a few of the most common types. One such approach to repeated measures analysis is the use of time-series models. These models generally consist of a long sequence of repeated measurements (≧ 50-100 measurements) on a single (or small number of) variable of interest. Time-series analysis is often used to predict/forecast temporal trends and cyclic patterns and is geared toward making inferences about prospective outcomes within a population (with relatively with less focus on inferring individual-level mechanisms and risk factors). A related type of repeated measures analysis is Intensive Longitudinal Data (ILD). Similar to time-series analysis, ILD models involve frequent measurements (~ 30-40 measurements) of the same individuals in a relatively circumspect period of time (e.g., experience sampling to obtain time series on many individuals). Although ILD models may include slightly fewer measurement occasions than time-series data, ILD models tend to have more subjects than time series models (~ 50-100 subjects). This allows ILD models to examine short-term patterns by incorporating a time series model that can sometimes fit parameter estimates to each individual’s data in order to model individual difference outcomes. The final type of repeated measures analysis that we will primarily focus on is the longitudinal panel model. These models follow a group of individuals— a panel (also referred to as a cohort) — across relatively fewer measurement occassions (~ 5-15), and are often interested in examining change across both, within-individuals and between-individuals.

While other longitudinal designs have their own unique strengths and applications, the longitudinal panel design is particularly well-suited for investigating developmental processes in the context of the ABCD Study. In the following sections, we will discuss various analytic methods commonly used to analyze longitudinal panel data, including growth models, mixed models, and a number of additional trajectory models. These methods provide valuable insights into within- and between-individual differences and are highly relevant for researchers working with the ABCD Study dataset. By focusing on these methods, we aim to equip readers with the knowledge necessary to conduct longitudinal research and perform analyses using the rich, longitudinal, and publicly available data from the ABCD Study.

# Longitudinal Analysis

\label{sec:headings}

### Types of longitudinal panel models

With the large and continually expanding body of research on statistical methods for longitudinal analyses, determining which longitudinal model to implement can be challenging. The aim of this section is to help researchers navigate these many options to identify the statistical approach most appropriate to their unique research question when deciding on how to measure change over time. Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.

Common examples include grouping by linear vs nonlinear models [@collins2006], the number of measurement occasions [@king2018], and statistical equivalency [e.g., change scores vs. residualized change; see @castro-schilo2018]. The organization we use below overlaps in a number of ways with these examples, and in particular with @bauer2019. However, it is important to note that in each case, the chosen way of grouping is primarily intended to allow the reader to compare and contrast various analytical approaches. In the following sections, we briefly summarize the advantages/disadvantages of a series of longitudinal models organized into the following groupings: Traditional Models, Modern GLM Extensions, SEM, and Advanced SEM. We note that this is not an exhaustive review of each of these methods, and for more in-depth detail we do provide the reader with relevant resources. As aptly summarized by @bauer2019, “there are many exceptions, alternatives, nuances, ‘what ifs’, and ‘but couldn’t you’s that aren’t addressed here.”

![Longitudinal Models/Data Structures](./figures/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"} <br>

#### Traditional Models

Traditional methods for longitudinal analysis primarily focus on modeling mean-level change, and how these changes may differ across groups or levels of some other variable. For example, is there a difference in average internalizing symptoms obtained across multiple assessments between boys and girls? Longitudinal models that focus on mean-level change are also referred to as marginal models and examples of specific methods include repeated measures ANOVA (and MANOVA), ANCOVA, and Generalized Estimating Equations (GEEs). Mean-level change models are commonly used when data is only available from 2 measurement occasions. For example, computing a difference score (e.g., mean internalizing scores at time 2 - mean internalizing scores at time 1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups. Additionally, the longitudinal signed-rank test, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data. Another common approach, often used in pre-/post-design studies but can be used with ABCD Study data, is to use residualized change score analysis to assess the degree of change in a variable, while controlling for its initial level [@castro-schilo2018].

For example, to examine change in cortico-limbic connectivity among ABCD participants, @brieant2021 regressed cortico-limbic connectivity at the year 2 follow-up on baseline cortico-limbic connectivity, which allowed the authors to examine the associations between negative life events and the variance of cortico-limbic connectivity unexplained by baseline connectivity. Similarly, @romer2021 used a residualized-change model to examine the bidirectional influences of executive functioning and a general psychopathology factor ‘p’ across the first two years of the ABCD Study. Both studies were able to conclude associations between their constructs of interest that could not be accounted for by prior frequencies at baseline.

Traditional longitudinal, such as residualized chance score models, can be useful in some contexts (e.g., two measurement occasions), but overall, their practical utility for answering questions about developmental processes is limited. Perhaps most notably, these models do not allow for characterizing patterns of within-person change. This is a particularly important limitation since most psychological theories posit within-person processes (i.e., what will happen within a given individual). As such, traditional approaches often correspond poorly with most theoretical models of change and a failure to disaggregate between-person and within-person effects can result in consequential errors of inference [e.g., ecological fallacy, @curran2011]. Moreover, even determining which of these procedures to use for comparing change over two time points across groups can be surprisingly complicated. A particularly vexing example is that of imbalanced baseline scores (i.e., when baseline scores are correlated with a covariate of interest), which can produce different conclusions across methods [e.g., **X versus X**; see @littlefield2023, for a review]. Given these shortcomings, and the complexity of the issues surrounding some of these methods, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data (and preferably make use of data collected across three or more time points, as is currently true for many ABCD assessments).

#### Modern GLM Extensions

Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability. Generalized Estimating Equations (GEE), Linear Mixed Models (LMM), Generalized Linear Mixed Models (GLMM), and Autoregressive Cross-Lagged Panel Models (ARCL) are examples of such contemporary techniques. GEE, an extension of Generalized Linear Models, combines the generalized linear model for non-normal outcomes with repeated measures (marginal) model and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects. For example, @vandijk2021 used GEE to obtain relative risks for psychiatric diagnoses among children in the ABCD Study with a family history of depression and used the ABCD Study sampling weights to generalize prevalence rates among 9 and 10 year-olds across the US. LMMs, also known as multilevel or hierarchical linear models, facilitate the simultaneous analysis of within-person and between-person variability, making them ideal for nested data structures or repeated measures. Within the ABCD Study, reseachers may want to consider nesting by individual, family (i.e., siblings or twins), school or district, and/or site. GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data, such as the use of ABCD data on substance use [e.g., @martz2022] screen media use [@lees2020], and microstructure of the brain [@palmer2022]. Finally, ARCL models are used to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects [although, ARCL models are relatively less useful for teasing apart between-person and within-person sources of variances; see @curran2021].

The strengths of these modern methods lie in their ability to account for individual differences, within-person change, and time-varying predictors, thereby providing a more comprehensive understanding of complex relationships in longitudinal data. Despite these advantages, modern approaches may require more complex modeling assumptions and higher computational demands compared to traditional methods. Additionally, proper model specification and the interpretation of results can be more challenging, especially in cases of high multicollinearity or missing data. However, modern longitudinal analysis methods have generally surpassed traditional methods in addressing a wider range of research questions, accommodating diverse data structures, and elucidating the intricate dynamics of developmental processes.

#### SEM

Structural equation modeling (SEM) approaches have gained prominence in longitudinal data analysis due to their ability to estimate complex relationships among observed and latent variables while accounting for measurement error. SEM techniques can be categorized into variable-centered, person-centered, and hybrid approaches, each with unique strengths and limitations. The choice of method depends on the research question, data structure, and underlying assumptions.

Variable-centered approaches, such as latent change scores, latent growth curve models, multivariate (parallel process) latent growth curve models, and latent state-trait models, focus on examining relationships among variables and population-level patterns. These models offer a powerful means to estimate individual change trajectories and the relationships between growth parameters of different variables, while also decomposing observed measurements into latent state and trait components. However, these approaches may not adequately capture distinct subgroups of individuals who share similar patterns of change over time, which can be crucial for understanding heterogeneous developmental processes.

Person-centered approaches, including latent transition analysis and latent class growth analysis, address this limitation by identifying subgroups of individuals who share similar patterns of change. These models can reveal meaningful subpopulations and help researchers understand the factors that contribute to differences in developmental trajectories. For example, taking advantage of the large sample size of the ABCD Study, @xiang2022 found evidence of four subgroups of youth with unique longitudinal patterns of depressive symptoms over time and identified risk factors that were differentially associated with the various trajectories. The use of such models allows for a more nuanced understanding of the associations between risk factors and change in symptomatology, as opposed to a snapshot of symptomatology at a single time point. Despite a range of potential model specifications for longitudinal mixture modeling, person-centered approaches tend to use parameterizations that default to settings found in popular software packages (e.g., Mplus). It has recently been demonstrated (see @mcneish2021) that the use of such specifications tends to identify the so-called “cat’s cradle” solution (see @sher2011) that consists of “…(a) a consistently ‘low’ group, (b) an ‘increase’ group, (c) a ‘decrease’ group, and (d) a consistently ‘high’ group” (@sher2011, p. 322). Indeed, @xiang2022 describe their four-group solution as follows: “Of all participants, 536 (10.80%) were classified as increasing, 269 (5.42%) as persistently high, 433 (8.73%) as decreasing, and 3724 (75.05%) as persistently low” (@xiang2022, p. 162). Although @sher2011 cautioned that groups from these trajectory-based approaches should not be over-reified, this practice also remains common (e.g., @hawes2016; @hawes2018). Thus, though person-centered approaches can, in theory, help researchers understand the factors that contribute to differences in developmental trajectories, researchers should more thoughtfully consider alternative specifications (see @littlefield2010, as an example) and be especially skeptical when default specifications identify these four prototypic groups.

Hybrid approaches, such as growth mixture modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters. This combination provides a more comprehensive understanding of longitudinal data by capturing both within- and between-person variability. However, hybrid models can be more complex, necessitating careful model specification, selection, and interpretation. Additionally, these methods may require larger sample sizes to ensure the stability and accuracy of results.

In summary, SEM approaches offer powerful tools for longitudinal data analysis, enabling researchers to investigate complex relationships, individual differences, and change dynamics over time. The choice between variable-centered, person-centered, and hybrid approaches depends on the research objectives and the nature of the data. Despite their limitations, these models have greatly advanced our understanding of developmental processes and the factors that contribute to individual differences in change trajectories.

#### Advanced SEM

Advanced structural equation modeling (SEM) approaches, such as the RI-CLPM and LCM-SR models, have emerged to address more complex research questions and data structures in longitudinal analysis. These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.

The RI-CLPM enhances the traditional cross-lagged panel model by incorporating random intercepts, which allow for the separation of stable individual differences from the dynamic within-person associations between variables over time. Within-person variance in these models are captured by a series of latent variables that reflect time specific variance (i.e., the residual variance from the random intercept). These time-specific variables are referred to as structured residuals. Distinguishing between-person variance subsumed by the random intercept from the structured residuals is particularly valuable for understanding the time-specific effects of one variable on another, while accounting for the influence of individual differences. However, RI-CLPM may require larger sample sizes to ensure stability and accuracy of the estimates and can be computationally demanding. Using three waves of ABCD Study data, @kulisch2022 found a prospective association between psychopathology and childhood obesity as well as between childhood obesity and later eating behavior. The authors also showed that reciprocal associations were overestimated when stable, interindividual trait differences was not included in the model (i.e., via the random intercept).

LCM-SR, on the other hand, extends the RI-CLPM by including additional growth factors, such as a random linear slope. That is, the LCM-SR is a hybrid between a latent growth model and CLPM. This approach allows for a more comprehensive understanding of within-person change dynamics and factors influencing change over time. By including structured residuals, LCM-SR can capture additional time-specific relationships that are not explained by the latent growth factors. However, even more so than the RI-CLPM, LCM-SR comes with increased model complexity and requires careful specification and interpretation.

In conclusion, advanced SEM approaches for longitudinal data analysis provide valuable tools for addressing complex research questions and data structures. While they offer more nuanced insights into within-person change dynamics and the influence of individual differences, these models also come with certain limitations, such as the necessity of multiple assessments (e.g., four or more for LCM-SR), increased complexity, computational demands, and the need for careful model specification and interpretation. As with any statistical method, researchers should carefully consider their research objectives, data characteristics, and the assumptions of each model when selecting the most appropriate advanced SEM approach for longitudinal analysis. Given that these modeling approaches necessitate more waves of data, they are not yet commonly used with ABCD Study data. We anticipate that as more waves of ABCD data are publically released, these models can be used to address some of the pitfalls of the more traditional methods.

![LDA Table](./figures/Tabletemp.png)

# Discussion

\label{sec:headings}

As we enter the era of large-scale longitudinal investigations, it is essential to critically examine the various analytical methods that can be employed to glean insights from these rich datasets. The complex nature of longitudinal data demands sophisticated and well-suited methodologies to accurately address research questions and minimize biases. This paper aimed to provide an overview of diverse longitudinal analysis techniques, with a particular emphasis on their application to extensive longitudinal studies such as the ABCD Study. Beyond contributing to the ever-growing body of knowledge on longitudinal data analysis, we hope this manuscript also serves as a valuable resource for researchers seeking to optimize the use of large-scale longitudinal investigations in advancing our understanding of human development and behavior. In this discussion, we will focus on the key findings and recommendations of our review and discuss potential innovations that can further enhance the utility of these methods.

We began by addressing fundamental concepts and considerations in longitudinal research that are essential for generating accurate and meaningful insights into developmental processes. Concepts such as vulnerable periods, developmental disturbances and snares, or cascade and experience effects (among many others), are instrumental in shaping the design, analysis, and interpretation of longitudinal studies. Together, these concepts provide a framework for understanding the mechanisms underlying the course of development, while also accounting for the complex interplay between individual development and the influence of environmental factors. By considering the intricate relationships among these factors, researchers can better identify the critical time periods, situations, and contexts that contribute to individual differences in developmental outcomes. This awareness enables more precise inferences regarding the causal relationships between exposures and outcomes, ultimately leading to more robust and meaningful findings that can help facilitate the translation of research findings into practical applications in clinical and public health settings.

We also discussed some of the opportunities, challenges, and pitfalls that arise when working with longitudinal data. Key issues include selecting appropriate methods to account for the intricacies of longitudinal data, addressing missing data in a way that minimizes biases, and determining suitable longitudinal data structures that align with research questions and context. To address these challenges, researchers should carefully consider issues such as study design, selection of methods that account for both within- and between-person sources of variance, and employing modern techniques, (e.g., FIML, multiple imputation) for handling missing data. By adhering to best practices in longitudinal research and remaining vigilant of potential pitfalls, researchers can effectively harness the power of longitudinal data to maximize the potential of their investigations and gain valuable insights into complex developmental processes, individual differences, and the underlying mechanisms that drive change over time.

The final section, along with associated code and additional resources made available as online supplements (), aims to serve as a resource for researchers seeking to understand and implement various longitudinal panel models. By providing an overview of different approaches, their strengths and limitations, and key considerations for their use, we hope to facilitate the selection of appropriate models tailored to specific research questions and data structures. It is essential for researchers to consider their research objectives, the characteristics of their data, and the assumptions underlying each model when choosing the most suitable approach for longitudinal analysis. We encourage researchers to consult the cited literature and online supplements for further guidance in selecting and implementing longitudinal models when using the ABCD Study dataset. As the field continues to advance, we anticipate the emergence of new methods and refinements to existing approaches, further expanding the toolkit available to researchers for the analysis of longitudinal data. By staying informed about developments in this area and critically evaluating the appropriateness of different models for their research questions, researchers can ensure that their longitudinal analyses are both rigorous and informative. Notably, in this vast and continually evolving field, with numerous models and approaches available to address a wide range of research questions, no single model is universally applicable or without limitations. The diversity of methods ensures that researchers can find an appropriate tool for their specific needs. By familiarizing themselves with the various types of longitudinal models, researchers can more effectively navigate the complexities of longitudinal data and contribute valuable insights into the developmental processes and individual differences that shape human experience.

{{< pagebreak >}}

# Part V: References

### 2024-03-02 posts/1b_Examples_DifferenceScores_SimpleRegression.qmd
img uploads

---
title: "Difference Scores: Simple Regression"
author: "Biostatistics Working Group"

reader-mode: true
---

# Simple Regression {.unnumbered}

## Overview

<p>

This example assesses whether growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1) differs significantly based on handedness, using height ("anthroheightcalc") as a representative metric of growth. The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 height measurements for each participant; 2) a simple regression analysis is used to test whether sex (boy, girl) predicts the average difference value in participants height from baseline to the 1-Year follow-up. Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between difference scores and sex. The ensuing analysis and interpretations are detailed in the subsequent sections.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# rm(list = ls())

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page

```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

:::
:::

## Results

::: panel-tabset
### Build Model: Step 1 {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to compute a difference score by subtracting each participant’s height at T1 from their height at T0. Relevant summary statistics are also provided.

**STEP 1: Compute Difference Score**
```{r build-model-regression-1}
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
#| class.source: 'fold-hide'

# Define the function to compute difference scores for a given variable and provide a summary
# Function to compute difference scores for a given variable and provide a summary
compute_difference_and_summary <- function(df, variable_name) {
    # Define the event names of interest
    baseline_event <- "baseline_year_1_arm_1"
    followup_event <- "1_year_follow_up_y_arm_1"

    # Compute the difference between Baseline and Year 1 data for the given variable
    diff_data <- df %>%
        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
        spread(eventname, variable_name) %>% # Convert data from long to wide format
        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
        drop_na(diff) # Exclude rows with NA in the computed difference

    # Summarize the computed difference scores
    diff_summary <- summary(diff_data$diff)

    # Return the difference data and its summary
    list(data = diff_data, summary = diff_summary)
}

# List of variables for which difference scores are to be computed
variables_of_interest <- c("anthroheightcalc")

# Compute and store difference scores and summaries for each variable in a list
difference_and_summary_list <- lapply(variables_of_interest, function(var) {
    compute_difference_and_summary(df, var)
})

# Extract the difference data for the 'anthroheightcalc' variable
height_diff_data <- difference_and_summary_list[[1]]$data

# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
df <- left_join(df, height_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")

```

**Descriptive Statistics for the Difference Score**
```{r model-checking}
# Compute statistical summaries for the difference score variable
lapply(difference_and_summary_list, function(item) {
    print(item$summary)
})
```

This summary of the difference score variable indicates that the differences range from a decrease of `r min(height_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(height_diff_data$diff, na.rm = TRUE)` units. The median difference is `r median(height_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(height_diff_data$diff, digits = 2, na.rm = TRUE)` units. There are `r sum(is.na(height_diff_data$diff))` missing values in this difference score dataset.

**Descriptive Statistics for the Height Variable**
```{r}
## Summary statistics
# Compute summary statistics for Height by eventname
summary <- df %>%
    group_by(eventname) %>%
    get_summary_stats(anthroheightcalc, type = "mean_sd")

data.frame(summary)
```

The summary statistics provide insights into the Height variable across different `eventname` categories. At baseline, the average height is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`. Over the years, there’s a noticeable increase in average height: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year. The standard deviation remains relatively consistent over the years, suggesting similar variability in height measurements across these time points.

:::

### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}

::: blue
A simple regression analyses is conducted to examine whether a grouping variable (sex) significantly predicts the difference score value (indicating significant group differences in the average difference score).

**STEP 2: Conduct regression on Difference Score**
```{r build-model-regression-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
merged_data <- height_diff_data %>%
  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")

# Ensure 'sex' is a factor
merged_data$sex <- as.factor(merged_data$sex)

# Run the regression
model <- lm(diff ~ sex, data = merged_data)

# Get the summary of the regression model
model_summary <- summary(model)

model_summary

```

This regression analysis evaluates whether sex (boy, girl) predicts the average difference in participants' height from baseline to the 1-Year follow-up. The output from our model provides:

i. an F-statistic of `r model_summary$fstatistic[1]`; 
ii. degrees of freedom of `r model_summary$df[2]`; 
iii. a parameter estimate of `r model_summary$coefficients["sexMale or other", "Estimate"]`; 
iv. standard error of `r model_summary$coefficients["sexMale or other", "Std. Error"]`; 
v. p-value of `r model_summary$coefficients["sexMale or other", "Pr(>|t|)"]`.

Compared to boys (the reference group), girls have an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units. This effect was marginally significant with a p-value of `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`. Overall, the model explained a very small portion of the variance in height difference, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 6)`, and the overall model significance was not statistically significant with a p-value of `r pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)`.

:::

### Model Plots

::: blue
```{r violin-plots}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Visualize the difference scores across different levels of sex

# Create a violin plot to show the distribution of difference scores by sex
# Jittered points are added to provide a more granular view of individual observations
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
merged_data <- height_diff_data %>%
  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")

# Plotting using ggplot2
ggplot(merged_data, aes(x = sex, y = diff, fill = sex)) +
  geom_violin() +
  geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") + 
  labs(
    title = "Difference Scores by Sex",
    x = "Sex",
    y = "Difference Score"
  ) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

```

The violin plot visualizes the distribution of difference scores in children's heights by their sex: Right, Left, and Mixed. Each violin shape provides insight into the density of the data at different height differences, with wider sections representing higher densities of data points. Superimposed jittered points offer a granular view of individual observations. From the plot, it appears that the distributions of height differences across the three handedness categories are somewhat similar, though there might be subtle variations in median and spread.

:::
:::

## Wrapping Up

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue
Differences in children's heights between baseline and a subsequent 1-year follow-up, we sought to understand the potential influence of sex on this difference. Descriptive statistics revealed that height differences ranged from a decrease of `r min(merged_data$diff, na.rm = TRUE)` units to an increase of `r max(merged_data$diff, na.rm = TRUE)` units, with a median difference of `r median(merged_data$diff, na.rm = TRUE)` units and an average difference of approximately `r mean(merged_data$diff, na.rm = TRUE)` units. Subsequent regression analysis indicated that, when compared to boys, girls had an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units, though this effect was marginally significant (p = `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`). Complementary to these findings, violin plots visually underscored the subtle variations in height differences across sex categories, suggesting broadly similar distributions but with nuanced variations in central tendency and spread. In conclusion, while sex exhibited a potential influence on height difference over the year, the observed effect was notably minor.

:::
:::

### 2024-03-02 posts/3b_Examples_LinearMixedModels.qmd
img uploads

---
title: "Linear Mixed Models: Random Intercept and Slope"
author: "Biostatistics Working Group"

---

# LMM: Intercept & Slope {.unnumbered}

## Overview

<p>

The Linear Mixed Model with a random intercept and slope (LMM:ris) extends the traditional fixed-effect linear regression by incorporating both a subject-specific random intercept and a random slope. This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.

In this example, we will utilize the LMM:ris to analyze height trajectories obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand the stability and evolution in height measurements, while factoring in the clustered nature of observations within individuals over time. The LMM:ris facilitates this by concurrently modeling an overarching sample mean trajectory (fixed effect) and individual variations (random effects) in both starting points (intercepts) and growth rates (slopes) around this mean trajectory.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT","lme4","report","broom","gridExtra","easystats","gtsummary")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(lme4)         # Linear mixed-effects models
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page
library(easystats)    #
library(gtsummary)    #
```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

:::
:::

## Results 
<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue

The code fits a linear mixed model to examine the 'Height' variable across time points ('eventname'). It incorporates both random intercepts and slopes for the time points ('eventname') within each participant ('src_subject_id') to capture individual-level variability in both starting values and rates of change over time. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 1: Compute LMM with Random Intercepts and Slopes**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

## Linear Mixed Model with a random intercept and random slope (LMM-ris)
random_intercepts_slopes <- lmer(anthroheightcalc ~ 1 + eventname + (1|src_subject_id), data = df, REML=T)

print(random_intercepts_slopes)

```

```{r model-output}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Output and reports extending from the LMM-ris analyses
summary(random_intercepts_slopes)
confint(random_intercepts_slopes, level = 0.95, method = "Wald")
report_performance(random_intercepts_slopes)

```

The code fits a Linear Mixed Model (LMM:ris) to predict children's height based on different time points (denoted as eventname). Individual variability is captured through random intercepts associated with each participant (src_subject_id). The results suggest a significant change in height across the time points, with the model accounting for 89% (r round(conditional_R2, 2)) of the total variation and 37% (r round(marginal_R2, 2)) attributed to the fixed effects alone.

In the model, the linear term for eventname has an estimated effect of r round(model_summary$coefficients["eventname.L", "Estimate"], 2), indicating a significant increase in height across time points. Additionally, there's variability in initial height across participants, reflected by a random intercept standard deviation of r round(random_effects["src_subject_id", "Std.Dev."], 2).

### Model Plots

The following set of plots are used to facilitate model diagnostics. The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects. The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity. The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.

::: blue
```{r diagnosticplots}

# Assuming your model is named `random_intercepts_slope`
# 1. Extract the random effects
random_effects <- ranef(random_intercepts_slopes)[[1]]

# 2. Convert to dataframe
random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)

# Plot 1: Histogram
# Plot 1: Histogram
hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
  theme_minimal()

print(hist_plot)

```

```{r trajectoryplot}

# Extract the data frame used in the model
model_data <- random_intercepts_slopes@frame

# Extract unique subject IDs from the model's data
original_subject_ids <- unique(model_data$src_subject_id)

# Subset the original data to include only those subjects
df_subset <- df %>% filter(src_subject_id %in% original_subject_ids)

eventname_map <- c(
  "baseline_year_1_arm_1" = "Baseline",
  "1_year_follow_up_y_arm_1" = "Year_1",
  "2_year_follow_up_y_arm_1" = "Year_2",
  "3_year_follow_up_y_arm_1" = "Year_3",
  "4_year_follow_up_y_arm_1" = "Year_4"
)

# Plot
ggplot(df, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +

# Individual estimated height trajectories in faded lines
geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +

# Overall group-mean trajectory in blue with increased thickness
stat_summary(aes(group = 1), fun = mean, geom = "line", color = "blue", linewidth = 1) +
labs(title = "Individual and Group-Mean Height Trajectories",
     x = "Event Name",
     y = "Height") +
  theme_minimal()

```

The provided code visualizes individual and group-mean height trajectories over different event names. Individual height trajectories for each subject are depicted as faded gray lines, allowing for a clear view of the variability among subjects. In contrast, the overall group-mean trajectory, which represents the average trend across all individuals for each event name, is highlighted in blue. The average height at r mean(df_descriptable$Height) has increased by r mean(df_descriptable$Height_followup - df_descriptable$Height) units from baseline.

## Wrapping Up
<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4). The eventname predictor was statistically significant with a p-value of r format.pval(fixed_effects$Pr(>|t|)[2], digits = 3). The model's overall ability to explain the variance in height was xxxxx, with a conditional R^2 of r report_performance(random_intercepts)$Conditional R2[1], indicating that it accounted for this proportion of the variability in height when considering both fixed and random effects. The marginal R^2 was r report_performance(random_intercepts)$Marginal R2[1], meaning that the fixed effects alone explained this proportion of the variability.

:::
:::

### 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
img uploads

---
title: "Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

# XXXXXXXXXXX {.unnumbered}

## Overview

<p>

XXXXXXXXXXX Models xxxxxxx. This xxxxxxx.

In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time. XXXX models facilitate this by xxxxx.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables

```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

# Define a function to compute descriptives
compute_descriptives <- function(data, event_name) {
  # For factor variables
  sex_desc <- paste0(table(data$sex), " (", round(100 * prop.table(table(data$sex)), 1), "%)")
  race_desc <- paste0(table(data$race.4level), " (", round(100 * prop.table(table(data$race.4level)), 1), "%)")

  # For numeric variables
  age_desc <- paste0(round(mean(data$age, na.rm = TRUE), 2), " (", round(sd(data$age, na.rm = TRUE), 2), ")")
  weight_desc <- paste0(round(mean(data$anthroweightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroweightcalc, na.rm = TRUE), 2), ")")
  height_desc <- paste0(round(mean(data$anthroheightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroheightcalc, na.rm = TRUE), 2), ")")

  # Combine into a data frame
  desc_df <- data.frame(
    Variable = c("Sex - Female", "Sex - Male or other", "Race - Asian", "Race - Black", "Race - Other/Mixed", "Race - White", "Age", "Weight", "Height"),
    Value = c(sex_desc, race_desc, age_desc, weight_desc, height_desc)
  )

  # Rename the Value column based on event name
  colnames(desc_df)[2] <- event_name
  return(desc_df)
}

# Compute descriptives for each event
baseline_desc <- compute_descriptives(subset(df, eventname == "baseline_year_1_arm_1"), "baseline_year_1_arm_1")
one_year_desc <- compute_descriptives(subset(df, eventname == "1_year_follow_up_y_arm_1"), "1_year_follow_up_y_arm_1")
two_year_desc <- compute_descriptives(subset(df, eventname == "2_year_follow_up_y_arm_1"), "2_year_follow_up_y_arm_1")
three_year_desc <- compute_descriptives(subset(df, eventname == "3_year_follow_up_y_arm_1"), "3_year_follow_up_y_arm_1")

# Join all data frames side-by-side
final_table <- baseline_desc %>%
  left_join(one_year_desc, by = "Variable") %>%
  left_join(two_year_desc, by = "Variable") %>%
  left_join(three_year_desc, by = "Variable")

# Adjust for the required format
final_table[1:6, 3:5] <- NA

# Round numeric values to two decimal places
numeric_cols <- sapply(final_table, is.numeric)
final_table[numeric_cols] <- lapply(final_table[numeric_cols], round, 2)

# Create heading rows with the same columns as final_table
heading_rows <- data.frame(
  Variable = c("Sex", "Race"),
  baseline_year_1_arm_1 = NA_real_,
  `1_year_follow_up_y_arm_1` = NA_real_,
  `2_year_follow_up_y_arm_1` = NA_real_,
  `3_year_follow_up_y_arm_1` = NA_real_
)

# Set column names of heading_rows to match final_table
colnames(heading_rows) <- colnames(final_table)

# Introduce group labels and adjust the "Variable" column
final_table <- rbind(
                    heading_rows[1,], 
                    final_table[1:2,],
                    heading_rows[2,], 
                    final_table[3:6,], 
                    final_table[7:9,]
                    )

# Update the Variable column to remove redundant factor variable name
final_table$Variable <- gsub("Sex - ", "", final_table$Variable)
final_table$Variable <- gsub("Race - ", "", final_table$Variable)
final_table$Variable[final_table$Variable == "Male or other"] <- "Male"

# Add non-breaking spaces for increased indentation
final_table$Variable[final_table$Variable %in% c("Female", "Male", "Asian", "Black", "Other/Mixed", "White")] <- 
  paste0(rep("&nbsp;", 6), final_table$Variable[final_table$Variable %in% c("Female", "Male", "Asian", "Black", "Other/Mixed", "White")])

# Update column headers
colnames(final_table)[2:5] <- c("Baseline", "Year 1", "Year 2", "Year 3")

# Display the table interactively without row names, with updated column headers, and with HTML entities rendered
datatable(final_table, 
          colnames = c("", "Baseline", "Year 1", "Year 2", "Year 3"),
          options = list(pageLength = nrow(final_table), autoWidth = TRUE),
          rownames = FALSE, escape = FALSE,
          caption = "Descriptives Table")  # Add table title

```

:::
:::

## Results 
<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue

------

**STEP 1: Convert dataframe from long-to-wide format**
```{r long-to-wide}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Split the data into static (unchanging) and changing variables
# Static variables are taken only from the baseline measurement occasion
df_static <- df %>%
  filter(eventname == "baseline_year_1_arm_1") %>%
  select(src_subject_id, site_id_l, sex)

df_changing <- df %>%
  select(src_subject_id, eventname, anthroweightcalc, anthroheightcalc)

# Convert the changing variables from long format to wide format
df_wide_changing <- df_changing %>%
  gather(variable, value, -src_subject_id, -eventname) %>%
  unite("variable_event", variable, eventname) %>%
  spread(variable_event, value)

# Merge the static and changing data frames based on the 'src_subject_id' variable
df_wide <- df_static %>%
  left_join(df_wide_changing, by = "src_subject_id")

# Rename columns to shorter names for easier reference
df_wide <- df_wide %>% 
        rename("anthroweightcalc_0"="anthroweightcalc_baseline_year_1_arm_1",
               "anthroweightcalc_1"="anthroweightcalc_1_year_follow_up_y_arm_1",
               "anthroweightcalc_2"="anthroweightcalc_2_year_follow_up_y_arm_1",
               "anthroweightcalc_3"="anthroweightcalc_3_year_follow_up_y_arm_1",
               "anthroweightcalc_4"="anthroweightcalc_4_year_follow_up_y_arm_1",
               "anthroheightcalc_0"="anthroheightcalc_baseline_year_1_arm_1",
               "anthroheightcalc_1"="anthroheightcalc_1_year_follow_up_y_arm_1",
               "anthroheightcalc_2"="anthroheightcalc_2_year_follow_up_y_arm_1",
               "anthroheightcalc_3"="anthroheightcalc_3_year_follow_up_y_arm_1",
               "anthroheightcalc_4"="anthroheightcalc_4_year_follow_up_y_arm_1"
               )

# Reorder columns in the dataframe
df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, anthroweightcalc_0, anthroweightcalc_1, anthroweightcalc_2, anthroweightcalc_3, anthroweightcalc_4, anthroheightcalc_0, anthroheightcalc_1, anthroheightcalc_2, anthroheightcalc_3, anthroheightcalc_4)

```

The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname'). It incorporates xxxxxx. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 2: Compute xxxxx**

```{r}

# Define the Latent Growth Curve Model
model <- '
  # Latent variables
    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2

  # Estimate means of latent variables
    i ~ 1
    s ~ 1

  # Variances
    i ~~ i
    s ~~ s

  # Covariance
    i ~~ s
'
```

```{r}
# Fit the model
fit <- sem(model, data = df_wide, missing = "FIML")

summary(fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

```

### 2024-03-02 posts/_12_Examples_LatentTransitionAnalysis.qmd
img uploads

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Latent Transition Analysis"
 author: "Biostatistics Working Group"
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT")
+packages_required <- c("tidyverse","rstatix","DT", "tidyLPA")
 
@@
 library(DT)           # Rendering interactive data tables
+library(tidyLPA)      # XXXXXXXX
 
@@
 
-------
+**STEP 1: XXXXXXXXX **
 
-**STEP 1: Convert dataframe from long-to-wide format**
-```{r long-to-wide}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# Split the data into static (unchanging) and changing variables
-# Static variables are taken only from the baseline measurement occasion
-df_static <- df %>%
-  filter(eventname == "baseline_year_1_arm_1") %>%
-  select(src_subject_id, site_id_l, sex)
-
-df_changing <- df %>%
-  select(src_subject_id, eventname, anthroweightcalc, anthroheightcalc)
-
-# Convert the changing variables from long format to wide format
-df_wide_changing <- df_changing %>%
-  gather(variable, value, -src_subject_id, -eventname) %>%
-  unite("variable_event", variable, eventname) %>%
-  spread(variable_event, value)
-
-# Merge the static and changing data frames based on the 'src_subject_id' variable
-df_wide <- df_static %>%
-  left_join(df_wide_changing, by = "src_subject_id")
-
-# Rename columns to shorter names for easier reference
-df_wide <- df_wide %>% 
-        rename("anthroweightcalc_0"="anthroweightcalc_baseline_year_1_arm_1",
-               "anthroweightcalc_1"="anthroweightcalc_1_year_follow_up_y_arm_1",
-               "anthroweightcalc_2"="anthroweightcalc_2_year_follow_up_y_arm_1",
-               "anthroweightcalc_3"="anthroweightcalc_3_year_follow_up_y_arm_1",
-               "anthroweightcalc_4"="anthroweightcalc_4_year_follow_up_y_arm_1",
-               "anthroheightcalc_0"="anthroheightcalc_baseline_year_1_arm_1",
-               "anthroheightcalc_1"="anthroheightcalc_1_year_follow_up_y_arm_1",
-               "anthroheightcalc_2"="anthroheightcalc_2_year_follow_up_y_arm_1",
-               "anthroheightcalc_3"="anthroheightcalc_3_year_follow_up_y_arm_1",
-               "anthroheightcalc_4"="anthroheightcalc_4_year_follow_up_y_arm_1"
-               )
-
-# Reorder columns in the dataframe
-df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, anthroweightcalc_0, anthroweightcalc_1, anthroweightcalc_2, anthroweightcalc_3, anthroweightcalc_4, anthroheightcalc_0, anthroheightcalc_1, anthroheightcalc_2, anthroheightcalc_3, anthroheightcalc_4)
-
-```
-
-The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname').
-It incorporates xxxxxx.
-The results of the model are then printed to provide a detailed summary of the fitted model parameters.
-
-**STEP 2: Compute xxxxx**
+-------
 
 ```{r}
-
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
-
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
-
-  # Variances
-    i ~~ i
-    s ~~ s
-
-  # Covariance
-    i ~~ s
-'
-```
-
-```{r}
-# Fit the model
-fit <- sem(model, data = df_wide, missing = "FIML")
+# Estimate the LTA model
+fit <- tidyLPA::estimate_profiles(data, 2, variances = "equal", covariances = "zero")
 
@@
 
+plot(fit)
+
 ```

### 2024-03-02 posts/_14_Examples_StateTraitModels.qmd
img uploads

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "State-Trait Models"
 author: "Biostatistics Working Group"
@@
 ```{r}
+# Define the model
+model <- '
+  # Trait component
+  trait =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3
 
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
+  # State component
+  state =~ anthroheightcalc_0 + anthroheightcalc_1 + anthroheightcalc_2 + anthroheightcalc_3
 
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
-
-  # Variances
-    i ~~ i
-    s ~~ s
-
-  # Covariance
-    i ~~ s
+  # Residual variances
+  anthroheightcalc_0 ~~ anthroheightcalc_0
+  anthroheightcalc_1 ~~ anthroheightcalc_1
+  anthroheightcalc_2 ~~ anthroheightcalc_2
+  anthroheightcalc_3 ~~ anthroheightcalc_3
 '
@@
 ```{r}
-# Fit the model
+# Estimate the State-Trait Model
 fit <- sem(model, data = df_wide, missing = "FIML")

### 2024-03-02 posts/_4_Examples_SignedRankTest.qmd
img uploads

---
title: "Signed-Rank Test"
author: "Biostatistics Working Group"
---

# Signed-Rank Test {.unnumbered}

####
It is actually quite common that numeric depend variables need to be transformed or converted into ranks, i.e. ordinal variables, because the distribution of residuals does not allow the application of parametric tests such as t-tests or linear regression. In such cases, as we are dealing with rank (ordinal) data, the application of a chi-square test is unwarranted and we need to use another test. There are different alternatives depending on whether the data are paired (coming from the same individuals) or if all observations are independent.

The non-parametric alternative for independent t-tests, i.e. for data where we are dealing with two separate groups and a numeric dependent variable that violates parametric assumptions (or an ordinal dependent variable), is the Mann-Whitney U-test. In contrast, if the groups under investigation represent identical participants that are tested under two conditions, the appropriate alternative is a Wilcoxon Signed Rank test (which is thus the alternative for paired t-test).

Imagine we wanted to determine if two language families differed with respect to the size of their phoneme inventories. You have already ranked the inventory sizes and would now like to now if language family correlates with inventory size. As such, we are dealing with two independent groups and we want to implement a non-parametric alternative of a t-test. To answer this question, you create the table shown below.
####

## Overview

<p>

The Signed-Rank Test xxxxxxx. This xxxxxxx.

In this example, we will utilize the signed-rank test to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time. The signed-rank test facilitates this by xxxxx.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT", "gtsummary","easystats")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(easystats)    #
library(gtsummary)    #
```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "anthroweightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    anthroweightcalc = as.numeric(anthroweightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

#####
#Creating new ordinal variable for the example
df$education <- factor(
  with(df, ifelse(high.educ.bl %in% c("1st grade", "2nd grade", "3rd grade", "4th grade",
                                      "5th grade",
                                      "6th grade", "7th grade", "8th grade", "9th grade",
                                      "10th grade",
                                      "11th grade", "12th grade"), "lt_highschool",
                  ifelse(high.educ.bl == "High school graduate", "highschool",
                  ifelse(high.educ.bl == "GED or equivalent Diploma", "GED",
                  ifelse(high.educ.bl == "Some college", "some_college",
                  ifelse(high.educ.bl %in% c("Associate degree: Academic Program", 
                                              "Associate degree: Occupational"),
                         "associate_degree",
                  ifelse(high.educ.bl == "Bachelor's degree (ex. BA", "Bachelor's_degree",
                  ifelse(high.educ.bl == "Master's degree (ex. MA", "Master's_degree",
                  ifelse(high.educ.bl %in% c("Doctoral degree (ex. PhD", "Professional
                                             School degree (ex. MD"), 
                         "Advanced_degree", NA))))))))))
  levels = c("lt_highschool", "highschool", "GED", "some_college", "associate_degree",
             "Bachelor's_degree", "Master's_degree", "Advanced_degree")
  ordered = TRUE

####

####
df$income <- factor(
  with(df, ifelse(household.income.bl == "Less than $5,000", "<$5,000",
                  ifelse(household.income.bl == "$5,000 through $11,999", "$5,000-$11,999",
                  ifelse(household.income.bl == "$12,000 through $15,999", "$12,000-$15,999",
                  ifelse(household.income.bl == "$16,000 through $24,999", "$16,000-$24,999",
                  ifelse(household.income.bl == "$25,000 through $34,999", "$25,000-$34,999",
                  ifelse(household.income.bl == "$35,000 through $49,999", "$35,000-$49,999",
                  ifelse(household.income.bl == "$50,000 through $74,999", "$50,000-$74,999",
                  ifelse(household.income.bl == "$75,000 through $99,999", "$75,000-$99,999",
                  ifelse(household.income.bl == "$100,000 through $199,999", "$100,000-$199,999",
                  ifelse(household.income.bl == "$200,000 and greater", "≥$200,000",
                  NA))))))))))),
  levels = c("<$5,000", "$5,000-$11,999", "$12,000-$15,999", "$16,000-$24,999",
             "$25,000-$34,999", "$35,000-$49,999", "$50,000-$74,999",
             "$75,000-$99,999", "$100,000-$199,999", "≥$200,000"),
  ordered = TRUE
)

####

# Create a mapping from factor levels to numeric values
income_levels <- c("Less than $5,000", "$5,000 through $11,999", "$12,000 through $15,999", "$16,000 through $24,999", "$25,000 through $34,999", "$35,000 through $49,999", "$50,000 through $74,999", "$75,000 through $99,999", "$100,000 through $199,999", "$200,000 and greater", "Don't know", "Refuse to answer", "No deseo responder")

# We assign numbers 1 through 10, but we have 12 levels, including "Don't know", "Refuse to answer", and "No deseo responder"
# Let's assume "Don't know" and any refusal to answer will be treated as NA
income_numeric_values <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, NA, NA, NA)

# Create a named vector for easy lookup
income_values <- setNames(income_numeric_values, income_levels)

# Convert the `income` factor to numeric using the named vector
df$income_numeric <- income_values[as.character(df$household.income.bl)]

# Handling NAs and refused to answer
df$income_numeric[is.na(df$household.income.bl) | df$household.income.bl %in% c("Don't know", "Refuse to answer", "No deseo responder")] <- NA

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, income_numeric, education) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", income_numeric ~ "Income", 
                 education ~ "Education"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

:::
:::

## Results 
<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue

The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname'). It incorporates xxxxxx. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 1: Convert dataframe from long-to-wide format**
```{r long-to-wide}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Split the data into static (unchanging) and changing variables
# Static variables are taken only from the baseline measurement occasion
df_static <- df %>%
  filter(eventname == "baseline_year_1_arm_1") %>%
  select(src_subject_id, site_id_l, sex)

df_changing <- df %>%
  select(src_subject_id, eventname, income_numeric, education)

# Convert the changing variables from long format to wide format
df_wide_changing <- df_changing %>%
  gather(variable, value, -src_subject_id, -eventname) %>%
  unite("variable_event", variable, eventname) %>%
  spread(variable_event, value)

# Merge the static and changing data frames based on the 'src_subject_id' variable
df_wide <- df_static %>%
  left_join(df_wide_changing, by = "src_subject_id")

# Rename columns to shorter names for easier reference
df_wide <- df_wide %>% 
        rename("income_numeric_0"="income_numeric_baseline_year_1_arm_1",
               "income_numeric_1"="income_numeric_1_year_follow_up_y_arm_1",
               "income_numeric_2"="income_numeric_2_year_follow_up_y_arm_1",
               "income_numeric_3"="income_numeric_3_year_follow_up_y_arm_1",
               "income_numeric_4"="income_numeric_4_year_follow_up_y_arm_1"
               )

# Reorder columns in the dataframe
df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, income_numeric_0, income_numeric_1, income_numeric_2, income_numeric_3, income_numeric_4)

```

**Descriptive Statistics for the Income Variable**
```{r}
## Summary statistics
# Compute summary statistics for Income by eventname
summary <- df %>%
    group_by(eventname) %>%
    get_summary_stats(income_numeric, type = "mean_sd")

data.frame(summary)
```

The summary statistics provide insights into the Weight variable across different `eventname` categories. At baseline, the average weight is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`. Over the years, there’s a noticeable increase in average weight: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year. The standard deviation remains relatively consistent over the years, suggesting similar variability in weight measurements across these time points.

**STEP 2: Compute Signed Rank Test**
```{r}

#########
df_wide$income_numeric_0 <- as.numeric(as.character(df_wide$income_numeric_0))
df_wide$income_numeric_1 <- as.numeric(as.character(df_wide$income_numeric_1))
df_wide$income_numeric_4 <- as.numeric(as.character(df_wide$income_numeric_4))
df_wide_complete <- na.omit(df_wide[, c("income_numeric_0", "income_numeric_4")])

model <- wilcox.test(df_wide_complete$income_numeric_0, df_wide_complete$income_numeric_4, paired = TRUE)

report(wilcox.test(df_wide_complete$income_numeric_0, df_wide_complete$income_numeric_4))

```

**Descriptive Statistics for the Signed Rank Test result**
```{r model-output}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

library(gridExtra)

# Creating the first density plot with ggplot2
p1 <- ggplot(data.frame(x = na.omit(df_wide$income_numeric_0)), aes(x = x)) +
  geom_density(fill="red", color="blue") +
  labs(title = "Kernel Density of Income at Baseline")

# Creating the second density plot with ggplot2
p2 <- ggplot(data.frame(x = na.omit(df_wide$income_numeric_1)), aes(x = x)) +
  geom_density(fill="blue", color="red") +
  labs(title = "Kernel Density of Income at T1")

p3<- ggplot(df_wide, aes(x = income_numeric_0, y = income_numeric_1)) + geom_point(color="#04c24a")

# Create both plot and store in variable
p4<-ggplot(df_wide, aes(x=income_numeric_0, y=income_numeric_1))+
geom_segment( aes(x=income_numeric_0, xend=income_numeric_1, y=0, yend=income_numeric_2), color="grey") + geom_point( color="green", size=4)

p5<-ggplot(df_wide, aes(x = income_numeric_0)) + 
geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)

p6<-ggplot(df_wide, aes(x=income_numeric_0, y=income_numeric_1)) + 
geom_bar(fill="#69b3a2", stat = "identity")

# Arrange the ggplot2 objects on a grid
grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)

ggplot(df_wide, aes(x = sex, y = income_numeric_0, fill = sex)) +
  geom_boxplot() +
  scale_fill_manual(values = c("orange", "darkgrey")) +
  theme_bw() +
  theme(legend.position = "none")
```

This signed rank test evaluates xxxx. The model output provides:

i. xxxxx;
ii. xxxxx;
iii. xxxxx;
iv. xxxxx;
v. xxxxx.

With a xxxx of `xxxxx` and a p-value of `xxxxx`, we can xxxx. Furthermore, we can xxx that xxxxx. This indicates xxxxxx in weight from Baseline to Year 1.

### 2024-03-02 posts/_6_Examples_GeneralizedEstimatingEquations.qmd
img uploads

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Generalized Estimating Equations"
 author: "Biostatistics Working Group"
@@
 
-# XXXXXXXXXXX {.unnumbered}
+# Generalized Estimating Equations {.unnumbered}
 
@@
 
-XXXXXXXXXXX Models xxxxxxx. This xxxxxxx.
-
-In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
-Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time.
-XXXX models facilitate this by xxxxx.
+This example assesses xxxxxxxx.
 
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT")
+packages_required <- c("tidyverse","rstatix","DT", "geepack")
 
@@
 library(DT)           # Rendering interactive data tables
-
+library(geepack)      # Generalized Estimating Equations (GEE) analysis
 ```
@@
 
-------
-
-**STEP 1: Convert dataframe from long-to-wide format**
-```{r long-to-wide}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# Split the data into static (unchanging) and changing variables
-# Static variables are taken only from the baseline measurement occasion
-df_static <- df %>%
-  filter(eventname == "baseline_year_1_arm_1") %>%
-  select(src_subject_id, site_id_l, sex)
-
-df_changing <- df %>%
-  select(src_subject_id, eventname, anthroweightcalc, anthroheightcalc)
-
-# Convert the changing variables from long format to wide format
-df_wide_changing <- df_changing %>%
-  gather(variable, value, -src_subject_id, -eventname) %>%
-  unite("variable_event", variable, eventname) %>%
-  spread(variable_event, value)
-
-# Merge the static and changing data frames based on the 'src_subject_id' variable
-df_wide <- df_static %>%
-  left_join(df_wide_changing, by = "src_subject_id")
-
-# Rename columns to shorter names for easier reference
-df_wide <- df_wide %>% 
-        rename("anthroweightcalc_0"="anthroweightcalc_baseline_year_1_arm_1",
-               "anthroweightcalc_1"="anthroweightcalc_1_year_follow_up_y_arm_1",
-               "anthroweightcalc_2"="anthroweightcalc_2_year_follow_up_y_arm_1",
-               "anthroweightcalc_3"="anthroweightcalc_3_year_follow_up_y_arm_1",
-               "anthroweightcalc_4"="anthroweightcalc_4_year_follow_up_y_arm_1",
-               "anthroheightcalc_0"="anthroheightcalc_baseline_year_1_arm_1",
-               "anthroheightcalc_1"="anthroheightcalc_1_year_follow_up_y_arm_1",
-               "anthroheightcalc_2"="anthroheightcalc_2_year_follow_up_y_arm_1",
-               "anthroheightcalc_3"="anthroheightcalc_3_year_follow_up_y_arm_1",
-               "anthroheightcalc_4"="anthroheightcalc_4_year_follow_up_y_arm_1"
-               )
-
-# Reorder columns in the dataframe
-df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, anthroweightcalc_0, anthroweightcalc_1, anthroweightcalc_2, anthroweightcalc_3, anthroweightcalc_4, anthroheightcalc_0, anthroheightcalc_1, anthroheightcalc_2, anthroheightcalc_3, anthroheightcalc_4)
-
-```
-
 The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname').
@@
 
-**STEP 2: Compute xxxxx**
-
+**STEP 1: Compute xxxxx**
 ```{r}
-
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
-
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
-
-  # Variances
-    i ~~ i
-    s ~~ s
-
-  # Covariance
-    i ~~ s
-'
+model <- geeglm(anthroweightcalc ~ eventname, data = df, id = src_subject_id, family = gaussian(), corstr = "exchangeable")
 ```
@@
 ```{r}
-# Fit the model
-fit <- sem(model, data = df_wide, missing = "FIML")
-
-summary(fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)
-
+summary(model)
 ```

### 2024-03-02 posts/_7_Examples_GeneralizedLinearMixedEffectsModels.qmd
img uploads

--- 2024-03-02 posts/_6_Examples_GeneralizedEstimatingEquations.qmd
@@
 ---
-title: "Generalized Estimating Equations"
+title: "Generalized Linear Mixed Effects Models"
 author: "Biostatistics Working Group"
@@
 
-# Generalized Estimating Equations {.unnumbered}
+# Generalized Linear Mixed Effects Models {.unnumbered}
 
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT", "geepack")
+packages_required <- c("tidyverse","rstatix","DT", "lme4")
 
@@
 library(DT)           # Rendering interactive data tables
-library(geepack)      # Generalized Estimating Equations (GEE) analysis
+library(lme4)         # Linear mixed-effects models
 ```
@@
 # Subset the nonimaging data to include desired variables
-selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
+selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "tlfb_alc_sip_l")
 subset_data <- data_nonimaging[, selected_vars]
@@
     anthroweightcalc = as.numeric(anthroweightcalc),
-    anthroheightcalc = as.numeric(anthroheightcalc)
+    tlfb_alc_sip_l = as.factor(tlfb_alc_sip_l)
   ) %>%
@@
   race_desc <- paste0(table(data$race.4level), " (", round(100 * prop.table(table(data$race.4level)), 1), "%)")
+  tlfb_alc_sip_l_desc <- paste0(table(data$tlfb_alc_sip_l), " (", round(100 * prop.table(table(data$tlfb_alc_sip_l)), 1), "%)")
 
@@
   weight_desc <- paste0(round(mean(data$anthroweightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroweightcalc, na.rm = TRUE), 2), ")")
-  height_desc <- paste0(round(mean(data$anthroheightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroheightcalc, na.rm = TRUE), 2), ")")
 
@@
   desc_df <- data.frame(
-    Variable = c("Sex - Female", "Sex - Male or other", "Race - Asian", "Race - Black", "Race - Other/Mixed", "Race - White", "Age", "Weight", "Height"),
-    Value = c(sex_desc, race_desc, age_desc, weight_desc, height_desc)
+    Variable = c("Sex - Female", "Sex - Male or other", "Race - Asian", "Race - Black", "Race - Other/Mixed", "Race - White", "tlfb_alc_sip_l - No", "tlfb_alc_sip_l - Yes", "Age", "Weight"),
+    Value = c(sex_desc, tlfb_alc_sip_l_desc, race_desc, age_desc, weight_desc)
   )
@@
 
+:::
+:::
+
 ## Results 
@@
 ```{r}
-model <- geeglm(anthroweightcalc ~ eventname, data = df, id = src_subject_id, family = gaussian(), corstr = "exchangeable")
+model <- glmer(tlfb_alc_sip_l ~ eventname + (1 | src_subject_id), data = df, family = binomial(link = "logit"))
 ```

### 2024-03-02 posts/ABCD_WorkedExample_CBCLADHD.R

@@ -1,0 +1,175 @@
+
+#ABCD worked example when the outcome variable is highly skewed.
+#CBCL ADHD summary score was the outcome of interest. This variable is both 
+#overdispersed and zero-inflated. 
+
+#overdispersment and zero-inflation can be verified using packages like the 
+#DHARMa package in R.
+
+#glmmTMB was used to model CBCL ADHD. Alternative packages that can also be 
+#used include the brms and glmmAdapative package.
+
+#examples include a fixed effect for interview age and sex (reference = female). 
+#interview age was scaled to improve model run time. 
+
+#a random intercept was included for the family structure. A random intercept
+#was not included for study site after review of the intraclass correlation 
+#coefficient. 
+
+#a random slope was included for the study visit. A fixed effect was also 
+#included for study visit. 
+
+#zero inflated models included the same covariates.
+
+#parallel processing was used to speed up the model run time. 
+
+#comparison of different distributions is not included here.
+
+#dataset annual_ABCD includes only participants that completed all five
+#annual visits included in this example.
+
+library(glmmTMB)
+
+#Generalized Poisson Model
+model_GenPois <- glmmTMB(CBCL_ADHD ~ 1 +  interview_age_scaled+ sex+ 
+                                 factor(visit)+(1+visit| rel_family_id.bl), 
+                               family=genpois(link="log"), data=annual_ABCD, 
+                               REML=FALSE, control = glmmTMBControl(parallel = 16))
+summary(model_GenPois)
+
+# AIC      BIC   logLik deviance df.resid 
+# 79022.9  79111.3 -39500.5  79000.9    22686 
+# 
+# Random effects:
+#   
+# Conditional model:
+# Groups           Name        Variance Std.Dev. Corr 
+# rel_family_id.bl (Intercept) 1.131451 1.06370       
+# visit       0.003188 0.05646  0.52 
+# Number of obs: 22697, groups:  rel_family_id.bl, 3839
+# 
+# Dispersion parameter for genpois family (): 1.35 
+# 
+# Conditional model:
+#   Estimate Std. Error z value Pr(>|z|)    
+# (Intercept)           0.17217    0.04665   3.691 0.000223 ***
[…]

### 2024-03-02 posts/ABCD_WorkedExample_CBCLADHD[54].R

@@ -1,0 +1,219 @@
+library(dplyr)
+
+#below requires the non_imaging_excluding_nt_5.0 dataframe
+annual_ABCD <-  non_imaging_excluding_nt_5.0 %>%
+  filter(eventname == "baseline_year_1_arm_1" | eventname == "1_year_follow_up_y_arm_1"
+         | eventname == "2_year_follow_up_y_arm_1" | eventname == "3_year_follow_up_y_arm_1" 
+         |eventname == "4_year_follow_up_y_arm_1")
+####
+#below requires the demo5.0 dataframe to extract the sex variable
+xy1 <- demo5.0[,c("src_subject_id", "eventname", "sex")]
+annual_ABCD <-  merge(annual_ABCD, xy1,
+                       all.x =TRUE, by=c("eventname", "src_subject_id"))
+
+#################################
+
+#Manual creation of the CBCL ADHD variable. May need to convert these to numeric first.These
+#were originally factor variables that I manually mutated using dplyr.
+
+annual_ABCD$CBCL_ADHD <- annual_ABCD$cbcl_q04_p_num1 + annual_ABCD$cbcl_q08_p_num1 + annual_ABCD$cbcl_q10_p_num1 + annual_ABCD$cbcl_q41_p_num1 +
+  annual_ABCD$cbcl_q78_p_num1 + annual_ABCD$cbcl_q93_p_num1 + annual_ABCD$cbcl_q104_p_num1
+
+
+#########
+#making a numeric study visit variable
+
+annual_ABCD<- annual_ABCD %>% 
+  mutate(visit = case_when(
+    eventname == "baseline_year_1_arm_1" ~ 1,
+    eventname == "1_year_follow_up_y_arm_1" ~ 2,
+    eventname == "2_year_follow_up_y_arm_1" ~3,
+    eventname == "3_year_follow_up_y_arm_1" ~ 4,
+    eventname == "4_year_follow_up_y_arm_1" ~ 5))
+
+####################
+
+###removing any participants with fewer than 5 CBCL ADHD datapoints
+annual_ABCD <- annual_ABCD %>% 
+  group_by(src_subject_id) %>% 
+  filter(n() >= 5)
+
+#the resulting dataframe should have 22810 observations (4562 participants)
+#####################
+#####################
+
+#ABCD worked example when the outcome variable is highly skewed.
+#CBCL ADHD summary score was the outcome of interest. This variable is both 
+#overdispersed and zero-inflated. 
+
[…]

### 2024-03-04 manuscript.qmd
testing render

---
title: "midFEMA Manuscript"
authors:
  - name: Samuel W. Hawes
    email: "shawes@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Mohammadreza Bayat
    email: "mbayat@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Anthony Dick
    email: "adick@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Raul Gonzalez
    email: "gonzara@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Additional Co-authors
    email: "tbd@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States

abstract: |
  The midFEMA research project is aimed at identifying and analyzing neurodevelopmental biomarkers for severe behavioral and emotional problems in youth. By leveraging Fast and Efficient Mixed‐Effects Analysis (FEMA) in conjunction with voxelwise task-based reward activation measurements, we aim to uncover insights into brain development and functioning across a large, prospectively followed sample of approximately 9,000 youth aged 9–16 years from the Adolescent Brain and Cognitive Development (ABCD) study.
index-terms: 
  - Longitudinal
  - fMRI
  - DBD
  - FEMA
  - MID
format:
  ieee-typst: default
bibliography: references.bib
---

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(lorem)
```

```{r}
library(tidyverse)
library(ggplot2)

df <- mtcars
str(df)
```

```{r}
# Set the URL for the CSV file
file_url <- "https://drive.google.com/uc?export=download&id=1QFdAFz09DQEeFVOj8bRWh4g6hK4b4Qj1"

# Set the destination filename
dest_file <- "df_preclean.csv"

# Download the file
download.file(file_url, destfile = dest_file, mode = "wb")

# Read the CSV file into R
my_data <- read.csv(dest_file)

# Display the first few rows of the data
head(my_data)

```

## Introduction
_Introduce the research topic, state the research problem, and outline the study's objectives. Provide background information to set the context. This is an exmaple in-text citation_ @netwok2020.
`r lorem::ipsum(paragraphs = 3)`

### Background
_Provide a brief review of the pertinent literature, highlighting the gaps your study aims to fill._
`r lorem::ipsum(paragraphs = 2)`

### Objectives
_State the specific research questions or hypotheses your study addresses._
`r lorem::ipsum(paragraphs = 2)`

## Methods
_Describe the methodological approach, detailing how the study was conducted, how data was collected, and how analyses were performed._
`r lorem::ipsum(paragraphs = 1)`

### Participants
_Describe the demographic characteristics of the study participants, including how they were selected, recruited, and any inclusion or exclusion criteria._
`r lorem::ipsum(paragraphs = 1)`

### Measures
_List the instruments or tools used to collect data, describing each measure's purpose and any relevant psychometric properties._
`r lorem::ipsum(paragraphs = 2)`

### Analytic Plan
_Detail the statistical tests or models employed to analyze the data, including any software used._
`r lorem::ipsum(paragraphs = 1)`

## Results
Present the findings of the study, supported by statistical analysis, tables, and figures as necessary.
`r lorem::ipsum(paragraphs = 3)`

### [Subsection Title]
_Present specific findings related to a particular aspect of the study, supported by data._
`r lorem::ipsum(paragraphs = 1)`

### [Subsection Title]
_Discuss another set of findings, providing details on the analysis and interpretation._
`r lorem::ipsum(paragraphs = 1)`

## Discussion
_Interpret the results, discussing their implications in the context of existing literature. Address the study's limitations and suggest directions for future research._
`r lorem::ipsum(paragraphs = 3)`

### Limitations
_Acknowledge any limitations of your study and their potential impact on the findings._
`r lorem::ipsum(paragraphs = 1)`

### Future Directions
_Suggest areas for future research that could build on your findings or address limitations._
`r lorem::ipsum(paragraphs = 1)`

## Conclusion
_Summarize the main findings and their implications. Conclude with a statement that reflects the significance of the study and its contributions to the field._
`r lorem::ipsum(paragraphs = 1)`

## Acknowledgments
_Acknowledge anyone who contributed to the research but does not meet authorship criteria, along with any funding sources._
`r lorem::ipsum(paragraphs = 1)`

## Appendices
_Include any additional material that is pertinent to the study but would be too distracting or bulky to include in the main body of the manuscript._
`r lorem::ipsum(paragraphs = 1)`

### 2024-03-04 index.qmd
update index

@@
 
-<link href="output.css" rel="stylesheet">
-
-```{=html}
-<div class="bg-gray-900">
-  <main>
-    <!-- Hero section -->
-    <div class="relative isolate overflow-hidden">
-      <svg class="absolute inset-0 -z-10 h-full w-full stroke-white/10 [mask-image:radial-gradient(100%_100%_at_top_right,white,transparent)]" aria-hidden="true">
-        <defs>
-          <pattern id="983e3e4c-de6d-4c3f-8d64-b9761d1534cc" width="200" height="200" x="50%" y="-1" patternUnits="userSpaceOnUse">
-            <path d="M.5 200V.5H200" fill="none" />
-          </pattern>
-        </defs>
-        <svg x="50%" y="-1" class="overflow-visible fill-gray-800/20">
-          <path d="M-200 0h201v201h-201Z M600 0h201v201h-201Z M-400 600h201v201h-201Z M200 800h201v201h-201Z" stroke-width="0" />
-        </svg>
-        <rect width="100%" height="100%" stroke-width="0" fill="url(#983e3e4c-de6d-4c3f-8d64-b9761d1534cc)" />
-      </svg>
-      <div class="absolute left-[calc(50%-4rem)] top-10 -z-10 transform-gpu blur-3xl sm:left-[calc(50%-18rem)] lg:left-48 lg:top-[calc(50%-30rem)] xl:left-[calc(50%-24rem)]" aria-hidden="true">
-        <div class="aspect-[1108/632] w-[69.25rem] bg-gradient-to-r from-[#80caff] to-[#4f46e5] opacity-20" style="clip-path: polygon(73.6% 51.7%, 91.7% 11.8%, 100% 46.4%, 97.4% 82.2%, 92.5% 84.9%, 75.7% 64%, 55.3% 47.5%, 46.5% 49.4%, 45% 62.9%, 50.3% 87.2%, 21.3% 64.1%, 0.1% 100%, 5.4% 51.1%, 21.4% 63.9%, 58.9% 0.2%, 73.6% 51.7%)"></div>
-      </div>
-      <div class="mx-auto max-w-7xl px-6 pb-24 pt-10 sm:pb-40 lg:flex lg:px-8 lg:pt-40">
-        <div class="mx-auto max-w-2xl flex-shrink-0 lg:mx-0 lg:max-w-xl lg:pt-8">
-          <img class="h-11" src="https://tailwindui.com/img/logos/mark.svg?color=indigo&shade=500" alt="Your Company">
-          <div class="mt-24 sm:mt-32 lg:mt-16">
-            <a href="#" class="inline-flex space-x-6">
-              <span class="rounded-full bg-indigo-500/10 px-3 py-1 text-sm font-semibold leading-6 text-indigo-400 ring-1 ring-inset ring-indigo-500/20">Latest updates</span>
-              <span class="inline-flex items-center space-x-2 text-sm font-medium leading-6 text-gray-300">
-                <span>Just shipped v1.0</span>
-                <svg class="h-5 w-5 text-gray-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
-                  <path fill-rule="evenodd" d="M7.21 14.77a.75.75 0 01.02-1.06L11.168 10 7.23 6.29a.75.75 0 111.04-1.08l4.5 4.25a.75.75 0 010 1.08l-4.5 4.25a.75.75 0 01-1.06-.02z" clip-rule="evenodd" />
-                </svg>
-              </span>
-            </a>
-          </div>
-          <h1 class="mt-10 text-4xl font-bold tracking-tight text-center  text-white sm:text-6xl">Longitudinal Analysis Project</h1>
-          <p class="mt-6 text-lg leading-8 text-gray-300">An open-source and collaborative effort guided by the Biostatistics Working Group of the <b> <a href="https://abcdstudy.org/">Adolescent Brain Cognitive Development (ABCD) Study&reg</a></b></p>
-          <div class="mt-10 flex items-center gap-x-6">
-            <a href="#" class="rounded-md bg-indigo-500 px-3.5 py-2.5 text-sm font-semibold text-white shadow-sm hover:bg-indigo-400 focus-visible:outline focus-visible:outline-2 focus-visible:outline-offset-2 focus-visible:outline-indigo-400">Get started</a>
-            <a href="#" class="text-sm font-semibold leading-6 text-white">Live demo <span aria-hidden="true">→</span></a>
-          </div>
-        </div>
-        <div class="mx-auto mt-16 flex max-w-2xl sm:mt-24 lg:ml-10 lg:mr-0 lg:mt-0 lg:max-w-none lg:flex-none xl:ml-32">
-          <div class="max-w-3xl flex-none sm:max-w-5xl lg:max-w-none">
-            <img src="https://tailwindui.com/img/component-images/dark-project-app-screenshot.png" alt="App screenshot" width="2432" height="1442" class="w-[76rem] rounded-md bg-white/5 shadow-2xl ring-1 ring-white/10">
-          </div>
-        </div>
-      </div>
-    </div>
-
-    <!-- Logo cloud -->
-    <div class="mx-auto mt-8 max-w-7xl px-6 sm:mt-16 lg:px-8">
-      <h2 class="text-center text-lg font-semibold leading-8 text-white">The world’s most innovative companies use our app</h2>
-      <div class="mx-auto mt-10 grid max-w-lg grid-cols-4 items-center gap-x-8 gap-y-10 sm:max-w-xl sm:grid-cols-6 sm:gap-x-10 lg:mx-0 lg:max-w-none lg:grid-cols-5">
-        <img class="col-span-2 max-h-12 w-full object-contain lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/transistor-logo-white.svg" alt="Transistor" width="158" height="48">
-        <img class="col-span-2 max-h-12 w-full object-contain lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/reform-logo-white.svg" alt="Reform" width="158" height="48">
-        <img class="col-span-2 max-h-12 w-full object-contain lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/tuple-logo-white.svg" alt="Tuple" width="158" height="48">
-        <img class="col-span-2 max-h-12 w-full object-contain sm:col-start-2 lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/savvycal-logo-white.svg" alt="SavvyCal" width="158" height="48">
-        <img class="col-span-2 col-start-2 max-h-12 w-full object-contain sm:col-start-auto lg:col-span-1" src="https://tailwindui.com/img/logos/158x48/statamic-logo-white.svg" alt="Statamic" width="158" height="48">
-      </div>
-    </div>
-
-    <!-- Feature section -->
-    <div class="mx-auto mt-32 max-w-7xl px-6 sm:mt-56 lg:px-8">
-      <div class="mx-auto max-w-2xl text-center">
-        <h2 class="text-base font-semibold leading-7 text-indigo-400">Deploy faster</h2>
-        <p class="mt-2 text-3xl font-bold tracking-tight text-white sm:text-4xl">Everything you need to deploy your app</p>
-        <p class="mt-6 text-lg leading-8 text-gray-300">Lorem ipsum dolor sit amet consect adipisicing elit. Possimus magnam voluptatum cupiditate veritatis in accusamus quisquam.</p>
-      </div>
-      <div class="mx-auto mt-16 max-w-2xl sm:mt-20 lg:mt-24 lg:max-w-none">
-        <dl class="grid max-w-xl grid-cols-1 gap-x-8 gap-y-16 lg:max-w-none lg:grid-cols-3">
-          <div class="flex flex-col">
-            <dt class="text-base font-semibold leading-7 text-white">
-              <div class="mb-6 flex h-10 w-10 items-center justify-center rounded-lg bg-indigo-500">
-                <svg class="h-6 w-6 text-white" fill="none" viewBox="0 0 24 24" stroke-width="1.5" stroke="currentColor" aria-hidden="true">
-                  <path stroke-linecap="round" stroke-linejoin="round" d="M3.75 13.5l10.5-11.25L12 10.5h8.25L9.75 21.75 12 13.5H3.75z" />
-                </svg>
-              </div>
-              Server monitoring
-            </dt>
-            <dd class="mt-1 flex flex-auto flex-col text-base leading-7 text-gray-300">
-              <p class="flex-auto">Non quo aperiam repellendus quas est est. Eos aut dolore aut ut sit nesciunt. Ex tempora quia. Sit nobis consequatur dolores incidunt.</p>
-              <p class="mt-6">
-                <a href="#" class="text-sm font-semibold leading-6 text-indigo-400">Learn more <span aria-hidden="true">→</span></a>
-              </p>
-            </dd>
-          </div>
-          <div class="flex flex-col">
-            <dt class="text-base font-semibold leading-7 text-white">
-              <div class="mb-6 flex h-10 w-10 items-center justify-center rounded-lg bg-indigo-500">
-                <svg class="h-6 w-6 text-white" fill="none" viewBox="0 0 24 24" stroke-width="1.5" stroke="currentColor" aria-hidden="true">
-                  <path stroke-linecap="round" stroke-linejoin="round" d="M15 19.128a9.38 9.38 0 002.625.372 9.337 9.337 0 004.121-.952 4.125 4.125 0 00-7.533-2.493M15 19.128v-.003c0-1.113-.285-2.16-.786-3.07M15 19.128v.106A12.318 12.318 0 018.624 21c-2.331 0-4.512-.645-6.374-1.766l-.001-.109a6.375 6.375 0 0111.964-3.07M12 6.375a3.375 3.375 0 11-6.75 0 3.375 3.375 0 016.75 0zm8.25 2.25a2.625 2.625 0 11-5.25 0 2.625 2.625 0 015.25 0z" />
-                </svg>
-              </div>
-              Collaborate
-            </dt>
-            <dd class="mt-1 flex flex-auto flex-col text-base leading-7 text-gray-300">
-              <p class="flex-auto">Vero eum voluptatem aliquid nostrum voluptatem. Vitae esse natus. Earum nihil deserunt eos quasi cupiditate. A inventore et molestiae natus.</p>
-              <p class="mt-6">
-                <a href="#" class="text-sm font-semibold leading-6 text-indigo-400">Learn more <span aria-hidden="true">→</span></a>
-              </p>
-            </dd>
-          </div>
-          <div class="flex flex-col">
-            <dt class="text-base font-semibold leading-7 text-white">
-              <div class="mb-6 flex h-10 w-10 items-center justify-center rounded-lg bg-indigo-500">
-                <svg class="h-6 w-6 text-white" fill="none" viewBox="0 0 24 24" stroke-width="1.5" stroke="currentColor" aria-hidden="true">
-                  <path stroke-linecap="round" stroke-linejoin="round" d="M6.75 3v2.25M17.25 3v2.25M3 18.75V7.5a2.25 2.25 0 012.25-2.25h13.5A2.25 2.25 0 0121 7.5v11.25m-18 0A2.25 2.25 0 005.25 21h13.5A2.25 2.25 0 0021 18.75m-18 0v-7.5A2.25 2.25 0 015.25 9h13.5A2.25 2.25 0 0121 11.25v7.5m-9-6h.008v.008H12v-.008zM12 15h.008v.008H12V15zm0 2.25h.008v.008H12v-.008zM9.75 15h.008v.008H9.75V15zm0 2.25h.008v.008H9.75v-.008zM7.5 15h.008v.008H7.5V15zm0 2.25h.008v.008H7.5v-.008zm6.75-4.5h.008v.008h-.008v-.008zm0 2.25h.008v.008h-.008V15zm0 2.25h.008v.008h-.008v-.008zm2.25-4.5h.008v.008H16.5v-.008zm0 2.25h.008v.008H16.5V15z" />
-                </svg>
-              </div>
-              Task scheduling
-            </dt>
-            <dd class="mt-1 flex flex-auto flex-col text-base leading-7 text-gray-300">
-              <p class="flex-auto">Et quod quaerat dolorem quaerat architecto aliquam accusantium. Ex adipisci et doloremque autem quia quam. Quis eos molestiae at iure impedit.</p>
-              <p class="mt-6">
-                <a href="#" class="text-sm font-semibold leading-6 text-indigo-400">Learn more <span aria-hidden="true">→</span></a>
-              </p>
-            </dd>
-          </div>
-        </dl>
-      </div>
-    </div>
-
-    <!-- Feature section -->
-    <div class="mt-32 sm:mt-56">
-      <div class="mx-auto max-w-7xl px-6 lg:px-8">
-        <div class="mx-auto max-w-2xl sm:text-center">
-          <h2 class="text-base font-semibold leading-7 text-indigo-400">Everything you need</h2>
-          <p class="mt-2 text-3xl font-bold tracking-tight text-white sm:text-4xl">No server? No problem.</p>
-          <p class="mt-6 text-lg leading-8 text-gray-300">Lorem ipsum, dolor sit amet consectetur adipisicing elit. Maiores impedit perferendis suscipit eaque, iste dolor cupiditate blanditiis.</p>
-        </div>
-      </div>
-      <div class="relative overflow-hidden pt-16">
-        <div class="mx-auto max-w-7xl px-6 lg:px-8">
-          <img src="https://tailwindui.com/img/component-images/dark-project-app-screenshot.png" alt="App screenshot" class="mb-[-12%] rounded-xl shadow-2xl ring-1 ring-white/10" width="2432" height="1442">
-          <div class="relative" aria-hidden="true">
-            <div class="absolute -inset-x-20 bottom-0 bg-gradient-to-t from-gray-900 pt-[7%]"></div>
-          </div>
-        </div>
-      </div>
-      <div class="mx-auto mt-16 max-w-7xl px-6 sm:mt-20 md:mt-24 lg:px-8">
-        <dl class="mx-auto grid max-w-2xl grid-cols-1 gap-x-6 gap-y-10 text-base leading-7 text-gray-300 sm:grid-cols-2 lg:mx-0 lg:max-w-none lg:grid-cols-3 lg:gap-x-8 lg:gap-y-16">
-          <div class="relative pl-9">
-            <dt class="inline font-semibold text-white">
-              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
-                <path fill-rule="evenodd" d="M5.5 17a4.5 4.5 0 01-1.44-8.765 4.5 4.5 0 018.302-3.046 3.5 3.5 0 014.504 4.272A4 4 0 0115 17H5.5zm3.75-2.75a.75.75 0 001.5 0V9.66l1.95 2.1a.75.75 0 101.1-1.02l-3.25-3.5a.75.75 0 00-1.1 0l-3.25 3.5a.75.75 0 101.1 1.02l1.95-2.1v4.59z" clip-rule="evenodd" />
-              </svg>
-              Push to deploy.
-            </dt>
-            <dd class="inline">Lorem ipsum, dolor sit amet consectetur adipisicing elit aute id magna.</dd>
-          </div>
-          <div class="relative pl-9">
-            <dt class="inline font-semibold text-white">
-              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
-                <path fill-rule="evenodd" d="M10 1a4.5 4.5 0 00-4.5 4.5V9H5a2 2 0 00-2 2v6a2 2 0 002 2h10a2 2 0 002-2v-6a2 2 0 00-2-2h-.5V5.5A4.5 4.5 0 0010 1zm3 8V5.5a3 3 0 10-6 0V9h6z" clip-rule="evenodd" />
-              </svg>
-              SSL certificates.
-            </dt>
-            <dd class="inline">Anim aute id magna aliqua ad ad non deserunt sunt. Qui irure qui lorem cupidatat commodo.</dd>
-          </div>
-          <div class="relative pl-9">
-            <dt class="inline font-semibold text-white">
-              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
-                <path fill-rule="evenodd" d="M15.312 11.424a5.5 5.5 0 01-9.201 2.466l-.312-.311h2.433a.75.75 0 000-1.5H3.989a.75.75 0 00-.75.75v4.242a.75.75 0 001.5 0v-2.43l.31.31a7 7 0 0011.712-3.138.75.75 0 00-1.449-.39zm1.23-3.723a.75.75 0 00.219-.53V2.929a.75.75 0 00-1.5 0V5.36l-.31-.31A7 7 0 003.239 8.188a.75.75 0 101.448.389A5.5 5.5 0 0113.89 6.11l.311.31h-2.432a.75.75 0 000 1.5h4.243a.75.75 0 00.53-.219z" clip-rule="evenodd" />
-              </svg>
-              Simple queues.
-            </dt>
-            <dd class="inline">Ac tincidunt sapien vehicula erat auctor pellentesque rhoncus.</dd>
-          </div>
-          <div class="relative pl-9">
-            <dt class="inline font-semibold text-white">
-              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
-                <path fill-rule="evenodd" d="M10 2.5c-1.31 0-2.526.386-3.546 1.051a.75.75 0 01-.82-1.256A8 8 0 0118 9a22.47 22.47 0 01-1.228 7.351.75.75 0 11-1.417-.49A20.97 20.97 0 0016.5 9 6.5 6.5 0 0010 2.5zM4.333 4.416a.75.75 0 01.218 1.038A6.466 6.466 0 003.5 9a7.966 7.966 0 01-1.293 4.362.75.75 0 01-1.257-.819A6.466 6.466 0 002 9c0-1.61.476-3.11 1.295-4.365a.75.75 0 011.038-.219zM10 6.12a3 3 0 00-3.001 3.041 11.455 11.455 0 01-2.697 7.24.75.75 0 01-1.148-.965A9.957 9.957 0 005.5 9c0-.028.002-.055.004-.082a4.5 4.5 0 018.996.084V9.15l-.005.297a.75.75 0 11-1.5-.034c.003-.11.004-.219.005-.328a3 3 0 00-3-2.965zm0 2.13a.75.75 0 01.75.75c0 3.51-1.187 6.745-3.181 9.323a.75.75 0 11-1.186-.918A13.687 13.687 0 009.25 9a.75.75 0 01.75-.75zm3.529 3.698a.75.75 0 01.584.885 18.883 18.883 0 01-2.257 5.84.75.75 0 11-1.29-.764 17.386 17.386 0 002.078-5.377.75.75 0 01.885-.584z" clip-rule="evenodd" />
-              </svg>
-              Advanced security.
-            </dt>
-            <dd class="inline">Lorem ipsum, dolor sit amet consectetur adipisicing elit aute id magna.</dd>
-          </div>
-          <div class="relative pl-9">
-            <dt class="inline font-semibold text-white">
-              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
-                <path fill-rule="evenodd" d="M7.84 1.804A1 1 0 018.82 1h2.36a1 1 0 01.98.804l.331 1.652a6.993 6.993 0 011.929 1.115l1.598-.54a1 1 0 011.186.447l1.18 2.044a1 1 0 01-.205 1.251l-1.267 1.113a7.047 7.047 0 010 2.228l1.267 1.113a1 1 0 01.206 1.25l-1.18 2.045a1 1 0 01-1.187.447l-1.598-.54a6.993 6.993 0 01-1.929 1.115l-.33 1.652a1 1 0 01-.98.804H8.82a1 1 0 01-.98-.804l-.331-1.652a6.993 6.993 0 01-1.929-1.115l-1.598.54a1 1 0 01-1.186-.447l-1.18-2.044a1 1 0 01.205-1.251l1.267-1.114a7.05 7.05 0 010-2.227L1.821 7.773a1 1 0 01-.206-1.25l1.18-2.045a1 1 0 011.187-.447l1.598.54A6.993 6.993 0 017.51 3.456l.33-1.652zM10 13a3 3 0 100-6 3 3 0 000 6z" clip-rule="evenodd" />
-              </svg>
-              Powerful API.
-            </dt>
-            <dd class="inline">Anim aute id magna aliqua ad ad non deserunt sunt. Qui irure qui lorem cupidatat commodo.</dd>
-          </div>
-          <div class="relative pl-9">
-            <dt class="inline font-semibold text-white">
-              <svg class="absolute left-1 top-1 h-5 w-5 text-indigo-500" viewBox="0 0 20 20" fill="currentColor" aria-hidden="true">
-                <path d="M4.632 3.533A2 2 0 016.577 2h6.846a2 2 0 011.945 1.533l1.976 8.234A3.489 3.489 0 0016 11.5H4c-.476 0-.93.095-1.344.267l1.976-8.234z" />
-                <path fill-rule="evenodd" d="M4 13a2 2 0 100 4h12a2 2 0 100-4H4zm11.24 2a.75.75 0 01.75-.75H16a.75.75 0 01.75.75v.01a.75.75 0 01-.75.75h-.01a.75.75 0 01-.75-.75V15zm-2.25-.75a.75.75 0 00-.75.75v.01c0 .414.336.75.75.75H13a.75.75 0 00.75-.75V15a.75.75 0 00-.75-.75h-.01z" clip-rule="evenodd" />
-              </svg>
-              Database backups.
-            </dt>
-            <dd class="inline">Ac tincidunt sapien vehicula erat auctor pellentesque rhoncus.</dd>
-          </div>
-        </dl>
-      </div>
-    </div>
-
-    <!-- Stats -->
-    <div class="mx-auto mt-32 max-w-7xl px-6 sm:mt-56 lg:px-8">
-      <div class="mx-auto max-w-2xl lg:mx-0 lg:max-w-xl">
-        <h2 class="text-base font-semibold leading-8 text-indigo-400">Our track record</h2>
-        <p class="mt-2 text-3xl font-bold tracking-tight text-white sm:text-4xl">Trusted by thousands of developers&nbsp;worldwide</p>
-        <p class="mt-6 text-lg leading-8 text-gray-300">Lorem ipsum, dolor sit amet consectetur adipisicing elit.
-Maiores impedit perferendis suscipit eaque, iste dolor cupiditate blanditiis ratione.</p>
-      </div>
-      <dl class="mx-auto mt-16 grid max-w-2xl grid-cols-1 gap-x-8 gap-y-10 text-white sm:mt-20 sm:grid-cols-2 sm:gap-y-16 lg:mx-0 lg:max-w-none lg:grid-cols-4">
-        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
-          <dt class="text-sm leading-6">Developers on the platform</dt>
-          <dd class="order-first text-3xl font-semibold tracking-tight">8,000+</dd>
-        </div>
-        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
-          <dt class="text-sm leading-6">Daily requests</dt>
-          <dd class="order-first text-3xl font-semibold tracking-tight">900m+</dd>
-        </div>
-        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
-          <dt class="text-sm leading-6">Uptime guarantee</dt>
-          <dd class="order-first text-3xl font-semibold tracking-tight">99.9%</dd>
-        </div>
-        <div class="flex flex-col gap-y-3 border-l border-white/10 pl-6">
-          <dt class="text-sm leading-6">Projects deployed</dt>
-          <dd class="order-first text-3xl font-semibold tracking-tight">12m</dd>
-        </div>
-      </dl>
-    </div>
-
-    <!-- CTA section -->
-    <div class="relative isolate mt-32 px-6 py-32 sm:mt-56 sm:py-40 lg:px-8">
-      <svg class="absolute inset-0 -z-10 h-full w-full stroke-white/10 [mask-image:radial-gradient(100%_100%_at_top_right,white,transparent)]" aria-hidden="true">
-        <defs>
-          <pattern id="1d4240dd-898f-445f-932d-e2872fd12de3" width="200" height="200" x="50%" y="0" patternUnits="userSpaceOnUse">
-            <path d="M.5 200V.5H200" fill="none" />
-          </pattern>
-        </defs>
-        <svg x="50%" y="0" class="overflow-visible fill-gray-800/20">
-          <path d="M-200 0h201v201h-201Z M600 0h201v201h-201Z M-400 600h201v201h-201Z M200 800h201v201h-201Z" stroke-width="0" />
-        </svg>
-        <rect width="100%" height="100%" stroke-width="0" fill="url(#1d4240dd-898f-445f-932d-e2872fd12de3)" />
-      </svg>
-      <div class="absolute inset-x-0 top-10 -z-10 flex transform-gpu justify-center overflow-hidden blur-3xl" aria-hidden="true">
-        <div class="aspect-[1108/632] w-[69.25rem] flex-none bg-gradient-to-r from-[#80caff] to-[#4f46e5] opacity-20" style="clip-path: polygon(73.6% 51.7%, 91.7% 11.8%, 100% 46.4%, 97.4% 82.2%, 92.5% 84.9%, 75.7% 64%, 55.3% 47.5%, 46.5% 49.4%, 45% 62.9%, 50.3% 87.2%, 21.3% 64.1%, 0.1% 100%, 5.4% 51.1%, 21.4% 63.9%, 58.9% 0.2%, 73.6% 51.7%)"></div>
-      </div>
-      <div class="mx-auto max-w-2xl text-center">
-        <h2 class="text-3xl font-bold tracking-tight text-white sm:text-4xl">Boost your productivity.<br>Start using our app today.</h2>
-        <p class="mx-auto mt-6 max-w-xl text-lg leading-8 text-gray-300">Incididunt sint fugiat pariatur cupidatat consectetur sit cillum anim id veniam aliqua proident excepteur commodo do ea.</p>
-        <div class="mt-10 flex items-center justify-center gap-x-6">
-          <a href="#" class="rounded-md bg-white px-3.5 py-2.5 text-sm font-semibold text-gray-900 shadow-sm hover:bg-gray-100 focus-visible:outline focus-visible:outline-2 focus-visible:outline-offset-2 focus-visible:outline-white">Get started</a>
-          <a href="#" class="text-sm font-semibold leading-6 text-white">Learn more <span aria-hidden="true">→</span></a>
-        </div>
-      </div>
-    </div>
-  </main>
-
-  <!-- Footer -->
-  <footer aria-labelledby="footer-heading" class="relative">
-    <h2 id="footer-heading" class="sr-only">Footer</h2>
-    <div class="mx-auto max-w-7xl px-6 pb-8 pt-4 lg:px-8">
-      <div class="border-t border-white/10 pt-8 md:flex md:items-center md:justify-between">
-        <div class="flex space-x-6 md:order-2">
-          <a href="#" class="text-gray-500 hover:text-gray-400">
-            <span class="sr-only">Facebook</span>
-            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
-              <path fill-rule="evenodd" d="M22 12c0-5.523-4.477-10-10-10S2 6.477 2 12c0 4.991 3.657 9.128 8.438 9.878v-6.987h-2.54V12h2.54V9.797c0-2.506 1.492-3.89 3.777-3.89 1.094 0 2.238.195 2.238.195v2.46h-1.26c-1.243 0-1.63.771-1.63 1.562V12h2.773l-.443 2.89h-2.33v6.988C18.343 21.128 22 16.991 22 12z" clip-rule="evenodd" />
-            </svg>
-          </a>
-          <a href="#" class="text-gray-500 hover:text-gray-400">
-            <span class="sr-only">Instagram</span>
-            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
-              <path fill-rule="evenodd" d="M12.315 2c2.43 0 2.784.013 3.808.06 1.064.049 1.791.218 2.427.465a4.902 4.902 0 011.772 1.153 4.902 4.902 0 011.153 1.772c.247.636.416 1.363.465 2.427.048 1.067.06 1.407.06 4.123v.08c0 2.643-.012 2.987-.06 4.043-.049 1.064-.218 1.791-.465 2.427a4.902 4.902 0 01-1.153 1.772 4.902 4.902 0 01-1.772 1.153c-.636.247-1.363.416-2.427.465-1.067.048-1.407.06-4.123.06h-.08c-2.643 0-2.987-.012-4.043-.06-1.064-.049-1.791-.218-2.427-.465a4.902 4.902 0 01-1.772-1.153 4.902 4.902 0 01-1.153-1.772c-.247-.636-.416-1.363-.465-2.427-.047-1.024-.06-1.379-.06-3.808v-.63c0-2.43.013-2.784.06-3.808.049-1.064.218-1.791.465-2.427a4.902 4.902 0 011.153-1.772A4.902 4.902 0 015.45 2.525c.636-.247 1.363-.416 2.427-.465C8.901 2.013 9.256 2 11.685 2h.63zm-.081 1.802h-.468c-2.456 0-2.784.011-3.807.058-.975.045-1.504.207-1.857.344-.467.182-.8.398-1.15.748-.35.35-.566.683-.748 1.15-.137.353-.3.882-.344 1.857-.047 1.023-.058 1.351-.058 3.807v.468c0 2.456.011 2.784.058 3.807.045.975.207 1.504.344 1.857.182.466.399.8.748 1.15.35.35.683.566 1.15.748.353.137.882.3 1.857.344 1.054.048 1.37.058 4.041.058h.08c2.597 0 2.917-.01 3.96-.058.976-.045 1.505-.207 1.858-.344.466-.182.8-.398 1.15-.748.35-.35.566-.683.748-1.15.137-.353.3-.882.344-1.857.048-1.055.058-1.37.058-4.041v-.08c0-2.597-.01-2.917-.058-3.96-.045-.976-.207-1.505-.344-1.858a3.097 3.097 0 00-.748-1.15 3.098 3.098 0 00-1.15-.748c-.353-.137-.882-.3-1.857-.344-1.023-.047-1.351-.058-3.807-.058zM12 6.865a5.135 5.135 0 110 10.27 5.135 5.135 0 010-10.27zm0 1.802a3.333 3.333 0 100 6.666 3.333 3.333 0 000-6.666zm5.338-3.205a1.2 1.2 0 110 2.4 1.2 1.2 0 010-2.4z" clip-rule="evenodd" />
-            </svg>
-          </a>
-          <a href="#" class="text-gray-500 hover:text-gray-400">
-            <span class="sr-only">Twitter</span>
-            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
-              <path d="M8.29 20.251c7.547 0 11.675-6.253 11.675-11.675 0-.178 0-.355-.012-.53A8.348 8.348 0 0022 5.92a8.19 8.19 0 01-2.357.646 4.118 4.118 0 001.804-2.27 8.224 8.224 0 01-2.605.996 4.107 4.107 0 00-6.993 3.743 11.65 11.65 0 01-8.457-4.287 4.106 4.106 0 001.27 5.477A4.072 4.072 0 012.8 9.713v.052a4.105 4.105 0 003.292 4.022 4.095 4.095 0 01-1.853.07 4.108 4.108 0 003.834 2.85A8.233 8.233 0 012 18.407a11.616 11.616 0 006.29 1.84" />
-            </svg>
-          </a>
-          <a href="#" class="text-gray-500 hover:text-gray-400">
-            <span class="sr-only">GitHub</span>
-            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
-              <path fill-rule="evenodd" d="M12 2C6.477 2 2 6.484 2 12.017c0 4.425 2.865 8.18 6.839 9.504.5.092.682-.217.682-.483 0-.237-.008-.868-.013-1.703-2.782.605-3.369-1.343-3.369-1.343-.454-1.158-1.11-1.466-1.11-1.466-.908-.62.069-.608.069-.608 1.003.07 1.531 1.032 1.531 1.032.892 1.53 2.341 1.088 2.91.832.092-.647.35-1.088.636-1.338-2.22-.253-4.555-1.113-4.555-4.951 0-1.093.39-1.988 1.029-2.688-.103-.253-.446-1.272.098-2.65 0 0 .84-.27 2.75 1.026A9.564 9.564 0 0112 6.844c.85.004 1.705.115 2.504.337 1.909-1.296 2.747-1.027 2.747-1.027.546 1.379.202 2.398.1 2.651.64.7 1.028 1.595 1.028 2.688 0 3.848-2.339 4.695-4.566 4.943.359.309.678.92.678 1.855 0 1.338-.012 2.419-.012 2.747 0 .268.18.58.688.482A10.019 10.019 0 0022 12.017C22 6.484 17.522 2 12 2z" clip-rule="evenodd" />
-            </svg>
-          </a>
-          <a href="#" class="text-gray-500 hover:text-gray-400">
-            <span class="sr-only">YouTube</span>
-            <svg class="h-6 w-6" fill="currentColor" viewBox="0 0 24 24" aria-hidden="true">
-              <path fill-rule="evenodd" d="M19.812 5.418c.861.23 1.538.907 1.768 1.768C21.998 8.746 22 12 22 12s0 3.255-.418 4.814a2.504 2.504 0 0 1-1.768 1.768c-1.56.419-7.814.419-7.814.419s-6.255 0-7.814-.419a2.505 2.505 0 0 1-1.768-1.768C2 15.255 2 12 2 12s0-3.255.417-4.814a2.507 2.507 0 0 1 1.768-1.768C5.744 5 11.998 5 11.998 5s6.255 0 7.814.418ZM15.194 12 10 15V9l5.194 3Z" clip-rule="evenodd" />
-            </svg>
-          </a>
-        </div>
-        <p class="mt-8 text-xs leading-5 text-gray-400 md:order-1 md:mt-0">&copy; 2020 Your Company, Inc. All rights reserved.</p>
-      </div>
-    </div>
-  </footer>
-</div>
-
-```
+Test

### 2024-03-06 _posts/2_Examples_ResidualizedChangeScores.qmd
revised manuscript docs

--- 2024-03-02 posts/1b_Examples_DifferenceScores_SimpleRegression.qmd
@@
 ---
-title: "Difference Scores: Simple Regression"
+title: "Residualized Change Scores"
 author: "Biostatistics Working Group"
-
-reader-mode: true
 ---
 
-# Simple Regression {.unnumbered}
+# Mulitple Regression {.unnumbered}
 
@@
 
-This example assesses whether growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1) differs significantly based on handedness, using height ("anthroheightcalc") as a representative metric of growth.
-The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 height measurements for each participant; 2) a simple regression analysis is used to test whether sex (boy, girl) predicts the average difference value in participants height from baseline to the 1-Year follow-up.
-Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between difference scores and sex.
+This example assesses whether height, measured at the 1-Year follow-up (T1) in a subsample of ABCD participants, is predicted by weight and height measured at baseline (T0).
+A visual inspection is further conducted via a scatterplot to graphically represent the relationship between height at Year_1 and weight at baseline, controlling for height at baseline.
 The ensuing analysis and interpretations are detailed in the subsequent sections.
@@
 #| output: FALSE
-
-# rm(list = ls())
 
@@
 ::: panel-tabset
-### Build Model: Step 1 {.tabset .tabset-fade .tabset-pills}
+### Build Model {.tabset .tabset-fade .tabset-pills}
 
 ::: blue
-The code snippet below tells R to compute a difference score by subtracting each participant’s height at T1 from their height at T0. Relevant summary statistics are also provided.
+The code snippet below tells R to conduct a multiple regression analysis by subtracting each participant's height at T1 from their height at T0. Relevant summary statistics are also provided.
 
-**STEP 1: Compute Difference Score**
-```{r build-model-regression-1}
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-#| class.source: 'fold-hide'
-
-# Define the function to compute difference scores for a given variable and provide a summary
-# Function to compute difference scores for a given variable and provide a summary
-compute_difference_and_summary <- function(df, variable_name) {
-    # Define the event names of interest
-    baseline_event <- "baseline_year_1_arm_1"
-    followup_event <- "1_year_follow_up_y_arm_1"
-
-    # Compute the difference between Baseline and Year 1 data for the given variable
-    diff_data <- df %>%
-        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
-        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
-        spread(eventname, variable_name) %>% # Convert data from long to wide format
-        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
-        drop_na(diff) # Exclude rows with NA in the computed difference
-
-    # Summarize the computed difference scores
-    diff_summary <- summary(diff_data$diff)
-
-    # Return the difference data and its summary
-    list(data = diff_data, summary = diff_summary)
-}
-
-# List of variables for which difference scores are to be computed
-variables_of_interest <- c("anthroheightcalc")
-
-# Compute and store difference scores and summaries for each variable in a list
-difference_and_summary_list <- lapply(variables_of_interest, function(var) {
-    compute_difference_and_summary(df, var)
-})
-
-# Extract the difference data for the 'anthroheightcalc' variable
-height_diff_data <- difference_and_summary_list[[1]]$data
-
-# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
-df <- left_join(df, height_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")
-
-```
-
-**Descriptive Statistics for the Difference Score**
-```{r model-checking}
-# Compute statistical summaries for the difference score variable
-lapply(difference_and_summary_list, function(item) {
-    print(item$summary)
-})
-```
-
-This summary of the difference score variable indicates that the differences range from a decrease of `r min(height_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(height_diff_data$diff, na.rm = TRUE)` units.
-The median difference is `r median(height_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(height_diff_data$diff, digits = 2, na.rm = TRUE)` units.
-There are `r sum(is.na(height_diff_data$diff))` missing values in this difference score dataset.
-
-**Descriptive Statistics for the Height Variable**
-```{r}
-## Summary statistics
-# Compute summary statistics for Height by eventname
-summary <- df %>%
-    group_by(eventname) %>%
-    get_summary_stats(anthroheightcalc, type = "mean_sd")
-
-data.frame(summary)
-```
-
-The summary statistics provide insights into the Height variable across different `eventname` categories.
-At baseline, the average height is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`.
-Over the years, there’s a noticeable increase in average height: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year.
-The standard deviation remains relatively consistent over the years, suggesting similar variability in height measurements across these time points.
-
-:::
-
-### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-A simple regression analyses is conducted to examine whether a grouping variable (sex) significantly predicts the difference score value (indicating significant group differences in the average difference score).
-
-**STEP 2: Conduct regression on Difference Score**
-```{r build-model-regression-2}
+**Compute Multiple Regression Model**
+```{r build-model-resid-regression-1}
 #| class.source: 'fold-hide'
@@
 
-# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
-merged_data <- height_diff_data %>%
-  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")
+# Split data using the correct eventname values
+baseline_data <- df %>% 
+  filter(eventname == "baseline_year_1_arm_1") %>% 
+  select(src_subject_id, Height_baseline = anthroheightcalc, Weight_baseline = anthroweightcalc)
 
-# Ensure 'sex' is a factor
-merged_data$sex <- as.factor(merged_data$sex)
+followup_data <- df %>% 
+  filter(eventname == "1_year_follow_up_y_arm_1") %>% 
+  select(src_subject_id, Height_followup = anthroheightcalc)
 
-# Run the regression
-model <- lm(diff ~ sex, data = merged_data)
+# Merge baseline and follow-up data
+merged_data <- baseline_data %>%
+  left_join(followup_data, by = "src_subject_id")
 
-# Get the summary of the regression model
-model_summary <- summary(model)
+# Check the first few rows of the merged data
+#head(merged_data)
 
-model_summary
+# Fit the regression model excluding NA values
+model <- lm(Height_followup ~ Height_baseline + Weight_baseline, data = merged_data, na.action = na.exclude)
 
@@
 
-This regression analysis evaluates whether sex (boy, girl) predicts the average difference in participants' height from baseline to the 1-Year follow-up. The output from our model provides:
+:::
+
+### Model Summary {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+
+A plot to show xxxxx.
+```{r model-checking}
+model_summary <- summary(model)
+print(model_summary)
+
+par(mfrow = c(2, 2))  # Set up a 2x2 plotting area
+plot(model)  # Generate diagnostic plots
+
+```
+
+This regression analysis evaluates how baseline height and weight predict children's height at a 1-year follow-up. The output from our model provides several key statistics:
 
@@
 ii. degrees of freedom of `r model_summary$df[2]`;
-iii. a parameter estimate of `r model_summary$coefficients["sexMale or other", "Estimate"]`; 
-iv. standard error of `r model_summary$coefficients["sexMale or other", "Std. Error"]`; 
-v. p-value of `r model_summary$coefficients["sexMale or other", "Pr(>|t|)"]`.
+iii. parameter estimates for baseline height and weight;
+iv. standard errors for these estimates; and
+v. p-values for the significance of these estimates.
 
-Compared to boys (the reference group), girls have an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units.
-This effect was marginally significant with a p-value of `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`.
-Overall, the model explained a very small portion of the variance in height difference, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 6)`, and the overall model significance was not statistically significant with a p-value of `r pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)`.
+For every one unit increase in baseline height, there's an associated increase of approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units in the follow-up height, and this effect was statistically significant with a p-value of `r model_summary$coefficients["Height_baseline", "Pr(>|t|)"]`.
+Similarly, for every one unit increase in baseline weight, there's an associated increase of approximately `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units in the follow-up height, which is also statistically significant with a p-value of `r model_summary$coefficients["Weight_baseline", "Pr(>|t|)"]`.
+Overall, this model explains a substantial portion of the variance in follow-up height, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 4)`.
+The overall model is highly significant with a p-value less than `r model_summary$coefficients[1,4]`, indicating that the predictors, collectively, have a significant relationship with the dependent variable.
 
@@
 
-### Model Plots
+### Model Plots  {.tabset .tabset-fade .tabset-pills}
 
 ::: blue
-```{r violin-plots}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
+```{r scatterplot}
 
-# Visualize the difference scores across different levels of sex
-
-# Create a violin plot to show the distribution of difference scores by sex
-# Jittered points are added to provide a more granular view of individual observations
-# Load necessary libraries
-library(dplyr)
-library(ggplot2)
-library(RColorBrewer)
-
-# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
-merged_data <- height_diff_data %>%
-  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")
-
-# Plotting using ggplot2
-ggplot(merged_data, aes(x = sex, y = diff, fill = sex)) +
-  geom_violin() +
-  geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
-  scale_fill_brewer(palette = "Set2") + 
-  labs(
-    title = "Difference Scores by Sex",
-    x = "Sex",
-    y = "Difference Score"
-  ) +
-  theme_minimal() + 
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    legend.position = "none"
-  )
+# Scatterplot with regression line for Weight_baseline
+ggplot(merged_data, aes(x = Weight_baseline, y = Height_followup)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", color = "black") + 
+  labs(title = "Relationship between Baseline Weight and Follow-up Height",
+       x = "Baseline Weight",
+       y = "Follow-up Height") +
+  theme_minimal()
 
@@
 
-The violin plot visualizes the distribution of difference scores in children's heights by their sex: Right, Left, and Mixed.
-Each violin shape provides insight into the density of the data at different height differences, with wider sections representing higher densities of data points.
-Superimposed jittered points offer a granular view of individual observations.
-From the plot, it appears that the distributions of height differences across the three handedness categories are somewhat similar, though there might be subtle variations in median and spread.
+The scatterplot visually depicts the relationship between children's weights at baseline (`r names(merged_data)[3]`) and their heights at Year_1 (`r names(merged_data)[4]`).
+Each point on the plot represents a child, with their baseline weight plotted on the x-axis and their Year_1 height on the y-axis.
+The clear positive linear trend, as illustrated by the black regression line, indicates that children who had higher weights at baseline generally had higher heights at Year_1.
+While further statistical analyses can quantify the strength and direction of this relationship, visually, the data suggests a pronounced positive association between baseline weight and follow-up height.
 
@@
 ## Wrapping Up
+<div class = "blue">
 
@@
 ::: blue
-Differences in children's heights between baseline and a subsequent 1-year follow-up, we sought to understand the potential influence of sex on this difference.
-Descriptive statistics revealed that height differences ranged from a decrease of `r min(merged_data$diff, na.rm = TRUE)` units to an increase of `r max(merged_data$diff, na.rm = TRUE)` units, with a median difference of `r median(merged_data$diff, na.rm = TRUE)` units and an average difference of approximately `r mean(merged_data$diff, na.rm = TRUE)` units.
-Subsequent regression analysis indicated that, when compared to boys, girls had an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units, though this effect was marginally significant (p = `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`).
-Complementary to these findings, violin plots visually underscored the subtle variations in height differences across sex categories, suggesting broadly similar distributions but with nuanced variations in central tendency and spread.
-In conclusion, while sex exhibited a potential influence on height difference over the year, the observed effect was notably minor.
+
+The regression analysis was conducted to predict children's height at the 1-Year follow-up using their baseline height (`r names(merged_data)[2]`) and weight (`r names(merged_data)[3]`).
+Both predictors were statistically significant.
+Specifically, for every one-unit increase in baseline height, the height at Year_1 increased by approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units, while holding weight constant.
+Similarly, for every one-unit increase in baseline weight, the height at Year_1 increased by about `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units, while holding baseline height constant.
+The overall model explained a substantial `r round(model_summary$adj.r.squared * 100, 2)`% of the variance in height at the 1-Year follow-up, as indicated by the adjusted R-squared value.
 

### 2024-03-06 _posts/3a_Examples_LinearMixedModels.qmd
revised manuscript docs

--- 2024-03-02 posts/3b_Examples_LinearMixedModels.qmd
@@
 ---
-title: "Linear Mixed Models: Random Intercept and Slope"
+title: "Linear Mixed Models: Random Intercept"
 author: "Biostatistics Working Group"
-
+execute:
+  freeze: auto  # re-render only when source changes
+  cache: true
 ---
-
-# LMM: Intercept & Slope {.unnumbered}
 
@@
 
-The Linear Mixed Model with a random intercept and slope (LMM:ris) extends the traditional fixed-effect linear regression by incorporating both a subject-specific random intercept and a random slope.
-This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.
+The LMM:ri is similar to traditional (fixed-effect) linear regression extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value zzzzt
 
-In this example, we will utilize the LMM:ris to analyze height trajectories obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
-Our primary objective is to understand the stability and evolution in height measurements, while factoring in the clustered nature of observations within individuals over time.
-The LMM:ris facilitates this by concurrently modeling an overarching sample mean trajectory (fixed effect) and individual variations (random effects) in both starting points (intercepts) and growth rates (slopes) around this mean trajectory.
+In this example, we will use the LMM:ri to analyze trajectories of height obtained across multiple measurement occasions in a sample of youth taking part in the ABCD Study.
+Our primary aim is to characterize stability and change in height assessments, while accounting for observations that are clustered within youth over time.
+To do so, we will use the LMM:ri to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory.
 
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT","lme4","report","broom","gridExtra","easystats","gtsummary")
+packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")
 
@@
 library(tidyverse)    # Collection of R packages for data science
-library(rstatix)      # Pipe-friendly framework for basic statistical tests
-library(DT)           # Rendering interactive data tables
+library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
 library(lme4)         # Linear mixed-effects models
+library(gtsummary)    # Publication-ready tables
 library(report)       # Easy reporting of regression analyses
@@
 library(gridExtra)    # Arrange multiple grid-based plots on a page
-library(easystats)    #
-library(gtsummary)    #
+
 ```
@@
 :::
-:::
-
-## Descriptives Overview
-
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
 
@@
 ```
-:::
+#:::
 
@@
 **This code creates a descriptives table**
+::: panel-tabset
+### Read and View Data {.tabset .tabset-fade .tabset-pills}
 
+:::
 ```{r descriptives}
@@
 :::
-:::
 
@@
 
-The code fits a linear mixed model to examine the 'Height' variable across time points ('eventname').
-It incorporates both random intercepts and slopes for the time points ('eventname') within each participant ('src_subject_id') to capture individual-level variability in both starting values and rates of change over time.
-The results of the model are then printed to provide a detailed summary of the fitted model parameters.
+The code fits a linear mixed model to predict the 'Height' variable based on time points ('eventname') and handedness ('Handedness'), while accounting for individual-level variability by including random intercepts for each participant ('src_subject_id').
+The results of the model are then printed to provide a summary of the fitted model parameters.
 
-**STEP 1: Compute LMM with Random Intercepts and Slopes**
+**STEP 1: Compute LMM Model with Random Intercepts**
 ```{r build-model-1}
@@
 #| echo: TRUE
+#| cache: true
 
-## Linear Mixed Model with a random intercept and random slope (LMM-ris)
-random_intercepts_slopes <- lmer(anthroheightcalc ~ 1 + eventname + (1|src_subject_id), data = df, REML=T)
-
-print(random_intercepts_slopes)
+## Linear Mixed Model with a random intercept (LMM-ri)
+model <- lmer(anthroweightcalc ~ 1 + eventname + sex + (1|src_subject_id), data = df, REML=T)
 
@@
 #| warning: FALSE
+#| code-summary: testing
 
-## Output and reports extending from the LMM-ris analyses
-summary(random_intercepts_slopes)
-confint(random_intercepts_slopes, level = 0.95, method = "Wald")
-report_performance(random_intercepts_slopes)
+## Output and reports extending from the LMM-ri analyses
+#summary(model)
+#confint(model, level = 0.95, method = "Wald")
+model_performance(model)
+model_parameters(model)
+model_parameters(model,standardize = "refit")
 
@@
 
-The code fits a Linear Mixed Model (LMM:ris) to predict children's height based on different time points (denoted as eventname).
-Individual variability is captured through random intercepts associated with each participant (src_subject_id).
-The results suggest a significant change in height across the time points, with the model accounting for 89% (r round(conditional_R2, 2)) of the total variation and 37% (r round(marginal_R2, 2)) attributed to the fixed effects alone.
-
-In the model, the linear term for eventname has an estimated effect of r round(model_summary$coefficients["eventname.L", "Estimate"], 2), indicating a significant increase in height across time points.
-Additionally, there's variability in initial height across participants, reflected by a random intercept standard deviation of r round(random_effects["src_subject_id", "Std.Dev."], 2).
+The code provided executes a linear mixed model (LMM) to predict children's 
+height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4), and also takes into account their sex.
+This model accounts for individual variability in height by including a random intercept for each subject (src_subject_id).
+The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.
 
@@
 
-# Assuming your model is named `random_intercepts_slope`
-# 1. Extract the random effects
-random_effects <- ranef(random_intercepts_slopes)[[1]]
+# Extract the random effects
+random_effects <- ranef(model)[[1]]
 
-# 2. Convert to dataframe
+# Convert to dataframe
 random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)
@@
 # Plot 1: Histogram
-# Plot 1: Histogram
 hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") +
-  labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
+  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") + labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
   theme_minimal()
 
-print(hist_plot)
+# Plot 2: Residuals vs Fitted Values
+resid_plot <- ggplot(NULL, aes(x = fitted_values, y = residuals)) +
+  geom_point(alpha = 0.5) +
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
+  theme_minimal()
+
+```
+
+```{r}
+plot(parameters(model)) +
+  ggplot2::labs(title = "A Dot-and-Whisker Plot")
+
+plot(model)
+theme_set(theme_modern())
+result <- estimate_expectation(model, data = "grid")
+
+plot(result)
+
+plot(result, type = "qq")
+plot(result, type = "pp")
+
+normality <- check_normality(model, effects = "random")
+plot(normality)
+
+heteroscedasticity <- check_heteroscedasticity(model)
+plot(heteroscedasticity)
+
+homogeneity <- check_homogeneity(model)
+plot(homogeneity)
+plot(homogeneity, type = "density")
+
+check_model(model)
+
+check_model(model, panel = FALSE)
+
+distribution <- check_distribution(model)
+distribution
+plot(distribution)
+
+contrasts <- estimate_contrasts(model, contrast = "sex")
+means <- estimate_means(model)
+plot(means)
+plot(contrasts, means)
+
+params <- model_parameters(model, effects = "fixed")
+plot(params)
+
+plot(params, show_labels = TRUE, size_text = 4)
 
@@
 # Extract the data frame used in the model
-model_data <- random_intercepts_slopes@frame
+model_data <- model@frame
 
@@
 
-# Plot
-ggplot(df, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +
+# Apply the recoding to the eventname variable
+df_subset$eventname <- factor(df_subset$eventname, levels = names(eventname_map), labels = eventname_map)
 
+# Verify the recoding
+table(df_subset$eventname)
+
+# Generate the plot
+ggplot(df_subset, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +
   # Individual estimated height trajectories in faded lines
   geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +
-
   # Overall group-mean trajectory in blue with increased thickness
@@
 
-The provided code visualizes individual and group-mean height trajectories over different event names.
-Individual height trajectories for each subject are depicted as faded gray lines, allowing for a clear view of the variability among subjects.
-In contrast, the overall group-mean trajectory, which represents the average trend across all individuals for each event name, is highlighted in blue.
-The average height at r mean(df_descriptable$Height) has increased by r mean(df_descriptable$Height_followup - df_descriptable$Height) units from baseline.
+```{r}
+#| echo: false
+
+## Creating variable (names) from output to be used for inline text
+
+library(broom.mixed)
+library(easystats)
+
+tidyfit <- tidy(model)
+tidyaugment <- augment(model)
+tidyglance <- glance(model)
+
+Intercept_estimate <- round(tidyfit[tidyfit$term == "(Intercept)", "estimate"], 3)
+Intercept_SE <- round(tidyfit[tidyfit$term == "(Intercept)", "std.error"], 3)
+
+sex_effect <- round(tidyfit[tidyfit$term == "sexMale or other", "estimate"], 3)
+
+# Extract values from the tibble for inline text
+Intercept_estimate <- Intercept_estimate$estimate
+Intercept_SE <- Intercept_SE$std.error
+sex_effect <-sex_effect$estimate
+
+#############
+
+#r eventname_estimate
+#r sex_effect
+#r conditional_R2
+#r conditional_R2*100**%
+#r marginal_R2
+
+```
+
+```{r}
+report(model)
+```
 
@@
 
-The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4).
-The eventname predictor was statistically significant with a p-value of r format.pval(fixed_effects$Pr(>|t|)[2], digits = 3).
-The model's overall ability to explain the variance in height was xxxxx, with a conditional R^2 of r report_performance(random_intercepts)$Conditional R2[1], indicating that it accounted for this proportion of the variability in height when considering both fixed and random effects.
-The marginal R^2 was r report_performance(random_intercepts)$Marginal R2[1], meaning that the fixed effects alone explained this proportion of the variability.
+The estimate of the random intercept was `r Intercept_estimate` (*SE* `r Intercept_SE`). The fixed effects section provides the coefficients for the intercept, eventname, and the sex variable.
+
+The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4) using the event name (eventname) and height (Height) variables.
+The intercept, corresponding to the reference levels of the predictors is estimated at `r Intercept_estimate` units, and this effect is highly significant.
+Regarding sex, boys show an increase in height of about `r sex_effect` (*SE* `r Intercept_SE`) units compared to girls.
+
+```{r}
+report_performance(model)
+```
 

### 2024-03-06 _posts/_11_Examples_MultivariateLatentGrowthCurveModels.qmd
revised manuscript docs

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Multivariate Latent Growth Curve Models"
 author: "Biostatistics Working Group"
 ---
+
+```{r}
+x <- 7;
+x;
+```
 
@@
 
-------
+--------
 
@@
 ```{r}
+# Specify the multivariate latent growth curve model
+model <- '
+  # Intercept and slope factors for variable 1
+  i_var1 =~ 1*anthroheightcalc_0 + 1*anthroheightcalc_1 + 1*anthroheightcalc_2
+  s_var1 =~ 0*anthroheightcalc_0 + 1*anthroheightcalc_1 + 2*anthroheightcalc_2
 
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
-
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
-
-  # Variances
-    i ~~ i
-    s ~~ s
-
-  # Covariance
-    i ~~ s
+  # Intercept and slope factors for variable 2
+  i_var1 =~ 1*anthroweightcalc_0 + 1*anthroweightcalc_1 + 1*anthroweightcalc_2
+  s_var1 =~ 0*anthroweightcalc_0 + 1*anthroweightcalc_1 + 2*anthroweightcalc_2
 '
@@
 ```{r}
-# Fit the model
-fit <- sem(model, data = df_wide, missing = "FIML")
+# Estimate the model
+fit <- lavaan::sem(model, data = df_wide, missing = "FIML")
 

### 2024-03-06 _posts/_8_Examples_AutoregressiveCrosslaggedPanelModels.qmd
revised manuscript docs

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Autoregressive Crosslagged Panel Models"
 author: "Biostatistics Working Group"
@@
 
-# XXXXXXXXXXX {.unnumbered}
+# Autoregressive Crosslagged Panel Models {.unnumbered}
 
@@
 
-XXXXXXXXXXX Models xxxxxxx. This xxxxxxx.
+Autoregressive Crosslagged Panel (ARCL) Models xxxxxxx. This xxxxxxx.
 
-In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
+In this example, we will utilize ARCL models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
 Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time.
-XXXX models facilitate this by xxxxx.
+ARCL models facilitate this by xxxxx.
 
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT")
+packages_required <- c("tidyverse","rstatix","DT","lavaan")
 
@@
 library(DT)           # Rendering interactive data tables
+library(lavaan)       # Structural equation modeling package
 
@@
 
-------
-
 **STEP 1: Convert dataframe from long-to-wide format**
@@
 **STEP 2: Compute xxxxx**
+```{r}
+
+model <- '
+  # Autoregressive paths
+  anthroweightcalc_4 ~ a1 * anthroweightcalc_3
+  anthroweightcalc_3 ~ a2 * anthroweightcalc_2
+  anthroweightcalc_2 ~ a1 * anthroweightcalc_1
+  anthroweightcalc_1 ~ a2 * anthroweightcalc_0
+  anthroheightcalc_4 ~ a1 * anthroheightcalc_3
+  anthroheightcalc_3 ~ a2 * anthroheightcalc_2
+  anthroheightcalc_2 ~ a1 * anthroheightcalc_1
+  anthroheightcalc_1 ~ a2 * anthroheightcalc_0
+
+# Cross-lagged paths
+anthroheightcalc_4 ~ b1 * anthroweightcalc_3
+anthroheightcalc_3 ~ b1 * anthroweightcalc_2
+anthroheightcalc_2 ~ b1 * anthroweightcalc_1
+anthroheightcalc_1 ~ b2 * anthroweightcalc_0
+
+anthroweightcalc_4 ~ b1 * anthroheightcalc_3
+anthroweightcalc_3 ~ b1 * anthroheightcalc_2
+anthroweightcalc_2 ~ b1 * anthroheightcalc_1
+anthroweightcalc_1 ~ b2 * anthroheightcalc_0
+'
+
+```
+
+Fit the ACPM using the lavaan function:
 
 ```{r}
+# Fit the model to your data (replace 'your_data' with the name of your data frame)
+fit <- sem(model, data = df_wide)
 
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
-
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
-
-  # Variances
-    i ~~ i
-    s ~~ s
-
-  # Covariance
-    i ~~ s
-'
-```
-
-```{r}
-# Fit the model
-fit <- sem(model, data = df_wide, missing = "FIML")
-
-summary(fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)
+# View the results
+summary(fit)
 

### 2024-03-06 _posts/_9_Examples_LatentChangeScoresModels.qmd
revised manuscript docs

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Latent Change Score Models"
 author: "Biostatistics Working Group"
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT")
+packages_required <- c("tidyverse","rstatix","DT", "lcsm","lavaan")
 
@@
 library(DT)           # Rendering interactive data tables
-
+library(lcsm)         #
+library(lavaan)       #
 ```
@@
 
-------
+----------
 
@@
 
+###########
+###########
+###########
+
 ```{r}
 
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
+uni_lavaan_syntax <- fit_uni_lcsm(data = df_wide, 
+             var =  c("anthroweightcalc_0", "anthroweightcalc_1",
+                      "anthroweightcalc_2","anthroweightcalc_3"),
+             model = list(alpha_constant = TRUE, 
+                          beta = TRUE, 
+                          phi = TRUE),
+                          return_lavaan_syntax = TRUE)
 
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
+# Return lavaan syntax in easy to read format
+cat(uni_lavaan_syntax)
 
-  # Variances
-    i ~~ i
-    s ~~ s
+# Now extract parameter estimates
+# Only extract first 7 columns for this example by adding [ , 1:7]
+param_bi_lcsm_01 <- extract_param(fit_uni_lcsm, printp = TRUE)[ , 1:4]
 
-  # Covariance
-    i ~~ s
-'
+# Print table of parameter estimates
+kable(param_bi_lcsm_01, digits = 3)
+
 ```
@@
 ```{r}
-# Fit the model
-fit <- sem(model, data = df_wide, missing = "FIML")
 
-summary(fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)
+library(knitr)
+
+# Fit bivariate lcsm and save the results
+uni_lavaan_results <- fit_uni_lcsm(data = df_wide, 
+             var =  c("anthroweightcalc_0", "anthroweightcalc_1",
+                    "anthroweightcalc_2","anthroweightcalc_3",
+                    "anthroweightcalc_4"),
+             model = list(alpha_constant = TRUE, 
+                          beta = TRUE, 
+                          phi = TRUE)
+             )
+
+# Save the lavaan syntax that is used to create the layout matrix for semPlot
+uni_lavaan_syntax <- fit_uni_lcsm(data = df_wide, 
+             var =  c("anthroweightcalc_0", "anthroweightcalc_1",
+                    "anthroweightcalc_2","anthroweightcalc_3",
+                    "anthroweightcalc_4"),
+             model = list(alpha_constant = TRUE, 
+                          beta = TRUE, 
+                          phi = TRUE),
+             return_lavaan_syntax = TRUE)
+
+# Plot the results
+plot_lcsm(lavaan_object = uni_lavaan_results,
+          lavaan_syntax = uni_lavaan_syntax,
+          edge.label.cex = .9,  
+          lcsm_colours = TRUE,
+          lcsm = "univariate")
+
+# Now extract parameter estimates
+# Only extract first 7 columns for this example by adding [ , 1:7]
+param_bi_lcsm_01 <- extract_param(uni_lavaan_results, printp = TRUE)[ , 1:6]
+
+# Print table of parameter estimates
+kable(param_bi_lcsm_01, digits = 3)
+
+# Create plot for construct x
+plot_x <- plot_trajectories(data = df_wide,
+                            id_var = "src_subject_id", 
+                            var_list = c("anthroweightcalc_0",
+                                         "anthroweightcalc_1",
+                                         "anthroweightcalc_2",
+                                         "anthroweightcalc_3",
+                                         "anthroweightcalc_4"),
+                            xlab = "Time", ylab = "X Score",
+                            connect_missing = FALSE, 
+                            random_sample_frac = 0.3)
+
+plot_x
 

### 2024-03-06 examples/examples.qmd
revised manuscript docs

--- 2024-03-02 examples.qmd
@@
 ::: {.g-col-4}
-```{=html}
-<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css">
 
-<div class="section_our_solution">
-  <div class="row">
-    <div class="col-lg-12 col-md-12 col-sm-12">
-      <div class="our_solution_category">
-        <div class="solution_cards_box">
-          <div class="solution_card">
-            <div class="hover_color_bubble"></div>
-            <div class="so_top_icon">
-              <svg id="Layer_1" enable-background="new 0 0 512 512" height="50" viewBox="0 0 512 512" width="40" xmlns="http://www.w3.org/2000/svg">
-                <g>
-                  <g>
-                    <g>
-                      <g>
-                        <path d="m47.478 452.317 295.441 19.76c5.511.369 10.277-3.8 10.645-9.31l28.393-424.517c.369-5.511-3.8-10.276-9.31-10.645l-295.441-19.76c-5.511-.369-10.276 3.8-10.645 9.31l-28.394 424.517c-.368 5.511 3.8 10.277 9.311 10.645z" fill="#fae19e" />
-                      </g>
-                      <g>
-                        <g>
-                          <g>
-                            <g>
-                              <g>
-                                <path d="m17.5 504.177h226.14l79.96-79.605v-355.86c0-5.523-4.477-10-10-10h-296.1c-5.523 0-10 4.477-10 10v425.466c0 5.522 4.477 9.999 10 9.999z" fill="#fff9e9" />
-                              </g>
-                              <path d="m313.601 58.712h-40c5.523 0 10 4.477 10 10v355.861l-.258 40.078 40.258-40.078v-355.861c0-5.523-4.477-10-10-10z" fill="#fff4d6" />
-                            </g>
-                          </g>
-                        </g>
-                        <path d="m243.64 504.177v-70.253c0-5.523 4.477-10 10-10h69.96z" fill="#ffeec2" />
-                      </g>
-                    </g>
-                    <g>
-                      <path d="m468.636 248.58-33.372.165v-50.826c0-9.183 7.463-16.662 16.673-16.708h.007c9.217-.046 16.693 7.371 16.693 16.562v50.807z" fill="#fed23a" />
-                      <path d="m451.96 504.177c-10.362-10.277-16.196-24.263-16.208-38.857l-.062-73.973c0-.644.524-1.169 1.171-1.173l30.038-.149c.647-.003 1.171.517 1.171 1.161l.062 74.079c.012 14.531-5.749 28.472-16.015 38.756z" fill="#54b1ff" />
-                      <path d="m451.959 469.333h-.01c-14.434.072-26.14-11.542-26.14-25.935v-213.527c0-6.778 5.477-12.283 12.255-12.316l27.626-.137c6.826-.034 12.378 5.49 12.378 12.316v213.436c0 14.38-11.687 26.091-26.109 26.163z" fill="#fdf385" />
-                      <path d="m465.69 217.417-23.769.118c6.037.79 10.708 5.94 10.708 12.198v213.437c0 9.823-5.455 18.397-13.507 22.87 3.79 2.115 8.164 3.317 12.826 3.293h.01c14.422-.072 26.109-11.783 26.109-26.163v-213.436c.001-6.826-5.551-12.351-12.377-12.317z" fill="#faee6e" />
-                      <path d="m491.274 247.925-71.615.355c-7.305.036-13.226 5.968-13.226 13.248 0 7.281 5.921 13.153 13.226 13.117l58.389-.29v77.489c0 7.281 5.921 13.153 13.226 13.117 7.305-.036 13.226-5.968 13.226-13.248v-90.672c0-7.28-5.922-13.152-13.226-13.116z" fill="#54b1ff" />
-                      <g>
-                        <path d="m491.274 247.925-38.441.188-.167 26.311 25.381-.067v77.489c0 7.281 5.921 13.153 13.226 13.117 7.305-.036 13.226-5.968 13.226-13.248v-90.672c.001-7.282-5.921-13.154-13.225-13.118z" fill="#3da7ff" />
-                      </g>
-                    </g>
-                  </g>
-                  <g fill="#060606">
-                    <path d="m373.147 20.122-295.44-19.761c-9.631-.638-17.984 6.665-18.629 16.293l-2.311 34.557h-39.267c-9.649 0-17.5 7.851-17.5 17.5v425.466c0 9.649 7.851 17.5 17.5 17.5h226.141c1.96 0 3.902-.801 5.292-2.185l34.138-33.987c.347.074.701.133 1.065.157l58.282 3.898c9.302.614 18.005-6.952 18.629-16.293l28.393-424.515c.639-9.528-6.766-17.993-16.293-18.63zm-122.006 465.902v-52.1c0-1.378 1.122-2.5 2.5-2.5h51.9zm94.939-23.757c-.244 1.51-1.131 2.286-2.66 2.327l-46.28-3.096 31.752-31.611c1.414-1.407 2.209-3.32 2.209-5.315v-355.86c0-9.649-7.851-17.5-17.5-17.5h-77.993c-9.697 0-9.697 15 0 15h77.993c1.379 0 2.5 1.122 2.5 2.5v347.712h-62.46c-9.649 0-17.5 7.851-17.5 17.5v62.753h-218.641c-1.378 0-2.5-1.122-2.5-2.5v-425.465c0-1.378 1.122-2.5 2.5-2.5h178.168c9.697 0 9.697-15 0-15h-123.868l2.244-33.556c.244-1.511 1.131-2.286 2.661-2.327l295.44 19.76c1.511.244 2.287 1.131 2.328 2.661z" />
-                    <path d="m267.827 237.047h-204.553c-4.142 0-7.5 3.358-7.5 7.5s3.358 7.5 7.5 7.5h204.553c4.143 0 7.5-3.358 7.5-7.5s-3.357-7.5-7.5-7.5z" />
-                    <path d="m267.827 289.332h-204.553c-4.142 0-7.5 3.358-7.5 7.5s3.358 7.5 7.5 7.5h204.553c4.143 0 7.5-3.358 7.5-7.5s-3.357-7.5-7.5-7.5z" />
-                    <path d="m55.774 192.262c0 4.142 3.358 7.5 7.5 7.5h204.553c4.143 0 7.5-3.358 7.5-7.5s-3.357-7.5-7.5-7.5h-204.553c-4.142 0-7.5 3.358-7.5 7.5z" />
-                    <path d="m91.807 139.977c0 4.142 3.358 7.5 7.5 7.5h132.487c4.142 0 7.5-3.358 7.5-7.5s-3.358-7.5-7.5-7.5h-132.487c-4.142 0-7.5 3.358-7.5 7.5z" />
-                    <path d="m194.755 438.787c-13.489.036-26.978.065-40.467.086-4.534.007-9.067.013-13.6.016-8.215.006-13.75-1.643-15.59-10.679-1.556-7.64-12.364-6.613-14.464 0-5.19 16.337-13.774 9.936-18.582-1.053-4.797-10.963-6.027-23.233-8.122-34.9-1.54-8.573-14.506-6.17-14.732 1.994-.298 10.751-1.302 21.331-4.031 31.758-2.815 10.758-7.034 21.097-11.222 31.376-3.651 8.961 10.867 12.816 14.464 3.988 3.711-9.108 7.427-18.266 10.193-27.714 5.14 12.36 15.774 26.34 30.927 18.101 2.819-1.533 5.452-3.712 7.763-6.253 7.88 9.106 19.609 8.388 30.584 8.375 15.627-.02 31.254-.054 46.881-.095 9.649-.025 9.667-15.025-.002-15z" />
-                    <path d="m505.932 246.439c-3.897-3.878-9.255-5.867-14.695-6.014l-5.668.028v-10.719c0-6.529-3.878-13.427-9.433-16.862v-15.098c0-31.069-48.372-30.934-48.372.146v15.1c-5.659 3.498-9.455 9.741-9.455 16.852v10.982c-24.966 1.7-25.037 39.745.028 41.232.16 33.575.152 66.6-.028 100.737-.049 9.414 14.949 9.966 15 .079.18-34.166.188-67.22.029-100.823l37.211-.185s-.048 110.848-.048 160.784c0 24.338-37.219 24.5-37.219-.253l.013-13.677c.585-9.68-14.387-10.583-14.973-.904v12.834c0 11 3.402 20.316 9.988 26.869.586 15.693 7.198 30.878 18.369 41.956 3.205 3.18 7.642 2.208 10.744-.182 11.365-11.385 17.769-26.394 18.169-42.414 4.951-4.931 9.908-9.896 9.908-26.896l.006-68.351c12.97 3.689 26.494-6.348 26.494-19.946v-90.672c0-5.523-2.155-10.709-6.068-14.603zm-72.623-5.727v-10.841c0-2.219 1.523-4.08 3.573-4.633l30.025-.149c.84.208 1.615.605 2.243 1.231.915.911 1.419 2.123 1.419 3.414v10.794zm18.671-52c4.604 0 9.155 4.514 9.155 9.062v12.166l-18.372.091v-12.111c.001-5.053 4.133-9.183 9.217-9.208zm-.011 303.901c-3.487-4.942-6.009-10.531-7.417-16.406 2.322.503 4.674.765 7.027.765 2.627 0 5.253-.326 7.839-.957-1.374 5.964-3.892 11.587-7.449 16.598zm45.031-140.899c0 7.101-11.452 7.66-11.452.131 0 0 .013-70.974.021-77.48.005-4.196-3.483-7.509-7.558-7.509l-58.389.29c-7.242 0-7.073-11.331.074-11.366l71.615-.355c3.463.295 5.359 2.168 5.688 5.617v90.672z" />
-                  </g>
-                </g>
-              </svg>
-            </div>
-            <div class="solu_description">
-              <p>
-                For an entry-level example of each longitudinal model
-              </p>
-              <button type="button" class="read_more_btn">Go Here</button>
-            </div>
-          </div>
-          </div>
-          </div>
-        </div>
-      </div>
-    </div>
-```
 :::
@@
 		<article class="postcard light blue">
-			<a class="postcard__img_link" href="#">
+			<!-- <a class="postcard__img_link" href="#">
 				<img class="postcard__img" src="https://picsum.photos/1000/1000" alt="Image Title" />
-			</a>
+			</a> -->
 			<div class="postcard__text t-dark">
@@
 		<article class="postcard light red">
-			<a class="postcard__img_link" href="#">
+			<!-- <a class="postcard__img_link" href="#">
 				<img class="postcard__img" src="https://picsum.photos/501/500" alt="Image Title" />	
-			</a>
+			</a> -->
 			<div class="postcard__text t-dark">
@@
 		<article class="postcard light red">
-			<a class="postcard__img_link" href="#">
+			<!-- <a class="postcard__img_link" href="#">
 				<img class="postcard__img" src="https://picsum.photos/601/600" alt="Image Title" />	
-			</a>
+			</a> -->
 			<div class="postcard__text t-dark">
@@
 		<article class="postcard light red">
-			<a class="postcard__img_link" href="#">
+			<!-- <a class="postcard__img_link" href="#">
 				<img class="postcard__img" src="https://picsum.photos/500/653" alt="Image Title" />	
-			</a>
+			</a> -->
 			<div class="postcard__text t-dark">
@@
 		<article class="postcard light red">
-			<a class="postcard__img_link" href="#">
+			<!-- <a class="postcard__img_link" href="#">
 				<img class="postcard__img" src="https://picsum.photos/801/800" alt="Image Title" />	
-			</a>
+			</a> -->
 			<div class="postcard__text t-dark">

### 2024-03-06 manuscript/manuscript-template.qmd
revised manuscript docs

--- 2024-03-04 manuscript.qmd
@@
 # Set the URL for the CSV file
-file_url <- "https://drive.google.com/uc?export=download&id=1QFdAFz09DQEeFVOj8bRWh4g6hK4b4Qj1"
+# file_url <- "https://drive.google.com/uc?export=download&id=1QFdAFz09DQEeFVOj8bRWh4g6hK4b4Qj1"
+
+file_url <- "https://fiudit-my.sharepoint.com/:x:/r/personal/shawes_fiu_edu/Documents/abcd-data/df_preclean.csv?d=wb8eb1be9fdce4edcbbbca8f8eb804633&csf=1&web=1&e=LiuHzu"
 
@@
 ## Introduction
-_Introduce the research topic, state the research problem, and outline the study's objectives. Provide background information to set the context. This is an exmaple in-text citation_ @netwok2020.
+_Introduce the research topic, state the research problem, and outline the study's objectives. Provide background information to set the context. This is an exmaple in-text citation_ @kenny2001.
 `r lorem::ipsum(paragraphs = 3)`

### 2024-03-06 examples/_queue/_13_Examples_GrowthMixtureModels.qmd
config updates

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Growth Mixture Models"
 author: "Biostatistics Working Group"
@@
 
-------
+**STEP 1: XXXXXXXXX **
 
-**STEP 1: Convert dataframe from long-to-wide format**
-```{r long-to-wide}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
+----
 
-# Split the data into static (unchanging) and changing variables
-# Static variables are taken only from the baseline measurement occasion
-df_static <- df %>%
-  filter(eventname == "baseline_year_1_arm_1") %>%
-  select(src_subject_id, site_id_l, sex)
-
-df_changing <- df %>%
-  select(src_subject_id, eventname, anthroweightcalc, anthroheightcalc)
-
-# Convert the changing variables from long format to wide format
-df_wide_changing <- df_changing %>%
-  gather(variable, value, -src_subject_id, -eventname) %>%
-  unite("variable_event", variable, eventname) %>%
-  spread(variable_event, value)
-
-# Merge the static and changing data frames based on the 'src_subject_id' variable
-df_wide <- df_static %>%
-  left_join(df_wide_changing, by = "src_subject_id")
-
-# Rename columns to shorter names for easier reference
-df_wide <- df_wide %>% 
-        rename("anthroweightcalc_0"="anthroweightcalc_baseline_year_1_arm_1",
-               "anthroweightcalc_1"="anthroweightcalc_1_year_follow_up_y_arm_1",
-               "anthroweightcalc_2"="anthroweightcalc_2_year_follow_up_y_arm_1",
-               "anthroweightcalc_3"="anthroweightcalc_3_year_follow_up_y_arm_1",
-               "anthroweightcalc_4"="anthroweightcalc_4_year_follow_up_y_arm_1",
-               "anthroheightcalc_0"="anthroheightcalc_baseline_year_1_arm_1",
-               "anthroheightcalc_1"="anthroheightcalc_1_year_follow_up_y_arm_1",
-               "anthroheightcalc_2"="anthroheightcalc_2_year_follow_up_y_arm_1",
-               "anthroheightcalc_3"="anthroheightcalc_3_year_follow_up_y_arm_1",
-               "anthroheightcalc_4"="anthroheightcalc_4_year_follow_up_y_arm_1"
-               )
-
-# Reorder columns in the dataframe
-df_wide <- df_wide %>% select(src_subject_id, site_id_l, sex, anthroweightcalc_0, anthroweightcalc_1, anthroweightcalc_2, anthroweightcalc_3, anthroweightcalc_4, anthroheightcalc_0, anthroheightcalc_1, anthroheightcalc_2, anthroheightcalc_3, anthroheightcalc_4)
-
+```{r}
+# Combine the data from all classes
+data <- rbind(class1, class2, class3)
 ```
 
-The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname').
-It incorporates xxxxxx.
-The results of the model are then printed to provide a detailed summary of the fitted model parameters.
-
-**STEP 2: Compute xxxxx**
-
-```{r}
-
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
-
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
-
-  # Variances
-    i ~~ i
-    s ~~ s
-
-  # Covariance
-    i ~~ s
-'
+### Model Specification and Estimation
+Now that we have our simulated data, we can fit a Growth Mixture Model.
+In this example, we will assume that there are three latent classes.
+However, in practice, the number of classes is usually unknown and must be determined through model comparisons or other methods.
+```{r, eval=FALSE}
+# Fit the Growth Mixture Model
+gmm <- hlme(value ~ time,
+            random = ~ time,
+            subject = "id",
+            mixture = ~ time,
+            data = data,
+            ng = 3,
+            B = list(value = c(1, 0, 0, 0)
 ```

### 2024-03-06 examples/_queue/_15_Examples_RandomInterceptCrosslaggedPanelModels.qmd
config updates

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Random-Intercept Crosslagged Panel Models"
 author: "Biostatistics Working Group"
@@
 
-```{r}
+```{r, eval=FALSE}
+# Define the model
+model <- '
+  # Random intercepts
+  i_X =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3 + 1 * anthroheightcalc_4
+  i_Y =~ 1 * anthroweightcalc_0 + 1 * anthroweightcalc_1 + 1 * anthroweightcalc_2 + 1 * anthroweightcalc_3 + 1 * anthroweightcalc_4
 
-# Define the Latent Growth Curve Model
-model <- '
-  # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
+  # Cross-lagged effects
+  anthroweightcalc_1 ~ beta_YX * anthroheightcalc_0
+  anthroweightcalc_2 ~ beta_YX * anthroheightcalc_1
+  anthroweightcalc_3 ~ beta_YX * anthroheightcalc_2
+  anthroweightcalc_4 ~ beta_YX * anthroheightcalc_3
 
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
+  anthroheightcalc_1 ~ beta_XY * anthroweightcalc_0
+  anthroheightcalc_2 ~ beta_XY * anthroweightcalc_1
+  anthroheightcalc_3 ~ beta_XY * anthroweightcalc_2
+  anthroheightcalc_4 ~ beta_XY * anthroweightcalc_3
 
-  # Variances
-    i ~~ i
-    s ~~ s
+  # Auto-regressive paths (optional but often included in CLPM)
+  anthroheightcalc_1 ~ alpha_X * anthroheightcalc_0
+  anthroheightcalc_2 ~ alpha_X * anthroheightcalc_1
+  anthroheightcalc_3 ~ alpha_X * anthroheightcalc_2
+  anthroheightcalc_4 ~ alpha_X * anthroheightcalc_3
 
-  # Covariance
-    i ~~ s
+  anthroweightcalc_1 ~ alpha_Y * anthroweightcalc_0
+  anthroweightcalc_2 ~ alpha_Y * anthroweightcalc_1
+  anthroweightcalc_3 ~ alpha_Y * anthroweightcalc_2
+  anthroweightcalc_4 ~ alpha_Y * anthroweightcalc_3
+
+  # Covariances (optional but can be included for model fit)
+  anthroheightcalc_0 ~~ anthroweightcalc_0
+  i_X ~~ i_Y
   '
@@
 ```{r}
-# Fit the model
+# Estimate the RI-CLPM 
 fit <- sem(model, data = df_wide, missing = "FIML")

### 2024-03-06 examples/_queue/_16_Examples_LatentCurveModelsStructuredResiduals.qmd
config updates

--- 2024-03-02 posts/_10_Examples_LatentGrowthCurveModels.qmd
@@
 ---
-title: "Latent Growth Curve Models"
+title: "Latent Curve Models with Structured Residuals"
 author: "Biostatistics Working Group"
@@
 
-# Define the Latent Growth Curve Model
 model <- '
   # Latent variables
-    i =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2
-    s =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2
+  intercept =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3 + 1 * anthroheightcalc_4
+  slope =~ 0 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 2 * anthroheightcalc_2 + 3 * anthroheightcalc_3 + 4 * anthroheightcalc_4
 
-  # Estimate means of latent variables
-    i ~ 1
-    s ~ 1
+  # Time-varying covariate effect on outcome
+  anthroheightcalc_0 ~ c1 * anthroweightcalc_0
+  anthroheightcalc_1 ~ c2 * anthroweightcalc_1
+  anthroheightcalc_2 ~ c3 * anthroweightcalc_2
+  anthroheightcalc_3 ~ c4 * anthroweightcalc_3
+  anthroheightcalc_4 ~ c5 * anthroweightcalc_4
 
-  # Variances
-    i ~~ i
-    s ~~ s
+  # Residual variances and covariances
+  anthroheightcalc_0 ~~ r1 * anthroheightcalc_0
+  anthroheightcalc_1 ~~ r2 * anthroheightcalc_1
+  anthroheightcalc_2 ~~ r3 * anthroheightcalc_2
+  anthroheightcalc_3 ~~ r4 * anthroheightcalc_3
+  anthroheightcalc_4 ~~ r5 * anthroheightcalc_4
 
-  # Covariance
-    i ~~ s
+  # Latent variable covariances
+  intercept ~~ i_s_cov * slope
 '
+
 ```
@@
 ```{r}
-# Fit the model
+# Estimate the LCM-SR Model
 fit <- sem(model, data = df_wide, missing = "FIML")

### 2024-03-06 examples/_queue/_1a_Examples_DifferenceScores_PairedTtests.qmd
config updates

---
title: "Difference Scores: Paired Samples T-test"
author: "Biostatistics Working Group"
---

## Overview

<p>

This example assesses growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1), using weight as a representative metric. The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 weight measurements for each participant; 2) a one-sample t-test is used to test whether the difference score is statistically different than a null hypothesis of zero change. Finally, a visual inspection is further conducted via a scatterplot to graphically represent the relationship between participant's weight at Baseline and Year_1. The ensuing analysis and interpretations are detailed in the subsequent sections.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page

```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroweightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroweightcalc = as.numeric(anthroweightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

:::
:::

## Results

::: panel-tabset
### Build Model: Step I {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to compute a difference score by subtracting each participant's weight at T1 from their weight at T0. Relevant summary statistics are also provided.

**STEP 1: Compute Difference Score**

```{r build-model-ttest-1}
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
#| class.source: 'fold-hide'

# Function to compute difference scores
compute_difference <- function(df, variable_name) {
    # Define the event names of interest
    baseline_event <- "baseline_year_1_arm_1"
    followup_event <- "1_year_follow_up_y_arm_1"

    # Compute the difference between Baseline and Year 1 data for the given variable
    diff_data <- df %>%
        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
        spread(eventname, variable_name) %>% # Convert data from long to wide format
        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
        mutate(group_resid = (diff) - mean(diff)) %>% # compute residuals (group_resid).
      drop_na(diff) # Exclude rows with NA in the computed difference

    # Summarize the computed difference scores
    diff_summary <- summary(diff_data$diff)

    # Return the difference data and its summary
    list(data = diff_data, summary = diff_summary)
}

# List of variables for which difference scores are to be computed
variables_of_interest <- c("anthroweightcalc")

# Compute and store difference scores and summaries for each variable in a list
difference_and_summary_list <- lapply(variables_of_interest, function(var) {
    compute_difference(df, var)
})

# Extract the difference data for the 'anthroweightcalc' variable
weight_diff_data <- difference_and_summary_list[[1]]$data

# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
df <- left_join(df, weight_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")

```

**Descriptive Statistics for the Difference Score**
```{r model-checking}
# Compute statistical summaries for the difference score variable
lapply(difference_and_summary_list, function(item) {
    print(item$summary)
})

gaze <- df %>%
  mutate(diff = baseline_event - followup_event) %>%
  mutate(group_resid = diff - mean(diff))

gaze <- weight_diff_data %>%
  mutate(diff1 = baseline_year_1_arm_1 - 1_year_follow_up_y_arm_1) %>%
  mutate(group_resid = diff - mean(diff))

dff.y

qqPlot(gaze_residual$group_resid)

shapiro.test(gaze_residual$group_resid)

```

This summary of the difference score variable indicates that the differences range from a decrease of `r min(weight_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(weight_diff_data$diff, na.rm = TRUE)` units. The median difference is `r median(weight_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(weight_diff_data$diff, digits = 2, na.rm = TRUE)` units. There are `r sum(is.na(weight_diff_data$diff))` missing values in this difference score dataset.

**Descriptive Statistics for the weight Variable**
```{r}
## Summary statistics
# Compute summary statistics for weight by eventname
summary <- df %>%
    group_by(eventname) %>%
    get_summary_stats(anthroweightcalc, type = "mean_sd")

data.frame(summary)
```

The summary statistics provide insights into the weight variable across different `eventname` categories. At baseline, the average weight is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`. Over the years, there’s a noticeable increase in average weight: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year. The standard deviation remains relatively consistent over the years, suggesting similar variability in weight measurements across these time points.

:::

### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}

::: blue

A one-sample t-test is computed to examine whether the average difference score is different than zero (indicating change).

**STEP 2: Conduct t-test on Difference Score**

```{r build-model-ttest-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Extract the difference scores for weight from the computed list
diff_data <- difference_and_summary_list[[1]]$data$diff

# Perform a one-sample t-test on the difference scores for weight
test_result <- t.test(diff_data, mu = 0, na.rm = TRUE)

test_result

gaze_test <- t.test(anthroweightcalc ~ eventname, 
                    paired = TRUE, 
                    data = df) %>% 
  tidy()

```

This one-sample t-test evaluates whether the average difference score deviates significantly from 0. The model output provides:

i. a t-statistic of `r test_result$statistic`;
ii. degrees of freedom of `r test_result$parameter`;
iii. a p-value of `r test_result$p.value`;
iv. a mean estimate of `r test_result$estimate["mean of x"]`;
v. 95% confidence intervals from `r test_result$conf.int[1]` to `r test_result$conf.int[2]`.

v. 95% confidence intervals from `r round(test_result$conf.int[1], 3)` to `r round(test_result$conf.int[2], 3)`.

With a t-value of `r test_result$statistic` and a p-value of `r test_result$p.value`, we can determine the significance of the difference. The average difference score across the sample is approximately `r round(test_result$estimate["mean of x"], 2)`. Furthermore, we can be 95% confident that the true mean difference score in the population lies between approximately `r round(test_result$conf.int[1], 2)` and `r round(test_result$conf.int[2], 2)`. This indicates the degree to which there's a change in weight from Baseline to Year 1.

:::

### Model Plots

::: blue
```{r scatterplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Filter data for Baseline
baseline_data <- df %>% filter(eventname == "Baseline")

# Filter data for Year_1 and join with Baseline data
joined_data <- baseline_data %>%
    left_join(df %>% filter(eventname == "Year_1"), by = "src_subject_id", suffix = c("_baseline", "_year1"))

# Create scatterplot
scatterplot <- ggplot(df, aes(x = anthroweightcalc, y = diff)) +
    geom_point(aes(color = sex), alpha = 0.5) + # Setting alpha to ensure points are visible
    geom_smooth(method = "lm", color = "red") + # Adding linear regression line
    labs(
        title = "Scatterplot of anthroweightcalc vs. Difference Score",
        x = "anthroweightcalc",
        y = "Difference Score"
    ) +
    theme_minimal()

scatterplot

```

The scatterplot visually depicts the relationship between individuals' weights at baseline and their weights at Year_1. Each point on the plot represents an individual, with their baseline weight plotted on the x-axis and their Year_1 weight on the y-axis. A noticeable positive linear trend can be observed, as highlighted by the blue regression line, indicating that those who were taller at baseline generally remained taller at Year_1. The strength and direction of this relationship can be further quantified with correlation coefficients, but visually, the data suggests a strong positive association between baseline and Year_1 weights.

:::
:::

## Wrapping Up

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue
In the study sample, the average weight at baseline was approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`, which showed a noticeable increase to an average of `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units by Year_1. A one-sample t-test was conducted to determine if the mean difference in weight from baseline to Year_1 significantly deviated from zero. The results indicated a statistically significant increase with a mean difference of approximately `r round(test_result$estimate, 2)` units (`r round(test_result$conf.int[1], 2)` to `r round(test_result$conf.int[2], 2)`), t(`r test_result$parameter`) = `r round(test_result$statistic, 3)`, p < `r test_result$p.value`. Further, a scatterplot visualizing the relationship between baseline and Year_1 weights showed a strong positive linear trend. This suggests that participants who were taller at baseline generally remained taller at Year_1, reaffirming the consistent growth trend observed in the data.

:::
:::

### 2024-03-06 examples/_queue/_5_Examples_MarginalModels.qmd
config updates

--- 2024-03-02 posts/3b_Examples_LinearMixedModels.qmd
@@
 ---
-title: "Linear Mixed Models: Random Intercept and Slope"
+title: "Marginal Models"
 author: "Biostatistics Working Group"
-
 ---
 
-# LMM: Intercept & Slope {.unnumbered}
+# Marginal Models {.unnumbered}
 
@@
 
-The Linear Mixed Model with a random intercept and slope (LMM:ris) extends the traditional fixed-effect linear regression by incorporating both a subject-specific random intercept and a random slope.
-This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.
+Marginal models xxxxxxx. This xxxxxxx.
 
-In this example, we will utilize the LMM:ris to analyze height trajectories obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
-Our primary objective is to understand the stability and evolution in height measurements, while factoring in the clustered nature of observations within individuals over time.
-The LMM:ris facilitates this by concurrently modeling an overarching sample mean trajectory (fixed effect) and individual variations (random effects) in both starting points (intercepts) and growth rates (slopes) around this mean trajectory.
+In this example, we will utilize marginal models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
+Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time.
+Marginal models facilitate this by xxxxx.
 
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT","lme4","report","broom","gridExtra","easystats","gtsummary")
+packages_required <- c("tidyverse","lme4","geepack", "report",
+                       "broom","gridExtra","easystats","gtsummary",
+                       "marginaleffects")
 
@@
 library(tidyverse)    # Collection of R packages for data science
-library(rstatix)      # Pipe-friendly framework for basic statistical tests
-library(DT)           # Rendering interactive data tables
 library(lme4)         # Linear mixed-effects models
+library(geepack)      # Generalized Estimating Equations (GEE) analysis
 library(report)       # Easy reporting of regression analyses
@@
 library(gtsummary)    #
+library(marginaleffects) #
 ```
@@
 
-The code fits a linear mixed model to examine the 'Height' variable across time points ('eventname').
-It incorporates both random intercepts and slopes for the time points ('eventname') within each participant ('src_subject_id') to capture individual-level variability in both starting values and rates of change over time.
+The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname').
+It incorporates xxxxxx.
 The results of the model are then printed to provide a detailed summary of the fitted model parameters.
 
-**STEP 1: Compute LMM with Random Intercepts and Slopes**
+**STEP 1: Compute xxxxx**
 ```{r build-model-1}
@@
 
-## Linear Mixed Model with a random intercept and random slope (LMM-ris)
-random_intercepts_slopes <- lmer(anthroheightcalc ~ 1 + eventname + (1|src_subject_id), data = df, REML=T)
-
-print(random_intercepts_slopes)
-
+# Fitting the longitudinal marginal model
+model <- geeglm(anthroweightcalc ~ eventname, data = df, id = src_subject_id, family = gaussian(), corstr = "exchangeable")
+#model <- geeglm(Job_Satisfaction ~ TimePoint, data = data, id = Individual, family = #binomial(link = "logit"), corstr = "exchangeable")
 ```
 
-```{r model-output}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-## Output and reports extending from the LMM-ris analyses
-summary(random_intercepts_slopes)
-confint(random_intercepts_slopes, level = 0.95, method = "Wald")
-report_performance(random_intercepts_slopes)
-
+```{r}
+summary(model)
 ```
-
-The code fits a Linear Mixed Model (LMM:ris) to predict children's height based on different time points (denoted as eventname).
-Individual variability is captured through random intercepts associated with each participant (src_subject_id).
-The results suggest a significant change in height across the time points, with the model accounting for 89% (r round(conditional_R2, 2)) of the total variation and 37% (r round(marginal_R2, 2)) attributed to the fixed effects alone.
-
-In the model, the linear term for eventname has an estimated effect of r round(model_summary$coefficients["eventname.L", "Estimate"], 2), indicating a significant increase in height across time points.
-Additionally, there's variability in initial height across participants, reflected by a random intercept standard deviation of r round(random_effects["src_subject_id", "Std.Dev."], 2).
-
-### Model Plots
-
-The following set of plots are used to facilitate model diagnostics.
-The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects.
-The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity.
-The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.
-
-::: blue
-```{r diagnosticplots}
-
-# Assuming your model is named `random_intercepts_slope`
-# 1. Extract the random effects
-random_effects <- ranef(random_intercepts_slopes)[[1]]
-
-# 2. Convert to dataframe
-random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)
-
-# Plot 1: Histogram
-# Plot 1: Histogram
-hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") +
-  labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
-  theme_minimal()
-
-print(hist_plot)
-
-```
-
-```{r trajectoryplot}
-
-# Extract the data frame used in the model
-model_data <- random_intercepts_slopes@frame
-
-# Extract unique subject IDs from the model's data
-original_subject_ids <- unique(model_data$src_subject_id)
-
-# Subset the original data to include only those subjects
-df_subset <- df %>% filter(src_subject_id %in% original_subject_ids)
-
-eventname_map <- c(
-  "baseline_year_1_arm_1" = "Baseline",
-  "1_year_follow_up_y_arm_1" = "Year_1",
-  "2_year_follow_up_y_arm_1" = "Year_2",
-  "3_year_follow_up_y_arm_1" = "Year_3",
-  "4_year_follow_up_y_arm_1" = "Year_4"
-)
-
-# Plot
-ggplot(df, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +
-
-# Individual estimated height trajectories in faded lines
-geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +
-
-# Overall group-mean trajectory in blue with increased thickness
-stat_summary(aes(group = 1), fun = mean, geom = "line", color = "blue", linewidth = 1) +
-labs(title = "Individual and Group-Mean Height Trajectories",
-     x = "Event Name",
-     y = "Height") +
-  theme_minimal()
-
-```
-
-The provided code visualizes individual and group-mean height trajectories over different event names.
-Individual height trajectories for each subject are depicted as faded gray lines, allowing for a clear view of the variability among subjects.
-In contrast, the overall group-mean trajectory, which represents the average trend across all individuals for each event name, is highlighted in blue.
-The average height at r mean(df_descriptable$Height) has increased by r mean(df_descriptable$Height_followup - df_descriptable$Height) units from baseline.
-
-## Wrapping Up
-<div class = "blue">
-
-::: panel-tabset
-### Write-up {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4).
-The eventname predictor was statistically significant with a p-value of r format.pval(fixed_effects$Pr(>|t|)[2], digits = 3).
-The model's overall ability to explain the variance in height was xxxxx, with a conditional R^2 of r report_performance(random_intercepts)$Conditional R2[1], indicating that it accounted for this proportion of the variability in height when considering both fixed and random effects.
-The marginal R^2 was r report_performance(random_intercepts)$Marginal R2[1], meaning that the fixed effects alone explained this proportion of the variability.
-
-:::
-:::

### 2024-03-06 examples/_queue/_example_CBCLADHD.qmd
config updates

ABCD worked example when the outcome variable is highly skewed. CBCL ADHD summary score was the outcome of interest. This variable is both overdispersed and zero-inflated.

overdispersment and zero-inflation can be verified using packages like the DHARMa package in R.

glmmTMB was used to model CBCL ADHD. Alternative packages that can also be used include the brms and glmmAdapative package.

examples include a fixed effect for interview age and sex (reference = female). interview age was scaled to improve model run time.

a random intercept was included for the family structure. A random intercept was not included for study site after review of the intraclass correlation coefficient.

a random slope was included for the study visit. A fixed effect was also included for study visit.

zero inflated models included the same covariates.

parallel processing was used to speed up the model run time.

comparison of different distributions is not included here.

dataset annual_ABCD includes only participants that completed all five annual visits included in this example.

```{r}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

library(dplyr)

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

data_nonimaging_r <-  data_nonimaging %>%
  filter(eventname == "baseline_year_1_arm_1" | 
           eventname == "1_year_follow_up_y_arm_1" |
           eventname == "2_year_follow_up_y_arm_1" | 
           eventname == "3_year_follow_up_y_arm_1" |
           eventname == "4_year_follow_up_y_arm_1")

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname",
                   "cbcl_q04_p","cbcl_q08_p","cbcl_q10_p",
                   "cbcl_q41_p","cbcl_q78_p","cbcl_q93_p","cbcl_q104_p")

subset_data <- data_nonimaging_r[, selected_vars]

#below requires the demo5.0 dataframe to extract the sex variable
xy1 <- data_demographics[,c("src_subject_id", "eventname", "sex", "age","rel_family_id.bl")]

annual_ABCD <-  merge(subset_data, xy1, all.x =TRUE, by=c("eventname", "src_subject_id"))

#Manual creation of the CBCL ADHD variable. May need to convert these to numeric first.These
#were originally factor variables that I manually mutated using dplyr.

# Assuming annual_ABCD is your data frame
annual_ABCD$CBCL_ADHD <- rowSums(annual_ABCD[c("cbcl_q04_p", "cbcl_q08_p",
                                               "cbcl_q10_p", "cbcl_q41_p", 
                                               "cbcl_q78_p", "cbcl_q93_p", 
                                               "cbcl_q104_p")], na.rm = TRUE)

#making a numeric study visit variable
annual_ABCD<- annual_ABCD %>% 
  mutate(visit = case_when(
    eventname == "baseline_year_1_arm_1" ~ 1,
    eventname == "1_year_follow_up_y_arm_1" ~ 2,
    eventname == "2_year_follow_up_y_arm_1" ~3,
    eventname == "3_year_follow_up_y_arm_1" ~ 4,
    eventname == "4_year_follow_up_y_arm_1" ~ 5))

###removing any participants with fewer than 5 CBCL ADHD datapoints
annual_ABCD <- annual_ABCD %>% 
  group_by(src_subject_id) %>% 
  filter(n() >= 5)

```

## 
```{r}

library(glmmTMB)

#Generalized Poisson Model
model_GenPois <- glmmTMB(CBCL_ADHD ~ 1 +  age + sex + 
                           factor(visit)+(1+visit| rel_family_id.bl),
                         family=genpois(link="log"), data=annual_ABCD, 
                         REML=FALSE, control = glmmTMBControl(parallel = 16))

summary(model_GenPois)

```

```{r}

confint(model_GenPois)
```

Interpretation:In the model without the ZI term, a 1-SD increase in interview age was associated with a -0.03 (95% CI: -0.08, 0.024, p-value=0.27) decrease in CBCL ADHD score. This was non-significant. Being male (reference=female) was associated with a 0.445-point increase in CBCL ADHD score (95% CI: 0.39, 0.5, p-value <0.0001).

## Truncated Generalized Poisson Model
```{r}

system.time(Model_ZIGenPois <- glmmTMB(CBCL_ADHD ~ 1 +  age + sex +
                                         factor(visit) + (1+visit| rel_family_id.bl),
                                       family=truncated_genpois(link="log"),
                                       ziformula=~1 + age + sex + factor(visit) +
                                         (1+visit| rel_family_id.bl),
                                       data=annual_ABCD, REML=FALSE, control =
                                         glmmTMBControl(parallel = 16)))

summary(Model_ZIGenPois)
```

```{r}
confint(model_GenPois)
```

Interpretation: In the model with the ZI term: In those with a CBCL ADHD score > 0, a 1-SD increase in interview age was associated with -0.002-point decrease in CBCL ADHD score (95% -0.05, 0.044, p value = 0.94). In those with a CBCL ADHD score > 0, being male was associated with a 0.27-point increase in CBCL ADHD score (95% CI: 0.22, 0.31, p-value < 0.0001). In the model with the ZI term: In those with a CBCL ADHD score = 0, a 1-SD increase in interview age was associated with 0.17-point increase in CBCL ADHD score (95% 0.018, 0.32, p value = 0.0282). In those with a CBCL ADHD score = 0, being male was associated with a -0.86-point decrease in CBCL ADHD score (95% CI: -1.00, -0.71, p-value < 0.0001).

### 2024-03-06 examples/_queue/_example_lgcmtwins.qmd
config updates

---
title: "Dynamic growth of ADHD symptoms in the youth: estimating variance components of latent growth curve factors in the ABCD Study"
author: "Luis FS Castro-de-Araujo a,b,*,  Michael C Neale a"
#a. Virginia Institute for Psychiatric and Behavioral Genetics, Virginia Commonwealth University, # p: +1 804 502-4074 P.O. Box 980126, Richmond, VA 23298‑0126, USA. luis.araujo@vcuhealth.org . 
#b. Dept of Psychiatry, The University of Melbourne, Austin Health, Victoria, Australia.

reader-mode: true
---

## Overview

1. Introduction
Attention-Deficit/Hyperactivity Disorder (ADHD) is a common externalizing neurodevelopmental disorder that affects 5-7% of children and adolescents in the population (Sousa et al., 2012). It is characterized by persistent patterns of inattention, hyperactivity, and impulsivity.  Despite its high prevalence, studying ADHD during child development poses several challenges, including the need for longitudinal studies with large sample sizes and multiple waves of data collection to capture meaningful changes over time.

Longitudinal studies are essential for understanding the dynamic growth of ADHD symptoms and identifying potential risk and protective factors. However, conducting longitudinal studies can be logistically challenging and time-consuming. Fortunately, the Adolescent Brain Cognitive Development (ABCD) study provides a unique opportunity to examine the trajectory of ADHD symptoms over time (http://abcdstudy.org). The ABCD study is a large-scale, longitudinal study which aims to track the development of approximately 11,000 children across the United States. This study collects data on various domains, including neuroimaging, genetics, cognitive function, and environmental factors, allowing for a comprehensive analysis of the factors influencing the dynamic growth of ADHD.  A subcohort of twins is included in the ABCD study, which allows for analyses using classical twin designs.

ADHD's symptoms are known to be highly heritable (Faraone and Larsson, 2019) with environmental events also being associated, e.g., prenatal substance use by mothers, low birth weight, environmental toxins (Faraone and Biederman, 2005; Hinshaw, 2018). These influences likely interfere with growth patterns, such that genetic influences might happen earlier in development and environmental effects may be more present at later stages of development. However, this is currently not well understood. Twin designs are particularly informative in these cases, as they can distinguish genetic, environmental and causal effects in the dynamic changes that are occurring during development (Neale & Cardon 1992).  Therefore, individual differences in latent growth trajectories may be accounted for by these background factors.

Previous research has found sex differences in the prevalence and manifestation of ADHD symptoms, with males being more commonly diagnosed with the disorder (Hinshaw, 2018; Sousa et al., 2012). The effect of sex on ADHD is not well known beyond the differences in prevalence in the population. Twin designs can also be informative in these cases, as it is possible to identify whether the sources of variation differ according to sex. However, in the ABCD Study(r), the lack of male-female DZ pairs limits the analysis to whether the variance components are the same size in males and females, rather than whether the same factors operate in the two sexes.
To analyze the growth pattern of ADHD symptoms in the youth, we will use a non-linear latent growth model extended with the classic twin-design.  This approach allows us to estimate a growth curve model with variance components (intercept, linear, quadratic etc.) each of which may have  genetic, shared environmental and unique environmental contributions. Furthermore,  we will test whether there are  significant sex influences to the growth pattern, with the hypothesis that ADHD symptoms  reduce over time as kids grow, and that there is a noticeable sex difference in the sources of variation.

2. Methods
2.1. Data source
The data for this analysis were obtained from the Adolescent Brain Cognitive Development (ABCD) study, a large-scale, longitudinal study that aims to track the development of approximately 11,000 children across the United States. The study collects data on various domains, including neuroimaging, genetics, cognitive function, and environmental factors (Maes et al., 2022). This analysis will use the twins that participated in the ABCD study (n= 1,541), to estimate sources of variation in individuals’ growth curves.

The instrument used to assess ADHD symptoms and their trajectory over time is the Child Behavior Checklist (CBCL) ADHD subscale, which includes questions for inattention, hyperactivity, and impulsivity (Achenbach, 2001). CBCL is a widely used measure of child and adolescent behavior problems, its ADHD subscale tracks DSM diagnosis reliably (Achenbach, 2013). CBCL has been reported to have  a 0.77 sensitivity  and  0.73 specificity to diagnose ADHD in comparison to the reference standard (clinical evaluation by qualified professional) (Chang et al., 2016).  In general, the DSM-oriented CBCL scales present a high Cronbach's alpha of 0.92 and the CBCL ADHD subscale shows an area under curve (AUC) for diagnostic efficiency of 0.75 (Jiang et al., 2023)

2.3. Data analysis
We used R version 4.2.2 (https://cran.r-project.org/) in all further steps of our analyses. All analyses were performed using R, with the OpenMx and umx packages used for structural equation modeling (SEM) analysis (Bates et al., 2019; Neale et al., 2016). The estimation was based on full information maximum likelihood (FIML) with the SLSQP optimizer. We checked identification using an OpenMx utility (mxCheckIdentification) (Hunter et al., 2021). Descriptive information of the data is reported in Table 1.

2.4. Model Specification
A non-linear ACE twin growth curve model was specified. The model consists of allowing A, C, and E variances from the  classic twin model to affect each of latent growth factors in a Cholesky decomposition structure originally reported by (Neale and McArdle, 2000). However, for this study we introduced a quadratic latent factor equivalent to the specifications by (Pritikin et al., 2019; Schmitt et al., 2019). A diagram for one twin is presented in Figure 1. This model was locally identified with either the mean structure for the latent factors, or for the measured CBCL scores. All the models included controls for lifetime use of ADHD medications.

3. Results
The sample comprises 1,541 twins at baseline, with 867 DZs and 674 MZs. Missingness level for the analyzed data set was moderate (20%, Supplement). The sample is mostly white (66.2%), male (50.6%), with a mean age of 10.14 (SD 0.55) at baseline. Demographic characteristics stratified by study waves are shown in Table 1.

The full twin-model includes a dominance factor (ACDE), however due to identification issues either the C or the D factor must be removed for the analysis. In the literature of ADHD and twin modeling, ADHD is often treated as an AE model. In other words, neither C or D are significant and often are removed from the model (Boomsma et al., 2010), but the AE model was rejected for the ADHD data in the ABCD study.

A non-linear latent growth curve model was fitted to twin data from the ABCD study to characterize the growth dynamics of ADHD symptoms over the study waves. The model specification is presented in Figure 2. It consists of a latent growth curve with genetic, shared and exclusive environmental added as a Cholesky structure influencing each of the latent factors. Initially, only a mean structure for the latent variables was added, but is not shown in Figure 2, as the model is not identified with means for both latent factors and manifested variables. The estimated model fitted well (χ²(104) = 285.84, p < 0.001; CFI = 0.959; TLI = 0.965; RMSEA = 0.045 [95%CI 0.04-0.05]) – while the chi-squared is significant, the sample size is large. We found a negative linear component with a mean of -0.065 and  a small negative quadratic component with a mean of -0.003.  A useful parallel that can be made to understand the linear and quadratic components is to think of them as speed and acceleration of growth. Our findings suggest that the ADHD symptoms are reducing with age during the early waves, and that the reduction is slowly accelerating in later waves. Estimates for each parameter can be found in Table 2.

To obtain the modeled means, the model was fitted a second time with a mean structure for the measured variables (and removing the mean structure for the latent variables).  Due to identification issues, it is not possible to retain both the latent and the manifest means in this model. Model fit was good (χ²(102) =278.77, p < 0.001; CFI = 0.959; TLI = 0.965; RMSEA = 0.045 [95%CI 0.04-0.05]), and the modeled means were obtained and plotted as an overlay of a spaghetti plot with each subject score across all study waves (Figure 2).

In order to assess sex differences  in the growth curve the model was fitted to males and females separately, and twice the negative log-likelihood (-2lnL) was extracted for each model. The sum of the -2lnLs for each sex (males -2lnL = 12432, Fit:χ²(104) = 266.43, p < 0.001; CFI = 0.933; TLI = 0.942; RMSEA = 0.058; females -2lnL =  112, Fit: χ²(104) = 229.25, p < 0.001; CFI = 0.939; TLI = 0.947; RMSEA = 0.054) was lower than the fit statistics for when the model was fitted to the combined male and female data set (total -2lnL =  24109). Thus, the likelihood ratio test for heterogeneity between groups is 24109-(12432+11526)=151 with degrees of freedom equal to 26,  the difference between the numbers of free parameters in two models, which equals the number of parameters in the single-sex  models. This overall heterogeneity test of male-female differences in means, variances and covariances is highly significant (Table 4), which justifies further exploration to identify the primary sources of the difference. The latent variable means were equated across sexes, and a model comparison to the unrestricted model was performed.  Equating the means caused a substantial loss of fit of 31.446 -2lnL units for only three degrees of freedom (Table 5).

Finally, equating the paths from the A, C, and E components to the phenotype across sexes generated a model that fitted significantly worse than the unrestricted multiple groups model, with a difference in fit of 24062 -2lnL, 24093 AIC and 18 df (p < 0.001). The intercept, linear and quadratic means for females changed from 1.63, -0.02, and -0.005 in the multiple groups model to 1.63, -0.01, -0.006 in the multiple groups model with equated ACE paths. The change for males was from 2.586, -0.11, 0.0007 in the multiple groups model to 2.56, -0.02 0.0008 in the equated ACE paths model. Altogether, these results show significant sex effects on both the means and sources of variance of growth of the ADHD symptoms.  Figure 3 shows the estimates and standard errors for both models after equating the ACE paths between sexes.

4. Discussion
We present results from a non-linear latent growth model extended with twin design, which allows decomposing growth factor variances into genetic, shared and unique environmental variances. It was found that there were significant linear and quadratic elements in the growth pattern, suggesting reduction of ADHD symptoms over time, with a trend of stabilization in later waves. Furthermore, the model was fit to data on males and females subjects, then a multiple groups model was fit to the data. This approach allowed us to assess the statistical significance of sex differences on the model parameters. The differences found were highly significant and substantial, revealing sex moderation of both means and variance components of ADHD development.

This approach hints at the genetic, shared, and unique environmental effects on the linear latent factor  (0.3, 0.23,  0.02, respectively), suggesting an interplay between genetic liability and environmental factors  in the reduction found in ADHD during the development. We noticed that in our data there was support for a full ACE model during our ACE reduction tests, however in the literature ADHD is often considered an AE condition. Notwithstanding, the C influence was consistently small.

The prevalence of ADHD is very different between males and females in the literature (Sousa et al., 2012), thus suggesting a sex effect on this phenotype. The male to female prevalence ratio is about 3:1 according to (Green et al., 2019; Willcutt, 2012), with males showing more hyperactivity and impulsivity than females (Biederman et al., 1999).  Sex-chromosome aneuplodies like Turner syndrome often present with attention related symptoms, suggesting that these symptoms may be related to X chromosome loci (Green et al., 2019). Our findings suggest a sex moderation on the ADHD symptom levels and growth, extending what was found in previous literature. We did not test whether the sex differences are limited to either the level, the linear growth or the quadratic components, such analyses are feasible and worthwhile of future explorations.

Some limitations of this study should be considered. The ACE model assumes no dominance or epistasis and that all genetic effects are additive and there is no gene-environment covariation or interaction. It is likely that there is substantial G-E covariance for cognitive abilities (Abdellaoui and Verweij, 2021; Demange et al., 2021). Further modeling might include polygenic scores for cognitive ability to operate as instrumental variables, as has been shown for Markov-type autoregressive models by (Singh et al., 2023).  The latter is considered a strong assumption, although methods for GxE interaction were developed  and adjustments for these effects are becoming a possibility (Boker et al., 2023). The classical twin design also assumes equal environment (EEA), which predicts that MZs and DZs are equally correlated in their exposure to environmental events that affect ADHD. Additionally, the CBCL is not a diagnostic tool, although it is known that it is highly correlated with structured diagnostic instruments and closely matches the DSM (Achenbach, 2013). Finally, the model specification used a path coefficient specification, instead of the variance component style (Verhulst et al., 2019). The path analysis specification is limited, notably the fact that the variances cannot be negative. This has the potential to bias the estimation of the ACE variances in the Cholesky structure of this LCM, especially after equating them across sexes as we did here. The findings, however, clearly demonstrate the existence of sex moderation in liability to ADHD symptoms.

This study contributes to our understanding of the development of ADHD symptoms in youth. Further research should continue to explore the complex interplay between genetic and environmental factors in the development and trajectory of ADHD symptoms.  In particular the sex effect on growth could be further explored with a full sex limitation model allowing examination of whether the same factors influence males and females (Neale and Cardon, 1992).  To do so would require data from a large sample of opposite-sex pairs; unfortunately this is not the case for the ABCD Study®, which omitted opposite sex twin pairs to preserve statistical power for within-sex analyses. Nevertheless, the ABCD Study® data combined with genetically informative latent growth curve modeling represents an exceptional opportunity to model and understand the development of a huge array of physical and psychological characteristics. Perhaps the greatest opportunities lie with multivariate applications of the genetically informed latent growth curve model, to gain understanding of how brain, cognition, psychopathology and substance co-occur and  cause changes across these phenotypic boundaries.

Table 1: Main demographic characteristics of the ABCD sample, stratified by study wave. n, number of observations; SD, standard deviation.

Baseline
Follow-up year 1
Follow-up year 2
Follow-up year 3
Follow-up year 4
n
1541
1487
1437
1376
713
Cbcl adhd scores (mean (SD))
2.00 (2.57)
1.94 (2.54)
1.82 (2.53)
1.81 (2.45)
1.60 (2.22)
sex = M (%)
780 ( 50.6)
751 ( 50.5)
722 ( 50.2)
685 ( 49.8)
356 ( 49.9)
Twin zygosity = MZ (%)
674 ( 43.7)
645 ( 43.4)
624 ( 43.4)
589 ( 42.8)
317 ( 44.5)
Age (mean (SD))
10.14 (0.55)
11.17 (0.57)
12.23 (0.59)
13.11 (0.58)
14.48 (0.62)
Race/Ethnicity (%)

White
1020 ( 66.2)
994 ( 66.8)
972 ( 67.6)
937 ( 68.1)
503 ( 70.5)
Black
219 ( 14.2)
206 ( 13.9)
194 ( 13.5)
180 ( 13.1)
85 ( 11.9)

Hispanic
152 (  9.9)
145 (  9.8)
135 (  9.4)
131 (  9.5)
68 (  9.5)

Asian
3 (  0.2)
3 (  0.2)
3 (  0.2)
3 (  0.2)
1 (  0.1)

Other
145 (  9.4)
137 (  9.2)
132 (  9.2)
123 (  8.9)
56 (  7.9)

NA
2 (  0.1)
2 (  0.1)
1 (  0.1)
2 (  0.1)
0 (  0.0)

ADHD medication lifetime use = present (%)
171 ( 11.1)
168 ( 11.3)
162 ( 11.3)
146 ( 10.6)
82 ( 11.5)

Table 2.Whole twin sample means and standard deviations (SD), and MZ and DZ twin pair correlations for the CBCL ADHD scores in the ABCD Study® sample.

Var
Mean
SD
rMZ (403)
rDZ (1724)
base
2.16
2.70
0.65 (0.03)
0.28 (0.04)
flw1
2.01
2.60
0.67 (0.03)
0.24 (0.04)
flw2
1.89
2.54
0.61 (0.04)
0.25 (0.04)
flw3
1.86
2.47
0.6 (0.04)
0.25 (0.04)
flw4
1.68
2.31
0.57 (0.05)
0.24 (0.06)

Table 3. Estimates for the non-linear growth curve model. Model Fit: χ²(104) = 285.84, p < 0.001; CFI = 0.959; TLI = 0.965; RMSEA = 0.04 [95%CI 0.03,0.05]. Path diagram of the model specification is shown in Figure 1. This model was specified with path coefficients, and the naming consists of the origin of the arrow to the destination, so a1_to_intercept is the path from the genetic variance to the intercept latent factor. Paths marked “with” means it has two arrows and represent the residual for that observed variable. The observed are base, flw1, flw2, flw3 and flw4. These are coding to the baseline cbcl scores, and each follow-up year. Meani, meanl and mean q, marks the means of the latent factors for the intercept, linear and quadratic effects.

Name
Estimate
SE
type
a1_to_intercept
2.111
0.129
Factor to factor
a1_to_linear
-0.305
0.129
Factor to factor
a1_to_quadratic
0.044
0.029
Factor to factor
a2_to_linear
-0.534
0.197
Factor to factor
a2_to_quadratic
0.061
0.05
Factor to factor
a3_to_quadratic
0
0.069
Factor to factor
c1_to_intercept
0.777
0.318
Factor to factor
c1_to_linear
0.233
0.332
Factor to factor
c1_to_quadratic
-0.103
0.064
Factor to factor
c2_to_linear
0.433
0.373
Factor to factor
c2_to_quadratic
-0.089
0.093
Factor to factor
c3_to_quadratic
0
0.033
Factor to factor
e1_to_intercept
-0.879
0.085
Factor to factor
e1_to_linear
-0.023
0.115
Factor to factor
e1_to_quadratic
0.011
0.029
Factor to factor
e2_to_linear
-0.18
0.31
Factor to factor
e2_to_quadratic
0.052
0.086
Factor to factor
e3_to_quadratic
0
0.023
Factor to factor
meani
2.122
0.081
Mean
meanl
-0.066
0.047
Mean
meanq
-0.003
0.011
Mean
base_with_base
1.148
0.116
Residual
flw1_with_flw1
1.549
0.083
Residual
flw2_with_flw2
1.631
0.094
Residual
flw3_with_flw3
1.579
0.091
Residual
flw4_with_flw4
1.052
0.164
Residual

Table 4. Comparison of growth curve model-fitting results from fitting to data on males and females with separate set of parameters for the two groups vs. a model where all model parameters are equated e multiple groups model was significantly better than the single group one.

Model
EP
Δ Fit
Δ df
p
AIC
Δ AIC
Compare with Model
Fit units
Multigroup
52

24062
0

-2lnL
Single group
49
150.168
26
< 0.001
2411
98.168
Multigroup
-2lnL

Table 5. Comparison of multiple groups growth curve model with the same model but means of the intercept, linear effect and quadratic effects factors are equated across sexes.

Model
EP
Δ Fit
Δ df
p
AIC
Δ AIC
Compare with Model
Fit units
Multigroup
52

24062
0

-2lnL
Equated means
49
37.446
3
< 0.001
24094
31.446
Multigroup
-2lnL

Figure 1.

Diagram with model specification. This represents a non-linear growth curve model for one twin with linear and quadratic components across five time points of the study. This model is identified either with the growth factors means or the observed variables means (not shown). We estimated both. Also, not shown is the control variable for ADHD lifetime medication use, which was specified as a definition variable. Estimated values for this specification can be found in Table 3. Base, baseline; flw1, follow-up year 1; flw2, follow-up year 2; flw3, follow-up year 3; flw4, follow-up year 4.

Figure 2.

Spaghetti plot with each individual CBCL ADHD scores across waves. Values tend to reduce over time; red dots mark the maximum likelihood estimates of the mean at each wave, which also show slight reduction over time. The means were directly estimated as single parameters for the observed scores at each wave of measurement, instead of being specified at the latent factor level.
Figure 3. Estimates and standard errors for each parameter in the multiple groups model and after fixing ACE paths equal across sexes (ace_equated)

Declarations
Funding

There was no funding for this study.

Conflicts of interest
Authors report no conflicts of interest
Ethics approval
Not applicable
Consent for publication
Not applicable
Availability of data and material
Data sharing not applicable to this article as no datasets were generated or analyzed during the current study.
Code availability
Code is available in a repository for replication.
Author contributions

References
Abdellaoui, A., Verweij, K.J.H., 2021. Dissecting polygenic signals from genome-wide association studies on human behaviour. Nat. Hum. Behav. 5, 686–694. https://doi.org/10.1038/s41562-021-01110-y
Achenbach, T.M., 2013. DSM-Oriented Guide for the Achenbach System of Empirically Based Assessment (ASEBA®).
Achenbach, T.M., 2001. Manual for ASEBA school-age forms & profiles. Univ. Vt. Res. Cent. Child. Youth Fam.
Bates, T.C., Maes, H., Neale, M.C., 2019. umx: Twin and Path-Based Structural Equation Modeling in R. Twin Res. Hum. Genet. 22, 27–41. https://doi.org/10.1017/thg.2019.2
Biederman, J., Faraone, S.V., Mick, E., Williamson, S., Wilens, T.E., Spencer, T.J., Weber, W., Jetton, J., Kraus, I., Pert, J., Zallen, B., 1999. Clinical correlates of ADHD in females: findings from a large group of girls ascertained from pediatric and psychiatric referral sources. J. Am. Acad. Child Adolesc. Psychiatry 38, 966–975. https://doi.org/10.1097/00004583-199908000-00012
Boker, S.M., von Oertzen, T., Pritikin, J.N., Hunter, M.D., Brick, T.R., Brandmaier, A.M., Neale, M.C., 2023. Products of Variables in Structural Equation Models. Struct. Equ. Model. Multidiscip. J. 30, 708–718. https://doi.org/10.1080/10705511.2022.2141749
Boomsma, D.I., Saviouk, V., Hottenga, J.-J., Distel, M.A., de Moor, M.H.M., Vink, J.M., Geels, L.M., van Beek, J.H.D.A., Bartels, M., de Geus, E.J.C., Willemsen, G., 2010. Genetic Epidemiology of Attention Deficit Hyperactivity Disorder (ADHD Index) in Adults. PLoS ONE 5, e10621. https://doi.org/10.1371/journal.pone.0010621
Chang, L.-Y., Wang, M.-Y., Tsai, P.-S., 2016. Diagnostic Accuracy of Rating Scales for Attention-Deficit/Hyperactivity Disorder: A Meta-analysis. Pediatrics 137, e20152749. https://doi.org/10.1542/peds.2015-2749
Demange, P.A., Malanchini, M., Mallard, T.T., Biroli, P., Cox, S.R., Grotzinger, A.D., Tucker-Drob, E.M., Abdellaoui, A., Arseneault, L., van Bergen, E., Boomsma, D.I., Caspi, A., Corcoran, D.L., Domingue, B.W., Harris, K.M., Ip, H.F., Mitchell, C., Moffitt, T.E., Poulton, R., Prinz, J.A., Sugden, K., Wertz, J., Williams, B.S., de Zeeuw, E.L., Belsky, D.W., Harden, K.P., Nivard, M.G., 2021. Investigating the genetic architecture of noncognitive skills using GWAS-by-subtraction. Nat. Genet. 53, 35–44. https://doi.org/10.1038/s41588-020-00754-2
Faraone, S.V., Biederman, J., 2005. What is the prevalence of adult ADHD? Results of a population screen of 966 adults. J. Atten. Disord. 9, 384–91. https://doi.org/10.1177/1087054705281478
Faraone, S.V., Larsson, H., 2019. Genetics of attention deficit hyperactivity disorder. Mol. Psychiatry 24, 562–575. https://doi.org/10.1038/s41380-018-0070-0
Green, T., Flash, S., Reiss, A.L., 2019. Sex differences in psychiatric disorders: what we can learn from sex chromosome aneuploidies. Neuropsychopharmacology 44, 9–21. https://doi.org/10.1038/s41386-018-0153-2
Hinshaw, S.P., 2018. Attention Deficit Hyperactivity Disorder (ADHD): Controversy, Developmental Mechanisms, and Multiple Levels of Analysis. Annu. Rev. Clin. Psychol. 14, 291–316. https://doi.org/10.1146/annurev-clinpsy-050817-084917
Hunter, M.D., Garrison, S.M., Burt, S.A., Rodgers, J.L., 2021. The Analytic Identification of Variance Component Models Common to Behavior Genetics. Behav. Genet. 51, 425–437. https://doi.org/10.1007/s10519-021-10055-x
Jiang, Z., Wang, D., Xu, H., Zhang, A., Zhao, Q., Yan, J., Wang, X., Zhang, W., Li, Yanlin, Yang, K., Hu, S., Cui, Y., Li, Ying, 2023. Diagnostic efficiency and psychometric properties of CBCL DSM-oriented scales in a large sample of Chinese school-attending students aged 5-16. Asian J. Psychiatry 88, 103724. https://doi.org/10.1016/j.ajp.2023.103724
Maes, H.H.M., Lapato, D.M., Schmitt, J.E., Luciana, M., Banich, M.T., Bjork, J.M., Hewitt, J.K., Madden, P.A., Heath, A.C., Barch, D.M., Thompson, W.K., Iacono, W.G., Neale, M.C., 2022. Genetic and Environmental Variation in Continuous Phenotypes in the ABCD Study®. Behav. Genet. https://doi.org/10.1007/s10519-022-10123-w
Neale, M.C., Cardon, L.R., 1992. Methodology for genetic studies of twins and families, Methodology for genetic studies of twins and families. Kluwer Academic/Plenum Publishers, New York, NY, US. https://doi.org/10.1007/978-94-015-8018-2
Neale, M.C., Hunter, M.D., Pritikin, J.N., Zahery, M., Brick, T.R., Kirkpatrick, R.M., Estabrook, R., Bates, T.C., Maes, H.H., Boker, S.M., 2016. OpenMx 2.0: Extended Structural Equation and Statistical Modeling. Psychometrika 81, 535–549. https://doi.org/10.1007/s11336-014-9435-8
Neale, M.C., McArdle, J.J., 2000. Structured latent growth curves for twin data. Twin Res. 3, 165–177. https://doi.org/10.1375/twin.3.3.165
Pritikin, J.N., Schmitt, J.E., Neale, M.C., 2019. Cloud computing for voxel-wise SEM analysis of MRI data. Struct. Equ. Model. Multidiscip. J. 26, 470–480. https://doi.org/10.1080/10705511.2018.1521285
Schmitt, J.E., Raznahan, A., Clasen, L.S., Wallace, G.L., Pritikin, J.N., Lee, N.R., Giedd, J.N., Neale, M.C., 2019. The Dynamic Associations Between Cortical Thickness and General Intelligence are Genetically Mediated. Cereb. Cortex N. Y. N 1991 29, 4743–4752. https://doi.org/10.1093/cercor/bhz007
Singh, M., Dolan, C.V., Neale, M.C., 2023. Integrating Cross-Lagged Panel Models with Instrumental Variables to Extend the Temporal Generalizability of Causal Inference. Multivar. Behav. Res. 1–2. https://doi.org/10.1080/00273171.2022.2160954
Sousa, A. de F., Isabel, C., Castanho, J., Polanczyk, G., Rohde, L., 2012. Attention-defficit hyperactivity disorders, in: Rey, J. (Ed.), E-Textbook of Child and Adolescent Mental Health. International Association for Child and Adolescent Psychiatry and Allied Professions, Geneva.
Verhulst, B., Prom-Wormley, E., Keller, M., Medland, S., Neale, M.C., 2019. Type I Error Rates and Parameter Bias in Multivariate Behavioral Genetic Models. Behav. Genet. 49, 99. https://doi.org/10.1007/s10519-018-9942-y
Willcutt, E.G., 2012. The prevalence of DSM-IV attention-deficit/hyperactivity disorder: a meta-analytic review. Neurother. J. Am. Soc. Exp. Neurother. 9, 490–499. https://doi.org/10.1007/s13311-012-0135-8

Supplemental material
Figure S1.

Missingness levels for the analyzed variables.

### 2024-03-06 examples/_queue/_example_vize_longitudinalinvariance.qmd
config updates

---
title: "Untitled"
format: html
editor: visual
---

```{r}
install.packages(SDSDiff)
library(SDSDiff)
######
### Loading and Cleaning ABCD data
######

setwd()

##Demo/Descriptives
site_df<-read.csv('abcd_y_lt.csv', header=T)  
demo_df<-read.csv('abcd_p_demo.csv', header=T)

##CBCL/UPPS
cbcl_df<-read.csv("mh_p_cbcl.csv",header=T)  #CBCL
upps_df<-read.csv("mh_y_upps.csv",header=T) #UPPS

##Parent KSADS
ksads_p_symp_df<-read.csv("mh_p_ksads_ss.csv", header=T)    #Symptom present/absent across diagnoses

##Youth KSADS
ksads_y_symp_df<-read.csv("mh_y_ksads_ss.csv", header=T)    #Symptom present/absent across diagnoses

##Fluid Int. and Prosocial Behavior 
fluid_df<-read.csv("nc_y_nihtb.csv",header=T) #Fluid Int. Composite
pro_y_df<-read.csv("ce_y_psb.csv", header=T) #Prosocial Bx-Youth
pro_p_df<-read.csv("ce_p_psb.csv", header=T) #Prosocial Bx-Parent

###Subset CBCL and KSADS data 
library(tidyverse)

#CBCL

cbcl_ext<-cbcl_df %>% 
  select(src_subject_id, eventname,
cbcl_q01_p,   #Acts too young for their age
cbcl_q03_p,	  #Argues a lot
cbcl_q04_p,	  #Fails to finish things they start
cbcl_q07_p,	  #Bragging, boasting (note: incorrectly omitted in prereg)
cbcl_q08_p,	  #Can't concentrate, can't pay attention for long-'Distracted/Hyperactive' composite
cbcl_q09_p,	  #Can't get their mind off certain thoughts; obsessions
cbcl_q10_p,	  #Can't sit still, restless, or hyperactive-'Distracted/Hyperactive' composite
cbcl_q13_p,	  #Confused or seems to be in a fog
cbcl_q15_p,	  #Cruel to animals
cbcl_q16_p,	  #Cruelty, bullying, or meanness to others
cbcl_q17_p,	  #Daydreams or gets lost in their thoughts
cbcl_q19_p,	  #Demands a lot of attention
cbcl_q20_p,	  #Destroys their own things-'Destroys' composite
cbcl_q21_p,	  #Destroys things belonging to their family or others-'Destroys' composite
cbcl_q22_p,	  #Disobedient at home-'Disobeys Rules' composite
cbcl_q23_p,	  #Disobedient at school-'Disobeys Rules' composite
cbcl_q25_p,	  #Doesn't get along with other kids-'Peer Problems' composite 
cbcl_q26_p,	  #Doesn't seem to feel guilty after misbehaving
cbcl_q27_p,	  #Easily jealous
cbcl_q28_p,	  #Breaks rules at home, school or elsewhere-'Disobeys Rules' composite
cbcl_q34_p,	  #Feels others are out to get them
cbcl_q36_p,	  #Gets hurt a lot, accident prone 
cbcl_q37_p,	  #Gets in many fights
cbcl_q39_p,	  #Hangs around with others who get in trouble
cbcl_q41_p,	  #Impulsive or acts without thinking
cbcl_q43_p,	  #Lying or cheating
cbcl_q46_p,	  #Nervous movements or twitching
cbcl_q48_p,	  #Not liked by other kids-'Peer Problems' composite 
cbcl_q57_p,	  #Physically attacks people-'Attacks/threatens' composite 
cbcl_q61_p,	  #Poor school work
cbcl_q62_p,	  #Poorly coordinated or clumsy
cbcl_q64_p,	  #Prefers being with younger kids
cbcl_q66_p,	  #Repeats certain acts over and over; compulsions
cbcl_q67_p,	  #Runs away from home
cbcl_q68_p,	  #Screams a lot
cbcl_q72_p,	  #Sets fires
cbcl_q74_p,	  #Showing off or clowning
cbcl_q78_p,	  #Inattentive or easily distracted-'Distracted/Hyperactive' composite 
cbcl_q80_p,	  #Stares blankly
cbcl_q81_p,	  #Steals at home-'Steals' composite
cbcl_q82_p,	  #Steals outside the home-'Steals' composite
cbcl_q85_p,	  #Strange ideas
cbcl_q86_p,	  #Stubborn, sullen, or irritable 
cbcl_q87_p,	  #Sudden changes in mood or feelings
cbcl_q88_p,	  #Sulks a lot
cbcl_q89_p,	  #Suspicious
cbcl_q90_p,	  #Swearing or obscene language
cbcl_q93_p,	  #Talks too much
cbcl_q94_p,	  #Teases a lot
cbcl_q95_p,	  #Temper tantrums or hot temper
cbcl_q96_p,	  #Thinks about sex too much
cbcl_q97_p,	  #Threatens people-'Attacks/threatens' composite 
cbcl_q106_p,	#Vandalism-'Destroys' composite
cbcl_q109_p)	#Whining

#Create composites 
cbcl_ext<- cbcl_ext %>%
  rowwise() %>% 
  mutate(AttThreat_comp = round(mean(c(cbcl_q57_p, cbcl_q97_p), na.rm = T),0), #round() with 0 ensures ratings for composites also range from 0-2
         Destroy_comp = round(mean(c(cbcl_q20_p, cbcl_q21_p, cbcl_q106_p), na.rm = T),0),
         Disobey_comp =  round(mean(c(cbcl_q22_p, cbcl_q23_p, cbcl_q28_p), na.rm = T),0),
         Steal_comp =  round(mean(c(cbcl_q81_p, cbcl_q82_p), na.rm = T),0),
         PeerProb_comp =  round(mean(c(cbcl_q25_p, cbcl_q48_p), na.rm = T),0),
         DistrHyp_comp =  round(mean(c(cbcl_q08_p, cbcl_q10_p, cbcl_q78_p), na.rm = T),0)) %>% 
  select(-c(cbcl_q57_p, cbcl_q97_p,   #drop individual items from df used to create composites 
            cbcl_q20_p, cbcl_q21_p, cbcl_q106_p, 
            cbcl_q22_p, cbcl_q23_p, cbcl_q28_p, 
            cbcl_q81_p, cbcl_q82_p, 
            cbcl_q25_p, cbcl_q48_p,
            cbcl_q08_p, cbcl_q10_p, cbcl_q78_p))

###KSADS-Parent 
ksads_p_symp_df[ksads_p_symp_df==888]<-NA #Recode 888 to be NA

##Create symptom count variables 
ksads_SymCount_p<- ksads_p_symp_df %>% 
  rowwise() %>% 
  ##Conduct disorder
  mutate(CD_count_p=sum(c(
 ksads_16_98_p,	  #Symptom - Often Lies, Present			
 ksads_16_100_p,	#Symptom - Truancy, Present			
 ksads_16_102_p,	#Symptom - Often initiates physical fights, Present			
 ksads_16_104_p,	#Symptom - Often bullies others, Present			
 ksads_16_106_p,	#Symptom - Stealing, Present			
 ksads_16_447_p,	#Symptom - Vandalism, Present			
 ksads_16_449_p,	#Symptom - Breaking and entering, Present			
 ksads_16_451_p,	#Symptom - Stealing while confronting victim, Present			
 ksads_16_453_p,	#Symptom - Fire setting, Present			
 ksads_16_455_p,	#Symptom - Stays out late at night despite parental prohibition, Present			
 ksads_16_457_p,	#Symptom - Ran away overnight, Present			
 ksads_16_459_p,	#Symptom - Use of a weapon that can cause serious harm, Present			
 ksads_16_461_p,	#Symptom - Has been physically cruel to people, Present			
 ksads_16_463_p,	#Symptom - Forced someone into sexual activity, Present			
 ksads_16_465_p),na.rm=T),	#Symptom - Has been physically cruel to animals, Present

##Oppositional defiant disorder
ODD_count_p=sum(c(
ksads_15_91_p,   #Symptom - Often loses temper Present			
ksads_15_93_p,   #Symptom - Often argues with adults/authority Present			
ksads_15_95_p,   #Symptom - Often disobeys rules/requests Present			
ksads_15_432_p,	#Symptom - Often touchy or easily annoyed, Present			
ksads_15_433_p,	#Symptom - Often angry or resentful, Present			
ksads_15_434_p,	#Symptom - Spiteful or vindictive, Present			
ksads_15_435_p,	#Symptom - Often deliberatively annoys people, Present			
ksads_15_436_p,	#Symptom - Often blames others for own mistakes, Present			
ksads_15_438_p),na.rm=T),	#Symptom - Duration of at least 6 months, Present

##ADHD
ADHD_count_p=sum(c(
ksads_14_76_p,	  #Symptom - Difficulty sustaining attention Present			
ksads_14_80_p,	  #Symptom - Easily distracted Present			
ksads_14_84_p,	  #Symptom - Difficulty remaining seated Present			
ksads_14_88_p,	  #Symptom - Impulsivity Present			
ksads_14_394_p,	#Symptom - Often makes careless mistakes, Present			
ksads_14_395_p,	#Symptom - Often doesn't seem to listen when spoken to, Present			
ksads_14_396_p,	#Symptom - Difficulty following Instructions, Present			
ksads_14_397_p,	#Symptom - Difficulty organizing tasks, Present			
ksads_14_398_p,	#Symptom - Avoids tasks requiring attention, Present			
ksads_14_399_p,	#Symptom - Often loses things, Present			
ksads_14_400_p,	#Symptom - Forgetful in daily activities, Present			
ksads_14_401_p,	#Symptom - Fidgets, Present			
ksads_14_402_p,	#Symptom - Runs or climbs excessively, Present			
ksads_14_403_p,	#Symptom - Acts like driven by a motor, Present			
ksads_14_404_p,	#Symptom - Difficulty playing quietly, Present			
ksads_14_405_p,	#Symptom - Blurts out answers, Present			
ksads_14_406_p,	#Symptom - Difficulty waiting turn, Present			
ksads_14_407_p,	#Symptom - Often Interrupts or intrudes on others, Present			
ksads_14_408_p), na.rm=T),	#Symptom - Often talks excessively, Present

##Major depressive disorder
MDD_count_p=sum(c(
ksads_1_1_p,	  #Symptom - Depressed Mood, Present			
ksads_1_3_p,	  #Symptom - Irritability, Present			
ksads_1_5_p,	  #Symptom - Anhedonia, Present			
ksads_1_157_p,	#Symptom - Hypersomnia, Present			
ksads_1_159_p,	#Symptom - Fatigue, Present			
ksads_1_161_p,	#Symptom - Concentration Disturbance, Present			
ksads_1_163_p,	#Symptom - Indecision, Present			
ksads_1_165_p,	#Symptom - Decreased Appetite, Present			
ksads_1_167_p,	#Symptom - Weight Loss, Present			
ksads_1_169_p,	#Symptom - Increased Appetite, Present			
ksads_1_171_p,	#Symptom - Weight Gain, Present			
ksads_1_175_p,	#Symptom - Psychomotor Retardation, Present			
ksads_1_177_p,	#Symptom - Guilt, Present			
ksads_1_179_p,	#Symptom - Hopeless, Present			
ksads_1_181_p),na.rm=T),	#Symptom - Decreased Self-Esteem, Present

##Suicidality/self-harm
SSH_count_p=sum(c(
ksads_23_143_p,	#Symptom - Self injurious behavior, Present			
ksads_23_145_p,	#Symptom - Wishes/Better off dead, Present			
ksads_23_147_p,	#Symptom - Suicidal Ideation, Present			
ksads_23_149_p,	#Symptom - Suicidal Attempt, Present			
ksads_23_807_p,	#Symptom - Self-injury, intent to die, Present			
ksads_23_808_p,	#Symptom - Self-Injury, thought could die from behavior, Present			
ksads_23_809_p,	#Symptom - Suicidal ideation thought of method, Present			
ksads_23_810_p,	#Symptom - Suicidal ideation, intent to act, Present			
ksads_23_811_p,	#Symptom - Suicidal ideation, specific plan, Present			
ksads_23_812_p,	#Symptom - Suicidal behavior, made preparations, Present			
ksads_23_813_p,	#Symptom - Aborted or interrupted suicide attempts, Present			
ksads_23_814_p,	#Symptom - Method of actual suicide attempt, Present			
ksads_23_815_p),na.rm=T),	#Symptom - Suicide attempt, thought could die, Present

##Generalized anxiety disorder
GAD_count_p=sum(c(
ksads_10_45_p,	  #Symptom - Excessive worries more days than not Present			
ksads_10_320_p,	#Symptom - Excessive worries across breadth of domains, Present			
ksads_10_322_p,	#Symptom - Worry associated with defined symptom(s), Present			
ksads_10_324_p),na.rm=T),	#Symptom - Difficulty controlling worries, Present

##Social anxiety disorder
SAD_count_p=sum(c(
ksads_8_29_p,	  #Symptom - Fear of Social Situations Present			
ksads_8_30_p,	  #Symptom - Duration (at least 6 months) Present			
ksads_8_301_p,	  #Symptom - Social situations invariably provoke anxiety, Present			
ksads_8_303_p,	  #Symptom - Social situations avoided or endured with distress, Present			
ksads_8_307_p), na.rm=T)) %>%  	  #Symptom - Social fear is excessive given threat or sociocultural context, Present			
select(src_subject_id, eventname, CD_count_p:SAD_count_p)

###KSADS-Youth
ksads_y_symp_df[ksads_y_symp_df==888]<-NA #Recode 888 to be NA

##Create symptom count variables 
ksads_SymCount_y<- ksads_y_symp_df %>% 
  rowwise() %>%
  mutate(
    ##Major depressive disorder
    MDD_count_y=sum(c(
      ksads_1_1_t,	  #Symptom - Depressed Mood, Present			
      ksads_1_3_t,	  #Symptom - Irritability, Present			
      ksads_1_5_t,	  #Symptom - Anhedonia, Present			
      ksads_1_157_t,	#Symptom - Hypersomnia, Present			
      ksads_1_159_t,	#Symptom - Fatigue, Present			
      ksads_1_161_t,	#Symptom - Concentration Disturbance, Present			
      ksads_1_163_t,	#Symptom - Indecision, Present			
      ksads_1_165_t,	#Symptom - Decreased Appetite, Present			
      ksads_1_167_t,	#Symptom - Weight Loss, Present			
      ksads_1_169_t,	#Symptom - Increased Appetite, Present			
      ksads_1_171_t,	#Symptom - Weight Gain, Present			
      ksads_1_175_t,	#Symptom - Psychomotor Retardation, Present			
      ksads_1_177_t,	#Symptom - Guilt, Present			
      ksads_1_179_t,	#Symptom - Hopeless, Present			
      ksads_1_181_t),na.rm=T),	#Symptom - Decreased Self-Esteem, Present

    ##Suicidality/self-harm
    SSH_count_y=sum(c(
      ksads_23_143_t,	#Symptom - Self injurious behavior, Present			
      ksads_23_145_t,	#Symptom - Wishes/Better off dead, Present			
      ksads_23_147_t,	#Symptom - Suicidal Ideation, Present			
      ksads_23_149_t,	#Symptom - Suicidal Attempt, Present			
      ksads_23_807_t,	#Symptom - Self-injury, intent to die, Present			
      ksads_23_808_t,	#Symptom - Self-Injury, thought could die from behavior, Present			
      ksads_23_809_t,	#Symptom - Suicidal ideation thought of method, Present			
      ksads_23_810_t,	#Symptom - Suicidal ideation, intent to act, Present			
      ksads_23_811_t,	#Symptom - Suicidal ideation, specific plan, Present			
      ksads_23_812_t,	#Symptom - Suicidal behavior, made preparations, Present			
      ksads_23_813_t,	#Symptom - Aborted or interrupted suicide attempts, Present			
      ksads_23_814_t,	#Symptom - Method of actual suicide attempt, Present			
      ksads_23_815_t),na.rm=T),	#Symptom - Suicide attempt, thought could die, Present

    ##Generalized anxiety disorder
    GAD_count_y=sum(c(
      ksads_10_45_t,	  #Symptom - Excessive worries more days than not Present			
      ksads_10_320_t,	#Symptom - Excessive worries across breadth of domains, Present			
      ksads_10_322_t,	#Symptom - Worry associated with defined symptom(s), Present			
      ksads_10_324_t),na.rm=T),	#Symptom - Difficulty controlling worries, Present

    ##Social anxiety disorder
    SAD_count_y=sum(c(
      ksads_8_29_t,	  #Symptom - Fear of Social Situations Present			
      ksads_8_30_t,	  #Symptom - Duration (at least 6 months) Present			
      ksads_8_301_t,	  #Symptom - Social situations invariably provoke anxiety, Present			
      ksads_8_303_t,	  #Symptom - Social situations avoided or endured with distress, Present			
      ksads_8_307_t), na.rm=T)) %>%	  #Symptom - Social fear is excessive given threat or sociocultural context, Present			
  select(src_subject_id, eventname, MDD_count_y:SAD_count_y)

######
######
### Analyses for Aim 1: Identify Hierarchical Structure of CBCL Externalizing Symptoms and Examine Discriminant Validity  
######
######

library(psych)
library(GPArotation)

#Subset to only focus on baseline CBCL data
cbcl_ext_base <- cbcl_ext %>% 
  subset(., eventname == 'baseline_year_1_arm_1')

#Parallel analysis and MAP
fa.parallel(cbcl_ext_base[,3:47], cor = 'poly')                  #14 factors optimal
vss(cbcl_ext_base[,3:47], rotate="promax", fm="pa", cor='poly')  #4 factors optimal based on MAP

#Run Forbes (2023) modified bass-ackward analyses on CBCL items 
##NOTE: these functions are contained in the "ExtendedBassAckwards functions with annotation.R" 
#file available on the OSF page for the paper: https://osf.io/ec36x/?view_only=d63a1452f133421b9e1bef34a41675f1

cbcl_poly<-polychoric(cbcl_ext_base[,3:47]) #first get polychoric corrs among cbcl items

CBCL_4_ExBA<-ExtendedBassAckwards(cbcl_poly$rho,  #use polychoic corr matrix as input 
                             num.comp = 4, #Extract 4 factors at lowest level of hierarchy
                             fm = "minres", #Factoring method
                             rotate = "promax", #Rotation
                             scores = "tenBerge") #Method for computing factor scores

CBCL_4_ExBA$comp.load #Loadings at each level of hierarchy 
CBCL_4_ExBA$comp.corr #Corrs between factors across levels 
hierCorrs<-HierarchicalCorrs(CBCL_4_ExBA$comp.corr, 'd4')
negCorrs<-NegativeCorrs(CBCL_4_ExBA$comp.corr, 'd4')

#Get loadings and export (used for Excel tables in supplement)

CBCL_4_ExBA_loadings<-as.data.frame(unclass(CBCL_4_ExBA$comp.load[[4]]))
CBCL_3_ExBA_loadings<-as.data.frame(unclass(CBCL_4_ExBA$comp.load[[3]]))
CBCL_2_ExBA_loadings<-as.data.frame(unclass(CBCL_4_ExBA$comp.load[[2]]))
CBCL_1_ExBA_loadings<-as.data.frame(unclass(CBCL_4_ExBA$comp.load[[1]]))

write.csv(CBCL_4_ExBA_loadings, "Extended_BassAck_4.csv")
write.csv(CBCL_3_ExBA_loadings, "Extended_BassAck_3.csv")
write.csv(CBCL_2_ExBA_loadings, "Extended_BassAck_2.csv")
write.csv(CBCL_1_ExBA_loadings, "Extended_BassAck_1.csv")

#3-factor solution 
CBCL_3_ExBA<-ExtendedBassAckwards(cbcl_poly$rho,  #use polychoic corr matrix as input 
                                  num.comp = 3, #Extract 4 factors at lowest level of hierarchy
                                  fm = "minres", #Factoring method
                                  rotate = "promax", #Rotation
                                  scores = "tenBerge") #Method for computing factor scores

#Identify any redundant factors in solution 
redund_CBCL_3_ExBA<-FindRedundantComp(CBCL_3_ExBA$comp.corr, CBCL_3_ExBA$cong, 'c3') 
redund_CBCL_3_ExBA$cong.chase 
redund_CBCL_3_ExBA$corr.chase
CBCL_3_ExBA$comp.corr #Corrs between factors across levels

#Examine 14-factor solution 
CBCL_14_ExBA<-ExtendedBassAckwards(cbcl_poly$rho, num.comp = 14,fm = "minres",rotate = "promax",scores = "tenBerge") #Extract 14 factors
CBCL_14_ExBA$comp.load

redund_CBCL_14_ExBA<-FindRedundantComp(CBCL_14_ExBA$comp.corr, CBCL_14_ExBA$cong, 'n14')

#Extended Bass-ackwards object doesn't contain easily accessible factor scores, so run fa.poly() to get these scores for 1- and 3-factor solutions 
#using same extraction, rotation, and factor scoring methods

CBCL_ext_EFA_1<-fa.poly(cbcl_ext_base[,3:47], nfactors = 1, fm="minres", rotate = "promax", scores = 'tenBerge')
CBCL_ext_EFA_3<-fa.poly(cbcl_ext_base[,3:47], nfactors = 3, fm="minres", rotate = "promax", scores = "tenBerge")

CBCL_ext_EFA_1scores<-as.data.frame(CBCL_ext_EFA_1[["scores"]][["scores"]]) #Factor score for broad-based externalizing
CBCL_ext_EFA_3scores<-as.data.frame(CBCL_ext_EFA_3[["scores"]][["scores"]]) #Factor score for 3-factor solution, which includes Ant and Dis

###################
### NOTE: In the subsequent code, what is called "narrow externalizing" in the manuscript is labeled as "Ant" or "ANT".
## What is called "neurodevelopmental problems" in the manuscript is labeled as "Dis" or "DIS". Full disclosure: we changed our minds about
## the labels in the manuscript pretty late in the game. 
##################

colnames(CBCL_ext_EFA_1scores) <- 'Broad_Ext'
colnames(CBCL_ext_EFA_3scores) <- c('Ant_Ext', 'Irr_Ext', 'Dis_Ext')

cbcl_ext_base<-bind_cols(cbcl_ext_base, CBCL_ext_EFA_1scores, CBCL_ext_EFA_3scores) #Add factor scores to our dataframe

######
##Examine Relations between Factors and External Correlates (Table 1)
######

#Merge CBCL results with other DFs
df_list<-list(site_df,
              demo_df,
              cbcl_ext_base,
              upps_df,
              ksads_SymCount_p,
              ksads_SymCount_y,
              fluid_df,
              pro_y_df,
              pro_p_df)

abcd_merged <- df_list %>% 
  reduce(left_join, by=c('src_subject_id','eventname'))

abcd_merged_baseline <- abcd_merged %>% 
subset(., eventname == 'baseline_year_1_arm_1') #only focus on baseline

abcd_corrs<-abcd_merged_baseline %>% 
  select(., Broad_Ext, Ant_Ext, Irr_Ext, Dis_Ext, 
         MDD_count_y:SAD_count_y, CD_count_p:SAD_count_p,
         upps_y_ss_lack_of_perseverance, upps_y_ss_lack_of_planning, upps_y_ss_negative_urgency, upps_y_ss_positive_urgency, upps_y_ss_sensation_seeking,
         psb_p_ss_mean, psb_y_ss_mean,
         nihtbx_fluidcomp_agecorrected) %>% 
  corr.test(alpha=.01)

print(abcd_corrs, short=F) #include short=F to get CIs

###
#Tests of dependent correlations
###

##Parent-report K-SADS outcomes

#CD
r.test(n=11790, r12=.60, r13 =.39, r23 = .62,     #r12 equals the correlations between Antagonism factor and Parent-report CD; 
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   #r13 is correlation with Irritability factor; r23 is corr between Ant. and Irr.

r.test(n=11790, r12=.60, r13 =.31, r23 = .52,     #r23 equals the correlation between Antagonism and Disinhibition/Neurodevelopmental problems
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

r.test(n=11790, r12=.39, r13 =.31, r23 = .58,     #r23 equals the correlation between Irritability and Disinhibition/Neurodevelopmental problems
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#ODD
r.test(n=11790, r12=.45, r13 =.53, r23 = .62,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.45, r13 =.32, r23 = .52,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.53, r13 =.32, r23 = .58,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#ADHD
r.test(n=11790, r12=.41, r13 =.38, r23 = .62,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.41, r13 =.57, r23 = .52,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.38, r13 =.57, r23 = .58,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#MDD
r.test(n=11790, r12=.26, r13 =.32, r23 = .62,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.26, r13 =.28, r23 = .52,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.32, r13 =.28, r23 = .58,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#Suicidality
r.test(n=11790, r12=.20, r13 =.20, r23 = .62,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.20, r13 =.19, r23 = .52,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.20, r13 =.19, r23 = .58,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#GAD
r.test(n=11790, r12=.09, r13 =.23, r23 = .62,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.09, r13 =.22, r23 = .52,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.23, r13 =.22, r23 = .58,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#SAD
r.test(n=11790, r12=.04, r13 =.08, r23 = .62,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.04, r13 =.13, r23 = .52,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.08, r13 =.13, r23 = .58,  
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

##Youth-report K-SADS outcomes

#MDD
r.test(n=11790, r12=.13, r13 =.09, r23 = .62,     #r12 equals the correlations between Antagonism factor and Youth-report MDD; 
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   #r13 is correlation with Irritability factor; r23 is corr between Ant. and Irr.

r.test(n=11790, r12=.13, r13 =.10, r23 = .52,     #r23 equals the correlation between Antagonism and Disinhibition/Neurodevelopmental problems
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

r.test(n=11790, r12=.09, r13 =.10, r23 = .58,     #r23 equals the correlation between Irritability and Disinhibition/Neurodevelopmental problems
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#Suicidality
r.test(n=11790, r12=.11, r13 =.09, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.11, r13 =.11, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.09, r13 =.11, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#GAD
r.test(n=11790, r12=.04, r13 =.05, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.04, r13 =.06, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.05, r13 =.06, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#SAD
r.test(n=11790, r12=.04, r13 =.04, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.04, r13 =.05, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.04, r13 =.05, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#UPPS
r.test(n=11790, r12=.16, r13 =.15, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.16, r13 =.11, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.15, r13 =.11, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

r.test(n=11790, r12=.15, r13 =.10, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.15, r13 =.12, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.10, r13 =.12, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

r.test(n=11790, r12=.15, r13 =.12, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.15, r13 =.14, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.12, r13 =.14, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

r.test(n=11790, r12=.12, r13 =.10, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.12, r13 =.18, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.10, r13 =.18, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

r.test(n=11790, r12=.06, r13 =.02, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=.06, r13 =.02, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=.02, r13 =.02, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#Prosocial Behavior-Parent
r.test(n=11790, r12=-.33, r13 =-.32, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=-.33, r13 =-.22, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=-.32, r13 =-.22, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#Prosocial Behavior-Youth
r.test(n=11790, r12=-.09, r13 =-.07, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=-.09, r13 =-.06, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=-.07, r13 =-.06, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

#Fluid intelligence
r.test(n=11790, r12=-.14, r13 =-.08, r23 = .62,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)   
r.test(n=11790, r12=-.14, r13 =-.17, r23 = .52,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)
r.test(n=11790, r12=-.08, r13 =-.17, r23 = .58,     
       n2 = NULL,pooled=TRUE, twotailed = TRUE)

######
######
### Analyses for Aim 2: Examine Longitudinal Measurement Invariance   
######
######

#####
##Prep Data for MI 
#####

##Check corrs between CBCL items used for composites at follow-up assessments

abcd_cbcl_compitems<-cbcl_df %>% 
  select(., src_subject_id, eventname, 
         cbcl_q08_p,	    #Can't concentrate, can't pay attention for long-'Distracted/Hyperactive' composite
         cbcl_q10_p,	    #Can't sit still, restless, or hyperactive-'Distracted/Hyperactive' composite
         cbcl_q78_p,	    #Inattentive or easily distracted-'Distracted/Hyperactive' composite
         cbcl_q20_p,	    #Destroys their own things-'Destroys' composite
         cbcl_q21_p,	    #Destroys things belonging to their family or others-'Destroys' composite
         cbcl_q106_p, 	  #Vandalism-'Destroys' composite
         cbcl_q22_p,	    #Disobedient at home-'Disobeys Rules' composite
         cbcl_q23_p,	    #Disobedient at school-'Disobeys Rules' composite
         cbcl_q28_p,	    #Breaks rules at home, school or elsewhere-'Disobeys Rules' composite
         cbcl_q25_p,	    #Doesn't get along with other kids-'Peer Problems' composite
         cbcl_q48_p,	    #Not liked by other kids-'Peer Problems' composite 
         cbcl_q57_p,	    #Physically attacks people-'Attacks/threatens' composite
         cbcl_q97_p,	    #Threatens people-'Attacks/threatens' composite
         cbcl_q81_p,	    #Steals at home-'Steals' composite
         cbcl_q82_p)      #Steals outside the home-'Steals' composite

#1 yr follow up
abcd_cbcl_compitems %>% 
  subset(., eventname=='1_year_follow_up_y_arm_1') %>%
  select(., 
         cbcl_q08_p,	    #Can't concentrate, can't pay attention for long-'Distracted/Hyperactive' composite
         cbcl_q10_p,	    #Can't sit still, restless, or hyperactive-'Distracted/Hyperactive' composite
         cbcl_q78_p,	    #Inattentive or easily distracted-'Distracted/Hyperactive' composite
         cbcl_q20_p,	    #Destroys their own things-'Destroys' composite
         cbcl_q21_p,	    #Destroys things belonging to their family or others-'Destroys' composite
         cbcl_q106_p, 	  #Vandalism-'Destroys' composite
         cbcl_q22_p,	    #Disobedient at home-'Disobeys Rules' composite
         cbcl_q23_p,	    #Disobedient at school-'Disobeys Rules' composite
         cbcl_q28_p,	    #Breaks rules at home, school or elsewhere-'Disobeys Rules' composite
         cbcl_q25_p,	    #Doesn't get along with other kids-'Peer Problems' composite
         cbcl_q48_p,	    #Not liked by other kids-'Peer Problems' composite 
         cbcl_q57_p,	    #Physically attacks people-'Attacks/threatens' composite
         cbcl_q97_p,	    #Threatens people-'Attacks/threatens' composite
         cbcl_q81_p,	    #Steals at home-'Steals' composite
         cbcl_q82_p) %>%       #Steals outside the home-'Steals' composite
  polychoric()

#2-yr follow-up
abcd_cbcl_compitems %>% 
  subset(., eventname=='2_year_follow_up_y_arm_1') %>%
  select(., 
         cbcl_q08_p,	    #Can't concentrate, can't pay attention for long-'Distracted/Hyperactive' composite
         cbcl_q10_p,	    #Can't sit still, restless, or hyperactive-'Distracted/Hyperactive' composite
         cbcl_q78_p,	    #Inattentive or easily distracted-'Distracted/Hyperactive' composite
         cbcl_q20_p,	    #Destroys their own things-'Destroys' composite
         cbcl_q21_p,	    #Destroys things belonging to their family or others-'Destroys' composite
         cbcl_q106_p, 	  #Vandalism-'Destroys' composite
         cbcl_q22_p,	    #Disobedient at home-'Disobeys Rules' composite
         cbcl_q23_p,	    #Disobedient at school-'Disobeys Rules' composite
         cbcl_q28_p,	    #Breaks rules at home, school or elsewhere-'Disobeys Rules' composite
         cbcl_q25_p,	    #Doesn't get along with other kids-'Peer Problems' composite
         cbcl_q48_p,	    #Not liked by other kids-'Peer Problems' composite 
         cbcl_q57_p,	    #Physically attacks people-'Attacks/threatens' composite
         cbcl_q97_p,	    #Threatens people-'Attacks/threatens' composite
         cbcl_q81_p,	    #Steals at home-'Steals' composite
         cbcl_q82_p) %>%       #Steals outside the home-'Steals' composite
  polychoric()

#3-yr follow-up
abcd_cbcl_compitems %>% 
  subset(., eventname=='3_year_follow_up_y_arm_1') %>%
  select(., 
         cbcl_q08_p,	    #Can't concentrate, can't pay attention for long-'Distracted/Hyperactive' composite
         cbcl_q10_p,	    #Can't sit still, restless, or hyperactive-'Distracted/Hyperactive' composite
         cbcl_q78_p,	    #Inattentive or easily distracted-'Distracted/Hyperactive' composite
         cbcl_q20_p,	    #Destroys their own things-'Destroys' composite
         cbcl_q21_p,	    #Destroys things belonging to their family or others-'Destroys' composite
         cbcl_q106_p, 	  #Vandalism-'Destroys' composite
         cbcl_q22_p,	    #Disobedient at home-'Disobeys Rules' composite
         cbcl_q23_p,	    #Disobedient at school-'Disobeys Rules' composite
         cbcl_q28_p,	    #Breaks rules at home, school or elsewhere-'Disobeys Rules' composite
         cbcl_q25_p,	    #Doesn't get along with other kids-'Peer Problems' composite
         cbcl_q48_p,	    #Not liked by other kids-'Peer Problems' composite 
         cbcl_q57_p,	    #Physically attacks people-'Attacks/threatens' composite
         cbcl_q97_p,	    #Threatens people-'Attacks/threatens' composite
         cbcl_q81_p,	    #Steals at home-'Steals' composite
         cbcl_q82_p) %>%       #Steals outside the home-'Steals' composite
  polychoric()

######
##Assigning CBCL Items to Parcels for MI Testing
######

##Broad-based Externalizing

CBCL_ITEMS_EXT<-as.data.frame(colnames(cbcl_ext_base[,3:47])) #Note that this df is CBCL items at baseline 
set.seed(111); Item_assign_EXT<-as.data.frame(sample(1:45, 45, replace=F)) #set seed for reproducibility and randomly generate numbers between 1-45 without replacement
                                                                           #Important Note: the item assignments are based on the first sampling iteration! Following iterations will produce different orders of items 
colnames(CBCL_ITEMS_EXT)<-'CBCL_Item'
colnames(Item_assign_EXT)<-'Assign_N'

Parcel_assignments_EXT<-bind_cols(CBCL_ITEMS_EXT, Item_assign_EXT) %>% 
  arrange(Assign_N) #Arrange in order from 1-45

#Create parcels using item assignments from above
print(Parcel_assignments_EXT)

cbcl_ext_base<-cbcl_ext_base %>%
  rowwise() %>% 
  mutate(
    EXT_parc1=sum(c(cbcl_q66_p, Disobey_comp, cbcl_q72_p, cbcl_q80_p, cbcl_q09_p, cbcl_q85_p, cbcl_q61_p, cbcl_q15_p, cbcl_q67_p), na.rm=T), 
    EXT_parc2=sum(c(cbcl_q17_p, cbcl_q94_p, cbcl_q96_p, cbcl_q87_p, cbcl_q01_p, cbcl_q64_p, cbcl_q41_p, cbcl_q43_p, cbcl_q88_p), na.rm=T),
    EXT_parc3=sum(c(cbcl_q04_p, cbcl_q03_p, cbcl_q62_p, cbcl_q86_p, AttThreat_comp, cbcl_q37_p, cbcl_q07_p, cbcl_q16_p, cbcl_q39_p), na.rm=T),
    EXT_parc4=sum(c(cbcl_q90_p, Steal_comp, cbcl_q46_p, cbcl_q109_p, DistrHyp_comp, cbcl_q74_p, cbcl_q89_p, cbcl_q13_p, cbcl_q19_p), na.rm=T),
    EXT_parc5=sum(c(cbcl_q95_p, cbcl_q34_p, cbcl_q26_p, cbcl_q68_p, cbcl_q93_p, cbcl_q36_p, Destroy_comp, PeerProb_comp, cbcl_q27_p), na.rm=T))

##Antagonistic Externalizing

cbcl_ant<-select(cbcl_ext_base, 
                 Steal_comp,	       #Steal Composite	
                 cbcl_q72_p,	       #Sets fires	
                 cbcl_q43_p,	       #Lying or cheating	
                 cbcl_q16_p,	       #Cruelty, bullying, or meanness to others	
                 cbcl_q15_p,	       #Cruel to animals	
                 Disobey_comp,	     #Disobey Composite	
                 cbcl_q37_p,	       #Gets in many fights	
                 cbcl_q90_p,	       #Swearing or obscene language
                 cbcl_q39_p,	       #Hangs around with others who get in trouble	
                 Destroy_comp,	     #Destroy Composite	
                 AttThreat_comp,     #Attack/Threaten Composite	
                 cbcl_q96_p,	       #Thinks about sex too much	
                 cbcl_q26_p,	       #Doesn't seem to feel guilty after misbehaving	
                 cbcl_q94_p,	       #Teases a lot	
                 cbcl_q74_p,	       #Showing off or clowning	
                 cbcl_q67_p)	       #Runs away from home	
                 #cbcl_q41_p	       #Impulsive or acts without thinking; due to cross-loading on disinhibition factor, excluded from scale 	
                 #PeerProb_comp	     #Peer Problems Composite; due to cross-loading on irritability factor, excluded from scale

CBCL_ITEMS_ANT_EXT<-as.data.frame(colnames(cbcl_ant))
set.seed(222); Item_assign_ANT_EXT<-as.data.frame(sample(1:16, 16, replace=F)) #set seed for reproducibility and randomly generate numbers between 1-17 without replacement 
colnames(CBCL_ITEMS_ANT_EXT)<-'CBCL_Item'
colnames(Item_assign_ANT_EXT)<-'Assign_N'

Parcel_assignments_ANT_EXT<-bind_cols(CBCL_ITEMS_ANT_EXT, Item_assign_ANT_EXT) %>% 
  arrange(Assign_N)

#Create parcels using item assignments from above
print(Parcel_assignments_ANT_EXT)

#4 parcels 
cbcl_ext_base<-cbcl_ext_base %>%
  rowwise() %>% 
  mutate(
    ANT_parc1.2=sum(c(cbcl_q39_p,cbcl_q43_p,cbcl_q74_p,cbcl_q90_p), na.rm=T), 
    ANT_parc2.2=sum(c(cbcl_q94_p, Disobey_comp, cbcl_q16_p, cbcl_q15_p), na.rm=T),
    ANT_parc3.2=sum(c(cbcl_q67_p, cbcl_q37_p, cbcl_q26_p, AttThreat_comp), na.rm=T),
    ANT_parc4.2=sum(c(Steal_comp, cbcl_q72_p, Destroy_comp, cbcl_q96_p), na.rm=T))

##Irritability Externalizing

cbcl_irr<-select(cbcl_ext_base,
                 cbcl_q88_p,	  #Sulks a lot
                 cbcl_q86_p,	  #Stubborn, sullen, or irritable 
                 cbcl_q95_p,	  #Temper tantrums or hot temper
                 cbcl_q87_p,	  #Sudden changes in mood or feelings
                 cbcl_q03_p,	  #Argues a lot
                 cbcl_q109_p, 	#Whining
                 cbcl_q68_p,	  #Screams a lot
                 cbcl_q27_p,	  #Easily jealous
                 cbcl_q34_p,	  #Feels others are out to get them
                 cbcl_q19_p,	  #Demands a lot of attention
                 cbcl_q89_p)	  #Suspicious

CBCL_ITEMS_IRR_EXT<-as.data.frame(colnames(cbcl_irr))
set.seed(444); Item_assign_IRR_EXT<-as.data.frame(sample(1:11, 11, replace=F)) #set seed for reproducibility and randomly generate numbers between 1-17 without replacement 
colnames(CBCL_ITEMS_IRR_EXT)<-'CBCL_Item'
colnames(Item_assign_IRR_EXT)<-'Assign_N'

Parcel_assignments_IRR_EXT<-bind_cols(CBCL_ITEMS_IRR_EXT, Item_assign_IRR_EXT) %>% 
  arrange(Assign_N)

#Create parcels using item assignments from above
print(Parcel_assignments_IRR_EXT)

#4 parcels 
cbcl_ext_base<-cbcl_ext_base %>%
  rowwise() %>% 
  mutate(
    IRR_parc1=sum(c(cbcl_q87_p, cbcl_q109_p, cbcl_q86_p), na.rm=T), 
    IRR_parc2=sum(c(cbcl_q19_p, cbcl_q68_p, cbcl_q88_p), na.rm=T),
    IRR_parc3=sum(c(cbcl_q27_p, cbcl_q34_p, cbcl_q03_p), na.rm=T),
    IRR_parc4=sum(c(cbcl_q95_p, cbcl_q89_p ), na.rm=T))

##Neurodevelopmental/Disinhibition

cbcl_dis<-select(cbcl_ext_base, 
                 cbcl_q17_p,	  #Daydreams or gets lost in their thoughts
                 cbcl_q13_p,	  #Confused or seems to be in a fog
                 cbcl_q80_p,	  #Stares blankly
                 DistrHyp_comp,	#Distracted/Hyperactive Composite
                 cbcl_q62_p,	  #Poorly coordinated or clumsy
                 cbcl_q04_p,	  #Fails to finish things they start
                 cbcl_q66_p,	  #Repeats certain acts over and over; compulsions
                 cbcl_q46_p,	  #Nervous movements or twitching
                 cbcl_q09_p,	  #Can't get their mind off certain thoughts; obsessions
                 cbcl_q01_p,	  #Acts too young for their age
                 cbcl_q85_p,	  #Strange ideas
                 cbcl_q61_p,	  #Poor school work
                 cbcl_q64_p,	  #Prefers being with younger kids
                 cbcl_q36_p,	  #Gets hurt a lot, accident prone 
                 cbcl_q93_p)	  #Talks too much

CBCL_ITEMS_DIS_EXT<-as.data.frame(colnames(cbcl_dis))
set.seed(333); Item_assign_DIS_EXT<-as.data.frame(sample(1:15, 15, replace=F)) #set seed for reproducibility and randomly generate numbers between 1-15 without replacement 
colnames(CBCL_ITEMS_DIS_EXT)<-'CBCL_Item'
colnames(Item_assign_DIS_EXT)<-'Assign_N'

Parcel_assignments_DIS_EXT<-bind_cols(CBCL_ITEMS_DIS_EXT, Item_assign_DIS_EXT) %>% 
  arrange(Assign_N)

#Create parcels using item assignments from above

print(Parcel_assignments_DIS_EXT)

#4 parcel solution
cbcl_ext_base<-cbcl_ext_base %>%
  rowwise() %>% 
  mutate(
    DIS_parc1.2=sum(c(cbcl_q01_p, DistrHyp_comp, cbcl_q46_p, cbcl_q09_p), na.rm=T), 
    DIS_parc2.2=sum(c(cbcl_q64_p, cbcl_q66_p, cbcl_q80_p, cbcl_q85_p), na.rm=T),
    DIS_parc3.2=sum(c(cbcl_q13_p, cbcl_q61_p, cbcl_q36_p, cbcl_q04_p), na.rm=T),
    DIS_parc4.2=sum(c(cbcl_q62_p, cbcl_q17_p, cbcl_q93_p), na.rm=T))

##Create Item parcels in longitudinal CBCL data

cbcl_ext <-cbcl_ext %>%
  rowwise() %>% 
  mutate(
    EXT_parc1=sum(c(cbcl_q66_p, Disobey_comp, cbcl_q72_p, cbcl_q80_p, cbcl_q09_p, cbcl_q85_p, cbcl_q61_p, cbcl_q15_p, cbcl_q67_p), na.rm=T), 
    EXT_parc2=sum(c(cbcl_q17_p, cbcl_q94_p, cbcl_q96_p, cbcl_q87_p, cbcl_q01_p, cbcl_q64_p, cbcl_q41_p, cbcl_q43_p, cbcl_q88_p), na.rm=T),
    EXT_parc3=sum(c(cbcl_q04_p, cbcl_q03_p, cbcl_q62_p, cbcl_q86_p, AttThreat_comp, cbcl_q37_p, cbcl_q07_p, cbcl_q16_p, cbcl_q39_p), na.rm=T),
    EXT_parc4=sum(c(cbcl_q90_p, Steal_comp, cbcl_q46_p, cbcl_q109_p, DistrHyp_comp, cbcl_q74_p, cbcl_q89_p, cbcl_q13_p, cbcl_q19_p), na.rm=T),
    EXT_parc5=sum(c(cbcl_q95_p, cbcl_q34_p, cbcl_q26_p, cbcl_q68_p, cbcl_q93_p, cbcl_q36_p, Destroy_comp, PeerProb_comp, cbcl_q27_p), na.rm=T))

cbcl_ext <-cbcl_ext %>%
  rowwise() %>% 
  mutate(
    ANT_parc1.2=sum(c(cbcl_q39_p,cbcl_q43_p,cbcl_q74_p,cbcl_q90_p), na.rm=T), 
    ANT_parc2.2=sum(c(cbcl_q94_p, Disobey_comp, cbcl_q16_p, cbcl_q15_p), na.rm=T),
    ANT_parc3.2=sum(c(cbcl_q67_p, cbcl_q37_p, cbcl_q26_p, AttThreat_comp), na.rm=T),
    ANT_parc4.2=sum(c(Steal_comp, cbcl_q72_p, Destroy_comp, cbcl_q96_p), na.rm=T))

cbcl_ext <-cbcl_ext %>%
  rowwise() %>% 
  mutate(
    IRR_parc1=sum(c(cbcl_q87_p, cbcl_q109_p, cbcl_q86_p), na.rm=T), 
    IRR_parc2=sum(c(cbcl_q19_p, cbcl_q68_p, cbcl_q88_p), na.rm=T),
    IRR_parc3=sum(c(cbcl_q27_p, cbcl_q34_p, cbcl_q03_p), na.rm=T),
    IRR_parc4=sum(c(cbcl_q95_p, cbcl_q89_p ), na.rm=T))

cbcl_ext <-cbcl_ext %>%
  rowwise() %>% 
  mutate(
    DIS_parc1.2=sum(c(cbcl_q01_p, DistrHyp_comp, cbcl_q46_p, cbcl_q09_p), na.rm=T), 
    DIS_parc2.2=sum(c(cbcl_q64_p, cbcl_q66_p, cbcl_q80_p, cbcl_q85_p), na.rm=T),
    DIS_parc3.2=sum(c(cbcl_q13_p, cbcl_q61_p, cbcl_q36_p, cbcl_q04_p), na.rm=T),
    DIS_parc4.2=sum(c(cbcl_q62_p, cbcl_q17_p, cbcl_q93_p), na.rm=T))

##Prep for MPLUS/Lavaan

#Requires wide format data

#Recode time to make code easier to work with 
cbcl_ext <-cbcl_ext %>%
  mutate(eventname_r=as_factor(case_match(eventname, 
                                          'baseline_year_1_arm_1'~'t1',
                                          "1_year_follow_up_y_arm_1"~'t2',
                                          "2_year_follow_up_y_arm_1"~'t3',
                                          "3_year_follow_up_y_arm_1"~'t4',
                                          "4_year_follow_up_y_arm_1"~'t5'))) #won't use 4 year follow up
#Subset parcels
cbcl_long<-cbcl_ext %>% 
  select(src_subject_id, eventname_r, EXT_parc1:DIS_parc4.2)

#Convert to wide         
cbcl_wide<- cbcl_long %>% 
  pivot_wider(id_cols = "src_subject_id", 
              names_from = "eventname_r",
              values_from=c(EXT_parc1, EXT_parc2, EXT_parc3, EXT_parc4, EXT_parc5,
                            ANT_parc1.2, ANT_parc2.2, ANT_parc3.2, ANT_parc4.2,
                            IRR_parc1, IRR_parc2, IRR_parc3, IRR_parc4,
                            DIS_parc1.2, DIS_parc2.2, DIS_parc3.2, DIS_parc4.2),
              names_glue = "{.value}_{eventname_r}")

#Get descriptives across time points

#Externalizing
cbcl_wide %>% 
  select(EXT_parc1_t1,
         EXT_parc2_t1,
         EXT_parc3_t1,
         EXT_parc4_t1,
         EXT_parc5_t1,
         EXT_parc1_t2,
         EXT_parc2_t2,
         EXT_parc3_t2,
         EXT_parc4_t2,
         EXT_parc5_t2,
         EXT_parc1_t3,
         EXT_parc2_t3,
         EXT_parc3_t3,
         EXT_parc4_t3,
         EXT_parc5_t3,
         EXT_parc1_t4,
         EXT_parc2_t4,
         EXT_parc3_t4,
         EXT_parc4_t4,
         EXT_parc5_t4) %>% 
  describe()

cbcl_wide %>% 
  select(EXT_parc1_t1,
         EXT_parc2_t1,
         EXT_parc3_t1,
         EXT_parc4_t1,
         EXT_parc5_t1,
         EXT_parc1_t2,
         EXT_parc2_t2,
         EXT_parc3_t2,
         EXT_parc4_t2,
         EXT_parc5_t2,
         EXT_parc1_t3,
         EXT_parc2_t3,
         EXT_parc3_t3,
         EXT_parc4_t3,
         EXT_parc5_t3,
         EXT_parc1_t4,
         EXT_parc2_t4,
         EXT_parc3_t4,
         EXT_parc4_t4,
         EXT_parc5_t4) %>% 
  corr.test()

#Antagonism

cbcl_wide %>% 
  select(ANT_parc1.2_t1,
         ANT_parc2.2_t1,
         ANT_parc3.2_t1,
         ANT_parc4.2_t1,
         ANT_parc1.2_t2,
         ANT_parc2.2_t2,
         ANT_parc3.2_t2,
         ANT_parc4.2_t2,
         ANT_parc1.2_t3,
         ANT_parc2.2_t3,
         ANT_parc3.2_t3,
         ANT_parc4.2_t3,
         ANT_parc1.2_t4,
         ANT_parc2.2_t4,
         ANT_parc3.2_t4,
         ANT_parc4.2_t4) %>% 
  describe()

cbcl_wide %>% 
  select(ANT_parc1.2_t1,
         ANT_parc2.2_t1,
         ANT_parc3.2_t1,
         ANT_parc4.2_t1,
         ANT_parc1.2_t2,
         ANT_parc2.2_t2,
         ANT_parc3.2_t2,
         ANT_parc4.2_t2,
         ANT_parc1.2_t3,
         ANT_parc2.2_t3,
         ANT_parc3.2_t3,
         ANT_parc4.2_t3,
         ANT_parc1.2_t4,
         ANT_parc2.2_t4,
         ANT_parc3.2_t4,
         ANT_parc4.2_t4) %>% 
  corr.test()

#Irritability
cbcl_wide %>% 
  select(IRR_parc1_t1,
         IRR_parc2_t1,
         IRR_parc3_t1,
         IRR_parc4_t1,
         IRR_parc1_t2,
         IRR_parc2_t2,
         IRR_parc3_t2,
         IRR_parc4_t2,
         IRR_parc1_t3,
         IRR_parc2_t3,
         IRR_parc3_t3,
         IRR_parc4_t3,
         IRR_parc1_t4,
         IRR_parc2_t4,
         IRR_parc3_t4,
         IRR_parc4_t4) %>% 
  describe()

cbcl_wide %>% 
  select(IRR_parc1_t1,
         IRR_parc2_t1,
         IRR_parc3_t1,
         IRR_parc4_t1,
         IRR_parc1_t2,
         IRR_parc2_t2,
         IRR_parc3_t2,
         IRR_parc4_t2,
         IRR_parc1_t3,
         IRR_parc2_t3,
         IRR_parc3_t3,
         IRR_parc4_t3,
         IRR_parc1_t4,
         IRR_parc2_t4,
         IRR_parc3_t4,
         IRR_parc4_t4) %>% 
  corr.test()

#Disinhibition

cbcl_wide %>% 
  select(DIS_parc1.2_t1,
         DIS_parc2.2_t1,
         DIS_parc3.2_t1,
         DIS_parc4.2_t1,
         DIS_parc1.2_t2,
         DIS_parc2.2_t2,
         DIS_parc3.2_t2,
         DIS_parc4.2_t2,
         DIS_parc1.2_t3,
         DIS_parc2.2_t3,
         DIS_parc3.2_t3,
         DIS_parc4.2_t3,
         DIS_parc1.2_t4,
         DIS_parc2.2_t4,
         DIS_parc3.2_t4,
         DIS_parc4.2_t4) %>% 
  describe()

cbcl_wide %>% 
  select(DIS_parc1.2_t1,
         DIS_parc2.2_t1,
         DIS_parc3.2_t1,
         DIS_parc4.2_t1,
         DIS_parc1.2_t2,
         DIS_parc2.2_t2,
         DIS_parc3.2_t2,
         DIS_parc4.2_t2,
         DIS_parc1.2_t3,
         DIS_parc2.2_t3,
         DIS_parc3.2_t3,
         DIS_parc4.2_t3,
         DIS_parc1.2_t4,
         DIS_parc2.2_t4,
         DIS_parc3.2_t4,
         DIS_parc4.2_t4) %>% 
  corr.test()

#################
#################
### MI using Lavaan
#################
#################

library(lavaan)
library(SBSDiff) #used for scaled chi square tests
library(semTools)

##Run our MI models, from configural to strict

###################
#Broad-based Externalizing 
###################

#Configural

EXT_config<-'
#EXT Factors at baseline through 3 year follow up

EXT_t1 =~ EXT_parc1_t1+EXT_parc2_t1+EXT_parc3_t1+EXT_parc4_t1+EXT_parc5_t1 
EXT_t2 =~ EXT_parc1_t2+EXT_parc2_t2+EXT_parc3_t2+EXT_parc4_t2+EXT_parc5_t2
EXT_t3 =~ EXT_parc1_t3+EXT_parc2_t3+EXT_parc3_t3+EXT_parc4_t3+EXT_parc5_t3
EXT_t4 =~ EXT_parc1_t4+EXT_parc2_t4+EXT_parc3_t4+EXT_parc4_t4+EXT_parc5_t4

#Estimate observed intercepts at each wave
EXT_parc1_t1~1
EXT_parc2_t1~1
EXT_parc3_t1~1
EXT_parc4_t1~1
EXT_parc5_t1~1

EXT_parc1_t2~1
EXT_parc2_t2~1
EXT_parc3_t2~1
EXT_parc4_t2~1
EXT_parc5_t2~1

EXT_parc1_t3~1
EXT_parc2_t3~1
EXT_parc3_t3~1
EXT_parc4_t3~1
EXT_parc5_t3~1

EXT_parc1_t4~1
EXT_parc2_t4~1
EXT_parc3_t4~1
EXT_parc4_t4~1
EXT_parc5_t4~1

#Residual Covariances

EXT_parc1_t1 ~~ EXT_parc1_t2
EXT_parc1_t1 ~~ EXT_parc1_t3
EXT_parc1_t1 ~~ EXT_parc1_t4
EXT_parc1_t2 ~~ EXT_parc1_t3
EXT_parc1_t2 ~~ EXT_parc1_t4
EXT_parc1_t3 ~~ EXT_parc1_t4

EXT_parc2_t1 ~~ EXT_parc2_t2
EXT_parc2_t1 ~~ EXT_parc2_t3
EXT_parc2_t1 ~~ EXT_parc2_t4
EXT_parc2_t2 ~~ EXT_parc2_t3
EXT_parc2_t2 ~~ EXT_parc2_t4
EXT_parc2_t3 ~~ EXT_parc2_t4

EXT_parc3_t1 ~~ EXT_parc3_t2
EXT_parc3_t1 ~~ EXT_parc3_t3
EXT_parc3_t1 ~~ EXT_parc3_t4
EXT_parc3_t2 ~~ EXT_parc3_t3
EXT_parc3_t2 ~~ EXT_parc3_t4
EXT_parc3_t3 ~~ EXT_parc3_t4

EXT_parc4_t1 ~~ EXT_parc4_t2
EXT_parc4_t1 ~~ EXT_parc4_t3
EXT_parc4_t1 ~~ EXT_parc4_t4
EXT_parc4_t2 ~~ EXT_parc4_t3
EXT_parc4_t2 ~~ EXT_parc4_t4
EXT_parc4_t3 ~~ EXT_parc4_t4

EXT_parc5_t1 ~~ EXT_parc5_t2
EXT_parc5_t1 ~~ EXT_parc5_t3
EXT_parc5_t1 ~~ EXT_parc5_t4
EXT_parc5_t2 ~~ EXT_parc5_t3
EXT_parc5_t2 ~~ EXT_parc5_t4
EXT_parc5_t3 ~~ EXT_parc5_t4
'

EXT_config.fit<-cfa(EXT_config, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(EXT_config.fit, fit.measures=T, standardized=T)

#McDonald's NCI
exp(-0.5*((767.72- 134)/(11866-1))) #Can compute by hand using exp(-0.5*((Chisquare of the target model - DF of the target model)/(N-1))); note scaled chi square is used
                                      #Cutoff for change in McDonald's NCI is .0065 for 1 factor w/ 5 indicators; .0067 for 1 factor w/ 4 indicators

#Weak/Metric

EXT_weak<-'
#EXT Factors at baseline through 3 year follow up, with constraints on factor loadings
#Note: while it looks like EXT_parc1 (t1-t5) is included twice, NA* tells lavaan to free the parameter, and then the l1 constraint is applied

EXT_t1 =~ NA*EXT_parc1_t1+l1*EXT_parc1_t1+l2*EXT_parc2_t1+l3*EXT_parc3_t1+l4*EXT_parc4_t1+l5*EXT_parc5_t1 
EXT_t2 =~ NA*EXT_parc1_t2+l1*EXT_parc1_t2+l2*EXT_parc2_t2+l3*EXT_parc3_t2+l4*EXT_parc4_t2+l5*EXT_parc5_t2
EXT_t3 =~ NA*EXT_parc1_t3+l1*EXT_parc1_t3+l2*EXT_parc2_t3+l3*EXT_parc3_t3+l4*EXT_parc4_t3+l5*EXT_parc5_t3
EXT_t4 =~ NA*EXT_parc1_t4+l1*EXT_parc1_t4+l2*EXT_parc2_t4+l3*EXT_parc3_t4+l4*EXT_parc4_t4+l5*EXT_parc5_t4

#Latent Intercepts
EXT_t1~1*0
EXT_t2~1*0
EXT_t3~1*0
EXT_t4~1*0

#Latent Variances
EXT_t1~~1*EXT_t1
EXT_t2~~EXT_t2
EXT_t3~~EXT_t3
EXT_t4~~EXT_t4

#Estimate observed intercepts at each wave
EXT_parc1_t1~1
EXT_parc2_t1~1
EXT_parc3_t1~1
EXT_parc4_t1~1
EXT_parc5_t1~1

EXT_parc1_t2~1
EXT_parc2_t2~1
EXT_parc3_t2~1
EXT_parc4_t2~1
EXT_parc5_t2~1

EXT_parc1_t3~1
EXT_parc2_t3~1
EXT_parc3_t3~1
EXT_parc4_t3~1
EXT_parc5_t3~1

EXT_parc1_t4~1
EXT_parc2_t4~1
EXT_parc3_t4~1
EXT_parc4_t4~1
EXT_parc5_t4~1

#Residual Covariances

EXT_parc1_t1 ~~ EXT_parc1_t2
EXT_parc1_t1 ~~ EXT_parc1_t3
EXT_parc1_t1 ~~ EXT_parc1_t4
EXT_parc1_t2 ~~ EXT_parc1_t3
EXT_parc1_t2 ~~ EXT_parc1_t4
EXT_parc1_t3 ~~ EXT_parc1_t4

EXT_parc2_t1 ~~ EXT_parc2_t2
EXT_parc2_t1 ~~ EXT_parc2_t3
EXT_parc2_t1 ~~ EXT_parc2_t4
EXT_parc2_t2 ~~ EXT_parc2_t3
EXT_parc2_t2 ~~ EXT_parc2_t4
EXT_parc2_t3 ~~ EXT_parc2_t4

EXT_parc3_t1 ~~ EXT_parc3_t2
EXT_parc3_t1 ~~ EXT_parc3_t3
EXT_parc3_t1 ~~ EXT_parc3_t4
EXT_parc3_t2 ~~ EXT_parc3_t3
EXT_parc3_t2 ~~ EXT_parc3_t4
EXT_parc3_t3 ~~ EXT_parc3_t4

EXT_parc4_t1 ~~ EXT_parc4_t2
EXT_parc4_t1 ~~ EXT_parc4_t3
EXT_parc4_t1 ~~ EXT_parc4_t4
EXT_parc4_t2 ~~ EXT_parc4_t3
EXT_parc4_t2 ~~ EXT_parc4_t4
EXT_parc4_t3 ~~ EXT_parc4_t4

EXT_parc5_t1 ~~ EXT_parc5_t2
EXT_parc5_t1 ~~ EXT_parc5_t3
EXT_parc5_t1 ~~ EXT_parc5_t4
EXT_parc5_t2 ~~ EXT_parc5_t3
EXT_parc5_t2 ~~ EXT_parc5_t4
EXT_parc5_t3 ~~ EXT_parc5_t4
'

EXT_weak.fit<-cfa(EXT_weak, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(EXT_weak.fit, fit.measures=T, standardized=T)
EXT_config.weak_test<-compareFit(EXT_config.fit, EXT_weak.fit)
summary(EXT_config.weak_test)

sbs.chi(910.07, 767.72, 146, 134, 1.748, 1.704) #provides chi square test using scaled chi square indices 
exp(-0.5*((767.72 - 134)/(11866-1)))-exp(-0.5*((910.07 - 146)/(11866-1))) #change in McDonald's NCI

#Strong/Scalar

EXT_strong<-'
#EXT Factors at baseline through 3 year follow up, with constraints on factor loadings
EXT_t1 =~ NA*EXT_parc1_t1+l1*EXT_parc1_t1+l2*EXT_parc2_t1+l3*EXT_parc3_t1+l4*EXT_parc4_t1+l5*EXT_parc5_t1 
EXT_t2 =~ NA*EXT_parc1_t2+l1*EXT_parc1_t2+l2*EXT_parc2_t2+l3*EXT_parc3_t2+l4*EXT_parc4_t2+l5*EXT_parc5_t2
EXT_t3 =~ NA*EXT_parc1_t3+l1*EXT_parc1_t3+l2*EXT_parc2_t3+l3*EXT_parc3_t3+l4*EXT_parc4_t3+l5*EXT_parc5_t3
EXT_t4 =~ NA*EXT_parc1_t4+l1*EXT_parc1_t4+l2*EXT_parc2_t4+l3*EXT_parc3_t4+l4*EXT_parc4_t4+l5*EXT_parc5_t4

#Baseline latent mean fixed at zero, others freed 
EXT_t1 ~ 0*1
EXT_t2 ~ 1
EXT_t3 ~ 1
EXT_t4 ~ 1

#Latent Variances
EXT_t1~~1*EXT_t1
EXT_t2~~EXT_t2
EXT_t3~~EXT_t3
EXT_t4~~EXT_t4

#Constrain Parcel Intercepts

EXT_parc1_t1~t1*1
EXT_parc2_t1~t2*1
EXT_parc3_t1~t3*1
EXT_parc4_t1~t4*1
EXT_parc5_t1~t5*1

EXT_parc1_t2~t1*1
EXT_parc2_t2~t2*1
EXT_parc3_t2~t3*1
EXT_parc4_t2~t4*1
EXT_parc5_t2~t5*1

EXT_parc1_t3~t1*1
EXT_parc2_t3~t2*1
EXT_parc3_t3~t3*1
EXT_parc4_t3~t4*1
EXT_parc5_t3~t5*1

EXT_parc1_t4~t1*1
EXT_parc2_t4~t2*1
EXT_parc3_t4~t3*1
EXT_parc4_t4~t4*1
EXT_parc5_t4~t5*1

#Residual Covariances

EXT_parc1_t1 ~~ EXT_parc1_t2
EXT_parc1_t1 ~~ EXT_parc1_t3
EXT_parc1_t1 ~~ EXT_parc1_t4
EXT_parc1_t2 ~~ EXT_parc1_t3
EXT_parc1_t2 ~~ EXT_parc1_t4
EXT_parc1_t3 ~~ EXT_parc1_t4

EXT_parc2_t1 ~~ EXT_parc2_t2
EXT_parc2_t1 ~~ EXT_parc2_t3
EXT_parc2_t1 ~~ EXT_parc2_t4
EXT_parc2_t2 ~~ EXT_parc2_t3
EXT_parc2_t2 ~~ EXT_parc2_t4
EXT_parc2_t3 ~~ EXT_parc2_t4

EXT_parc3_t1 ~~ EXT_parc3_t2
EXT_parc3_t1 ~~ EXT_parc3_t3
EXT_parc3_t1 ~~ EXT_parc3_t4
EXT_parc3_t2 ~~ EXT_parc3_t3
EXT_parc3_t2 ~~ EXT_parc3_t4
EXT_parc3_t3 ~~ EXT_parc3_t4

EXT_parc4_t1 ~~ EXT_parc4_t2
EXT_parc4_t1 ~~ EXT_parc4_t3
EXT_parc4_t1 ~~ EXT_parc4_t4
EXT_parc4_t2 ~~ EXT_parc4_t3
EXT_parc4_t2 ~~ EXT_parc4_t4
EXT_parc4_t3 ~~ EXT_parc4_t4

EXT_parc5_t1 ~~ EXT_parc5_t2
EXT_parc5_t1 ~~ EXT_parc5_t3
EXT_parc5_t1 ~~ EXT_parc5_t4
EXT_parc5_t2 ~~ EXT_parc5_t3
EXT_parc5_t2 ~~ EXT_parc5_t4
EXT_parc5_t3 ~~ EXT_parc5_t4
'

EXT_strong.fit<-cfa(EXT_strong, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(EXT_strong.fit, fit.measures=T, standardized=T)
EXT_weak.strong_test<-compareFit(EXT_strong.fit, EXT_weak.fit)
summary(EXT_weak.strong_test)

sbs.chi(1569.46, 910.07, 158, 146, 1.691, 1.748)
exp(-0.5*((1569.46 - 158)/(11866-1)))-exp(-0.5*((910.07 - 146)/(11866-1))) #change in McDonald's NCI

#Mod Indices
lavTestScore(EXT_strong.fit, epc=T)
modificationindices(EXT_strong.fit)

##Test partial strong invariance

EXT_strong.mod<-'
#EXT Factors at baseline through 3 year follow up, with constraints on factor loadings
EXT_t1 =~ NA*EXT_parc1_t1+l1*EXT_parc1_t1+l2*EXT_parc2_t1+l3*EXT_parc3_t1+l4*EXT_parc4_t1+l5*EXT_parc5_t1 
EXT_t2 =~ NA*EXT_parc1_t2+l1*EXT_parc1_t2+l2*EXT_parc2_t2+l3*EXT_parc3_t2+l4*EXT_parc4_t2+l5*EXT_parc5_t2
EXT_t3 =~ NA*EXT_parc1_t3+l1*EXT_parc1_t3+l2*EXT_parc2_t3+l3*EXT_parc3_t3+l4*EXT_parc4_t3+l5*EXT_parc5_t3
EXT_t4 =~ NA*EXT_parc1_t4+l1*EXT_parc1_t4+l2*EXT_parc2_t4+l3*EXT_parc3_t4+l4*EXT_parc4_t4+l5*EXT_parc5_t4

#Baseline latent mean fixed at zero, others freed 
EXT_t1 ~ 0*1
EXT_t2 ~ 1
EXT_t3 ~ 1
EXT_t4 ~ 1

#Latent Variances
EXT_t1~~1*EXT_t1
EXT_t2~~EXT_t2
EXT_t3~~EXT_t3
EXT_t4~~EXT_t4

#Constrain Parcel Intercepts

EXT_parc1_t1~t1*1
EXT_parc2_t1~t2*1
EXT_parc3_t1~t3*1
EXT_parc4_t1~1
EXT_parc5_t1~1

EXT_parc1_t2~t1*1
EXT_parc2_t2~t2*1
EXT_parc3_t2~t3*1
EXT_parc4_t2~1
EXT_parc5_t2~1

EXT_parc1_t3~t1*1
EXT_parc2_t3~t2*1
EXT_parc3_t3~t3*1
EXT_parc4_t3~1
EXT_parc5_t3~1

EXT_parc1_t4~t1*1
EXT_parc2_t4~t2*1
EXT_parc3_t4~t3*1
EXT_parc4_t4~1
EXT_parc5_t4~1

#Residual Covariances

EXT_parc1_t1 ~~ EXT_parc1_t2
EXT_parc1_t1 ~~ EXT_parc1_t3
EXT_parc1_t1 ~~ EXT_parc1_t4
EXT_parc1_t2 ~~ EXT_parc1_t3
EXT_parc1_t2 ~~ EXT_parc1_t4
EXT_parc1_t3 ~~ EXT_parc1_t4

EXT_parc2_t1 ~~ EXT_parc2_t2
EXT_parc2_t1 ~~ EXT_parc2_t3
EXT_parc2_t1 ~~ EXT_parc2_t4
EXT_parc2_t2 ~~ EXT_parc2_t3
EXT_parc2_t2 ~~ EXT_parc2_t4
EXT_parc2_t3 ~~ EXT_parc2_t4

EXT_parc3_t1 ~~ EXT_parc3_t2
EXT_parc3_t1 ~~ EXT_parc3_t3
EXT_parc3_t1 ~~ EXT_parc3_t4
EXT_parc3_t2 ~~ EXT_parc3_t3
EXT_parc3_t2 ~~ EXT_parc3_t4
EXT_parc3_t3 ~~ EXT_parc3_t4

EXT_parc4_t1 ~~ EXT_parc4_t2
EXT_parc4_t1 ~~ EXT_parc4_t3
EXT_parc4_t1 ~~ EXT_parc4_t4
EXT_parc4_t2 ~~ EXT_parc4_t3
EXT_parc4_t2 ~~ EXT_parc4_t4
EXT_parc4_t3 ~~ EXT_parc4_t4

EXT_parc5_t1 ~~ EXT_parc5_t2
EXT_parc5_t1 ~~ EXT_parc5_t3
EXT_parc5_t1 ~~ EXT_parc5_t4
EXT_parc5_t2 ~~ EXT_parc5_t3
EXT_parc5_t2 ~~ EXT_parc5_t4
EXT_parc5_t3 ~~ EXT_parc5_t4
'

EXT_strong.mod.fit<-cfa(EXT_strong.mod, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(EXT_strong.mod.fit, fit.measures=T, standardized=T)
EXT_weak.strong.mod_test<-compareFit(EXT_strong.mod.fit, EXT_weak.fit)
summary(EXT_weak.strong.mod_test)

sbs.chi(947.21, 910.07, 152, 146, 1.718, 1.748)
exp(-0.5*((947.21 - 152)/(11866-1)))-exp(-0.5*((910.07 - 146)/(11866-1))) #change in McDonald's NCI

#Get effect sizes for MI using dmacs package 
devtools::install_github("ddueber/dmacs") 
library(dmacs)
lavaan_dmacs(EXT_strong.mod.fit, MEtype='long') #Input is the partial strong invariance model

#Get Mean/Variance values for dMACS table in manuscript 
EXT_descript<-cbcl_wide %>% 
  rowwise() %>% 
  mutate(EXT_t1=sum(c(EXT_parc1_t1, EXT_parc2_t1, EXT_parc3_t1, EXT_parc4_t1, EXT_parc5_t1), na.rm=T),
          EXT_t2=sum(c(EXT_parc1_t2, EXT_parc2_t2, EXT_parc3_t2, EXT_parc4_t2, EXT_parc5_t2), na.rm=T),
          EXT_t3=sum(c(EXT_parc1_t3, EXT_parc2_t3, EXT_parc3_t3, EXT_parc4_t3, EXT_parc5_t3), na.rm=T),
          EXT_t4=sum(c(EXT_parc1_t4, EXT_parc2_t4, EXT_parc3_t4, EXT_parc4_t4, EXT_parc5_t4), na.rm=T))

describe(select(EXT_descript, EXT_t1, EXT_t2, EXT_t3, EXT_t4 ))

###################
#Antagonism
###################

#Configural

ANT_config<-'
#ANT Factors at baseline through 3 year follow up 
ANT_t1 =~ ANT_parc1.2_t1+ANT_parc2.2_t1+ANT_parc3.2_t1+ANT_parc4.2_t1
ANT_t2 =~ ANT_parc1.2_t2+ANT_parc2.2_t2+ANT_parc3.2_t2+ANT_parc4.2_t2
ANT_t3 =~ ANT_parc1.2_t3+ANT_parc2.2_t3+ANT_parc3.2_t3+ANT_parc4.2_t3
ANT_t4 =~ ANT_parc1.2_t4+ANT_parc2.2_t4+ANT_parc3.2_t4+ANT_parc4.2_t4

#Estimate observed intercepts at each wave
ANT_parc1.2_t1~1
ANT_parc2.2_t1~1
ANT_parc3.2_t1~1
ANT_parc4.2_t1~1

ANT_parc1.2_t2~1
ANT_parc2.2_t2~1
ANT_parc3.2_t2~1
ANT_parc4.2_t2~1

ANT_parc1.2_t3~1
ANT_parc2.2_t3~1
ANT_parc3.2_t3~1
ANT_parc4.2_t3~1

ANT_parc1.2_t4~1
ANT_parc2.2_t4~1
ANT_parc3.2_t4~1
ANT_parc4.2_t4~1

#Residual Covariances

ANT_parc1.2_t1 ~~ ANT_parc1.2_t2
ANT_parc1.2_t1 ~~ ANT_parc1.2_t3
ANT_parc1.2_t1 ~~ ANT_parc1.2_t4
ANT_parc1.2_t2 ~~ ANT_parc1.2_t3
ANT_parc1.2_t2 ~~ ANT_parc1.2_t4
ANT_parc1.2_t3 ~~ ANT_parc1.2_t4

ANT_parc2.2_t1 ~~ ANT_parc2.2_t2
ANT_parc2.2_t1 ~~ ANT_parc2.2_t3
ANT_parc2.2_t1 ~~ ANT_parc2.2_t4
ANT_parc2.2_t2 ~~ ANT_parc2.2_t3
ANT_parc2.2_t2 ~~ ANT_parc2.2_t4
ANT_parc2.2_t3 ~~ ANT_parc2.2_t4

ANT_parc3.2_t1 ~~ ANT_parc3.2_t2
ANT_parc3.2_t1 ~~ ANT_parc3.2_t3
ANT_parc3.2_t1 ~~ ANT_parc3.2_t4
ANT_parc3.2_t2 ~~ ANT_parc3.2_t3
ANT_parc3.2_t2 ~~ ANT_parc3.2_t4
ANT_parc3.2_t3 ~~ ANT_parc3.2_t4

ANT_parc4.2_t1 ~~ ANT_parc4.2_t2
ANT_parc4.2_t1 ~~ ANT_parc4.2_t3
ANT_parc4.2_t1 ~~ ANT_parc4.2_t4
ANT_parc4.2_t2 ~~ ANT_parc4.2_t3
ANT_parc4.2_t2 ~~ ANT_parc4.2_t4
ANT_parc4.2_t3 ~~ ANT_parc4.2_t4
'

ANT_config.fit<-cfa(ANT_config, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(ANT_config.fit, fit.measures=T, standardized=T)
exp(-0.5*((312.54 - 74)/(11866-1))) #McDonald's NCI

#Weak/Metric

ANT_weak<-'
#ANT Factors at baseline through 3 year follow up, with constraints on factor loadings
ANT_t1 =~ NA*ANT_parc1.2_t1+l1*ANT_parc1.2_t1+l2*ANT_parc2.2_t1+l3*ANT_parc3.2_t1+l4*ANT_parc4.2_t1 
ANT_t2 =~ NA*ANT_parc1.2_t2+l1*ANT_parc1.2_t2+l2*ANT_parc2.2_t2+l3*ANT_parc3.2_t2+l4*ANT_parc4.2_t2
ANT_t3 =~ NA*ANT_parc1.2_t3+l1*ANT_parc1.2_t3+l2*ANT_parc2.2_t3+l3*ANT_parc3.2_t3+l4*ANT_parc4.2_t3
ANT_t4 =~ NA*ANT_parc1.2_t4+l1*ANT_parc1.2_t4+l2*ANT_parc2.2_t4+l3*ANT_parc3.2_t4+l4*ANT_parc4.2_t4

#Latent Intercepts
ANT_t1~1*0
ANT_t2~1*0
ANT_t3~1*0
ANT_t4~1*0

#Latent Variances
ANT_t1~~1*ANT_t1
ANT_t2~~ANT_t2
ANT_t3~~ANT_t3
ANT_t4~~ANT_t4

#Estimate observed intercepts at each wave
ANT_parc1.2_t1~1
ANT_parc2.2_t1~1
ANT_parc3.2_t1~1
ANT_parc4.2_t1~1

ANT_parc1.2_t2~1
ANT_parc2.2_t2~1
ANT_parc3.2_t2~1
ANT_parc4.2_t2~1

ANT_parc1.2_t3~1
ANT_parc2.2_t3~1
ANT_parc3.2_t3~1
ANT_parc4.2_t3~1

ANT_parc1.2_t4~1
ANT_parc2.2_t4~1
ANT_parc3.2_t4~1
ANT_parc4.2_t4~1

#Residual Covariances
ANT_parc1.2_t1 ~~ ANT_parc1.2_t2
ANT_parc1.2_t1 ~~ ANT_parc1.2_t3
ANT_parc1.2_t1 ~~ ANT_parc1.2_t4
ANT_parc1.2_t2 ~~ ANT_parc1.2_t3
ANT_parc1.2_t2 ~~ ANT_parc1.2_t4
ANT_parc1.2_t3 ~~ ANT_parc1.2_t4

ANT_parc2.2_t1 ~~ ANT_parc2.2_t2
ANT_parc2.2_t1 ~~ ANT_parc2.2_t3
ANT_parc2.2_t1 ~~ ANT_parc2.2_t4
ANT_parc2.2_t2 ~~ ANT_parc2.2_t3
ANT_parc2.2_t2 ~~ ANT_parc2.2_t4
ANT_parc2.2_t3 ~~ ANT_parc2.2_t4

ANT_parc3.2_t1 ~~ ANT_parc3.2_t2
ANT_parc3.2_t1 ~~ ANT_parc3.2_t3
ANT_parc3.2_t1 ~~ ANT_parc3.2_t4
ANT_parc3.2_t2 ~~ ANT_parc3.2_t3
ANT_parc3.2_t2 ~~ ANT_parc3.2_t4
ANT_parc3.2_t3 ~~ ANT_parc3.2_t4

ANT_parc4.2_t1 ~~ ANT_parc4.2_t2
ANT_parc4.2_t1 ~~ ANT_parc4.2_t3
ANT_parc4.2_t1 ~~ ANT_parc4.2_t4
ANT_parc4.2_t2 ~~ ANT_parc4.2_t3
ANT_parc4.2_t2 ~~ ANT_parc4.2_t4
ANT_parc4.2_t3 ~~ ANT_parc4.2_t4
'

ANT_weak.fit<-cfa(ANT_weak, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(ANT_weak.fit, fit.measures=T, standardized=T)
ANT_config.weak_test<-compareFit(ANT_config.fit, ANT_weak.fit)
summary(ANT_config.weak_test)

sbs.chi(338.15, 312.54, 83, 74, 3.029, 2.809)
exp(-0.5*((338.15 - 83)/(11866-1)))-exp(-0.5*((312.54 - 74)/(11866-1))) #change in McDonald's NCI

#Strong/Scalar

ANT_strong<-'
#ANT Factors at baseline through 3 year follow up, with constraints on factor loadings
ANT_t1 =~ NA*ANT_parc1.2_t1+l1*ANT_parc1.2_t1+l2*ANT_parc2.2_t1+l3*ANT_parc3.2_t1+l4*ANT_parc4.2_t1 
ANT_t2 =~ NA*ANT_parc1.2_t2+l1*ANT_parc1.2_t2+l2*ANT_parc2.2_t2+l3*ANT_parc3.2_t2+l4*ANT_parc4.2_t2
ANT_t3 =~ NA*ANT_parc1.2_t3+l1*ANT_parc1.2_t3+l2*ANT_parc2.2_t3+l3*ANT_parc3.2_t3+l4*ANT_parc4.2_t3
ANT_t4 =~ NA*ANT_parc1.2_t4+l1*ANT_parc1.2_t4+l2*ANT_parc2.2_t4+l3*ANT_parc3.2_t4+l4*ANT_parc4.2_t4

#Latent Intercepts
ANT_t1~1*0
ANT_t2~1
ANT_t3~1
ANT_t4~1

#Latent Variances
ANT_t1~~1*ANT_t1
ANT_t2~~ANT_t2
ANT_t3~~ANT_t3
ANT_t4~~ANT_t4

#Constrain Parcel Intercepts
ANT_parc1.2_t1~t1*1
ANT_parc2.2_t1~t2*1
ANT_parc3.2_t1~t3*1
ANT_parc4.2_t1~t4*1

ANT_parc1.2_t2~t1*1
ANT_parc2.2_t2~t2*1
ANT_parc3.2_t2~t3*1
ANT_parc4.2_t2~t4*1

ANT_parc1.2_t3~t1*1
ANT_parc2.2_t3~t2*1
ANT_parc3.2_t3~t3*1
ANT_parc4.2_t3~t4*1

ANT_parc1.2_t4~t1*1
ANT_parc2.2_t4~t2*1
ANT_parc3.2_t4~t3*1
ANT_parc4.2_t4~t4*1

#Residual Covariances
ANT_parc1.2_t1 ~~ ANT_parc1.2_t2
ANT_parc1.2_t1 ~~ ANT_parc1.2_t3
ANT_parc1.2_t1 ~~ ANT_parc1.2_t4
ANT_parc1.2_t2 ~~ ANT_parc1.2_t3
ANT_parc1.2_t2 ~~ ANT_parc1.2_t4
ANT_parc1.2_t3 ~~ ANT_parc1.2_t4

ANT_parc2.2_t1 ~~ ANT_parc2.2_t2
ANT_parc2.2_t1 ~~ ANT_parc2.2_t3
ANT_parc2.2_t1 ~~ ANT_parc2.2_t4
ANT_parc2.2_t2 ~~ ANT_parc2.2_t3
ANT_parc2.2_t2 ~~ ANT_parc2.2_t4
ANT_parc2.2_t3 ~~ ANT_parc2.2_t4

ANT_parc3.2_t1 ~~ ANT_parc3.2_t2
ANT_parc3.2_t1 ~~ ANT_parc3.2_t3
ANT_parc3.2_t1 ~~ ANT_parc3.2_t4
ANT_parc3.2_t2 ~~ ANT_parc3.2_t3
ANT_parc3.2_t2 ~~ ANT_parc3.2_t4
ANT_parc3.2_t3 ~~ ANT_parc3.2_t4

ANT_parc4.2_t1 ~~ ANT_parc4.2_t2
ANT_parc4.2_t1 ~~ ANT_parc4.2_t3
ANT_parc4.2_t1 ~~ ANT_parc4.2_t4
ANT_parc4.2_t2 ~~ ANT_parc4.2_t3
ANT_parc4.2_t2 ~~ ANT_parc4.2_t4
ANT_parc4.2_t3 ~~ ANT_parc4.2_t4
'

ANT_strong.fit<-cfa(ANT_strong, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(ANT_strong.fit, fit.measures=T, standardized=T)
ANT_weak.strong_test<-compareFit(ANT_weak.fit, ANT_strong.fit)
summary(ANT_weak.strong_test)

sbs.chi(371.439, 338.15, 92, 83, 2.833, 3.029)
exp(-0.5*((371.439 - 92)/(11866-1)))-exp(-0.5*((338.15 - 83)/(11866-1))) #change in McDonald's NCI

#Since Antagonism shows strong invariance, test for strict invariance

#Strict

ANT_strict<-'
#ANT Factors at baseline through 3 year follow up, with constraints on factor loadings
ANT_t1 =~ NA*ANT_parc1.2_t1+l1*ANT_parc1.2_t1+l2*ANT_parc2.2_t1+l3*ANT_parc3.2_t1+l4*ANT_parc4.2_t1 
ANT_t2 =~ NA*ANT_parc1.2_t2+l1*ANT_parc1.2_t2+l2*ANT_parc2.2_t2+l3*ANT_parc3.2_t2+l4*ANT_parc4.2_t2
ANT_t3 =~ NA*ANT_parc1.2_t3+l1*ANT_parc1.2_t3+l2*ANT_parc2.2_t3+l3*ANT_parc3.2_t3+l4*ANT_parc4.2_t3
ANT_t4 =~ NA*ANT_parc1.2_t4+l1*ANT_parc1.2_t4+l2*ANT_parc2.2_t4+l3*ANT_parc3.2_t4+l4*ANT_parc4.2_t4

#Latent Intercepts
ANT_t1~1*0
ANT_t2~1
ANT_t3~1
ANT_t4~1

#Latent Variances
ANT_t1~~1*ANT_t1
ANT_t2~~ANT_t2
ANT_t3~~ANT_t3
ANT_t4~~ANT_t4

#Constrain Parcel Intercepts
ANT_parc1.2_t1~t1*1
ANT_parc2.2_t1~t2*1
ANT_parc3.2_t1~t3*1
ANT_parc4.2_t1~t4*1

ANT_parc1.2_t2~t1*1
ANT_parc2.2_t2~t2*1
ANT_parc3.2_t2~t3*1
ANT_parc4.2_t2~t4*1

ANT_parc1.2_t3~t1*1
ANT_parc2.2_t3~t2*1
ANT_parc3.2_t3~t3*1
ANT_parc4.2_t3~t4*1

ANT_parc1.2_t4~t1*1
ANT_parc2.2_t4~t2*1
ANT_parc3.2_t4~t3*1
ANT_parc4.2_t4~t4*1

#Constrain Parcel Residuals
ANT_parc1.2_t1 ~~ r1*ANT_parc1.2_t1
ANT_parc2.2_t1 ~~ r2*ANT_parc2.2_t1
ANT_parc3.2_t1 ~~ r3*ANT_parc3.2_t1
ANT_parc4.2_t1 ~~ r4*ANT_parc4.2_t1

ANT_parc1.2_t2 ~~ r1*ANT_parc1.2_t2
ANT_parc2.2_t2 ~~ r2*ANT_parc2.2_t2
ANT_parc3.2_t2 ~~ r3*ANT_parc3.2_t2
ANT_parc4.2_t2 ~~ r4*ANT_parc4.2_t2

ANT_parc1.2_t3 ~~ r1*ANT_parc1.2_t3
ANT_parc2.2_t3 ~~ r2*ANT_parc2.2_t3
ANT_parc3.2_t3 ~~ r3*ANT_parc3.2_t3
ANT_parc4.2_t3 ~~ r4*ANT_parc4.2_t3

ANT_parc1.2_t4 ~~ r1*ANT_parc1.2_t4
ANT_parc2.2_t4 ~~ r2*ANT_parc2.2_t4
ANT_parc3.2_t4 ~~ r3*ANT_parc3.2_t4
ANT_parc4.2_t4 ~~ r4*ANT_parc4.2_t4

#Residual Covariances
ANT_parc1.2_t1 ~~ ANT_parc1.2_t2
ANT_parc1.2_t1 ~~ ANT_parc1.2_t3
ANT_parc1.2_t1 ~~ ANT_parc1.2_t4
ANT_parc1.2_t2 ~~ ANT_parc1.2_t3
ANT_parc1.2_t2 ~~ ANT_parc1.2_t4
ANT_parc1.2_t3 ~~ ANT_parc1.2_t4

ANT_parc2.2_t1 ~~ ANT_parc2.2_t2
ANT_parc2.2_t1 ~~ ANT_parc2.2_t3
ANT_parc2.2_t1 ~~ ANT_parc2.2_t4
ANT_parc2.2_t2 ~~ ANT_parc2.2_t3
ANT_parc2.2_t2 ~~ ANT_parc2.2_t4
ANT_parc2.2_t3 ~~ ANT_parc2.2_t4

ANT_parc3.2_t1 ~~ ANT_parc3.2_t2
ANT_parc3.2_t1 ~~ ANT_parc3.2_t3
ANT_parc3.2_t1 ~~ ANT_parc3.2_t4
ANT_parc3.2_t2 ~~ ANT_parc3.2_t3
ANT_parc3.2_t2 ~~ ANT_parc3.2_t4
ANT_parc3.2_t3 ~~ ANT_parc3.2_t4

ANT_parc4.2_t1 ~~ ANT_parc4.2_t2
ANT_parc4.2_t1 ~~ ANT_parc4.2_t3
ANT_parc4.2_t1 ~~ ANT_parc4.2_t4
ANT_parc4.2_t2 ~~ ANT_parc4.2_t3
ANT_parc4.2_t2 ~~ ANT_parc4.2_t4
ANT_parc4.2_t3 ~~ ANT_parc4.2_t4
'

ANT_strict.fit<-cfa(ANT_strict, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(ANT_strict.fit, fit.measures=T, standardized=T)
ANT_strong.strict_test<-compareFit(ANT_strong.fit, ANT_strict.fit)
summary(ANT_strong.strict_test)

sbs.chi(418.21, 371.44, 104, 92, 3.073, 2.833) 
exp(-0.5*((418.21 - 104)/(11866-1)))-exp(-0.5*((371.44 - 92)/(11866-1))) #change in McDonald's NCI

###################
#Irritability
###################

#Configural

IRR_config<-'
#IRR Factors at baseline through 3 year follow up 
IRR_t1 =~ IRR_parc1_t1+IRR_parc2_t1+IRR_parc3_t1+IRR_parc4_t1
IRR_t2 =~ IRR_parc1_t2+IRR_parc2_t2+IRR_parc3_t2+IRR_parc4_t2
IRR_t3 =~ IRR_parc1_t3+IRR_parc2_t3+IRR_parc3_t3+IRR_parc4_t3
IRR_t4 =~ IRR_parc1_t4+IRR_parc2_t4+IRR_parc3_t4+IRR_parc4_t4

#Estimate observed intercepts at each wave
IRR_parc1_t1~1
IRR_parc2_t1~1
IRR_parc3_t1~1
IRR_parc4_t1~1

IRR_parc1_t2~1
IRR_parc2_t2~1
IRR_parc3_t2~1
IRR_parc4_t2~1

IRR_parc1_t3~1
IRR_parc2_t3~1
IRR_parc3_t3~1
IRR_parc4_t3~1

IRR_parc1_t4~1
IRR_parc2_t4~1
IRR_parc3_t4~1
IRR_parc4_t4~1

#Residual Covariances

IRR_parc1_t1 ~~ IRR_parc1_t2
IRR_parc1_t1 ~~ IRR_parc1_t3
IRR_parc1_t1 ~~ IRR_parc1_t4
IRR_parc1_t2 ~~ IRR_parc1_t3
IRR_parc1_t2 ~~ IRR_parc1_t4
IRR_parc1_t3 ~~ IRR_parc1_t4

IRR_parc2_t1 ~~ IRR_parc2_t2
IRR_parc2_t1 ~~ IRR_parc2_t3
IRR_parc2_t1 ~~ IRR_parc2_t4
IRR_parc2_t2 ~~ IRR_parc2_t3
IRR_parc2_t2 ~~ IRR_parc2_t4
IRR_parc2_t3 ~~ IRR_parc2_t4

IRR_parc3_t1 ~~ IRR_parc3_t2
IRR_parc3_t1 ~~ IRR_parc3_t3
IRR_parc3_t1 ~~ IRR_parc3_t4
IRR_parc3_t2 ~~ IRR_parc3_t3
IRR_parc3_t2 ~~ IRR_parc3_t4
IRR_parc3_t3 ~~ IRR_parc3_t4

IRR_parc4_t1 ~~ IRR_parc4_t2
IRR_parc4_t1 ~~ IRR_parc4_t3
IRR_parc4_t1 ~~ IRR_parc4_t4
IRR_parc4_t2 ~~ IRR_parc4_t3
IRR_parc4_t2 ~~ IRR_parc4_t4
IRR_parc4_t3 ~~ IRR_parc4_t4
'

IRR_config.fit<-cfa(IRR_config, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(IRR_config.fit, fit.measures=T, standardized=T)
exp(-0.5*((166.25 - 74)/(11866-1))) #McDonald's NCI

#Weak/Metric

IRR_weak<-'
#IRR Factors at baseline through 3 year follow up, with constraints on factor loadings
IRR_t1 =~ NA*IRR_parc1_t1+l1*IRR_parc1_t1+l2*IRR_parc2_t1+l3*IRR_parc3_t1+l4*IRR_parc4_t1 
IRR_t2 =~ NA*IRR_parc1_t2+l1*IRR_parc1_t2+l2*IRR_parc2_t2+l3*IRR_parc3_t2+l4*IRR_parc4_t2
IRR_t3 =~ NA*IRR_parc1_t3+l1*IRR_parc1_t3+l2*IRR_parc2_t3+l3*IRR_parc3_t3+l4*IRR_parc4_t3
IRR_t4 =~ NA*IRR_parc1_t4+l1*IRR_parc1_t4+l2*IRR_parc2_t4+l3*IRR_parc3_t4+l4*IRR_parc4_t4

#Latent Intercepts
IRR_t1~1*0
IRR_t2~1*0
IRR_t3~1*0
IRR_t4~1*0

#Latent Variances
IRR_t1~~1*IRR_t1
IRR_t2~~IRR_t2
IRR_t3~~IRR_t3
IRR_t4~~IRR_t4

#Estimate observed intercepts at each wave
IRR_parc1_t1~1
IRR_parc2_t1~1
IRR_parc3_t1~1
IRR_parc4_t1~1

IRR_parc1_t2~1
IRR_parc2_t2~1
IRR_parc3_t2~1
IRR_parc4_t2~1

IRR_parc1_t3~1
IRR_parc2_t3~1
IRR_parc3_t3~1
IRR_parc4_t3~1

IRR_parc1_t4~1
IRR_parc2_t4~1
IRR_parc3_t4~1
IRR_parc4_t4~1

#Residual Covariances
IRR_parc1_t1 ~~ IRR_parc1_t2
IRR_parc1_t1 ~~ IRR_parc1_t3
IRR_parc1_t1 ~~ IRR_parc1_t4
IRR_parc1_t2 ~~ IRR_parc1_t3
IRR_parc1_t2 ~~ IRR_parc1_t4
IRR_parc1_t3 ~~ IRR_parc1_t4

IRR_parc2_t1 ~~ IRR_parc2_t2
IRR_parc2_t1 ~~ IRR_parc2_t3
IRR_parc2_t1 ~~ IRR_parc2_t4
IRR_parc2_t2 ~~ IRR_parc2_t3
IRR_parc2_t2 ~~ IRR_parc2_t4
IRR_parc2_t3 ~~ IRR_parc2_t4

IRR_parc3_t1 ~~ IRR_parc3_t2
IRR_parc3_t1 ~~ IRR_parc3_t3
IRR_parc3_t1 ~~ IRR_parc3_t4
IRR_parc3_t2 ~~ IRR_parc3_t3
IRR_parc3_t2 ~~ IRR_parc3_t4
IRR_parc3_t3 ~~ IRR_parc3_t4

IRR_parc4_t1 ~~ IRR_parc4_t2
IRR_parc4_t1 ~~ IRR_parc4_t3
IRR_parc4_t1 ~~ IRR_parc4_t4
IRR_parc4_t2 ~~ IRR_parc4_t3
IRR_parc4_t2 ~~ IRR_parc4_t4
IRR_parc4_t3 ~~ IRR_parc4_t4
'

IRR_weak.fit<-cfa(IRR_weak, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(IRR_weak.fit, fit.measures=T, standardized=T)
IRR_config.weak_test<-compareFit(IRR_config.fit, IRR_weak.fit)
summary(IRR_config.weak_test)

sbs.chi(222.48, 166.25, 83, 74, 1.744, 1.703)
exp(-0.5*((222.483 - 83)/(11866-1)))-exp(-0.5*((166.25 - 74)/(11866-1))) #change in McDonald's NCI

#Strong/Scalar

IRR_strong<-'
#IRR Factors at baseline through 3 year follow up, with constraints on factor loadings
IRR_t1 =~ NA*IRR_parc1_t1+l1*IRR_parc1_t1+l2*IRR_parc2_t1+l3*IRR_parc3_t1+l4*IRR_parc4_t1 
IRR_t2 =~ NA*IRR_parc1_t2+l1*IRR_parc1_t2+l2*IRR_parc2_t2+l3*IRR_parc3_t2+l4*IRR_parc4_t2
IRR_t3 =~ NA*IRR_parc1_t3+l1*IRR_parc1_t3+l2*IRR_parc2_t3+l3*IRR_parc3_t3+l4*IRR_parc4_t3
IRR_t4 =~ NA*IRR_parc1_t4+l1*IRR_parc1_t4+l2*IRR_parc2_t4+l3*IRR_parc3_t4+l4*IRR_parc4_t4

#Latent Intercepts
IRR_t1~1*0
IRR_t2~1
IRR_t3~1
IRR_t4~1

#Latent Variances
IRR_t1~~1*IRR_t1
IRR_t2~~IRR_t2
IRR_t3~~IRR_t3
IRR_t4~~IRR_t4

#Constrain Parcel Intercepts
IRR_parc1_t1~t1*1
IRR_parc2_t1~t2*1
IRR_parc3_t1~t3*1
IRR_parc4_t1~t4*1

IRR_parc1_t2~t1*1
IRR_parc2_t2~t2*1
IRR_parc3_t2~t3*1
IRR_parc4_t2~t4*1

IRR_parc1_t3~t1*1
IRR_parc2_t3~t2*1
IRR_parc3_t3~t3*1
IRR_parc4_t3~t4*1

IRR_parc1_t4~t1*1
IRR_parc2_t4~t2*1
IRR_parc3_t4~t3*1
IRR_parc4_t4~t4*1

#Residual Covariances
IRR_parc1_t1 ~~ IRR_parc1_t2
IRR_parc1_t1 ~~ IRR_parc1_t3
IRR_parc1_t1 ~~ IRR_parc1_t4
IRR_parc1_t2 ~~ IRR_parc1_t3
IRR_parc1_t2 ~~ IRR_parc1_t4
IRR_parc1_t3 ~~ IRR_parc1_t4

IRR_parc2_t1 ~~ IRR_parc2_t2
IRR_parc2_t1 ~~ IRR_parc2_t3
IRR_parc2_t1 ~~ IRR_parc2_t4
IRR_parc2_t2 ~~ IRR_parc2_t3
IRR_parc2_t2 ~~ IRR_parc2_t4
IRR_parc2_t3 ~~ IRR_parc2_t4

IRR_parc3_t1 ~~ IRR_parc3_t2
IRR_parc3_t1 ~~ IRR_parc3_t3
IRR_parc3_t1 ~~ IRR_parc3_t4
IRR_parc3_t2 ~~ IRR_parc3_t3
IRR_parc3_t2 ~~ IRR_parc3_t4
IRR_parc3_t3 ~~ IRR_parc3_t4

IRR_parc4_t1 ~~ IRR_parc4_t2
IRR_parc4_t1 ~~ IRR_parc4_t3
IRR_parc4_t1 ~~ IRR_parc4_t4
IRR_parc4_t2 ~~ IRR_parc4_t3
IRR_parc4_t2 ~~ IRR_parc4_t4
IRR_parc4_t3 ~~ IRR_parc4_t4
'

IRR_strong.fit<-cfa(IRR_strong, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(IRR_strong.fit, fit.measures=T, standardized=T)
IRR_weak.strong_test<-compareFit(IRR_weak.fit, IRR_strong.fit)
summary(IRR_weak.strong_test)

sbs.chi(283.71, 222.483, 92, 83, 1.774, 1.744)
exp(-0.5*((283.71 - 92)/(11866-1)))-exp(-0.5*((222.483 - 83)/(11866-1))) #change in McDonald's NCI

#Since Irritability shows strong invariance, test for strict invariance

#Strict

IRR_strict<-'
#IRR Factors at baseline through 3 year follow up, with constraints on factor loadings
IRR_t1 =~ NA*IRR_parc1_t1+l1*IRR_parc1_t1+l2*IRR_parc2_t1+l3*IRR_parc3_t1+l4*IRR_parc4_t1 
IRR_t2 =~ NA*IRR_parc1_t2+l1*IRR_parc1_t2+l2*IRR_parc2_t2+l3*IRR_parc3_t2+l4*IRR_parc4_t2
IRR_t3 =~ NA*IRR_parc1_t3+l1*IRR_parc1_t3+l2*IRR_parc2_t3+l3*IRR_parc3_t3+l4*IRR_parc4_t3
IRR_t4 =~ NA*IRR_parc1_t4+l1*IRR_parc1_t4+l2*IRR_parc2_t4+l3*IRR_parc3_t4+l4*IRR_parc4_t4

#Latent Intercepts
IRR_t1~1*0
IRR_t2~1
IRR_t3~1
IRR_t4~1

#Latent Variances
IRR_t1~~1*IRR_t1
IRR_t2~~IRR_t2
IRR_t3~~IRR_t3
IRR_t4~~IRR_t4

#Constrain Parcel Intercepts
IRR_parc1_t1~t1*1
IRR_parc2_t1~t2*1
IRR_parc3_t1~t3*1
IRR_parc4_t1~t4*1

IRR_parc1_t2~t1*1
IRR_parc2_t2~t2*1
IRR_parc3_t2~t3*1
IRR_parc4_t2~t4*1

IRR_parc1_t3~t1*1
IRR_parc2_t3~t2*1
IRR_parc3_t3~t3*1
IRR_parc4_t3~t4*1

IRR_parc1_t4~t1*1
IRR_parc2_t4~t2*1
IRR_parc3_t4~t3*1
IRR_parc4_t4~t4*1

#Constrain Parcel Residuals
IRR_parc1_t1 ~~ r1*IRR_parc1_t1
IRR_parc2_t1 ~~ r2*IRR_parc2_t1
IRR_parc3_t1 ~~ r3*IRR_parc3_t1
IRR_parc4_t1 ~~ r4*IRR_parc4_t1

IRR_parc1_t2 ~~ r1*IRR_parc1_t2
IRR_parc2_t2 ~~ r2*IRR_parc2_t2
IRR_parc3_t2 ~~ r3*IRR_parc3_t2
IRR_parc4_t2 ~~ r4*IRR_parc4_t2

IRR_parc1_t3 ~~ r1*IRR_parc1_t3
IRR_parc2_t3 ~~ r2*IRR_parc2_t3
IRR_parc3_t3 ~~ r3*IRR_parc3_t3
IRR_parc4_t3 ~~ r4*IRR_parc4_t3

IRR_parc1_t4 ~~ r1*IRR_parc1_t4
IRR_parc2_t4 ~~ r2*IRR_parc2_t4
IRR_parc3_t4 ~~ r3*IRR_parc3_t4
IRR_parc4_t4 ~~ r4*IRR_parc4_t4

#Residual Covariances
IRR_parc1_t1 ~~ IRR_parc1_t2
IRR_parc1_t1 ~~ IRR_parc1_t3
IRR_parc1_t1 ~~ IRR_parc1_t4
IRR_parc1_t2 ~~ IRR_parc1_t3
IRR_parc1_t2 ~~ IRR_parc1_t4
IRR_parc1_t3 ~~ IRR_parc1_t4

IRR_parc2_t1 ~~ IRR_parc2_t2
IRR_parc2_t1 ~~ IRR_parc2_t3
IRR_parc2_t1 ~~ IRR_parc2_t4
IRR_parc2_t2 ~~ IRR_parc2_t3
IRR_parc2_t2 ~~ IRR_parc2_t4
IRR_parc2_t3 ~~ IRR_parc2_t4

IRR_parc3_t1 ~~ IRR_parc3_t2
IRR_parc3_t1 ~~ IRR_parc3_t3
IRR_parc3_t1 ~~ IRR_parc3_t4
IRR_parc3_t2 ~~ IRR_parc3_t3
IRR_parc3_t2 ~~ IRR_parc3_t4
IRR_parc3_t3 ~~ IRR_parc3_t4

IRR_parc4_t1 ~~ IRR_parc4_t2
IRR_parc4_t1 ~~ IRR_parc4_t3
IRR_parc4_t1 ~~ IRR_parc4_t4
IRR_parc4_t2 ~~ IRR_parc4_t3
IRR_parc4_t2 ~~ IRR_parc4_t4
IRR_parc4_t3 ~~ IRR_parc4_t4
'

IRR_strict.fit<-cfa(IRR_strict, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(IRR_strict.fit, fit.measures=T, standardized=T)
IRR_strong.strict_test<-compareFit(IRR_strong.fit, IRR_strict.fit)
summary(IRR_strong.strict_test)

#################
###Disinhibition 
#################

#Configural

DIS_config<-'
#DIS Factors at baseline through 3 year follow up 
DIS_t1 =~ DIS_parc1.2_t1+DIS_parc2.2_t1+DIS_parc3.2_t1+DIS_parc4.2_t1
DIS_t2 =~ DIS_parc1.2_t2+DIS_parc2.2_t2+DIS_parc3.2_t2+DIS_parc4.2_t2
DIS_t3 =~ DIS_parc1.2_t3+DIS_parc2.2_t3+DIS_parc3.2_t3+DIS_parc4.2_t3
DIS_t4 =~ DIS_parc1.2_t4+DIS_parc2.2_t4+DIS_parc3.2_t4+DIS_parc4.2_t4

#Estimate observed intercepts at each wave
DIS_parc1.2_t1~1
DIS_parc2.2_t1~1
DIS_parc3.2_t1~1
DIS_parc4.2_t1~1

DIS_parc1.2_t2~1
DIS_parc2.2_t2~1
DIS_parc3.2_t2~1
DIS_parc4.2_t2~1

DIS_parc1.2_t3~1
DIS_parc2.2_t3~1
DIS_parc3.2_t3~1
DIS_parc4.2_t3~1

DIS_parc1.2_t4~1
DIS_parc2.2_t4~1
DIS_parc3.2_t4~1
DIS_parc4.2_t4~1

#Residual Covariances

DIS_parc1.2_t1 ~~ DIS_parc1.2_t2
DIS_parc1.2_t1 ~~ DIS_parc1.2_t3
DIS_parc1.2_t1 ~~ DIS_parc1.2_t4
DIS_parc1.2_t2 ~~ DIS_parc1.2_t3
DIS_parc1.2_t2 ~~ DIS_parc1.2_t4
DIS_parc1.2_t3 ~~ DIS_parc1.2_t4

DIS_parc2.2_t1 ~~ DIS_parc2.2_t2
DIS_parc2.2_t1 ~~ DIS_parc2.2_t3
DIS_parc2.2_t1 ~~ DIS_parc2.2_t4
DIS_parc2.2_t2 ~~ DIS_parc2.2_t3
DIS_parc2.2_t2 ~~ DIS_parc2.2_t4
DIS_parc2.2_t3 ~~ DIS_parc2.2_t4

DIS_parc3.2_t1 ~~ DIS_parc3.2_t2
DIS_parc3.2_t1 ~~ DIS_parc3.2_t3
DIS_parc3.2_t1 ~~ DIS_parc3.2_t4
DIS_parc3.2_t2 ~~ DIS_parc3.2_t3
DIS_parc3.2_t2 ~~ DIS_parc3.2_t4
DIS_parc3.2_t3 ~~ DIS_parc3.2_t4

DIS_parc4.2_t1 ~~ DIS_parc4.2_t2
DIS_parc4.2_t1 ~~ DIS_parc4.2_t3
DIS_parc4.2_t1 ~~ DIS_parc4.2_t4
DIS_parc4.2_t2 ~~ DIS_parc4.2_t3
DIS_parc4.2_t2 ~~ DIS_parc4.2_t4
DIS_parc4.2_t3 ~~ DIS_parc4.2_t4
'

DIS_config.fit<-cfa(DIS_config, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(DIS_config.fit, fit.measures=T, standardized=T)
exp(-0.5*((264.19 - 74)/(11866-1))) #McDonald's NCI

#Weak/Metric

DIS_weak<-'
#DIS Factors at baseline through 3 year follow up, with constraints on factor loadings
DIS_t1 =~ NA*DIS_parc1.2_t1+l1*DIS_parc1.2_t1+l2*DIS_parc2.2_t1+l3*DIS_parc3.2_t1+l4*DIS_parc4.2_t1 
DIS_t2 =~ NA*DIS_parc1.2_t2+l1*DIS_parc1.2_t2+l2*DIS_parc2.2_t2+l3*DIS_parc3.2_t2+l4*DIS_parc4.2_t2
DIS_t3 =~ NA*DIS_parc1.2_t3+l1*DIS_parc1.2_t3+l2*DIS_parc2.2_t3+l3*DIS_parc3.2_t3+l4*DIS_parc4.2_t3
DIS_t4 =~ NA*DIS_parc1.2_t4+l1*DIS_parc1.2_t4+l2*DIS_parc2.2_t4+l3*DIS_parc3.2_t4+l4*DIS_parc4.2_t4

#Latent Intercepts
DIS_t1~1*0
DIS_t2~1*0
DIS_t3~1*0
DIS_t4~1*0

#Latent Variances
DIS_t1~~1*DIS_t1
DIS_t2~~DIS_t2
DIS_t3~~DIS_t3
DIS_t4~~DIS_t4

#Estimate observed intercepts at each wave
DIS_parc1.2_t1~1
DIS_parc2.2_t1~1
DIS_parc3.2_t1~1
DIS_parc4.2_t1~1

DIS_parc1.2_t2~1
DIS_parc2.2_t2~1
DIS_parc3.2_t2~1
DIS_parc4.2_t2~1

DIS_parc1.2_t3~1
DIS_parc2.2_t3~1
DIS_parc3.2_t3~1
DIS_parc4.2_t3~1

DIS_parc1.2_t4~1
DIS_parc2.2_t4~1
DIS_parc3.2_t4~1
DIS_parc4.2_t4~1

#Residual Covariances
DIS_parc1.2_t1 ~~ DIS_parc1.2_t2
DIS_parc1.2_t1 ~~ DIS_parc1.2_t3
DIS_parc1.2_t1 ~~ DIS_parc1.2_t4
DIS_parc1.2_t2 ~~ DIS_parc1.2_t3
DIS_parc1.2_t2 ~~ DIS_parc1.2_t4
DIS_parc1.2_t3 ~~ DIS_parc1.2_t4

DIS_parc2.2_t1 ~~ DIS_parc2.2_t2
DIS_parc2.2_t1 ~~ DIS_parc2.2_t3
DIS_parc2.2_t1 ~~ DIS_parc2.2_t4
DIS_parc2.2_t2 ~~ DIS_parc2.2_t3
DIS_parc2.2_t2 ~~ DIS_parc2.2_t4
DIS_parc2.2_t3 ~~ DIS_parc2.2_t4

DIS_parc3.2_t1 ~~ DIS_parc3.2_t2
DIS_parc3.2_t1 ~~ DIS_parc3.2_t3
DIS_parc3.2_t1 ~~ DIS_parc3.2_t4
DIS_parc3.2_t2 ~~ DIS_parc3.2_t3
DIS_parc3.2_t2 ~~ DIS_parc3.2_t4
DIS_parc3.2_t3 ~~ DIS_parc3.2_t4

DIS_parc4.2_t1 ~~ DIS_parc4.2_t2
DIS_parc4.2_t1 ~~ DIS_parc4.2_t3
DIS_parc4.2_t1 ~~ DIS_parc4.2_t4
DIS_parc4.2_t2 ~~ DIS_parc4.2_t3
DIS_parc4.2_t2 ~~ DIS_parc4.2_t4
DIS_parc4.2_t3 ~~ DIS_parc4.2_t4
'

DIS_weak.fit<-cfa(DIS_weak, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(DIS_weak.fit, fit.measures=T, standardized=T)
DIS_config.weak_test<-compareFit(DIS_config.fit, DIS_weak.fit)
summary(DIS_config.weak_test)

sbs.chi(387.31, 264.19, 83, 74, 1.707, 1.668) 
exp(-0.5*((387.31 - 83)/(11866-1)))-exp(-0.5*((264.19 - 74)/(11866-1))) #change in McDonald's NCI

#Strong/Scalar

DIS_strong<-'
#DIS Factors at baseline through 3 year follow up, with constraints on factor loadings
DIS_t1 =~ NA*DIS_parc1.2_t1+l1*DIS_parc1.2_t1+l2*DIS_parc2.2_t1+l3*DIS_parc3.2_t1+l4*DIS_parc4.2_t1 
DIS_t2 =~ NA*DIS_parc1.2_t2+l1*DIS_parc1.2_t2+l2*DIS_parc2.2_t2+l3*DIS_parc3.2_t2+l4*DIS_parc4.2_t2
DIS_t3 =~ NA*DIS_parc1.2_t3+l1*DIS_parc1.2_t3+l2*DIS_parc2.2_t3+l3*DIS_parc3.2_t3+l4*DIS_parc4.2_t3
DIS_t4 =~ NA*DIS_parc1.2_t4+l1*DIS_parc1.2_t4+l2*DIS_parc2.2_t4+l3*DIS_parc3.2_t4+l4*DIS_parc4.2_t4

#Baseline latent mean fixed at zero, others freed 
DIS_t1 ~ 0*1
DIS_t2 ~ 1
DIS_t3 ~ 1
DIS_t4 ~ 1

#Latent Variances
DIS_t1~~1*DIS_t1
DIS_t2~~DIS_t2
DIS_t3~~DIS_t3
DIS_t4~~DIS_t4

#Constrain Parcel Intercepts
DIS_parc1.2_t1~t1*1
DIS_parc2.2_t1~t2*1
DIS_parc3.2_t1~t3*1
DIS_parc4.2_t1~t4*1

DIS_parc1.2_t2~t1*1
DIS_parc2.2_t2~t2*1
DIS_parc3.2_t2~t3*1
DIS_parc4.2_t2~t4*1

DIS_parc1.2_t3~t1*1
DIS_parc2.2_t3~t2*1
DIS_parc3.2_t3~t3*1
DIS_parc4.2_t3~t4*1

DIS_parc1.2_t4~t1*1
DIS_parc2.2_t4~t2*1
DIS_parc3.2_t4~t3*1
DIS_parc4.2_t4~t4*1

#Residual Covariances
DIS_parc1.2_t1 ~~ DIS_parc1.2_t2
DIS_parc1.2_t1 ~~ DIS_parc1.2_t3
DIS_parc1.2_t1 ~~ DIS_parc1.2_t4
DIS_parc1.2_t2 ~~ DIS_parc1.2_t3
DIS_parc1.2_t2 ~~ DIS_parc1.2_t4
DIS_parc1.2_t3 ~~ DIS_parc1.2_t4

DIS_parc2.2_t1 ~~ DIS_parc2.2_t2
DIS_parc2.2_t1 ~~ DIS_parc2.2_t3
DIS_parc2.2_t1 ~~ DIS_parc2.2_t4
DIS_parc2.2_t2 ~~ DIS_parc2.2_t3
DIS_parc2.2_t2 ~~ DIS_parc2.2_t4
DIS_parc2.2_t3 ~~ DIS_parc2.2_t4

DIS_parc3.2_t1 ~~ DIS_parc3.2_t2
DIS_parc3.2_t1 ~~ DIS_parc3.2_t3
DIS_parc3.2_t1 ~~ DIS_parc3.2_t4
DIS_parc3.2_t2 ~~ DIS_parc3.2_t3
DIS_parc3.2_t2 ~~ DIS_parc3.2_t4
DIS_parc3.2_t3 ~~ DIS_parc3.2_t4

DIS_parc4.2_t1 ~~ DIS_parc4.2_t2
DIS_parc4.2_t1 ~~ DIS_parc4.2_t3
DIS_parc4.2_t1 ~~ DIS_parc4.2_t4
DIS_parc4.2_t2 ~~ DIS_parc4.2_t3
DIS_parc4.2_t2 ~~ DIS_parc4.2_t4
DIS_parc4.2_t3 ~~ DIS_parc4.2_t4
'

DIS_strong.fit<-cfa(DIS_strong, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(DIS_strong.fit, fit.measures=T, standardized=T)
DIS_weak.strong_test<-compareFit(DIS_weak.fit, DIS_strong.fit)
summary(DIS_weak.strong_test)

sbs.chi(840.08, 387.31, 92, 83, 1.637, 1.707) 
exp(-0.5*((840.08 - 92)/(11866-1)))-exp(-0.5*((387.31 - 83)/(11866-1))) #change in McDonald's NCI

#Mod Indices
lavTestScore(DIS_strong.fit, epc=T)
modificationindices(DIS_strong.fit)

#Partial Strong/Scalar

DIS_strong.mod<-'
#DIS Factors at baseline through 3 year follow up, with constraints on factor loadings
DIS_t1 =~ NA*DIS_parc1.2_t1+l1*DIS_parc1.2_t1+l2*DIS_parc2.2_t1+l3*DIS_parc3.2_t1+l4*DIS_parc4.2_t1 
DIS_t2 =~ NA*DIS_parc1.2_t2+l1*DIS_parc1.2_t2+l2*DIS_parc2.2_t2+l3*DIS_parc3.2_t2+l4*DIS_parc4.2_t2
DIS_t3 =~ NA*DIS_parc1.2_t3+l1*DIS_parc1.2_t3+l2*DIS_parc2.2_t3+l3*DIS_parc3.2_t3+l4*DIS_parc4.2_t3
DIS_t4 =~ NA*DIS_parc1.2_t4+l1*DIS_parc1.2_t4+l2*DIS_parc2.2_t4+l3*DIS_parc3.2_t4+l4*DIS_parc4.2_t4

#Baseline latent mean fixed at zero, others freed 
DIS_t1 ~ 0*1
DIS_t2 ~ 1
DIS_t3 ~ 1
DIS_t4 ~ 1

#Latent Variances
DIS_t1~~1*DIS_t1
DIS_t2~~DIS_t2
DIS_t3~~DIS_t3
DIS_t4~~DIS_t4

#Constrain Parcel Intercepts
DIS_parc1.2_t1~t1*1
DIS_parc2.2_t1~t2*1
DIS_parc3.2_t1~1
DIS_parc4.2_t1~1

DIS_parc1.2_t2~t1*1
DIS_parc2.2_t2~t2*1
DIS_parc3.2_t2~t3*1
DIS_parc4.2_t2~t4*1

DIS_parc1.2_t3~t1*1
DIS_parc2.2_t3~t2*1
DIS_parc3.2_t3~t3*1
DIS_parc4.2_t3~t4*1

DIS_parc1.2_t4~t1*1
DIS_parc2.2_t4~t2*1
DIS_parc3.2_t4~1
DIS_parc4.2_t4~1

#Residual Covariances
DIS_parc1.2_t1 ~~ DIS_parc1.2_t2
DIS_parc1.2_t1 ~~ DIS_parc1.2_t3
DIS_parc1.2_t1 ~~ DIS_parc1.2_t4
DIS_parc1.2_t2 ~~ DIS_parc1.2_t3
DIS_parc1.2_t2 ~~ DIS_parc1.2_t4
DIS_parc1.2_t3 ~~ DIS_parc1.2_t4

DIS_parc2.2_t1 ~~ DIS_parc2.2_t2
DIS_parc2.2_t1 ~~ DIS_parc2.2_t3
DIS_parc2.2_t1 ~~ DIS_parc2.2_t4
DIS_parc2.2_t2 ~~ DIS_parc2.2_t3
DIS_parc2.2_t2 ~~ DIS_parc2.2_t4
DIS_parc2.2_t3 ~~ DIS_parc2.2_t4

DIS_parc3.2_t1 ~~ DIS_parc3.2_t2
DIS_parc3.2_t1 ~~ DIS_parc3.2_t3
DIS_parc3.2_t1 ~~ DIS_parc3.2_t4
DIS_parc3.2_t2 ~~ DIS_parc3.2_t3
DIS_parc3.2_t2 ~~ DIS_parc3.2_t4
DIS_parc3.2_t3 ~~ DIS_parc3.2_t4

DIS_parc4.2_t1 ~~ DIS_parc4.2_t2
DIS_parc4.2_t1 ~~ DIS_parc4.2_t3
DIS_parc4.2_t1 ~~ DIS_parc4.2_t4
DIS_parc4.2_t2 ~~ DIS_parc4.2_t3
DIS_parc4.2_t2 ~~ DIS_parc4.2_t4
DIS_parc4.2_t3 ~~ DIS_parc4.2_t4
'

DIS_strong.mod.fit<-cfa(DIS_strong.mod, data=cbcl_wide, missing='FIML', estimator='MLR')
summary(DIS_strong.mod.fit, fit.measures=T, standardized=T)
DIS_weak.strong_test<-compareFit(DIS_weak.fit, DIS_strong.mod.fit)
summary(DIS_weak.strong_test)

sbs.chi(524.247, 387.31, 88, 83, 1.668, 1.637) 
exp(-0.5*((387.307 - 83)/(11866-1)))-exp(-0.5*((524.247 - 88)/(11866-1))) #change in McDonald's NCI

#Get effect sizes for MI using dmacs package 
lavaan_dmacs(DIS_strong.mod.fit, MEtype='long') #partial strong invariance model is input

#Get values for dMACS table in manuscript 
DIS_descript<-cbcl_wide %>% 
  rowwise() %>% 
  mutate(DIS_t1=sum(c(DIS_parc1.2_t1, DIS_parc2.2_t1, DIS_parc3.2_t1, DIS_parc4.2_t1), na.rm=T),
         DIS_t2=sum(c(DIS_parc1.2_t2, DIS_parc2.2_t2, DIS_parc3.2_t2, DIS_parc4.2_t2), na.rm=T),
         DIS_t3=sum(c(DIS_parc1.2_t3, DIS_parc2.2_t3, DIS_parc3.2_t3, DIS_parc4.2_t3), na.rm=T),
         DIS_t4=sum(c(DIS_parc1.2_t4, DIS_parc2.2_t4, DIS_parc3.2_t4, DIS_parc4.2_t4), na.rm=T))

describe(select(DIS_descript, DIS_t1, DIS_t2, DIS_t3, DIS_t4))

########
##For MPLUS Only
########

####Code to include site/family IDs

#Subset specific assessment periods 
site_df_r<-subset(site_df, eventname %in% c("baseline_year_1_arm_1",
                                          "1_year_follow_up_y_arm_1",
                                          "2_year_follow_up_y_arm_1",
                                          "3_year_follow_up_y_arm_1"))

#Recode time to make code easier to work with 
site_df_r <-site_df_r %>%
  mutate(eventname_r=as_factor(case_match(eventname, 
                                          'baseline_year_1_arm_1'~'t1',
                                          "1_year_follow_up_y_arm_1"~'t2',
                                          "2_year_follow_up_y_arm_1"~'t3',
                                          "3_year_follow_up_y_arm_1"~'t4')))

df_list.MPlus<-list(cbcl_long, #note that 'cbcl_long' is from previous code, prior to converting it to wide format
                    site_df_r)

abcd_merged.MPlus <- df_list.MPlus %>% 
  reduce(left_join, by=c('src_subject_id','eventname_r'))

abcd_merged.MPlus_subset<-abcd_merged.MPlus %>% 
  select(src_subject_id, eventname_r, site_id_l, rel_family_id, EXT_parc1:DIS_parc4.2)

#Convert to wide         
abcd_merged.MPlus_wide<- abcd_merged.MPlus_subset %>% 
  pivot_wider(id_cols = "src_subject_id",  
              names_from = "eventname_r",
              values_from=c(site_id_l, rel_family_id,
                            EXT_parc1, EXT_parc2, EXT_parc3, EXT_parc4, EXT_parc5,
                            ANT_parc1.2, ANT_parc2.2, ANT_parc3.2, ANT_parc4.2,
                            IRR_parc1, IRR_parc2, IRR_parc3, IRR_parc4,
                            DIS_parc1.2, DIS_parc2.2, DIS_parc3.2, DIS_parc4.2),
              names_glue = "{.value}_{eventname_r}")

#Subset site id, fam id, and parcels
MPlus_final<-abcd_merged.MPlus_wide %>% 
  select(site_id_l_t1, rel_family_id_t1,
         EXT_parc1_t1:EXT_parc1_t4,
         EXT_parc2_t1:EXT_parc2_t4,
         EXT_parc3_t1:EXT_parc3_t4,
         EXT_parc4_t1:EXT_parc4_t4,
         EXT_parc5_t1:EXT_parc5_t4,
         ANT_parc1.2_t1:ANT_parc1.2_t4,
         ANT_parc2.2_t1:ANT_parc2.2_t4,
         ANT_parc3.2_t1:ANT_parc3.2_t4,
         ANT_parc4.2_t1:ANT_parc4.2_t4,
         IRR_parc1_t1:IRR_parc1_t4,
         IRR_parc2_t1:IRR_parc2_t4,
         IRR_parc3_t1:IRR_parc3_t4,
         IRR_parc4_t1:IRR_parc4_t4,
         DIS_parc1.2_t1:DIS_parc1.2_t4,
         DIS_parc2.2_t1:DIS_parc2.2_t4,
         DIS_parc3.2_t1:DIS_parc3.2_t4,
         DIS_parc4.2_t1:DIS_parc4.2_t4)

MPlus_final$site_id_l_t1<-as.numeric(gsub("[a-zA-Z ]", "", MPlus_final$site_id_l_t1)) #remove string portion of site variable (Mplus can't have text in data)
MPlus_final$rel_family_id_t1<-replace(MPlus_final$rel_family_id_t1, MPlus_final$rel_family_id_t1==0, 2) #Mplus requires positive integers for clusters so 0 can't be used; 2 is not assigned so is used as substitute value

#Replace NA with 9999
MPlus_final[is.na(MPlus_final)] <- 9999 #Replace NAs w/ 9999

#Data for Mplus; top row will need to be deleted 
write.csv(MPlus_final, "ABCD_CBCL_wide.csv", row.names = F)

############
####Sensitivity Analyses 
############

#Bass-ackwards (Use Alternative Factoring Method; Weighted Least Squares)

#Run EFA with WLS extraction, but no rotation method
CBCL_wls<-fa.poly(cbcl_ext_base[,3:47], nfactors=4, fm="wls", rotate="none") #Factoring Method: WLS; Rotation=None
CBCL.loadings_wls<-as.matrix(unclass(CBCL_wls$fa$loadings)) #extract factor loadings

#Use cfQ() function from GPArotation package, which applies oblique equamax rotation to WLS factor loadings
CBCL_equmax_solution<-cfQ(CBCL.loadings_wls, 
                          Tmat=diag(ncol(CBCL.loadings_wls)),
                          kappa=0, 
                          normalize=FALSE, 
                          eps=1e-5, 
                          maxit=1000)

#Compare loadings
CBCL_4_minres_promax<-fa.poly(cbcl_ext_base[,3:47], nfactors = 4, fm="minres", rotate = "promax") #run fa.poly w/ four factors so item loadings are in order (needed for congruence coefficients)
original_loadings<-as.data.frame(unclass(CBCL_4_minres_promax$fa$loadings))
CBCL.loadings_wls_df<-as.data.frame(CBCL_equmax_solution$loadings)

#Factor Congruence (Tucker Congruence Coefficients)
factor.congruence(original_loadings, CBCL.loadings_wls_df) #Near perfect congruence

###Examine impact of variability in item-parcel assignments

library(lavaan)
library(semTools)

cbcl_ext <-cbcl_ext %>%
  mutate(eventname_r=as_factor(case_match(eventname, 
                                          'baseline_year_1_arm_1'~'t1',
                                          "1_year_follow_up_y_arm_1"~'t2',
                                          "2_year_follow_up_y_arm_1"~'t3',
                                          "3_year_follow_up_y_arm_1"~'t4',
                                          "4_year_follow_up_y_arm_1"~'t5'))) #won't use 4 year follow up

#Subset externalizing items
cbcl_long_item<-cbcl_ext %>% 
  select(src_subject_id, eventname_r, 
         cbcl_q66_p, Disobey_comp, cbcl_q72_p, cbcl_q80_p, cbcl_q09_p, cbcl_q85_p, cbcl_q61_p, cbcl_q15_p, cbcl_q67_p,
         cbcl_q17_p, cbcl_q94_p, cbcl_q96_p, cbcl_q87_p, cbcl_q01_p, cbcl_q64_p, cbcl_q41_p, cbcl_q43_p, cbcl_q88_p, 
         cbcl_q04_p, cbcl_q03_p, cbcl_q62_p, cbcl_q86_p, AttThreat_comp, cbcl_q37_p, cbcl_q07_p, cbcl_q16_p, cbcl_q39_p,
         cbcl_q90_p, Steal_comp, cbcl_q46_p, cbcl_q109_p, DistrHyp_comp, cbcl_q74_p, cbcl_q89_p, cbcl_q13_p, cbcl_q19_p,
         cbcl_q95_p, cbcl_q34_p, cbcl_q26_p, cbcl_q68_p, cbcl_q93_p, cbcl_q36_p, Destroy_comp, PeerProb_comp, cbcl_q27_p)

#Convert to wide         
cbcl_wide_item<- cbcl_long_item %>% 
  pivot_wider(id_cols = "src_subject_id", 
              names_from = "eventname_r",
              values_from=c(cbcl_q66_p, Disobey_comp, cbcl_q72_p, cbcl_q80_p, cbcl_q09_p, cbcl_q85_p, cbcl_q61_p, cbcl_q15_p, cbcl_q67_p,
                              cbcl_q17_p, cbcl_q94_p, cbcl_q96_p, cbcl_q87_p, cbcl_q01_p, cbcl_q64_p, cbcl_q41_p, cbcl_q43_p, cbcl_q88_p, 
                              cbcl_q04_p, cbcl_q03_p, cbcl_q62_p, cbcl_q86_p, AttThreat_comp, cbcl_q37_p, cbcl_q07_p, cbcl_q16_p, cbcl_q39_p,
                              cbcl_q90_p, Steal_comp, cbcl_q46_p, cbcl_q109_p, DistrHyp_comp, cbcl_q74_p, cbcl_q89_p, cbcl_q13_p, cbcl_q19_p,
                              cbcl_q95_p, cbcl_q34_p, cbcl_q26_p, cbcl_q68_p, cbcl_q93_p, cbcl_q36_p, Destroy_comp, PeerProb_comp, cbcl_q27_p ),
              names_glue = "{.value}_{eventname_r}")

#Broad-based Externalizing

#Need to specify lavaan model using items instead of parcels
EXT_base_model.item<-'
EXT_t1 =~ cbcl_q66_p_t1+ Disobey_comp_t1+ cbcl_q72_p_t1+ cbcl_q80_p_t1+ cbcl_q09_p_t1+ cbcl_q85_p_t1+ cbcl_q61_p_t1+ cbcl_q15_p_t1+ cbcl_q67_p_t1+
cbcl_q17_p_t1+ cbcl_q94_p_t1+ cbcl_q96_p_t1+ cbcl_q87_p_t1+ cbcl_q01_p_t1+ cbcl_q64_p_t1+ cbcl_q41_p_t1+ cbcl_q43_p_t1+ cbcl_q88_p_t1+ 
cbcl_q04_p_t1+ cbcl_q03_p_t1+ cbcl_q62_p_t1+ cbcl_q86_p_t1+ AttThreat_comp_t1+ cbcl_q37_p_t1+ cbcl_q07_p_t1+ cbcl_q16_p_t1+ cbcl_q39_p_t1+
cbcl_q90_p_t1+ Steal_comp_t1+ cbcl_q46_p_t1+cbcl_q109_p_t1+ DistrHyp_comp_t1+ cbcl_q74_p_t1+ cbcl_q89_p_t1+ cbcl_q13_p_t1+ cbcl_q19_p_t1+
cbcl_q95_p_t1+ cbcl_q34_p_t1+ cbcl_q26_p_t1+ cbcl_q68_p_t1+ cbcl_q93_p_t1+ cbcl_q36_p_t1+ Destroy_comp_t1+ PeerProb_comp_t1+ cbcl_q27_p_t1'

#Broad-based Externalizing Model using parcels

EXT_base_model.parc<-'
EXT_t1 =~ EXT_parc1_t1 + EXT_parc2_t1 + EXT_parc3_t1 + EXT_parc4_t1 + EXT_parc5_t1'

EXT_sens<-parcelAllocation(model=EXT_base_model.parc, 
                 data=cbcl_wide_item, 
                 parcel.names=c('EXT_parc1_t1', 'EXT_parc2_t1', 'EXT_parc3_t1', 'EXT_parc4_t1', 'EXT_parc5_t1'), 
                 item.syntax=EXT_base_model.item, 
                 nAlloc = 500, #run 500 different item-parcel allocations 
                 fun = "cfa", #apply cfa() function from lavaan to data
                 iseed=444, #set seed for reproducibility 
                 std.lv=T, #standardize latent factor and observed variables to get fully standardized solution 
                 std.ov=T,
                 missing='FIML',
                 estimator='MLR',
                 fit.measures = c("chisq.scaled", "rmsea.scaled", "tli.scaled", "cfi.scaled"),
                 alpha = 0.05) #sig. at alpha=.05

##Antagonism

#Item-based model 
ANT_base_model.item<-'
ANT_EXT =~ AttThreat_comp_t1+ cbcl_q72_p_t1+ cbcl_q94_p_t1+ cbcl_q37_p_t1+ 
cbcl_q67_p_t1+ cbcl_q15_p_t1+ cbcl_q43_p_t1+ cbcl_q16_p_t1+
cbcl_q90_p_t1+ Disobey_comp_t1+ cbcl_q39_p_t1+
cbcl_q96_p_t1+ cbcl_q26_p_t1+ Steal_comp_t1+ Destroy_comp_t1+ cbcl_q74_p_t1'

#Using parcels
ANT_base_model.parc<-'
ANT_EXT =~ ANT_parc1.2_t1 + ANT_parc2.2_t1 + ANT_parc3.2_t1 + ANT_parc4.2_t1 '

ANT_sens<-parcelAllocation(model=ANT_base_model.parc,  
                           data=cbcl_wide_item, 
                           parcel.names=c('ANT_parc1.2_t1', 'ANT_parc2.2_t1', 'ANT_parc3.2_t1', 'ANT_parc4.2_t1'), 
                           item.syntax=ANT_base_model.item, 
                           nAlloc = 500,
                           fun = "cfa", #apply cfa() function from lavaan to data
                           iseed=444,
                           std.lv=T, #standardize latent factor and observed variables to get fully standardized solution 
                           std.ov=T,
                           missing='FIML',
                           estimator='MLR',
                           fit.measures = c("chisq.scaled", "rmsea.scaled", "tli.scaled", "cfi.scaled"),
                           alpha = 0.05) #sig. at alpha=.05

##Irritability

#Item-based model 
IRR_base_model.item<-'
IRR_EXT =~ cbcl_q87_p_t1+ cbcl_q109_p_t1+ cbcl_q86_p_t1+
cbcl_q19_p_t1+ cbcl_q68_p_t1+ cbcl_q88_p_t1+
cbcl_q27_p_t1+ cbcl_q34_p_t1+ cbcl_q03_p_t1+
cbcl_q95_p_t1+ cbcl_q89_p_t1 '

#Using parcels
IRR_base_model.parc<-'
IRR_EXT =~ IRR_parc1_t1 + IRR_parc2_t1 + IRR_parc3_t1 + IRR_parc4_t1 '

IRR_sens<-parcelAllocation(model=IRR_base_model.parc,  
                           data=cbcl_wide_item, 
                           parcel.names=c('IRR_parc1_t1', 'IRR_parc2_t1', 'IRR_parc3_t1', 'IRR_parc4_t1'), 
                           item.syntax=IRR_base_model.item, 
                           nAlloc = 500,
                           fun = "cfa", #apply cfa() function from lavaan to data
                           iseed=555,
                           std.lv=T, #standardize latent factor and observed variables to get fully standardized solution 
                           std.ov=T,
                           missing='FIML',
                           estimator='MLR',
                           fit.measures = c("chisq.scaled", "rmsea.scaled", "tli.scaled", "cfi.scaled"),
                           alpha = 0.05) #sig. at alpha=.05

#Disinhibition

#Item-based model
DIS_base_model.item<-'
DIS_EXT =~ cbcl_q01_p_t1+ DistrHyp_comp_t1+ cbcl_q46_p_t1+ cbcl_q09_p_t1+
cbcl_q64_p_t1+ cbcl_q66_p_t1+ cbcl_q80_p_t1+ cbcl_q85_p_t1+ 
cbcl_q13_p_t1+ cbcl_q61_p_t1+ cbcl_q36_p_t1+ cbcl_q04_p_t1+
cbcl_q62_p_t1+ cbcl_q17_p_t1+ cbcl_q93_p_t1'

#Parcel model
DIS_base_model.parc<-'
DIS_EXT =~ DIS_parc1.2_t1 + DIS_parc2.2_t1 + DIS_parc3.2_t1 + DIS_parc4.2_t1 '

DIS_sens<-parcelAllocation(model=DIS_base_model.parc, 
                 data=cbcl_wide_item, 
                 parcel.names=c('DIS_parc1.2_t1', 'DIS_parc2.2_t1', 'DIS_parc3.2_t1', 'DIS_parc4.2_t1'), 
                 item.syntax=DIS_base_model.item, 
                 nAlloc = 500,
                 fun = "cfa", #apply cfa() function from lavaan to data
                 iseed=444,
                 std.lv=T, #standardize latent factor and observed variables to get fully standardized solution 
                 std.ov=T,
                 missing='FIML',
                 estimator='MLR',
                 fit.measures = c("chisq.scaled", "rmsea.scaled", "tli.scaled", "cfi.scaled"),
                 alpha = 0.05) #sig. at alpha=.05

###Examine the impact of clustering variables (Site and Family)

##See MPlus output on https://osf.io/ec36x/?view_only=d63a1452f133421b9e1bef34a41675f1

```

### 2024-03-06 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
update manuscript template

--- 2024-03-06 examples/_queue/_1a_Examples_DifferenceScores_PairedTtests.qmd
@@
 ```
-
-**Descriptive Statistics for the Difference Score**
-```{r model-checking}
-# Compute statistical summaries for the difference score variable
-lapply(difference_and_summary_list, function(item) {
-    print(item$summary)
-})
-
-gaze <- df %>%
-  mutate(diff = baseline_event - followup_event) %>%
-  mutate(group_resid = diff - mean(diff))
-
-gaze <- weight_diff_data %>%
-  mutate(diff1 = baseline_year_1_arm_1 - 1_year_follow_up_y_arm_1) %>%
-  mutate(group_resid = diff - mean(diff))
-
-dff.y
-
-qqPlot(gaze_residual$group_resid)
-
-shapiro.test(gaze_residual$group_resid)
-
-```
-
-This summary of the difference score variable indicates that the differences range from a decrease of `r min(weight_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(weight_diff_data$diff, na.rm = TRUE)` units.
-The median difference is `r median(weight_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(weight_diff_data$diff, digits = 2, na.rm = TRUE)` units.
-There are `r sum(is.na(weight_diff_data$diff))` missing values in this difference score dataset.
-
-**Descriptive Statistics for the weight Variable**
-```{r}
-## Summary statistics
-# Compute summary statistics for weight by eventname
-summary <- df %>%
-    group_by(eventname) %>%
-    get_summary_stats(anthroweightcalc, type = "mean_sd")
-
-data.frame(summary)
-```
-
-The summary statistics provide insights into the weight variable across different `eventname` categories.
-At baseline, the average weight is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`.
-Over the years, there’s a noticeable increase in average weight: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year.
-The standard deviation remains relatively consistent over the years, suggesting similar variability in weight measurements across these time points.
-
-:::
-
-### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-A one-sample t-test is computed to examine whether the average difference score is different than zero (indicating change).
-
-**STEP 2: Conduct t-test on Difference Score**
-
-```{r build-model-ttest-2}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# Extract the difference scores for weight from the computed list
-diff_data <- difference_and_summary_list[[1]]$data$diff
-
-# Perform a one-sample t-test on the difference scores for weight
-test_result <- t.test(diff_data, mu = 0, na.rm = TRUE)
-
-test_result
-
-gaze_test <- t.test(anthroweightcalc ~ eventname, 
-                    paired = TRUE, 
-                    data = df) %>% 
-  tidy()
-
-```
-
-This one-sample t-test evaluates whether the average difference score deviates significantly from 0. The model output provides:
-
-i. a t-statistic of `r test_result$statistic`;
-ii. degrees of freedom of `r test_result$parameter`;
-iii. a p-value of `r test_result$p.value`;
-iv. a mean estimate of `r test_result$estimate["mean of x"]`;
-v. 95% confidence intervals from `r test_result$conf.int[1]` to `r test_result$conf.int[2]`.
-
-v. 95% confidence intervals from `r round(test_result$conf.int[1], 3)` to `r round(test_result$conf.int[2], 3)`.
-
-With a t-value of `r test_result$statistic` and a p-value of `r test_result$p.value`, we can determine the significance of the difference.
-The average difference score across the sample is approximately `r round(test_result$estimate["mean of x"], 2)`.
-Furthermore, we can be 95% confident that the true mean difference score in the population lies between approximately `r round(test_result$conf.int[1], 2)` and `r round(test_result$conf.int[2], 2)`.
-This indicates the degree to which there's a change in weight from Baseline to Year 1.
-
-:::
-
-### Model Plots
-
-::: blue
-```{r scatterplot}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
-
-# Filter data for Baseline
-baseline_data <- df %>% filter(eventname == "Baseline")
-
-# Filter data for Year_1 and join with Baseline data
-joined_data <- baseline_data %>%
-    left_join(df %>% filter(eventname == "Year_1"), by = "src_subject_id", suffix = c("_baseline", "_year1"))
-
-# Create scatterplot
-scatterplot <- ggplot(df, aes(x = anthroweightcalc, y = diff)) +
-    geom_point(aes(color = sex), alpha = 0.5) + # Setting alpha to ensure points are visible
-    geom_smooth(method = "lm", color = "red") + # Adding linear regression line
-    labs(
-        title = "Scatterplot of anthroweightcalc vs. Difference Score",
-        x = "anthroweightcalc",
-        y = "Difference Score"
-    ) +
-    theme_minimal()
-
-scatterplot
-
-```
-
-The scatterplot visually depicts the relationship between individuals' weights at baseline and their weights at Year_1.
-Each point on the plot represents an individual, with their baseline weight plotted on the x-axis and their Year_1 weight on the y-axis.
-A noticeable positive linear trend can be observed, as highlighted by the blue regression line, indicating that those who were taller at baseline generally remained taller at Year_1.
-The strength and direction of this relationship can be further quantified with correlation coefficients, but visually, the data suggests a strong positive association between baseline and Year_1 weights.
-
-:::
-:::
-
-## Wrapping Up
-
-::: panel-tabset
-### Write-up {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-In the study sample, the average weight at baseline was approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`, which showed a noticeable increase to an average of `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units by Year_1.
-A one-sample t-test was conducted to determine if the mean difference in weight from baseline to Year_1 significantly deviated from zero.
-The results indicated a statistically significant increase with a mean difference of approximately `r round(test_result$estimate, 2)` units (`r round(test_result$conf.int[1], 2)` to `r round(test_result$conf.int[2], 2)`), t(`r test_result$parameter`) = `r round(test_result$statistic, 3)`, p < `r test_result$p.value`.
-Further, a scatterplot visualizing the relationship between baseline and Year_1 weights showed a strong positive linear trend.
-This suggests that participants who were taller at baseline generally remained taller at Year_1, reaffirming the consistent growth trend observed in the data.
-
-:::
-:::

### 2024-03-06 manuscript/_manuscript-template.qmd
update manuscript template

--- 2024-03-06 manuscript/manuscript-template.qmd
@@
+---
+title: "Manuscript Title"
+authors:
+  - name: Jane Doe
+    email: "janedoe@mockuniversity.edu"
+    affiliations:
+      - name: University of Mock 
+        department: Department of Something 
+        city: Miami
+        state: Florida
+        country: United States
+  - name: Author 2
+    email: "xxxxxx"
+    affiliations:
+      - name: xxxxx
+        department: xxxxxx
+        city: xxxxx
+        state: xxxxx
+        country: xxxxxx
+  - name: Additional Co-author(s)
+    email: "xxxxxxx"
+    affiliations:
+      - name: xxxxx
+        department: xxxxxx
+        city: xxxx
+        state: xxxxx
+        country: xxxxxx
+
+abstract: |
+  The is the abstract section.
+Ipsum leo porta – lacinia sed ac interdum nec vel ad.
+Elementum augue platea eleifend himenaeos nisl, viverra sem molestie.
+Hendrerit enim torquent litora fusce ultrices, magna, lacus id potenti!
+Imperdiet vulputate etiam conubia sem mi incep- tos.
+Quisque mollis viverra lacinia bibendum ultricies, dui est gravida tristique nascetur.
+Pellentesque id elementum com- modo pretium quisque purus mi taciti.
+
+index-terms: 
+  - Keyword 1
+  - Keyword 2 
+  - Keyword 3
+  - Keyword 4
+  - Keyword 5
+
+format:
+  ieee-typst: default
+bibliography: references.bib
+---
+```{r}
+#| echo: FALSE
+library(lorem)
+```
+
+## Introduction
+_Introduce the research topic, state the research problem, and outline the study's objectives. Provide background information to set the context. This is an exmaple in-text citation_ @netwok2020.
+`r lorem::ipsum(paragraphs = 3)`
+
+### Background
+_Provide a brief review of the pertinent literature, highlighting the gaps your study aims to fill._
+`r lorem::ipsum(paragraphs = 2)`
+
+### Objectives
+_State the specific research questions or hypotheses your study addresses._
+`r lorem::ipsum(paragraphs = 2)`
+
+## Methods
+_Describe the methodological approach, detailing how the study was conducted, how data was collected, and how analyses were performed._
+`r lorem::ipsum(paragraphs = 1)`
+
+### Participants
+_Describe the demographic characteristics of the study participants, including how they were selected, recruited, and any inclusion or exclusion criteria._
+`r lorem::ipsum(paragraphs = 1)`
+
+### Measures
+_List the instruments or tools used to collect data, describing each measure's purpose and any relevant psychometric properties._
+`r lorem::ipsum(paragraphs = 2)`
+
+### Analytic Plan
+_Detail the statistical tests or models employed to analyze the data, including any software used._
+`r lorem::ipsum(paragraphs = 1)`
+
+## Results
+
+```{r}
+#| echo: FALSE
+#| output: FALSE
+
+# This is an example of a basic analysis code chunk
+## Load necessary libraries
+library(ggplot2)
+
+## Simple linear regression analysis with 'mtcars' dataset
+lm_model <- lm(mpg ~ wt, data = mtcars)
+
+## Summary of the linear model
+summary(lm_model)
+
+```
+
+```{r, broomoutput}
+#| echo: FALSE
+
+## Using broom to tidy the regression model results for presentation in-text
+library(broom)
+
+library(broom)
+
+# Tidy the regression model output for easy access to coefficients
+tidy_lm <- broom::tidy(lm_model)
+
+# Round the coefficients to 2 decimal places
+tidy_lm$estimate <- round(tidy_lm$estimate, 2)
+tidy_lm$std.error <- round(tidy_lm$std.error, 2)
+tidy_lm$statistic <- round(tidy_lm$statistic, 2)
+tidy_lm$p.value <- round(tidy_lm$p.value, 2)
+
+# Extract the rounded coefficient and p-value for 'wt'
+wt_coefficient <- tidy_lm$estimate[2] # The coefficient of 'wt', rounded to 2 decimal places
+wt_pvalue <- tidy_lm$p.value[2] # The p-value for the 'wt' coefficient, rounded to 2 decimal places
+
+# Glance at the model's summary statistics to get R-squared and round it
+model_summary <- broom::glance(lm_model)
+
+# Round the summary statistics to 2 decimal places
+model_summary$r.squared <- round(model_summary$r.squared, 2)
+model_summary$statistic <- round(model_summary$statistic, 2)
+model_summary$p.value <- round(model_summary$p.value, 2)
+
+# Assign the rounded values to variables
+r_squared <- model_summary$r.squared # R-squared value, rounded to 2 decimal places
+f_statistic <- model_summary$statistic # F-statistic, rounded to 2 decimal places
+model_pvalue <- model_summary$p.value # Model p-value, rounded to 2 decimal places
+
+```
+
+The regression analysis revealed that for every additional 1,000 pounds of weight, the car's miles per gallon decreases by an average of `r wt_coefficient` units (p = `r wt_pvalue`).
+This model explains `r r_squared`% of the variance in fuel efficiency, indicating a strong relationship between car weight and mpg.
+The model's F-statistic is `r f_statistic`, with a p-value of `r model_pvalue`, demonstrating the statistical significance of the regression model.
+
+The following portion of code [geom_smooth()` using formula = 'y ~ x'] still shows in the rendered text even though I have set `#| echo: FALSE`.
+```{r, fig.width=4, fig.height=4}
+#| echo: FALSE
+
+# This is an example of a basic scatterplot of the regression analysis.
+
+ggplot(mtcars, aes(x = wt, y = mpg)) +
+    geom_point() +
+    geom_smooth(method = "lm", col = "blue") +
+    labs(title = "MPG vs. Car Weight", x = "Weight (1000 lbs)", y = "Miles per Gallon") +
+    theme_minimal()
+
+```
+
+```{r, fig.width=4, fig.height=4}
+#| echo: FALSE
+
+# This is an example of a diagnostic plot of the regression analysis.
+
+par(mfrow = c(2, 2)) # Arrange plots in a 2x2 grid
+plot(lm_model)
+
+```
+
+### [Subsection Title]
+_Present additional findings related to a particular aspect of the study, supported by data._
+`r lorem::ipsum(paragraphs = 1)`
+
+## Discussion
+_Interpret the results, discussing their implications in the context of existing literature. Address the study's limitations and suggest directions for future research._
+`r lorem::ipsum(paragraphs = 3)`
+
+### Limitations
+_Acknowledge any limitations of your study and their potential impact on the findings._
+`r lorem::ipsum(paragraphs = 1)`
+
+### Future Directions
+_Suggest areas for future research that could build on your findings or address limitations._
+`r lorem::ipsum(paragraphs = 1)`
+
+## Conclusion
+_Summarize the main findings and their implications. Conclude with a statement that reflects the significance of the study and its contributions to the field._
+`r lorem::ipsum(paragraphs = 1)`
+
+## Acknowledgments
+_Acknowledge anyone who contributed to the research but does not meet authorship criteria, along with any funding sources._
+`r lorem::ipsum(paragraphs = 1)`
+
+## Appendices
+_Include any additional material that is pertinent to the study but would be too distracting or bulky to include in the main body of the manuscript._
+`r lorem::ipsum(paragraphs = 1)`
+
+#########
+#########
+#########
+
 ---

### 2024-03-06 index.qmd
updates

@@
   html:
-    css: styles.css
+    #css: styles.css
     #theme: cosmo

### 2024-03-06 examples/examples.qmd
updated site output directory

@@
 #title: "Examples: A Brief Overview"
-format:
-  html:
-    css: styles.css
-    theme: cosmo
+# format:
+#   html:
+    # css: styles.css
+    # theme: cosmo
     #page-layout: full

### 2024-03-06 manuscript/manuscript.qmd
updated site output directory

--- 2024-03-02 manuscript/other_manuscript_draft.qmd
@@
     toc: false
-  elsevier-pdf:
-    journal:
-      cite-style: number
-      formatting: preprint
-      model: 3p
-      layout: twocolumn
-      #cite-style: number
-    include-in-header: 
-      text: |
-        \newpageafter{author}
+  # elsevier-pdf:
+  #   journal:
+  #     cite-style: number
+  #     formatting: preprint
+  #     model: 3p
+  #     layout: twocolumn
+  #     #cite-style: number
+  #   include-in-header: 
+  #     text: |
+  #       \newpageafter{author}
+  ieee-typst: default
   html:
@@
 ---
+
+[test](./img/lights.png)
+<!-- [Types of Stability and Change](img/StabilityChange.png) -->
 
@@
 
-![Types of Stability and Change](./figures/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}
+<!-- ![Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"} -->
 
@@
 While much has been written about “time-to-event” data (**see xxx; xxx; xxxx**), including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study (@brislin2023), our emphasis will be given to the modeling of “repeated measures” data.
-
-[![Longitudinal Data Analysis Structural Diagram](./figures/LDAStructure2.png)](https://centerstat.org/resources/)
+manuscript/
+<!-- [![Longitudinal Data Analysis Structural Diagram](../img/LDAStructure2.png)](https://centerstat.org/resources/) -->
 
@@
 
-![Longitudinal Models/Data Structures](./figures/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"} <br>
+<!-- ![Longitudinal Models/Data Structures](img/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"} <br> -->
 
@@
 
-![LDA Table](./figures/Tabletemp.png)
+<!-- ![LDA Table](img/Tabletemp.png) -->
 

### 2024-03-06 examples/data_shift_code.qmd
subsetting

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroweightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroweightcalc = as.numeric(anthroweightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))
```

-----

```{r}

Randomly drawn subsample
# Assuming your dataframe is named df

# Set seed for reproducibility
set.seed(123)

# Calculate the number of rows to sample (30% of the dataset)
sample_size <- floor(0.3 * nrow(df))

# Randomly sample row indices
sampled_indices <- sample(x = 1:nrow(df), size = sample_size)

# Subset the dataframe using the sampled indices
df_sampled <- df[sampled_indices, ]

# df_sampled now contains a random 30% cross-section of the full dataset

```

----

```{r}
# Generating random/unique participant IDs
set.seed(123) # For reproducibility
# Create a vector of random unique IDs
new_ids <- sample(10000:99999, size = nrow(df_sampled), replace = FALSE)

# Assign the new IDs to the participant_id column or to a new column
df_sampled$new_participant_id <- new_ids

# Optionally, if you want to replace the original ID column:
df_sampled$participant_id <- df_sampled$new_participant_id

```

### 2024-03-06 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
updates

@@
 
+----
+
+```{r}
+# Install and load the googledrive package
+if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
+library(googledrive)
+
+# Authenticate with Google Drive
+
+# Specify the file ID and the destination path
+file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
+dest_file <- "my_data.rds" # Specify your destination path
+
+# Download the file
+drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
+
+# Load the .rds file for analysis
+my_data <- readRDS(dest_file)
+
+# Begin your analysis (example: view the first few rows)
+head(my_data)
+
+```
+
+----
+
 ### Descriptives

### 2024-03-06 examples/data_shift_code.qmd
updates

@@
 # Define event names to be retained in the analysis and convert variables to appropriate data types
-eventnames_to_include <- c("baseline_year_1_arm_1",
+eventnames_to_include <- c(
+    "baseline_year_1_arm_1",
     "1_year_follow_up_y_arm_1",
@@
     "3_year_follow_up_y_arm_1",
-                           "4_year_follow_up_y_arm_1")
+    "4_year_follow_up_y_arm_1"
+)
 
-df <- merged_data %>%
+df_temp1 <- merged_data %>%
     filter(eventname %in% eventnames_to_include) %>%
@@
         acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
-    rel_family_id.bl = as.factor(rel_family_id.bl),
         site_id_l = as.factor(site_id_l),
@@
 ```{r}
-
-Randomly drawn subsample
-# Assuming your dataframe is named df
+# Creating a randomly drawn subsample
+df_sampled_temp <- df_temp1
 
@@
 # Calculate the number of rows to sample (30% of the dataset)
-sample_size <- floor(0.3 * nrow(df))
+sample_size <- floor(0.3 * nrow(df_sampled_temp))
 
@@
 # Subset the dataframe using the sampled indices
-df_sampled <- df[sampled_indices, ]
+df_sampled <- df_sampled_temp[sampled_indices, ]
 
@@
 set.seed(123) # For reproducibility
-# Create a vector of random unique IDs
-new_ids <- sample(10000:99999, size = nrow(df_sampled), replace = FALSE)
 
-# Assign the new IDs to the participant_id column or to a new column
-df_sampled$new_participant_id <- new_ids
+# Create a vector of random unique IDs and directly replace the original ID column
+df_sampled$src_subject_id <- sample(10000:99999, size = nrow(df_sampled), replace = FALSE)
 
-# Optionally, if you want to replace the original ID column:
-df_sampled$participant_id <- df_sampled$new_participant_id
+df <- df_sampled
+
+# Save df in the current working directory as my_dataframe.rds
+saveRDS(df, "FaKEbcd.rds")
 
 ```
+
+```{r}
+# Install and load the googledrive package
+if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
+library(googledrive)
+
+# Authenticate with Google Drive
+
+# Specify the file ID and the destination path
+file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
+dest_file <- "my_data.rds" # Specify your destination path
+
+# Download the file
+drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
+
+# Load the .rds file for analysis
+my_data <- readRDS(dest_file)
+
+# Begin your analysis (example: view the first few rows)
+head(my_data)
+
+```

### 2024-03-06 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
updates

@@
 
-```{r read-data}
+<!-- ```{r read-data}
 #| echo: TRUE
@@
 ```
-:::
+::: -->
 
@@
 
-```{r}
+<!-- ```{r}
 # Install and load the googledrive package
@@
 
+``` -->
+
+```{r}
+# Set the URL for the CSV file
+file_url <- "https://drive.google.com/file/d/1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr/view?usp=drive_link"
+
+# Set the destination filename
+dest_file <- "my_data.rds"
+
+# Download the file
+download.file(file_url, destfile = dest_file, mode = "wb")
+
+# Read the CSV file into R
+my_data <- readRDS(dest_file)
+
+# Display the first few rows of the data
+head(my_data)
 ```

### 2024-03-07 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
codespace updates

--- 2024-03-06 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
@@
 # Create a list of required packages
-packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")
+packages_required <- c("tidyverse", "easystats", "lme4", "gtsummary", "report", "broom", "gridExtra")
 
@@
 ```
+
 :::
@@
 
-knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
-                      comment=NA, cache=T, code_folding=T,
-                      R.options=list(width=220, digits = 3),
-                      fig.align='center', 
-                      out.width='75%', fig.asp=.75)
+knitr::opts_chunk$set(
+    echo = T, message = F, warning = F, error = F,
+    comment = NA, cache = T, code_folding = T,
+    R.options = list(width = 220, digits = 3),
+    fig.align = "center",
+    out.width = "75%", fig.asp = .75
+)
 ```
-:::
+
 :::
@@
 
-::: blue
+```{r}
+# Install and load the googledrive package
+if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
+library(googledrive)
+
+# Authenticate with Google Drive
+drive_auth()
+
+# Specify the file ID and the destination path
+file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
+dest_file <- "my_data.rds" # Specify your destination path
+
+# Download the file
+drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
+
+# Load the .rds file for analysis
+my_data <- readRDS(dest_file)
+
+# Begin your analysis (example: view the first few rows)
+head(my_data)
+
+```
+
+:::
+
+<!-- ::: blue
 **This code reads in and shows the data to be used in the current
-example**
+example** -->
 
-```{r read-data}
+<!-- ```{r read-data}
 #| echo: TRUE
@@
 ```
-:::
-
-----
-
-```{r}
-# Install and load the googledrive package
-if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
-library(googledrive)
-
-# Authenticate with Google Drive
-
-# Specify the file ID and the destination path
-file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
-dest_file <- "my_data.rds" # Specify your destination path
-
-# Download the file
-drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
-
-# Load the .rds file for analysis
-my_data <- readRDS(dest_file)
-
-# Begin your analysis (example: view the first few rows)
-head(my_data)
-
-```
-
-----
+::: -->
 
@@
 
+```
+
 ```{r descriptives}
@@
 
+df <- my_data
+
 descriptives_table <- df %>%
     select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
-  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
+    mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1", "Year 2", "Year 3", "Year 4"))) %>%
     mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
@@
         missing = "no",
-    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
-                 anthroweightcalc ~ "Weight"),
+        label = list(
+            sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic",
+            anthroweightcalc ~ "Weight"
+        ),
         statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
@@
 ```
+
+**Descriptive Statistics for the Difference Score**
+```{r model-checking}
+# Compute statistical summaries for the difference score variable
+lapply(difference_and_summary_list, function(item) {
+    print(item$summary)
+})
+
+gaze <- df %>%
+  mutate(diff = baseline_event - followup_event) %>%
+  mutate(group_resid = diff - mean(diff))
+
+gaze <- weight_diff_data %>%
+  mutate(diff1 = baseline_year_1_arm_1 - 1_year_follow_up_y_arm_1) %>%
+  mutate(group_resid = diff - mean(diff))
+
+dff.y
+
+qqPlot(gaze_residual$group_resid)
+
+shapiro.test(gaze_residual$group_resid)
+
+```
+
+This summary of the difference score variable indicates that the differences range from a decrease of `r min(weight_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(weight_diff_data$diff, na.rm = TRUE)` units.
+The median difference is `r median(weight_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(weight_diff_data$diff, digits = 2, na.rm = TRUE)` units.
+There are `r sum(is.na(weight_diff_data$diff))` missing values in this difference score dataset.
+
+**Descriptive Statistics for the weight Variable**
+```{r}
+## Summary statistics
+# Compute summary statistics for weight by eventname
+library(dplyr)
+summary <- df %>%
+    group_by(eventname) %>%
+    get_summary_stats(anthroweightcalc, type = "mean_sd")
+
+data.frame(summary)
+```
+
+The summary statistics provide insights into the weight variable across different `eventname` categories.
+At baseline, the average weight is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`.
+Over the years, there’s a noticeable increase in average weight: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year.
+The standard deviation remains relatively consistent over the years, suggesting similar variability in weight measurements across these time points.
+
+:::
+
+### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+
+A one-sample t-test is computed to examine whether the average difference score is different than zero (indicating change).
+
+**STEP 2: Conduct t-test on Difference Score**
+
+```{r build-model-ttest-2}
+#| class.source: 'fold-hide'
+#| message: FALSE
+#| warning: FALSE
+#| echo: TRUE
+
+# Extract the difference scores for weight from the computed list
+diff_data <- difference_and_summary_list[[1]]$data$diff
+
+# Perform a one-sample t-test on the difference scores for weight
+test_result <- t.test(diff_data, mu = 0, na.rm = TRUE)
+
+test_result
+
+gaze_test <- t.test(anthroweightcalc ~ eventname, 
+                    paired = TRUE, 
+                    data = df) %>% 
+  tidy()
+
+```
+
+This one-sample t-test evaluates whether the average difference score deviates significantly from 0. The model output provides:
+
+i. a t-statistic of `r test_result$statistic`;
+ii. degrees of freedom of `r test_result$parameter`;
+iii. a p-value of `r test_result$p.value`;
+iv. a mean estimate of `r test_result$estimate["mean of x"]`;
+v. 95% confidence intervals from `r test_result$conf.int[1]` to `r test_result$conf.int[2]`.
+
+v. 95% confidence intervals from `r round(test_result$conf.int[1], 3)` to `r round(test_result$conf.int[2], 3)`.
+
+With a t-value of `r test_result$statistic` and a p-value of `r test_result$p.value`, we can determine the significance of the difference.
+The average difference score across the sample is approximately `r round(test_result$estimate["mean of x"], 2)`.
+Furthermore, we can be 95% confident that the true mean difference score in the population lies between approximately `r round(test_result$conf.int[1], 2)` and `r round(test_result$conf.int[2], 2)`.
+This indicates the degree to which there's a change in weight from Baseline to Year 1.
+
+:::
+
+### Model Plots
+
+::: blue
+```{r scatterplot}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+#| code-summary: testing
+
+# Filter data for Baseline
+baseline_data <- df %>% filter(eventname == "Baseline")
+
+# Filter data for Year_1 and join with Baseline data
+joined_data <- baseline_data %>%
+    left_join(df %>% filter(eventname == "Year_1"), by = "src_subject_id", suffix = c("_baseline", "_year1"))
+
+# Create scatterplot
+scatterplot <- ggplot(df, aes(x = anthroweightcalc, y = diff)) +
+    geom_point(aes(color = sex), alpha = 0.5) + # Setting alpha to ensure points are visible
+    geom_smooth(method = "lm", color = "red") + # Adding linear regression line
+    labs(
+        title = "Scatterplot of anthroweightcalc vs. Difference Score",
+        x = "anthroweightcalc",
+        y = "Difference Score"
+    ) +
+    theme_minimal()
+
+scatterplot
+
+```
+
+The scatterplot visually depicts the relationship between individuals' weights at baseline and their weights at Year_1.
+Each point on the plot represents an individual, with their baseline weight plotted on the x-axis and their Year_1 weight on the y-axis.
+A noticeable positive linear trend can be observed, as highlighted by the blue regression line, indicating that those who were taller at baseline generally remained taller at Year_1.
+The strength and direction of this relationship can be further quantified with correlation coefficients, but visually, the data suggests a strong positive association between baseline and Year_1 weights.
+
+:::
+:::

### 2024-03-07 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
example testing

@@
 # Create a list of required packages
-packages_required <- c("tidyverse", "easystats", "lme4", "gtsummary", "report", "broom", "gridExtra")
+packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")
 
@@
 ```
-
 :::
@@
 
-knitr::opts_chunk$set(
-    echo = T, message = F, warning = F, error = F,
-    comment = NA, cache = T, code_folding = T,
-    R.options = list(width = 220, digits = 3),
-    fig.align = "center",
-    out.width = "75%", fig.asp = .75
-)
+knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
+                      comment=NA, cache=T, code_folding=T,
+                      R.options=list(width=220, digits = 3),
+                      fig.align='center', 
+                      out.width='75%', fig.asp=.75)
 ```
-
+:::
 :::
@@
 
-```{r}
-# Install and load the googledrive package
-if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
-library(googledrive)
-
-# Authenticate with Google Drive
-drive_auth()
-
-# Specify the file ID and the destination path
-file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
-dest_file <- "my_data.rds" # Specify your destination path
-
-# Download the file
-drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
-
-# Load the .rds file for analysis
-my_data <- readRDS(dest_file)
-
-# Begin your analysis (example: view the first few rows)
-head(my_data)
-
-```
-
-:::
-
-<!-- ::: blue
+::: blue
 **This code reads in and shows the data to be used in the current
-example** -->
+example**
 
@@
 
+----
+
+<!-- ```{r}
+# Install and load the googledrive package
+if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
+library(googledrive)
+
+# Authenticate with Google Drive
+
+# Specify the file ID and the destination path
+file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
+dest_file <- "my_data.rds" # Specify your destination path
+
+# Download the file
+drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
+
+# Load the .rds file for analysis
+my_data <- readRDS(dest_file)
+
+# Begin your analysis (example: view the first few rows)
+head(my_data)
+
+``` -->
+
+```{r}
+# Set the URL for the CSV file
+file_url <- "https://drive.google.com/file/d/1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr/view?usp=drive_link"
+
+# Set the destination filename
+dest_file <- "my_data.rds"
+
+# Download the file
+download.file(file_url, destfile = dest_file, mode = "wb")
+
+# Read the CSV file into R
+my_data <- readRDS(dest_file)
+
+# Display the first few rows of the data
+head(my_data)
+```
+
 ### Descriptives
@@
 **This code creates a descriptives table**
-
-```
 
@@
 
-df <- my_data
-
 descriptives_table <- df %>%
   select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
-    mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1", "Year 2", "Year 3", "Year 4"))) %>%
+  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
   mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
@@
     missing = "no",
-        label = list(
-            sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic",
-            anthroweightcalc ~ "Weight"
-        ),
+    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
+                 anthroweightcalc ~ "Weight"),
     statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
@@
 # Compute summary statistics for weight by eventname
-library(dplyr)
 summary <- df %>%
@@
 :::
+
+## Wrapping Up
+
+::: panel-tabset
+### Write-up {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+In the study sample, the average weight at baseline was approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`, which showed a noticeable increase to an average of `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units by Year_1.
+A one-sample t-test was conducted to determine if the mean difference in weight from baseline to Year_1 significantly deviated from zero.
+The results indicated a statistically significant increase with a mean difference of approximately `r round(test_result$estimate, 2)` units (`r round(test_result$conf.int[1], 2)` to `r round(test_result$conf.int[2], 2)`), t(`r test_result$parameter`) = `r round(test_result$statistic, 3)`, p < `r test_result$p.value`.
+Further, a scatterplot visualizing the relationship between baseline and Year_1 weights showed a strong positive linear trend.
+This suggests that participants who were taller at baseline generally remained taller at Year_1, reaffirming the consistent growth trend observed in the data.
+
+:::
+:::

### 2024-03-07 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd

--- 2024-03-06 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
@@
 
-::: blue
+```{r}
+# Install and load the googledrive package
+if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
+library(googledrive)
+
+# Authenticate with Google Drive
+drive_auth()
+
+# Specify the file ID and the destination path
+file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
+dest_file <- "my_data.rds" # Specify your destination path
+
+# Download the file
+drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
+
+# Load the .rds file for analysis
+my_data <- readRDS(dest_file)
+
+# Begin your analysis (example: view the first few rows)
+head(my_data)
+
+```
+
+:::
+
+<!-- ::: blue
 **This code reads in and shows the data to be used in the current
-example**
+example** -->
 

### 2024-03-12 examples/web-r/webr-test.qmd
updates

---
title: webR in Quarto HTML Documents
format: html
engine: knitr
# webr: 
#   show-startup-message: false    # Disable displaying status of webR initialization
#   packages: ['ggplot2', 'dplyr'] # Install R packages on document open
# 'tidyverse', 'easystats', 'lme4', 'gtsummary', 'report', 'broom', 'gridExtra')
filters:
  - webr
---

This is a webr-enabled code cell in a Quarto HTML document.

```{webr-r}
fit <- lm(mpg ~ am, data = mtcars)
summary(fit)
```

```{webr-r}

# Set the data paths
data_path_1 <- "FaKEbcd.rds"

# Read the data
df <- readRDS(data_path_1)

head(df)
str(df)

```

### 2024-03-13 manuscript/manuscript.qmd
manuscript updates

--- 2024-03-02 manuscript/other_manuscript_draft.qmd
@@
 
+abstract: |
+  The Adolescent Brain Cognitive Development^SM (ABCD) Study provides a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of youths, aged 9-10 at baseline and assessed annually for 10 years.
+Given the size and complexity of the ABCD Study©, researchers analyzing its data will encounter a myriad of methodological and analytical considerations.
+This review provides an examination of key concepts and techniques related to longitudinal analyses of the ABCD Study data.
+We discuss the importance of longitudinal data, focusing on the types of inferences that are possible when one assesses individuals across multiple time points, including: 1) characterization of the factors associated with variation in developmental trajectories; 2) assessment of how level and timing of exposures may impact subsequent development; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships.
+We emphasize the importance of selecting appropriate statistical models to address these research questions, e.g., accounting for correlation in repeated measurements, employing linear or non-linear models as indicated by the data, and using link functions that adequately model outcome distributions.
+By presenting the advantages and potential challenges of longitudinal analyses in the ABCD Study, this review seeks to equip researchers with foundational knowledge and tools to make informed decisions as they navigate and effectively analyze and interpret the multi-dimensional longitudinal data currently available.
+
+index-terms: 
+  - Longitudinal Analysis
+  - ABCD Study
+  - Development
+
 format:
-  docx:
-    toc: false
-  elsevier-pdf:
-    journal:
-      cite-style: number
-      formatting: preprint
-      model: 3p
-      layout: twocolumn
-      #cite-style: number
-    include-in-header: 
-      text: |
-        \newpageafter{author}
+  ieee-typst: default
   html:
@@
       margin-width: 20px
+bibliography: references.bib
 
-#fontfamily: libertinus
-#fontsize: 20pt
-#colorlinks: true
-#format-links: [html, pdf]
-#execute:
-  #echo: false
-#bibliography: "references.bib"
-#biblio-style: unsrt
-
-abstract: |
-  The Adolescent Brain Cognitive Development^SM (ABCD) Study provides a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of youths, aged 9-10 at baseline and assessed annually for 10 years.
-Given the size and complexity of the ABCD Study©, researchers analyzing its data will encounter a myriad of methodological and analytical considerations.
-This review provides an examination of key concepts and techniques related to longitudinal analyses of the ABCD Study data.
-We discuss the importance of longitudinal data, focusing on the types of inferences that are possible when one assesses individuals across multiple time points, including: 1) characterization of the factors associated with variation in developmental trajectories; 2) assessment of how level and timing of exposures may impact subsequent development; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships.
-We emphasize the importance of selecting appropriate statistical models to address these research questions, e.g., accounting for correlation in repeated measurements, employing linear or non-linear models as indicated by the data, and using link functions that adequately model outcome distributions.
-By presenting the advantages and potential challenges of longitudinal analyses in the ABCD Study, this review seeks to equip researchers with foundational knowledge and tools to make informed decisions as they navigate and effectively analyze and interpret the multi-dimensional longitudinal data currently available.
-
-keywords:
-- ABCD Study
-- longitudinal analyses
-- development
 ---
@@
 
-\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States.
+[The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (Volkow et al.
@@
 
-![Types of Stability and Change](./figures/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}
+[!Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}
 
@@
 There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2011; @hoffman2015; @littlefield2021].
-
 First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, [e.g., @curran2011]).
@@
 Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015; latent curve models with structured residuals [LCM-SR], @curran2014).
-It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].
+It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context (see @littlefield2022, for further discussion).
 
@@
 An extensive discussion of effect sizes and their relevance for ABCD is given in @dick2021.
-Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered (e.g., @morris2002].
-@wang2019 note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic [see @wang2019, for more details].
+Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered (e.g., @morris2002).
+@wang2019 note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic (see @wang2019, for more details).
 Thus, obtaining effect size metrics based on standardized estimates that are relatively simple in cross-sectional data (such as r) becomes more complex in the context of prospective longitudinal data.
-@feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from traditional analysis [see @feingold2009, for more details].
+@feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from traditional analysis (see @feingold2009, for more details).
 Given this issue, there have been various proposals for adjusting effect size measures in repeated assessments.
@@
 
-[![Longitudinal Data Analysis Structural Diagram](./figures/LDAStructure2.png)](https://centerstat.org/resources/)
+[![Longitudinal Data Analysis Structural Diagram](img/LDAStructure2.png)](https://centerstat.org/resources/)
 
@@
 
-![Longitudinal Models/Data Structures](./figures/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"} <br>
+[!Longitudinal Models/Data Structures](img/Mindmap1copy.png)]{#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"}
+
+[![Longitudinal Data Analysis Structural Diagram](img/LDAStructure2.png)](https://centerstat.org/resources/)
+<br>
 
 #### Traditional Models
-
 Traditional methods for longitudinal analysis primarily focus on modeling mean-level change, and how these changes may differ across groups or levels of some other variable.
@@
 #### Modern GLM Extensions
-
 Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability.
@@
 #### SEM
-
 Structural equation modeling (SEM) approaches have gained prominence in longitudinal data analysis due to their ability to estimate complex relationships among observed and latent variables while accounting for measurement error.
@@
 #### Advanced SEM
-
 Advanced structural equation modeling (SEM) approaches, such as the RI-CLPM and LCM-SR models, have emerged to address more complex research questions and data structures in longitudinal analysis.
@@
 
-![LDA Table](./figures/Tabletemp.png)
+<!-- ![LDA Table](img/Tabletemp.png) -->
 
 # Discussion
-
 \label{sec:headings}
@@
 By familiarizing themselves with the various types of longitudinal models, researchers can more effectively navigate the complexities of longitudinal data and contribute valuable insights into the developmental processes and individual differences that shape human experience.
-
-{{< pagebreak >}}
-
-# Part V: References

### 2024-03-15 manuscript/manuscript.qmd
manuscript updates

@@
 
-[The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States.
-Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (Volkow et al.
-2018).
+[The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest longitudinal investigation of neurodevelopment and child health in the United States.
+Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (@volkow2018).
 At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes.
-Central to achieving these goals is the commitment of the ABCD Study® and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods, by espousing practices that increase access, integrity, and reproducibility of scientific research.
-In this sense, the ABCD Study is a collaboration with the larger research community.
+Central to achieving these goals is the commitment of the ABCD Study© and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods by espousing practices that increase access, integrity, and reproducibility of scientific research.
+In this context, the ABCD Study is a collaboration with the broader research community.
 
@@
 We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.0 data with accompanying R scripts.
-Worked examples are available in Quarto files, accessible at **XXXX**.
+Worked examples are available in Quarto files, accessible in the project's [GitHub repository]( https://github.com/OpenDevSci/LongDev-ABCD/tree/main/examples)
 
@@
 
-The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged between 9-11 years at baseline, as well as their parents/guardians.
+The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged between 9-10 years at baseline, each with a parent/guardian.
 The study sample was recruited from household populations in defined catchment areas for each of the 21 (originally 22) study sites across the United States.
-Information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at https://abcdstudy.org/.
+Information regarding funding agencies, recruitment sites, investigators, and project organizations can be obtained at https://abcdstudy.org.
 The ABCD Study design is described in more detail in @garavan2018 and @dick2021.
 
-The ABCD Study is currently collecting longitudinal data on a rich variety of outcomes that will enable the construction of complex statistical models potentially incorporating factors from many domains.
+The ABCD Study is currently collecting longitudinal data on a rich variety of outcomes that will enable the construction of complex statistical models, potentially incorporating factors from many domains.
 Each new wave of data collection provides another building block for characterizing developmental trajectories and implementing longitudinal analyses that allow researchers to characterize normative development, to identify variables that presage deviations from normative development, and to assess a range of variables associated with biopsychosocial outcomes of interest.
-These data include: 1) a neurocognitive battery [@luciana2018a; @thompson2019]; 2) mental and physical health assessments [@barch2018]; 3) measures of culture and environment [@gonzalez2021; @zucker2018]; 4) substance use [@lisdahl2021]; 5) gender identity and sexual health [@potter2022]; 6) biospecimens [@uban2018]; 7) structural and functional brain imaging [@casey2018; @hagler2019]; 8) geolocation-based environmental exposure data [@fan2021]; 9) wearables, and mobile technology [@bagot2018]; and 10) whole genome genotyping [@loughnan2020].
+These data include: 1) a neurocognitive battery (@luciana2018; @thompson2019); 2) mental and physical health assessments (@barch2018); 3) measures of culture and environment (@gonzalez2021; @zucker2018); 4) substance use (@lisdahl2021); 5) gender identity and sexual health (@potter2022); 6) biospecimens (@uban2018); 7) structural and functional brain imaging (@casey2018; @hagler2019; @palmer2022); 8) geolocation-based environmental exposure data (@fan2021); 9) wearables and mobile technology (@bagot2018); and 10) whole-genome genotyping (@loughnan2020).
 Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward.
-A limited number of assessments are collected in semi-annual telephone interviews between in-person visits.
+A limited number of assessments are collected in semi-annual brief telephone or online assessments.
 
-Data are publicly released roughly annually, currently through the NIMH Data Archive (NDA).
-By necessity, the study’s earliest data releases consisted primarily of one or two visits per participant.
-However, the most recent public release as of the writing of this paper (Release 5.0) contains data collected across five annual visits, including three brain imaging assessments (baseline, year 2 follow-up, and year 4 follow-up visits) for at least a subset of the cohort.
-Hence, starting with Release 5.0, it is feasible for researchers to begin focusing on the estimation of neurodevelopmental and other trajectories.
+Data are publicly released approximately annually, currently through the NIMH Data Archive (NDA).
+The study’s earliest data releases consisted primarily of one or two visits per participant.
+However, the most recent public release as of the writing of this paper (Release 5.1) contains data collected across five annual visits, including three brain imaging assessments (baseline, year 2 follow-up, and year 4 follow-up visits) for at least a subset of the cohort.
+Hence, starting with Release 5.0, it is feasible for researchers to begin focusing on the characterization of neurodevelopmental and other trajectories.
 
@@
 
--   Identify fundamental concepts
+-   Fundamental concepts
 
@@
 
--   Highlight key challenges
+-   Method and analysis
 
@@
 There are several important concepts to consider when conducting longitudinal analyses in a developmental context.
-These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represents the same or different phenomena.
+These include different ways of thinking about the developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represents the same or different phenomena.
 
@@
 
-For example, collecting two or more within-person observations on the same construct at different times enables estimation of individual rates of change (slopes); more observations allow for more precise estimates of individual slopes, as well as characterization of non-linear development.
-Rate of change or other trajectory characteristics may be more informative about individuals than the simple snapshots of level differences that cross-sectional data are limited to informing.
-Cross-sectional age-related differences across individuals are poor substitutes for longitudinal trajectory estimates except under highly restrictive assumptions, e.g., parallel trajectories and lack of age cohort and experience effects [@thompson2011].
+For example, collecting two or more within-person observations on the same construct at different times enables estimation of individual rates of change (slopes) where more observations allow for more precise estimates of individual slopes (random slopes), as well as characterization of non-linear development.
+Rate of change or other trajectory characteristics may be more informative about individuals than the simple snapshots of level differences that cross-sectional data are limited to informing about.
+Cross-sectional age-related differences across individuals are poor substitutes for longitudinal trajectory estimates, except under highly restrictive assumptions, e.g., parallel trajectories and lack of age, cohort and experience effects (@thompson2011).
 Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.
@@
 However, unique epochs and experiences can alter the course of this idealized form of development.
-Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@arseneault2002; @bechtold2016; @hasan2020; @semple2005]; or similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development, corresponding to early adolescence in humans [@spear2016; @crews2000; @ji2018].
-In another example, longitudinal data from the National Consortium on Alcohol in Adolescence (NCANDA) show that binge drinking is associated more strongly with decrements in gray matter volume early in adolescence compared to later [@infante2022].
+Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development (@arseneault2002; @bechtold2016; @hasan2020; @semple2005).
+Similarly, rodent brains are especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development, corresponding to early adolescence in humans (@spear2016; @crews2000; @ji2018).
+In another example, longitudinal data from the National Consortium on Alcohol and Neurodevelopment in Adolescence (NCANDA) show that binge drinking is associated more strongly with decrements in gray matter volume early in adolescence compared to later (@infante2022).
 These examples highlight the importance of considering the role of vulnerable periods – e.g., temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.
@@
 
-Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times, environmental exposures will tend to suppress stability and disrupt the orderly stochastic process of normative development [e.g., @schulenberg2019].
+Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times, environmental exposures will tend to suppress stability and disrupt the orderly stochastic process of normative development (e.g., @schulenberg2019).
 This situation reflects a developmental disturbance in that the normal course of development is “altered” for a time by some time-limited process.
 In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period.
-That is, once the environmental presses are removed (or the individual is removed from the environment), patterns of individual differences (and autoregressive effects) recover to levels similar to those prior to entering the environment.
-For example, in @infante2022, recent binge drinking appears to be most predictive of gray matter volume trajectories, as opposed to more distal prior binge drinking or cumulative number of binge drinks, suggesting the potential for recovery of gray matter trajectories to prior levels of growth if binge drinking levels subside.
+That is, once the environmental pressures are removed (or the individual is removed from the environment), patterns of individual differences (and autoregressive effects) recover to levels similar to those prior to entering the environment.
 
@@
 Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next.
-For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior [@moffitt2015].
-Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified [e.g., @masten2005; @rogosch2010].
+For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior (@moffitt2015).
+Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified (e.g., @masten2005; @rogosch2010).
 Although conceptually distinct from vulnerable periods, both types of developmental considerations highlight the importance of viewing behavior in the context of development and attempting to determine how various developmental pathways unfold.
@@
 
-**2.1.4 Intermediate processes and feedback loops**
-**Something here on the concept of mediation and cross-lagged relationships?**
+### Mediational Processes.
+
+Questions regarding the biological mechanisms whereby exposures impact outcomes can often be framed in terms of mediation analyses (@mackinnon2007; @vanderweele2016).
+Mediation analyses can be implemented using the causal steps approach (@baron1986) and structural equation models (SEM) (@preacher2011).
+More recently, mediation models have been adapted for longitudinal exposures, mediators, and/or outcomes (@bind2016; @vanderweele2017).
+All of these modeling approaches decompose the total effects of an exposure on an outcome into direct and indirect effects, where indirect effects of an exposure flow through its impact on a mediating process.
+@vanderweele2017 details conditions under which the direct and indirect causal effects can be in a longitudinal setting.
+An important example of mediational analyses in the ABCD Study is the impact of exposures on behavioral outcomes (e.g., neurocognition, mental health, substance use) via their impact on the brain, as quantified by imaging-derived phenotypes (IDPs).
+Methods for mediational analyses using multi-dimensional IDPs have been developed and applied to functional MRI data (@lindquist2012; @zhao2018).
 
@@
 
-## Interpretation / Issues / Pitfalls & Assumption
+## Considerations and Challenges
 
-The hallmark characteristic of longitudinal data analysis is the administration of repeated measurements of the same constructs on assessment targets (e.g., individuals, families) across time.
-The primary rationale for collecting longitudinal data is to assess within-person change over time, allowing researchers to estimate individual developmental trajectories and the genetic and/or environmental factors that may impact these trajectories.
-Administering repeated measurements more frequently or over longer time spans enables researchers to ask more nuanced questions and to make stronger inferences.
+The hallmark characteristic of longitudinal data analysis (LDA) is the administration of repeated measurements of the same constructs on assessment targets (e.g., individuals, families) across time.
+The primary rationale for collecting longitudinal data is to assess within-person change over time, allowing researchers to estimate individual developmental trajectories and the genetic and person-level factors that may impact these trajectories.
+Administering repeated measurements more frequently or over longer periods enables researchers to ask more nuanced questions and to make stronger inferences.
 
@@
 Although the clear leap from cross-sectional to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions.
-Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1).
-At the same time, data based on two assessments are inherently limited on multiple fronts.
+Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when an exposure is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1).
+At the same time, analyses of data based on two assessments are inherently limited on multiple fronts.
 As @rogosa1982 noted over forty years ago, “Two waves of data are better than one, but maybe not much better” (p.
@@
 These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for trajectory estimation, model identification and accurate parameter inferences.
-It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (given that linear change is the only estimable form for two assessments waves; [see @duncan2009]).
-Research designs that include three (but preferably more) time points allow for non-linear trajectory estimation and increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of developmental research.
+This is also consistent with recommendations that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (given that linear change is the only estimable form for two assessment waves; (see @duncan2009).
+Research designs that include three (but preferably more) time points allow for non-linear trajectory estimation and increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments (@king2018)– a key aspect of developmental research.
 
 To illustrate, developmental theories are useful for understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level.
-This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” [@littlefield2021].
+This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” (@littlefield2021).
 It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling.
-Indeed, several contemporary longitudinal models designed to isolate within-individual variability (e.g., the Latent Curve Model with Structured Residuals; @curran2014) require at least four assessments to parameterize fully and, more generally, increasingly accurate and nuanced parameter estimates are obtained as more assessment occasions are used [@duncan2009].
+Indeed, several contemporary longitudinal models designed to isolate within-individual variability (e.g., the Latent Curve Model with Structured Residuals [LCM: SR]; @curran2014) require at least four assessments to parameterize fully and, more generally, increasingly accurate and nuanced parameter estimates are obtained as more assessment occasions are used (@duncan2009).
 
@@
 Ideally, t is a biologically meaningful temporal index (e.g., calendar age) as opposed to an exogenous progression of events (e.g., study visit number).
-Properties of interest might include rate of change over time, degree of smoothness (e.g., continuously differentiable), shape (e.g., polynomial or asymptotic behavior), how and how much fi(t) differs across individuals, and what factors predict either within-individual variation (at different times) or between-individual variation (either overall or at specific times).
+Properties of interest might include rate of change over time, degree of smoothness (e.g., continuously differentiable), shape (e.g., polynomial or asymptotic behavior), how and how much f(t) differs across individuals, and what factors predict either within-individual variation (at different times) or between-individual variation (either overall or at specific times).
 
 There are a few different ways to think about patterns of stability and change (see Figure 1).
-Consider measuring school disengagement at the start of middle school and the end of middle school .
+Consider measuring school disengagement at the start of middle school and the end of middle school.
 A common first step may be to compare sixth graders’ average disengagement values and eighth graders’ disengagement values.
 This comparison of the average scores for the same group of individuals at multiple time points is referred to as “mean-level”, as it provides information about change over time (or lack thereof) for an outcome of interest aggregated across members of a group.
-In contrast, “between-individual” stability could be assessed, e.g., by calculating the Spearman correlation between the values obtained at different time points (e.g., ‘disengagement in sixth grade’ with ‘disengagement in eighth grade).
+In contrast, “between-individual” stability could be assessed, e.g., by calculating the Spearman correlation between the values obtained at different time points (e.g., ‘disengagement in sixth grade’ with ’disengagement in eighth grade).
 This analysis focuses on the degree to which individuals retain their relative placement in a group across time.
@@
 Mean-level change in certain traits might help to explain why, in general, populations of individuals tend to be particularly vulnerable to the effects of environmental factors in specific age ranges; rank-order stability might help to quantify the extent to which certain characteristics of the individual are more or less trait-like compared to others.
-For example, in some areas of development, considerable mean-level change occurs over time (e.g., changes in Big 5 personality traits; @bleidorn2022), but exhibit relatively high rank-order stability, at least over shorter measurement intervals [@bleidorn2022; @roberts2000; @roberts2006].
-Despite the useful information afforded by examining mean-level and rank-order stability and change, these approaches are limited in that they provide little information about the overall patterns of within-individual change (i.e., trajectories) and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest [@curran2011].
+For example, in some areas of development, considerable mean-level change occurs over time (e.g., changes in Big 5 personality traits (@bleidorn2022), but exhibit relatively high rank-order stability, at least over shorter measurement intervals (@bleidorn2022; @roberts2000; @roberts2006).
+
+Despite the useful information afforded by examining mean-level and rank-order stability and change, these approaches are limited in that they provide little information about the overall patterns of within-individual change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest (@curran2011).
 For example, questions related to the impact of early-onset substance use on brain development focus on changes within a given individual (i.e., intraindividual differences).
-The longitudinal nature of the ABCD Study, which will provide researchers with over ten time points for some constructs (e.g., substance use) across a ten-year period, allows for the study of within-person processes.
+The ABCD Study will provide researchers with over ten time points for certain constructs (e.g., substance use) across a ten-year period, allowing for a detailed study of some within-person processes.
 
@@
 ### Use of appropriate longitudinal models
-There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2011; @hoffman2015; @littlefield2021].
-First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, [e.g., @curran2011]).
+There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess (e.g., @curran2011; @hoffman2015; @littlefield2021).
+First, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person processes (e.g., how phenotypes change or remain stable within individuals over time, (e.g., @curran2011)).
 Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures.
-Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021].
-Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015; latent curve models with structured residuals [LCM-SR], @curran2014).
-It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context (see @littlefield2022, for further discussion).
+Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time (see @littlefield2021; @orth2021).
+Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time (e.g., random-intercept cross-lagged panel model [RI-CLPM], @hamaker2015); latent curve models with structured residuals [LCM-SR], @curran2014).
+It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context (see @littlefield2021, for further discussion).
 
@@
 Common assumptions of parametric statistical models (e.g., linear mixed-effects models) include normality and equality of variances.
-These assumptions must be carefully considered before finalizing analytical approaches, so that valid inferences can be made from the data, as violation of a model’s assumptions can substantively invalidate the interpretation of results.
+These assumptions should be carefully considered before finalizing analytical approaches, so that valid inferences can be made from the data, as violation of a model’s assumptions can substantively invalidate the interpretation of results.
 For example, longitudinal data can exhibit heterogeneous variability (i.e., the variance of the response changes over the duration of the study) that may need to be accounted for within a model.
 Another pertinent modeling assumption is whether trajectories are linear or non-linear.
-With 2-3 assessments per individual, only a linear model of within-person change is usually feasible; with more time points, higher level polynomials or models with more flexible trajectory shapes (e.g., curve smoothing via cubic splines) becomes possible [REFs].
-Note, however, baseline age in the ABCD Study ranges over two full years; for some outcomes it may be desirable to include a (possibly nonlinear) effect of baseline age along with a linear effect of within-person change in age with only 2-3 assessment times [@thompson2013].
-As the study progresses and more times points are assessed, nonparametric curve estimation at the mean [GAMMs] and at the individual level may also become useful [@ramsay2002].
+With two or three assessments per individual, usually only a linear model of within-person change is feasible.
+
+As the study progresses and more time points are assessed, the potentially nonlinear aspects of trajectories can be assessed, for example using quadratic functions of time.
+Methods that make even fewer assumptions about trajectory shapes, such as nonparametric curve estimation at the mean (e.g., Generalized Additive Mixed Models [GAMMs]; (@wood2017)) and at the individual level (e.g., Functional Data Analysis [FDA]; @ramsay2002) may also become useful.
+Note, baseline age in the ABCD Study ranges over two full years; for some outcomes it may be feasible to include a possibly nonlinear effect of baseline age along with a linear effect of within-person change in age even with only two or three assessment times (@thompson2013).
 
 ### Continuous and Discrete Outcomes
-Implementing valid and efficient statistical models requires an understanding of the type of data being analyzed.
-For example, repeated assessments within the ABCD Study can be based on continuous or discrete measures.
+Repeated assessments within the ABCD Study can be based on continuous or discrete measures.
 Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., caregiver-reported items measuring emotional and behavioral concerns via the Child Behavior Checklist including the categories of “Not True”, “Somewhat True”, and “Very True”), and count variables (e.g., number of cigarettes smoked per day).
-In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016].
-For example, the Mplus manual [@muthen2017] includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, and d) linear growth models for a count outcome assuming a zero-inflated Poisson model.
-Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data [@ren2022].
-These models can account for issues that can occur when working with discrete outcomes, including overdispersion [when the variance is higher than would be expected based on a given distribution; see @lenz2016].
-Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous variables [e.g., @little2013].
-However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes [e.g., @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes].
+In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks (e.g., see @lenz2016).
+For example, the Mplus manual (@muthen2017) includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, and d) linear growth models for a count outcome assuming a zero-inflated Poisson model.
+Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data (@ren2022).
+These models account for issues that may occur when working with discrete outcomes, including overdispersion, i.e., when the variance is higher than would be expected based on a given parametric distribution (see @lenz2016).
+Given the sheer breadth of issues relevant to determining adequate models for discrete outcomes, it is not uncommon for texts on LDA to only cover models and approaches that assume continuous variables (e.g., @little2013).
+However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes: @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes.
 
 ### Issues in attributing longitudinal change to development
-One can observe systematic changes over time in a variable of interest and assume this change is attributable to development.
-However, various pitfalls with longitudinal data can complicate or even invalidate this conclusion.
+
+Systematic changes over time in a variable of interest are not always attributable to development: various pitfalls with longitudinal data can complicate or even invalidate this conclusion.
 For example, if data missingness or participant dropout are related to the values of the outcome, changing sample composition as the study progresses can bias mean trajectory estimates (we describe this in more detail in Section 3.1.7 below).
-Another prerequisite for valid developmental interpretations of longitudinal data is to establish whether a construct is measured consistently over time, i.e., longitudinal measurement invariance [@liu2017; @van2015; @willoughby2012].
-Establishing longitudinal measurement invariance provides researchers with greater confidence that any change over time identified for a construct is attributable to individual change rather than a measurement artifact.
-For example, one study using data from the ABCD Study [@brislin2023) found differential item functioning in two items from a brief delinquency measure, revealing significant bias in an arrest item across Black and White youth.
+Another prerequisite for valid developmental interpretations of longitudinal data is to establish whether a construct is measured consistently over time, i.e., longitudinal measurement invariance (@liu2017; @van2015; @willoughby2012).
+Establishing longitudinal measurement invariance ensures that change over time for a given construct is attributable to individual development rather than merely a measurement artifact.
+For instance, one study using data from the ABCD Study (@brislin2023) found differential item functioning in two items from a brief delinquency measure, revealing significant bias in an arrest item across Black and White youth.
 More specifically, Black youth were more likely to report being arrested compared to White youth with similar levels of delinquency.
 Prevalence rates of delinquent behavior would have been severely biased if measurement invariance had not been tested.
+Alternatively, @vize2023 showed partially strong to strong evidence of longitudinal measurement invariance across broad externalizing dimensions in youth taking in the ABCD Study, suggesting that changes observed over time in these constructs were not due to systematic measurement error, but likely reflect true developmental change.
 
 Observed patterns of growth and decline often differ between cross-sectional vs.
-longitudinal effects [@salthouse2014] where subjects gain increasing experience with the assessment with each successive measurement occasion.
-Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically.
-In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before).
-Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; @french2010] and this phenomenon has been well documented in research using structured diagnostic interviews [@robins1985].
-While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018].
+longitudinal effects (@salthouse2014) where subjects gain increasing experience with the assessment with each successive measurement occasion.
+Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD (@sullivan2017) and highlight the need to consider these effects and address them analytically.
+In the case of performance-based measures (e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014), this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before).
+Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments (e.g., @beck1961; @french2010) and this phenomenon has been well documented in research using structured diagnostic interviews (@robins1985).
+While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artifactually elevated (see @shrout2018).
 
-Some longitudinal studies, e.g., accelerated longitudinal designs [ALDs; @thompson2011] are especially well suited for discovering these effects and modeling them.
-While ABCD is not an ALD, the variability in age (and grade in school) at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an ALD (e.g., substance use; prosocial behavior; family conflict; screen time).
-It is possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments.
+Some longitudinal studies, e.g., accelerated longitudinal designs (ALDs; @thompson2011) are especially well suited for discovering these effects and modeling them.
+While ABCD is not an ALD, the variability in age (and grade in school) at the time of baseline recruitment (9 to 11 years) allows some measures, collected every year, to be conceptualized as an ALD (e.g., substance use; prosocial behavior; family conflict; screen time).
+It is also possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments.
 However, ABCD is fundamentally a single-cohort, longitudinal design, wherein number of prior assessments and age are mostly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.
 
-### Covariance Structures
+### Modeling Covariance
 
 A central issue for repeated measurements on an individual is how to account for the correlated nature of the data.
-Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess the same individuals across time.
-That is, given that residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results.
-Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic [@durbin1950], more commonly, independence is assumed to be violated in study designs with repeated assessments.
-Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.
+Lack of independence of residuals across time occurs for longitudinal data with repeated assessments on individuals and in other situations with nested data (e.g., visits nested within participants, children nested within schools; siblings nested within families).
+Note, the ABCD Study has multiple levels of nesting, depending on the analysis, including within-participant, within-family, within-school, within-MRI scanner, and within-site.
 
-Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (involving the covariance among pairs of repeated outcomes on an individual), coupled with a model for the mean response and its dependence on covariates (e.g., sex at birth).
-There exists a range of methods to model covariance structures, each with its own set of tradeoffs between model fit and parsimony and which may be more or less appropriate for each specific application [e.g., see @kincaid2005].
+Statistical models for nested data include two main components, coupling a model for the mean response and its dependence on covariates with a model for the covariance among repeated outcomes on an individual.
+In contrast, traditional methods, such as multiple regression and ANOVAs, assume residuals are independent and thus are generally inappropriate for designs that incorporate some type of nesting.
+Specifically, given that residuals are no longer independent in a repeated measures design, standard errors from these models are biased and can produce misleading inferences.
+Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model their covariance structure.
+A range of methods can be used to model covariance structures, each with its own set of tradeoffs between model fit and parsimony and which may be more or less appropriate for each specific application (e.g., see @kincaid2005).
 
-There are various approaches to adjust models for the lack of independence of residuals across time, be it for data with repeated assessments or other situations with so-called nested data (e.g., children nested within schools; siblings nested within families).
 The most common approach is to use random effects.
-Essentially, random effects allow for variance estimates around fixed effects.
+Essentially, random effects allow for covariance estimates around fixed effects.
 A classic example (from @bryk1992; @singer1998) involves math achievement measured among students nested within schools.
-In a basic, intercept-only model with no covariates, there would be one fixed effect (the grand mean, or intercept, of math achievement) and two random effects (one representing variation in the intercept, or the variation between schools, and another for the within-school residual, or how much variance in math achievement is left over after taking into account nesting in school).
-From this framework, each student’s score would be a function of the fixed effect (the overall mean), the variation in scores that exists between schools, and the variation that exists among students within the same school.
+In a basic, intercept-only model with no covariates, there would be one fixed effect (the grand mean, or intercept, of math achievement), one school random effect (representing variation in the intercept between schools) and the within-school student residuals (variation left over after accounting for fixed and random effects).
+In this framework, each student’s score would be the sum of the fixed effect (the grand mean), the school random effect and the student’s within-school residual.
 Assumptions about the variance and covariance components of this model dictate the form of the variance/covariance structure.
-For example, if we assume the random effects are independent (i.e., uncorrelated), the implied structure would be compound symmetry, where it is assumed the covariance of any two students in a single class is represented by a random effect (the variance between school means) and the covariance of any two students in different classes is zero.
-The assumptions of this relatively simple covariance structure can be relaxed, resulting in different covariance structures with additional parameters (see @singer1998).
+For example, if we assume the random effects are independent and identically distributed, the implied structure would be compound symmetry, where it is assumed the covariance of any two students in a single school is captured by a school random intercept and the covariance of any two students in different schools is zero.
+The assumptions of this relatively simple covariance structure can be relaxed depending on the nesting structure of the data, resulting in different covariance structures with additional parameters (see @singer1998).
 
-In a similar fashion, individual growth models could be fit to a series of repeated assessments, where assessments are nested within individuals.
-For example, an unconditional linear growth model would involve two fixed effects – one for the intercept (an estimate of the average score when time is coded zero) and one for the linear slope (an estimate of the change in scores for each unit increase in time).
-Random effects could include a random effect for intercept (or an estimate of the variation in scores when time is coded as zero), a random effect for the linear slope (or an estimate of the variation in linear change across time), and a random effect for the within-person residual (or an estimate of the left over variance, or residual, of a given assessment when accounting for the intercept and linear slope).
+In longitudinal studies, visits are nested within individuals.
+Mixed-effect models can be fitted to longitudinal data that couple a model for growth (development) at the mean level with a model for capturing within-individual covariance of assessments.
+For example, a linear growth model would involve two fixed effects – one for the intercept (the average score when time is coded zero) and one for the linear slope (the change in scores for each unit increase in time).
+Random effects could include a random effect for intercept, capturing individual variation in scores at time zero, and a random effect for the linear slope, capturing individual variation in linear change across time.
+Within-individual residuals account for the remaining variation in assessments after accounting for the fixed and random intercepts and slopes.
 Assumptions regarding the covariation among the random effects also indicate different covariance structures.
-For example, it is typical to assume the random intercept and random slope components covary (e.g., if time is coded such that the first time point = zero, the covariance estimate suggests that an individual’s score at baseline relates to the amount of change exhibited across time).
-Further, particularly in structural equation model forms of this model, it is typically assumed the random effect for the within-person residual varies across assessments (@curran2003).
+For example, it is typical to assume that the random intercept and slope components covary, i.e., an individual’s score at time zero relates to the amount of change exhibited across time.
+Further, particularly in structural equation model forms of this model, it is sometimes assumed that the variance of the residuals varies across assessments (@curran2003).
 
-One alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
-As the name implies, the structure assumes a subsequent measurement occasion (e.g., assessment at Wave 2) is regressed onto (that is, is predicted by) a prior measurement occasion (e.g., assessment at Wave 1).
-The most common type of autoregressive design is the AR(1), where assessments at time T + 1 are regressed on assessments at Time T.
-Identical to compound symmetry, this model assumes the variances are homogenous across time.
-Diverting from compound symmetry, this model assumes the correlations between repeated assessments decline exponentially across measurement occasions rather than remaining constant.
+An alternative to random effects is the autoregressive structure, which allows for correlations between repeated assessments to diminish across time.
+As the name suggests, the structure assumes the residual of a subsequent measurement occasion (e.g., visit 2) is regressed onto the residual of a prior measurement occasion (e.g., baseline visit).
+The most common type of autoregressive structure is the AR(1), where residuals at time t + 1 are regressed on residuals at time t.
+Identical to compound symmetry, this model assumes the variances are homogenous across time; however, it differs from compound symmetry in that the correlations between repeated assessments decline exponentially across visits rather than remaining constant.
 That is, we can think of the underlying process as a stochastic one that wears itself out over time.
-For example, per the AR(1) structure, if the correlation between Time 1 and Time 2 data is thought to be .5, then the correlation between Time 1 and Time 3 data would be assumed to be .5 × .5 = .25, and the correlation between Time 1 and Time 4 data would be assumed to be .5 × .5 × .5 = .125.
-As with compound symmetry, the basic AR(1) model is parsimonious in that it only requires two parameters (the variance of the assessments and the autoregressive coefficient).
-Notably, the assumption of constant autoregressive relations between assessments is often relaxed in commonly employed designs that use autoregressive modeling (e.g., cross-lagged panel models [CLPM]).
-These designs still typically assume an AR(1) process (e.g., it is sufficient to regress the Time 3 assessment onto the Time 2 assessment and is not necessary to also regress the Time 3 assessment onto the Time 1 assessment, which would result in an AR(2) process).
-However, the magnitude of these relations is often allowed to differ across different AR(1) pairs of assessment (e.g., the relation between Time 1 and Time 2 can be different from the relation between Time 2 and Time 3).
-These more commonly employed models also often relax the assumption of equal variances of the repeated assessments.
+For example, per the AR(1) structure, if the correlation between visit 1 and visit 2 data is thought to be .5, then the correlation between visit 1 and visit 3 data would be assumed to be .5 × .5 = .25, and the correlation between visit 1 and visit 4 data would be assumed to be .5 × .5 × .5 = .125.
+As with compound symmetry, the basic AR(1) model is parsimonious in that it only requires two parameters: the variance of the residuals and the autoregressive coefficient.
+
+Notably, the assumption of constant autoregressive relations between assessments is often relaxed in commonly employed designs that use autoregressive modeling (e.g., CLPM).
+These designs still typically assume an AR(1) process.
+However, the magnitude of these relations is often allowed to differ across different AR(1) pairs of assessment (e.g., the relation between visit 1 and visit 2 can be different from the relation between visit 2 and visit 3).
+These models also often relax the assumption of equal variances of the repeated assessments.
+
 Although the AR(1) structure may involve a more realistic set of assumptions compared to compound symmetry, in that the AR(1) model allows for diminishing correlations across time, the basic AR(1) model, as well as autoregressive models more generally, can also suffer from several limitations in contexts that are common in prospective designs.
 In particular, recent work demonstrates that if a construct being assessed prospectively across time is trait-like in nature, then a simple AR(1) process fail to adequately account for this trait-like structure, with the downstream consequence that estimates derived from models based on AR structures (such as the CLPM) can be misleading and fail to adequately demarcate between- vs.
-within-person sources of variance [@hamaker2015].
-Finally, it is also worth noting that in many longitudinal contexts, the time intervals between assessments are not equidistant and researchers need to consider carefully how to appropriately model time in their model and what this model implies.
+within-person sources of variance (@hamaker2015).
+Note also, discrete-time autoregressive structures such as AR(1) implicitly assumes relatively constant time gaps between visits; this may not be true in many applications using the ABCD Study data.
 
@@
 
-Attrition from a longitudinal panel study such as ABCD is inevitable and represents a potential threat to the validity of longitudinal analyses and cross-sectional analyses conducted at later time points, especially since attrition can only be expected to grow over time[@littlefield2022].
-The ABCD Retention Workgroup (RW) employs a data-driven approach to examine, track, and intervene in these issues and while preliminary findings show participant race and parent education level to be associated with late and missing visits, although to date, attrition in ABCD has been minimal [@ewing2022].
-Ideally, one tries to minimize attrition through good retention practices from the outset via strategies designed to maintain engagement in the project [@cotter2005; @hill2016; @watson2018].
+Attrition from a longitudinal study such as ABCD is inevitable and represents a potential threat to the external validity of analyses conducted at later visits, especially since attrition can only be expected to grow over time (@littlefield2022).
+The ABCD Retention Workgroup employs a data-driven approach to examine, track, and intervene in these issues and while preliminary findings show participant race and parent education level to be associated with late and missing visits, although to date, formal attrition in ABCD has been minimal (@ewing2022).
+Ideally, one tries to minimize attrition through good retention practices from the outset via strategies designed to maintain engagement in the project (@cotter2005; @hill2016; @watson2018).
 However, even the best-executed studies need to anticipate growing attrition over the length of the study and implement analytic strategies designed to provide the most valid inferences.
+
 Perhaps the most key concern when dealing with data that is missing due to attrition is determining the degree of bias in retained variables that is a consequence of attrition.
@@
 
-Three types of missingness are considered in the literature [see @little1989], namely: a) missing completely at random (MCAR), b) missing at random (MAR), and c) missing not at random (MNAR).
+Three types of missingness are considered in the literature (@little1989; @little2013), namely: a) missing completely at random (MCAR), b) missing at random (MAR), and c) missing not at random (MNAR).
 Data that are MCAR are a simple random sample of all data in a given dataset.
 MAR implies missing data are a random sample (i.e., does not hinge on some unmeasured variables) within strata of the measured covariates in a dataset (e.g., biological sex).
-Data that are MNAR are missing as a function of unobserved variables and may bias assocations even after conditioning on the observed covariates.
+Data that are MNAR are missing as a function of unobserved variables and may bias associations even after conditioning on the observed covariates.
 @graham2009 provides an excellent and easy-to-digest overview of further details involving missing data considerations.
 
-Modern approaches for handling missing data, such as full information maximum likelihood, propensity weighting, auxiliary variables and multiple imputation avoid the biases of older approaches [see @enders2010; @graham2009].
+Modern approaches for handling missing data, such as full-information maximum likelihood, propensity weighting, auxiliary variables and multiple imputation avoid the biases of older approaches (see @enders2010; @graham2009).
 @graham2009 noted several “myths” regarding missing data.
@@
 Violations of MAR impact both traditional and more modern data estimation procedures, though as noted by Graham, violations of MAR tend to have a greater effect on older methods.
-Graham thus suggests that imputing missing data is a better approach compared to listwise deletion in most circumstances, regardless of the model of missingness [i.e., MCAR, MAR, MNAR; see @graham2009].
+Graham thus suggests that imputing missing data is a better approach compared to listwise deletion in most circumstances, regardless of the model of missingness (i.e., MCAR, MAR, MNAR; see @graham2009).
+The ABCD Biostatistics Workgroup is currently implementing several missing data approaches which are being implemented and compared to each other (and listwise deletion) in the 5.0 data release, including, propensity score weighting, and multiple (multilevel) imputation.
 
@@
 
-Given that longitudinal data involve multiple sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data.
-An effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014].
-Common effect size metrics include the correlation r between two variables and the standardized difference between two means, Cohen’s *d* [@cohen1988].
-An extensive discussion of effect sizes and their relevance for ABCD is given in @dick2021.
+Given that longitudinal data involve multiple sources of variation, quantifying effect sizes longitudinally is more complex compared to deriving such estimates from cross-sectional data.
+An effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” (Kraemer 2014).
+Common effect size metrics include the Pearson correlation r between two variables and the standardized difference between two means, Cohen’s d (@cohen1988).
+An extensive discussion of cross-sectional effect sizes and their relevance for ABCD is given in @dick2021.
+
 Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered (e.g., @morris2002).
@@
 Thus, obtaining effect size metrics based on standardized estimates that are relatively simple in cross-sectional data (such as r) becomes more complex in the context of prospective longitudinal data.
-@feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from traditional analysis (see @feingold2009, for more details).
+@feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from cross-sectional analysis (see @feingold2009, for more details).
+
 Given this issue, there have been various proposals for adjusting effect size measures in repeated assessments.
-@feingold2019 reviews the approach for effect size metrics for analyses based on growth modeling, including when considering linear and non-linear (i.e., quadratic) growth factors.
+@feingold2019 reviews the approach for effect size metrics for analyses based on growth modeling, including when considering linear and non-linear (e.g., quadratic) growth factors.
 @morris2002 review various equations for effect size calculations relevant to combining estimates in meta-analysis with repeated measures and independent-groups designs.
 Other approaches to quantifying effect sizes longitudinally may be based on standardized estimates from models that more optimally disentangle between- and within-person sources of variance.
-As an example, within a random-intercept cross-lagged panel model (RI-CLPM) framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.
+As an example, within a random-intercept cross-lagged panel model (RI-CLPM) framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relationships.
 
@@
 
-An ideal longitudinal study integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model [@collins2006].
-To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure **X**).
-An understanding of these data structures is helpful, as they can warrant different types of longitudinal data analysis (LDA).
-Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and Bauer [2019; @bauer2019] provide a nice on-ramp in first distinguishing between the use of “time-to-event” and “repeated measures” data.
-Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change [@bauer2019].
-Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
+An ideal longitudinal analysis integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model (@collins2006).
+To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure 1).
+An understanding of these data structures is helpful, as they can warrant different types of LDA.
+Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and @bauer2019 provide a nice on-ramp in first distinguishing between the use of “time-to-event” and “repeated measures” data.
+Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change (@bauer2019 Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
 This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models) and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events.
-While much has been written about “time-to-event” data (**see xxx; xxx; xxxx**), including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study (@brislin2023), our emphasis will be given to the modeling of “repeated measures” data.
+While much has been written about “time-to-event” data (@hosmer2008; @rizopoulos2012), including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study (@brislin2023), our emphasis will be given to the modeling of “repeated measures” data.
 
@@
 One such approach to repeated measures analysis is the use of time-series models.
-These models generally consist of a long sequence of repeated measurements (≧ 50-100 measurements) on a single (or small number of) variable of interest.
-Time-series analysis is often used to predict/forecast temporal trends and cyclic patterns and is geared toward making inferences about prospective outcomes within a population (with relatively with less focus on inferring individual-level mechanisms and risk factors).
+These models generally consist of a long sequence of repeated measurements (≧ 50-100 measurements) on a single or small number of variables of interest.
+Time-series analysis is often used to predict temporal trends and cyclic patterns and is geared toward making inferences about prospective outcomes within a population (with relatively less focus on inferring individual-level mechanisms and risk factors).
+
 A related type of repeated measures analysis is Intensive Longitudinal Data (ILD).
-Similar to time-series analysis, ILD models involve frequent measurements (~ 30-40 measurements) of the same individuals in a relatively circumspect period of time (e.g., experience sampling to obtain time series on many individuals).
-Although ILD models may include slightly fewer measurement occasions than time-series data, ILD models tend to have more subjects than time series models (~ 50-100 subjects).
-This allows ILD models to examine short-term patterns by incorporating a time series model that can sometimes fit parameter estimates to each individual’s data in order to model individual difference outcomes.
-The final type of repeated measures analysis that we will primarily focus on is the longitudinal panel model.
-These models follow a group of individuals— a panel (also referred to as a cohort) — across relatively fewer measurement occassions (~ 5-15), and are often interested in examining change across both, within-individuals and between-individuals.
+Similar to time-series analysis, ILD models involve frequent measurements (~ 30-40 measurements) of the same individuals in a relatively circumspect period (e.g., experience sampling to obtain time series on many individuals).
+Although ILD models may include slightly fewer measurement occasions than time-series data, ILD models tend to have more subjects than time-series models (~ 50-100 subjects).
+This allows ILD models to examine short-term patterns by incorporating a time series model that can sometimes fit parameter estimates to each individual’s data to model individual difference outcomes.
+
+The final type of repeated measures analysis that we will primarily focus on is the longitudinal panel study.
+These models follow a group of individuals— a panel (also referred to as a cohort) — across relatively fewer measurement occasions (~ 5-15) and are often focused on examining both change within- and between-individuals.
+The ABCD Study is primarily a longitudinal panel study, though some data streams (e.g., functional brain imaging, FitBit data) could be analyzed as ILP or even time series methods.
 
@@
 With the large and continually expanding body of research on statistical methods for longitudinal analyses, determining which longitudinal model to implement can be challenging.
-The aim of this section is to help researchers navigate these many options to identify the statistical approach most appropriate to their unique research question when deciding on how to measure change over time.
+This section aims to help researchers navigate these many options to identify the statistical approach most appropriate to their unique research question when deciding on how to measure change over time.
 Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
-
-Common examples include grouping by linear vs nonlinear models [@collins2006], the number of measurement occasions [@king2018], and statistical equivalency [e.g., change scores vs.
-residualized change; see @castro-schilo2018].
-The organization we use below overlaps in a number of ways with these examples, and in particular with @bauer2019.
+Common examples include grouping by linear vs nonlinear models (@collins2006), the number of measurement occasions (@king2018), and statistical equivalency (e.g., change scores vs.
+residualized change; see @castro-schilo2018).
+The organization we use below overlaps in several ways with these examples, and in particular with @bauer2019.
 However, it is important to note that in each case, the chosen way of grouping is primarily intended to allow the reader to compare and contrast various analytical approaches.
-In the following sections, we briefly summarize the advantages/disadvantages of a series of longitudinal models organized into the following groupings: Traditional Models, Modern GLM Extensions, SEM, and Advanced SEM.
+In the following sections, we briefly summarize the advantages/disadvantages of a series of longitudinal models organized into the following groupings: Traditional Models, Modern GLM Extensions, Structural Equation Models (SEM), and Advanced SEM (see Figure 2).
 We note that this is not an exhaustive review of each of these methods, and for more in-depth detail we do provide the reader with relevant resources.
-As aptly summarized by @bauer2019, “there are many exceptions, alternatives, nuances, ‘what ifs’, and ‘but couldn’t you’s that aren’t addressed here.”
+As aptly summarized by @bauer2019, “…there are many exceptions, alternatives, nuances, ‘what ifs’, and ’but couldn’t you’s that aren’t addressed here.”
 
@@
 For example, is there a difference in average internalizing symptoms obtained across multiple assessments between boys and girls?
-Longitudinal models that focus on mean-level change are also referred to as marginal models and examples of specific methods include repeated measures ANOVA (and MANOVA), ANCOVA, and Generalized Estimating Equations (GEEs).
-Mean-level change models are commonly used when data is only available from 2 measurement occasions.
-For example, computing a difference score (e.g., mean internalizing scores at time 2 - mean internalizing scores at time 1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups.
+Longitudinal models that focus on mean-level change are also referred to as marginal models and examples of specific methods include repeated measures ANOVA, ANCOVA and Generalized Estimating Equations (GEEs).
+Mean-level change models are commonly used when data are only available from 2 measurement occasions.
+For example, computing a difference score (e.g., mean internalizing scores at visit 2 - mean internalizing scores at visit 1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups.
 Additionally, the longitudinal signed-rank test, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data.
-Another common approach, often used in pre-/post-design studies but can be used with ABCD Study data, is to use residualized change score analysis to assess the degree of change in a variable, while controlling for its initial level [@castro-schilo2018].
+Another common approach, often used in pre-/post-design studies but can be used with ABCD Study data, is to use residualized change score analysis to assess the degree of change in a variable, while controlling for its initial level (@castro-schilo2018).
 
@@
 
-Traditional longitudinal, such as residualized chance score models, can be useful in some contexts (e.g., two measurement occasions), but overall, their practical utility for answering questions about developmental processes is limited.
+Traditional longitudinal models, such as residualized change score models, can be useful in some contexts (e.g., two measurement occasions), but overall, their practical utility for answering questions about developmental processes is limited.
 Perhaps most notably, these models do not allow for characterizing patterns of within-person change.
 This is a particularly important limitation since most psychological theories posit within-person processes (i.e., what will happen within a given individual).
-As such, traditional approaches often correspond poorly with most theoretical models of change and a failure to disaggregate between-person and within-person effects can result in consequential errors of inference [e.g., ecological fallacy, @curran2011].
+As such, traditional approaches often correspond poorly with most theoretical models of change and a failure to disaggregate between-person and within-person effects can result in consequential errors of inference (e.g., ecological fallacy, @curran2011).
 Moreover, even determining which of these procedures to use for comparing change over two time points across groups can be surprisingly complicated.
-A particularly vexing example is that of imbalanced baseline scores (i.e., when baseline scores are correlated with a covariate of interest), which can produce different conclusions across methods [e.g., **X versus X**; see @littlefield2023, for a review].
-Given these shortcomings, and the complexity of the issues surrounding some of these methods, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data (and preferably make use of data collected across three or more time points, as is currently true for many ABCD assessments).
+A particularly vexing example is that of imbalanced baseline scores (i.e., when baseline scores are correlated with a covariate of interest), which can produce different conclusions across methods (e.g., see @littlefield2023, for a review).
+Given these shortcomings, and the complexity of the issues surrounding some of these methods, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data and preferably make use of data collected across three or more time points, as is currently true for many ABCD Study assessments.
 
 #### Modern GLM Extensions
-Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability.
+Modern approaches to LDA have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability.
 Generalized Estimating Equations (GEE), Linear Mixed Models (LMM), Generalized Linear Mixed Models (GLMM), and Autoregressive Cross-Lagged Panel Models (ARCL) are examples of such contemporary techniques.
-GEE, an extension of Generalized Linear Models, combines the generalized linear model for non-normal outcomes with repeated measures (marginal) model and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects.
-For example, @vandijk2021 used GEE to obtain relative risks for psychiatric diagnoses among children in the ABCD Study with a family history of depression and used the ABCD Study sampling weights to generalize prevalence rates among 9 and 10 year-olds across the US.
+GEE, an extension of Generalized Linear Models, combines the generalized linear model for non-normal outcomes with repeated measures and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects.
+For example, @vandijk2021 used GEE to obtain relative risks for psychiatric diagnoses among children in the ABCD Study with a family history of depression and used the ABCD Study sampling weights to generalize prevalence rates among 9 and 10-year-olds across the US.
+
 LMMs, also known as multilevel or hierarchical linear models, facilitate the simultaneous analysis of within-person and between-person variability, making them ideal for nested data structures or repeated measures.
-Within the ABCD Study, reseachers may want to consider nesting by individual, family (i.e., siblings or twins), school or district, and/or site.
-GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data, such as the use of ABCD data on substance use [e.g., @martz2022] screen media use [@lees2020], and microstructure of the brain [@palmer2022].
-Finally, ARCL models are used to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects [although, ARCL models are relatively less useful for teasing apart between-person and within-person sources of variances; see @curran2021].
+Within the ABCD Study, researchers may want to consider nesting by individual, family (i.e., siblings or twins), school or district, and/or site.
+GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data, such as the use of ABCD data on substance use (e.g., @martz2022) screen media use (@lees2020), and microstructure of the brain (@palmer2022).
+
+Finally, ARCL models are used to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects, although ARCL models are relatively less useful for teasing apart between-person and within-person sources of variances; see @curran2021.
 
@@
 
-#### SEM
-Structural equation modeling (SEM) approaches have gained prominence in longitudinal data analysis due to their ability to estimate complex relationships among observed and latent variables while accounting for measurement error.
-SEM techniques can be categorized into variable-centered, person-centered, and hybrid approaches, each with unique strengths and limitations.
-The choice of method depends on the research question, data structure, and underlying assumptions.
+#### Structural Equation Modeling (SEM)
+Structural Equation Modeling (SEM) is a flexible modeling framework that integrates elements of path analysis and confirmatory factor analysis (CFA) to examine complex relationships between a set of observable variables and latent constructs (@hair2021).
+The integration of structural (regression) and measurement (CFA) components within a unified framework supports a theory-driven approach that allows researchers to rigorously test hypothesized relationships among variables of interest and their underlying causes (@hair2021; @raykov2012).
+Over the years, the flexibility of the SEM framework has evolved to become particularly adept for modeling autoregressive processes (which often assume underlying stationarity) and growth processes which accommodate both, mean trajectories and individual differences in them (@mcardle2009; @little2013).
 
-Variable-centered approaches, such as latent change scores, latent growth curve models, multivariate (parallel process) latent growth curve models, and latent state-trait models, focus on examining relationships among variables and population-level patterns.
-These models offer a powerful means to estimate individual change trajectories and the relationships between growth parameters of different variables, while also decomposing observed measurements into latent state and trait components.
-However, these approaches may not adequately capture distinct subgroups of individuals who share similar patterns of change over time, which can be crucial for understanding heterogeneous developmental processes.
+Longitudinal SEM techniques share many similarities with mixed-effects methods and research demonstrates their mathematical equivalence in many situations (@curran2003; @mehta2005).
+However, these related approaches often cater to distinct theoretical and analytical needs.
+For instance, mixed-effects techniques are an extension of the regression framework and often excel when working with complex data structures such as multiple levels of nesting, small samples, and non-equidistant time points (@mcneish2018; @hedeker2006).
+Alternatively, applying SEM methods to longitudinal analysis provides a flexible means for modeling the underlying process of change.
+It also addresses several challenges faced by competing approaches, including the ability to accommodate intricate error structures and deal effectively with missing data, as well as the implementation of numerous modeling extensions (@mcneish2018; @curran2021).
+These models have grown increasingly popular for modeling longitudinal outcomes particularly due to their ability to build statistical models that match some particular underlying theory (@serang2019).
 
+Considering the variety of available techniques, it can be helpful to classify longitudinal SEM, broadly (if not coarsely), into variable-centered, person-centered, and hybrid analyses, each with unique strengths and limitations.
+Variable-centered analyses (e.g., latent growth curves @curran2003, latent change scores @mcardle2001, latent state-trait models @geiser2012) are primarily concerned with understanding covariation among variables at the group level and characterizing population-level patterns of change, while person-centered analyses (e.g., latent class and latent transition models) identify distinct subgroups or patterns within the data (@muthen2000; @woo2024; @howard2018).
+Hybrid models combine these perspectives to offer a comprehensive analysis of latent subgroups and growth parameter relationships (@morin2018; @lubke2005).
+The choice between these approaches is primarily driven by the research question, data structure, and relevant underlying assumptions.
+
+#### Variable-centered models
+
+One key application of the SEM framework to the analysis of longitudinal data is the latent growth curve model (LGCM).
+This is a variable-centered approach that characterizes average group trajectories and individual variations (random effects) in an outcome over time (@curran2003).
+These models are similar to their linear mixed effects counterpart in many ways, with the main conceptual difference being that LGCM includes a repeatedly measured outcome in the model as a function of time (closely resembling a standard CFA approach), rather than as an explanatory variable (as in a standard regression approach) (@mcneish2018).
+Specifically, observed scores at each time point are treated as indicator variables with their factors loading scaled to reflect a hypothesized pattern of change (e.g., loadings of 0, 1, and 2 would assume equidistant, linear change).
+Latent intercepts (initial levels) and slopes (rates of change) are estimated, along with their variances and covariance to capture common trends and individual deviations over time.
+This method was used in a recent study by @trevino2023 to show a decreasing trajectory of parent-reported externalizing behaviors from ages 9-12 among youth taking part in the ABCD Study.
+This study also examined hypothesized predictors of the growth trajectory intercept and slope factors, highlighting a particular strength of these models–– their flexibility and extensibility.
+As an example, @roy2024 used publicly available data from the ABCD Study and several other large-scale datasets to explore bivariate (parallel process) relationships between white matter pathways and literacy over time.
+Beyond these examples, LGCMs can be extended in numerous ways, including to compare rates of growth across groups, investigate the consequences of change, and incorporate time-invariant or time-varying covariates, to highlight only a few (for a more detailed treatment of LGCM applications and methodologies, refer to @preacher2010; @preacher2018; @curran2010).
+
+The latent change score model (LCSM) is a variable-centered approach uniquely tailored for analyzing temporal variations in how a construct changes over time (@mcardle1994; @mcardle2001).
+These models share many features with growth curve analysis, but with a more explicit focus on how change occurs between measurement occasions (@serang2019; @mcardle2009).
+Specifically, LCSM estimates a series of latent variables to model change in an outcome from one time point to the next, as a function of scores on that outcome at prior time points (@mcardle2001; @ghisletta2012).
+Some types of LCSM estimate two underlying latent factors: a constant change factor that remains fixed over time, and a proportional change factor that adjusts for previous scores.
+By disaggregating change into constant and proportional components, this approach facilitates a more nuanced understanding of whether prior changes in a given process are related to future changes in the same process (@serang2019; @kievit2018).
+Expanding upon the capabilities of this framework, LCSM also allows for comprehensive multivariate analyses that can facilitate investigations into how change in one construct is associated with change in another construct.
+The appeal of this approach is evidenced by several recent studies that have used data from the ABCD Study to explore bivariate associations between brain development and changes in several mental and physical health indicators (@wiker2023; @rapuano2022; @beck2023; @nweze2023; @mewton2023).
+
+Latent State-Trait Models (LSTM) offer another variable-centered approach to longitudinal analysis that also allows for the estimation of patterns of change over time.
+Unlike LGCM, which conceptualizes change as a function of time, and LCSM, which views change through sequential measurements, this approach disaggregates observed behaviors into distinct stable (trait) and occasion-specific (state) components (@kenny2001; @steyer2015).
+Based on LST theory (@steyer1999; @steyer1992), these models hold that scores on a repeated measures outcome can be partitioned into an enduring latent trait variable that reflects between-individual differences, and a transient latent state residual that represents situational influences (@stadtbaeumer2022; @geiser2012).
+Beyond parsing out these key variance components, LSTM can be extended in many ways, such as by incorporating autoregressive effects to capture relative stability and the influence of past states on future responses (i.e.
+carry-over effects; @cole2005; @eid2017; @geiser2012).
+The merits of this approach are highlighted in a recent review by @sanchez2020 focused on strategies for modeling neurobehavioral development.
+These study authors encourage researchers to leverage data from the ABCD Study and other large-scale longitudinal and publicly available datasets and to apply state-trait methods to map neural and behavioral trajectories in youth (for a more detailed overview of these models, see @kenny2001; @steyer2023; @steyer1999).
+In general, while many commonalities and important features are shared across different variable-centered approaches, selecting the most appropriate statistical model for assessing change hinges on the specific theoretical model of change and what is intended to be learned from the model (see @kievit2018; @mcardle2009; @ghisletta2012, for discussion), which is critical for informing the interpretation and applicability of the research findings.
+
+#### Person-centered models
+
+Despite the flexibility afforded by variable-centered analysis, these methods are not generally equipped to capture underlying developmental trajectories that are unique to distinct clusters of individuals.
+This limitation can be particularly notable for research that aims to characterize heterogeneous developmental processes.
 Person-centered approaches, including latent transition analysis and latent class growth analysis, address this limitation by identifying subgroups of individuals who share similar patterns of change.
@@
 For example, taking advantage of the large sample size of the ABCD Study, @xiang2022 found evidence of four subgroups of youth with unique longitudinal patterns of depressive symptoms over time and identified risk factors that were differentially associated with the various trajectories.
+
 The use of such models allows for a more nuanced understanding of the associations between risk factors and change in symptomatology, as opposed to a snapshot of symptomatology at a single time point.
@@
 
-Hybrid approaches, such as growth mixture modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
+Hybrid approaches, such as growth mixture (jung2008; @muthen2000) and factor-mixture (@lubke2005; @lubke2007) modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
 This combination provides a more comprehensive understanding of longitudinal data by capturing both within- and between-person variability.
@@
 
-In summary, SEM approaches offer powerful tools for longitudinal data analysis, enabling researchers to investigate complex relationships, individual differences, and change dynamics over time.
+In summary, SEM approaches offer powerful tools for LDA, enabling researchers to investigate complex relationships, individual differences, and change dynamics over time.
 The choice between variable-centered, person-centered, and hybrid approaches depends on the research objectives and the nature of the data.
@@
 
-#### Advanced SEM
-Advanced structural equation modeling (SEM) approaches, such as the RI-CLPM and LCM-SR models, have emerged to address more complex research questions and data structures in longitudinal analysis.
+#### Advanced Structural Equation Models
+Advanced SEM approaches, such as the RI-CLPM and LCM-SR models, have emerged to provide a clearer understanding of important research questions and data structures in longitudinal analysis.
 These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.
@@
 The RI-CLPM enhances the traditional cross-lagged panel model by incorporating random intercepts, which allow for the separation of stable individual differences from the dynamic within-person associations between variables over time.
-Within-person variance in these models are captured by a series of latent variables that reflect time specific variance (i.e., the residual variance from the random intercept).
+Within-person variance in these models is captured by a series of latent variables that reflect time specific variance (i.e., the residual variance from the random intercept).
 These time-specific variables are referred to as structured residuals.
@@
 Using three waves of ABCD Study data, @kulisch2022 found a prospective association between psychopathology and childhood obesity as well as between childhood obesity and later eating behavior.
-The authors also showed that reciprocal associations were overestimated when stable, interindividual trait differences was not included in the model (i.e., via the random intercept).
+The authors also showed that reciprocal associations were overestimated when stable, interindividual trait differences were not included in the model (i.e., via the random intercept).
 
@@
 
-In conclusion, advanced SEM approaches for longitudinal data analysis provide valuable tools for addressing complex research questions and data structures.
+In conclusion, advanced SEM approaches for LDA provide valuable tools for addressing complex research questions and data structures.
 While they offer more nuanced insights into within-person change dynamics and the influence of individual differences, these models also come with certain limitations, such as the necessity of multiple assessments (e.g., four or more for LCM-SR), increased complexity, computational demands, and the need for careful model specification and interpretation.
@@
 
-<!-- ![LDA Table](img/Tabletemp.png) -->
+![LDA Table](img/Tabletemp.png)
+
+#### Longitudinal Analysis of Neuroimaging Data
+
+Neuroimaging data, characterized by its large scale, spatial structure and binary data formats, requires the use of specialized software for effective analysis.
+Fortunately, there are now several freely available software packages that provide options for statistical modeling of brain imaging data, thus facilitating analysis of the brain's function or structure at every voxel or vertex in an image (see Fig.
+3).
+However, the most widely used packages have only rudimentary support for longitudinal data.
+Prominent software solutions such as SPM (https://www.fil.ion.ucl.ac.uk/spm) and FSL (https://fsl.fmrib.ox.ac.uk) packages offer limited support in analyzing longitudinal data due to their use of strong assumptions.
+SPM assumes a common longitudinal correlation structure over space, and FSL requires balanced designs and relies on the assumption of compound symmetry.
+Despite these limitations, there is a steady growth in the development of neuroimaging tools that provide for comprehensive longitudinal data analysis, typically via commonly adopted modeling approaches such as LMMs marginal models.
+These emerging tools are equipped to handle more complex scenarios, including unbalanced designs and random covariate effects, among others.
+
+AFNI (https://afni.nimh.nih.gov/), a well-established tool in neuroimaging, integrates 3dLMEr (@chen2013; https://afni.nimh.nih.gov/pub/dist/doc/program_help/3dLMEr.html) which adopts an LMM-based approach by providing access to the advanced capabilities of R’s lme4’s lmer function.
+For surface-based data, Freesurfer (https://surfer.nmr.mgh.harvard.edu/) provides the linear mixed effects (LME) package for modelling longitudinal data (@bernal2013statistical; https://surfer.nmr.mgh.harvard.edu/fswiki/LinearMixedEffectsModels).
+This tool is able to apply spatial regularization of LMM parameters with surface-based ROIs to improve stability (@bernal2013spatiotemporal).
+
+For imaging data, LMM’s present a significant computational challenge, not only because they require iterative optimization but also because the computations cannot be vectorized as efficiently compared to ordinary least squares.
+To overcome this challenge, The Big Linear Mixed Models (BLMM, https://github.com/NISOx-BDI/BLMM) software addresses this by using Python’s broadcasting operations to estimate LMM’s as efficiently as possible (@maullin2022; @maullin2021).
+BLMM further separates the computation of sufficient statistics and parameter estimation, allowing sensitive image data to remain private if needed.
+
+A different yet efficient approach is used with Fast and Efficient Mixed-effects Analysis (FEMA, https://github.com/cmig-research-group/cmig_tools), which uses a non-iterative regression estimator of the LMM variance components plus variance parameter quantization.
+This allows vectorization within groups of voxels that share the same variance parameters (@parekh2021).
+While this method uses different approximations, the authors have shown it provides results that closely match a traditional LMM implementation.
+
+An alternative method for modeling longitudinal data is the marginal model.
+This approach differs from others by modeling only the population-level factors and covariates rather than explicitly modeling individual intercepts and slopes.
+It employs a flexible intra-subject covariance model to account for residual dependence.
+Like the LMM, it allows for unbalanced designs and singleton subjects, and it implicitly measures any covariance that would otherwise be explained by random covariates.
+In a marginal model, a “working covariance” matrix is utilized.
+This matrix does not need to be accurately modelled and may even be constructed under the potentially incorrect assumption that the errors are independent.
+In this case, parameter estimation reduces to ordinary least squares, but remains consistent.
+A crucial component of this method is the use of a robust “sandwich estimator” for standard errors, which accounts for any dependence not captured by the working covariance.
+The Sandwich Estimator (SwE) is a toolbox for SPM (https://www.nisox.org/Software/SwE, including CIFTI support) and FSL (https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/Swe) that provides marginal model inference using an independence working covariance matrix (@guillaume2014).
+
+![Figure 3 Statistical modeling of brain imaging data](img/Picture1.png)
+![Figure 4 Statistical modeling of brain imaging data](img/Picture2.png)
 
@@
 This paper aimed to provide an overview of diverse longitudinal analysis techniques, with a particular emphasis on their application to extensive longitudinal studies such as the ABCD Study.
-Beyond contributing to the ever-growing body of knowledge on longitudinal data analysis, we hope this manuscript also serves as a valuable resource for researchers seeking to optimize the use of large-scale longitudinal investigations in advancing our understanding of human development and behavior.
+Beyond contributing to the ever-growing body of knowledge on LDA, we hope this manuscript also serves as a valuable resource for researchers seeking to optimize the use of large-scale longitudinal investigations in advancing our understanding of human development and behavior.
 In this discussion, we will focus on the key findings and recommendations of our review and discuss potential innovations that can further enhance the utility of these methods.
@@
 
-The final section, along with associated code and additional resources made available as online supplements (), aims to serve as a resource for researchers seeking to understand and implement various longitudinal panel models.
+The final section, along with associated code and additional resources made available as online supplements, aims to serve as a resource for researchers seeking to understand and implement various longitudinal panel models.
 By providing an overview of different approaches, their strengths and limitations, and key considerations for their use, we hope to facilitate the selection of appropriate models tailored to specific research questions and data structures.
 It is essential for researchers to consider their research objectives, the characteristics of their data, and the assumptions underlying each model when choosing the most suitable approach for longitudinal analysis.
+
 We encourage researchers to consult the cited literature and online supplements for further guidance in selecting and implementing longitudinal models when using the ABCD Study dataset.

### 2024-03-15 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
manuscript updates

@@
 ```
-
+:::
 :::
@@
 
-```{r}
-# Install and load the googledrive package
-if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
-library(googledrive)
+::: blue
 
-# Authenticate with Google Drive
-drive_auth()
+**This code reads in and shows the data to be used in the current example**
 
-# Specify the file ID and the destination path
-file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
-dest_file <- "my_data.rds" # Specify your destination path
-
-# Download the file
-drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
-
-# Load the .rds file for analysis
-my_data <- readRDS(dest_file)
-
-# Begin your analysis (example: view the first few rows)
-head(my_data)
-
-```
-
-:::
-
-<!-- ::: blue
-**This code reads in and shows the data to be used in the current
-example** -->
-
-<!-- ```{r read-data}
+```{r read-data}
 #| echo: TRUE
@@
     "3_year_follow_up_y_arm_1",
-                           "4_year_follow_up_y_arm_1")
+    "4_year_follow_up_y_arm_1"
+)
 
@@
 ```
-::: -->
-
-----
-
-<!-- ```{r}
-# Install and load the googledrive package
-if (!requireNamespace("googledrive", quietly = TRUE)) install.packages("googledrive")
-library(googledrive)
-
-# Authenticate with Google Drive
-
-# Specify the file ID and the destination path
-file_id <- "1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr" # Replace with your actual file ID
-dest_file <- "my_data.rds" # Specify your destination path
-
-# Download the file
-drive_download(as_id(file_id), path = dest_file, overwrite = TRUE)
-
-# Load the .rds file for analysis
-my_data <- readRDS(dest_file)
-
-# Begin your analysis (example: view the first few rows)
-head(my_data)
-
-``` -->
-
-```{r}
-# Set the URL for the CSV file
-file_url <- "https://drive.google.com/file/d/1_gnaEx0wJ28tT0CD_B6mPFwngdfSb6Dr/view?usp=drive_link"
-
-# Set the destination filename
-dest_file <- "my_data.rds"
-
-# Download the file
-download.file(file_url, destfile = dest_file, mode = "wb")
-
-# Read the CSV file into R
-my_data <- readRDS(dest_file)
-
-# Display the first few rows of the data
-head(my_data)
-```
-
-----
+:::
 

### 2024-03-15 examples/_queue/_4_Examples_SignedRankTest.qmd
manuscript updates

--- 2024-03-02 posts/_4_Examples_SignedRankTest.qmd
@@
 ::: blue
-**This code reads in and shows the data to be used in the current
-example**
+
+**This code reads in and shows the data to be used in the current example**
 
@@
 ```
-
-**Descriptive Statistics for the Signed Rank Test result**
-```{r model-output}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
-
-library(gridExtra)
-
-# Creating the first density plot with ggplot2
-p1 <- ggplot(data.frame(x = na.omit(df_wide$income_numeric_0)), aes(x = x)) +
-  geom_density(fill="red", color="blue") +
-  labs(title = "Kernel Density of Income at Baseline")
-
-# Creating the second density plot with ggplot2
-p2 <- ggplot(data.frame(x = na.omit(df_wide$income_numeric_1)), aes(x = x)) +
-  geom_density(fill="blue", color="red") +
-  labs(title = "Kernel Density of Income at T1")
-
-p3<- ggplot(df_wide, aes(x = income_numeric_0, y = income_numeric_1)) + geom_point(color="#04c24a")
-
-# Create both plot and store in variable
-p4<-ggplot(df_wide, aes(x=income_numeric_0, y=income_numeric_1))+
-geom_segment( aes(x=income_numeric_0, xend=income_numeric_1, y=0, yend=income_numeric_2), color="grey") + geom_point( color="green", size=4)
-
-p5<-ggplot(df_wide, aes(x = income_numeric_0)) + 
-geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)
-
-p6<-ggplot(df_wide, aes(x=income_numeric_0, y=income_numeric_1)) + 
-geom_bar(fill="#69b3a2", stat = "identity")
-
-# Arrange the ggplot2 objects on a grid
-grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
-
-ggplot(df_wide, aes(x = sex, y = income_numeric_0, fill = sex)) +
-  geom_boxplot() +
-  scale_fill_manual(values = c("orange", "darkgrey")) +
-  theme_bw() +
-  theme(legend.position = "none")
-```
-
-This signed rank test evaluates xxxx. The model output provides:
-
-i. xxxxx;
-ii. xxxxx;
-iii. xxxxx;
-iv. xxxxx;
-v. xxxxx.
-
-With a xxxx of `xxxxx` and a p-value of `xxxxx`, we can xxxx. Furthermore, we can xxx that xxxxx. This indicates xxxxxx in weight from Baseline to Year 1.

### 2024-03-15 manuscript/manuscript.qmd
manuscript updates

@@
 
-[!Longitudinal Models/Data Structures](img/Mindmap1copy.png)]{#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"}
+[!Longitudinal Models/Data Structures](img/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"}
 

### 2024-03-17 examples/_queue/1b_Examples_DifferenceScores_SimpleRegression.qmd
examples 1-4 updates

--- 2024-03-02 posts/1b_Examples_DifferenceScores_SimpleRegression.qmd
@@
 author: "Biostatistics Working Group"
-
-reader-mode: true
+theme: cosmo
 ---
@@
 library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
+library(rstatix)      # Pipe-friendly framework for performing common stats
 library(gtsummary)    # Publication-ready tables

### 2024-03-17 examples/_queue/3a_Examples_LinearMixedModels.qmd
examples 1-4 updates

--- 2024-03-06 _posts/3a_Examples_LinearMixedModels.qmd
@@
 :::
+:::
+
+## Descriptives Overview
+
+::: panel-tabset
+### Read and View Data {.tabset .tabset-fade .tabset-pills}
 
@@
 ```
-#:::
+:::
 
@@
 **This code creates a descriptives table**
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
 
-:::
 ```{r descriptives}
@@
 ## Results
-<div class = "blue">
 
@@
 ::: blue
-
 The code fits a linear mixed model to predict the 'Height' variable based on time points ('eventname') and handedness ('Handedness'), while accounting for individual-level variability by including random intercepts for each participant ('src_subject_id').
@@
 #confint(model, level = 0.95, method = "Wald")
+
 model_performance(model)
+
 model_parameters(model)
+
 model_parameters(model,standardize = "refit")
@@
 
+:::
+
 ### Model Plots
 
+::: blue
 The following set of plots are used to facilitate model diagnostics.
@@
 
-::: blue
 ```{r diagnosticplots}
@@
 ```{r}
+
+library(ggplot2)
+theme_set(theme_modern(base_size = 6))
+
 plot(parameters(model)) +
   ggplot2::labs(title = "A Dot-and-Whisker Plot")
+#plot(model)
 
-plot(model)
-theme_set(theme_modern())
 result <- estimate_expectation(model, data = "grid")
-
 plot(result)
-
-plot(result, type = "qq")
-plot(result, type = "pp")
+#plot(result, type = "qq")
+#plot(result, type = "pp")
 
 normality <- check_normality(model, effects = "random")
-plot(normality)
+#plot(normality)
 
 heteroscedasticity <- check_heteroscedasticity(model)
-plot(heteroscedasticity)
+#plot(heteroscedasticity)
 
 homogeneity <- check_homogeneity(model)
-plot(homogeneity)
-plot(homogeneity, type = "density")
-
-check_model(model)
-
-check_model(model, panel = FALSE)
+#plot(homogeneity, type = "density")
 
 distribution <- check_distribution(model)
-distribution
-plot(distribution)
+#plot(distribution)
 
-contrasts <- estimate_contrasts(model, contrast = "sex")
-means <- estimate_means(model)
-plot(means)
-plot(contrasts, means)
+#contrasts <- estimate_contrasts(model, contrast = "sex")
+#means <- estimate_means(model)
+#plot(means)
+#plot(contrasts, means)
 
 params <- model_parameters(model, effects = "fixed")
-plot(params)
+#plot(params, show_labels = TRUE, size_text = 4)
 
-plot(params, show_labels = TRUE, size_text = 4)
+#check_model(model)
 
 ```
+
+:::
 
@@
 ```
+:::
 

### 2024-03-18 examples/3b_Examples_LinearMixedModels.qmd
updates

--- 2024-03-02 posts/3b_Examples_LinearMixedModels.qmd
@@
 ::: blue
-
 The code fits a linear mixed model to examine the 'Height' variable across time points ('eventname').
@@
 ```
+
+:::
 
@@
 The average height at r mean(df_descriptable$Height) has increased by r mean(df_descriptable$Height_followup - df_descriptable$Height) units from baseline.
+:::
 

### 2024-03-18 examples/_Template_Example_Complete.qmd
updates

--- 2024-03-15 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
@@
 ---
-title: "Difference Scores: Paired Samples T-test"
-author: "Biostatistics Working Group"
+title: "Add Example Title Here"
+author: "Name Here"
 ---
@@
 
-This example assesses growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1), using weight as a representative metric.
-The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 weight measurements for each participant; 2) a one-sample t-test is used to test whether the difference score is statistically different than a null hypothesis of zero change.
-Finally, a visual inspection is further conducted via a scatterplot to graphically represent the relationship between participant's weight at Baseline and Year_1.
-The ensuing analysis and interpretations are detailed in the subsequent sections.
+[ Insert a brief description of the analysis and its purpose here. ]
 
@@
 ::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
+> **This code installs the r packages necessary for this example, if they are not already installed**
 
@@
 # Create a list of required packages
-packages_required <- c("tidyverse", "easystats", "lme4", "gtsummary", "report", "broom", "gridExtra")
+packages_required <- c("addpackage", "addpackage", "addpackage")
 
@@
 
-library(tidyverse) # Collection of R packages for data science
-library(easystats) # Collection of R packages statistical modeling, visualization, and reporting
-library(gtsummary) # Publication-ready tables
-library(report) # Easy reporting of regression analyses
-library(broom) # Tidy and augment statistical models output
-library(gridExtra) # Arrange multiple grid-based plots on a page
+library(addpackage) # snippet about the package
+library(addpackage) # snippet about the package
+library(addpackage) # snippet about the package
 
@@
 #| output: FALSE
-#| cache: FALSE
 
-# Set the data paths
-data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
-data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
-
-# Read the data
-data_demographics <- readRDS(data_path_1)
-data_nonimaging <- readRDS(data_path_2)
-
-# Subset the nonimaging data to include desired variables
-selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroweightcalc")
-subset_data <- data_nonimaging[, selected_vars]
-
-library(dplyr)
-# # Merge the datasets on 'src_subject_id' and 'eventname'
-merged_data <- data_demographics %>%
-    full_join(subset_data, by = c("src_subject_id", "eventname"))
-
-# Inspect the merged data structure
-str(merged_data)
-
-# Define event names to be retained in the analysis and convert variables to appropriate data types
-eventnames_to_include <- c(
-    "baseline_year_1_arm_1",
-    "1_year_follow_up_y_arm_1",
-    "2_year_follow_up_y_arm_1",
-    "3_year_follow_up_y_arm_1",
-    "4_year_follow_up_y_arm_1"
-)
-
-df <- merged_data %>%
-    filter(eventname %in% eventnames_to_include) %>%
-    mutate(
-        src_subject_id = as.factor(src_subject_id),
-        eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
-        age = as.numeric(age),
-        sex = as.factor(sex),
-        race.4level = as.factor(race.4level),
-        hisp = as.factor(hisp),
-        high.educ.bl = as.factor(high.educ.bl),
-        household.income.bl = as.factor(household.income.bl),
-        acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
-        rel_family_id.bl = as.factor(rel_family_id.bl),
-        site_id_l = as.factor(site_id_l),
-        nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
-        anthroweightcalc = as.numeric(anthroweightcalc),
-        anthroweightcalc = as.numeric(anthroweightcalc)
-    ) %>%
-    # Exclude cases from unused assessment waves
-    filter(!is.na(eventname))
+# Read and wrangle data
 
@@
 
-df <- my_data
+```
 
-descriptives_table <- df %>%
-    select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
-    mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1", "Year 2", "Year 3", "Year 4"))) %>%
-    mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
-    tbl_summary(
-        by = eventname,
-        missing = "no",
-        label = list(
-            sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic",
-            anthroweightcalc ~ "Weight"
-        ),
-        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
-    ) %>%
-    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
-    bold_labels() %>%
-    italicize_levels() %>%
-    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
-theme_gtsummary_compact()
+:::
 
-descriptives_table
+## Results
+
+::: blue
+[Brief explanation of the code]
+```{r build-model-ttest-1}
+#| message: FALSE
+#| warning: FALSE
+#| echo: TRUE
+
 ```
@@
 
-## Results
+:::
+
+### Model Summary {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+
+A plot to show xxxxx.
+```{r model-checking}
+
+```
+
+[brief write-up/explanation of the results]
+
+:::
+
+### Model Plots  {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+```{r scatterplot}
+
+```
+
+:::
+:::
+
+## Wrapping Up
 
 ::: panel-tabset
-### Build Model: Step I {.tabset .tabset-fade .tabset-pills}
+### Write-up {.tabset .tabset-fade .tabset-pills}
 
 ::: blue
-The code snippet below tells R to compute a difference score by subtracting each participant's weight at T1 from their weight at T0. Relevant summary statistics are also provided.
 
-**STEP 1: Compute Difference Score**
+[insert brief write-up here]
 
-```{r build-model-ttest-1}
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-#| class.source: 'fold-hide'
-
-# Function to compute difference scores
-compute_difference <- function(df, variable_name) {
-    # Define the event names of interest
-    baseline_event <- "baseline_year_1_arm_1"
-    followup_event <- "1_year_follow_up_y_arm_1"
-
-    # Compute the difference between Baseline and Year 1 data for the given variable
-    diff_data <- df %>%
-        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
-        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
-        spread(eventname, variable_name) %>% # Convert data from long to wide format
-        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
-        mutate(group_resid = (diff) - mean(diff)) %>% # compute residuals (group_resid).
-        drop_na(diff) # Exclude rows with NA in the computed difference
-
-    # Summarize the computed difference scores
-    diff_summary <- summary(diff_data$diff)
-
-    # Return the difference data and its summary
-    list(data = diff_data, summary = diff_summary)
-}
-
-# List of variables for which difference scores are to be computed
-variables_of_interest <- c("anthroweightcalc")
-
-# Compute and store difference scores and summaries for each variable in a list
-difference_and_summary_list <- lapply(variables_of_interest, function(var) {
-    compute_difference(df, var)
-})
-
-# Extract the difference data for the 'anthroweightcalc' variable
-weight_diff_data <- difference_and_summary_list[[1]]$data
-
-# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
-df <- left_join(df, weight_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")
-
-```
+:::
+:::

### 2024-03-18 manuscript/manuscript.qmd
update

@@
 
-[!Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}
+![Figure 3 Statistical modeling of brain imaging data](img/Picture1.png)
+
+<!-- ![Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"} -->
+
+<img src="img/StabilityChange.png" alt="Types of Stability and Change" width="50%" height="auto"/>
+
+![Types of Stability and Change](img/StabilityChange.png){width=50%}
+
+![Types of Stability and Change](img/StabilityChange.png){fig-width=50%}
+
+![Types of Stability and Change](img/StabilityChange.png)
+
+<!-- ![Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram."}
+
+[!Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"} -->
 
@@
 
-[!Longitudinal Models/Data Structures](img/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"}
+![Longitudinal Models/Data Structures](img/Mindmap1.png)
+
+<!-- [![Longitudinal Models/Data Structures](img/Mindmap1copy.png)]{#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="100" height="100" fig-align="left"}  -->
 

### 2024-03-18 index.qmd
update

--- 2024-03-04 index.qmd
@@
 ---
-#title: "Longitudinal Analysis"
-
+title: "Longitudinal Data Analysis with the ABCD Study"
 format:
   html:
-    css: styles.css
-    #theme: cosmo
-    page-layout: custom
-    grid:
-      sidebar-width: 200px
-      body-width: 2000px
-      margin-width: 0px
-      gutter-width: 0rem
+    #css: styles.css
+    theme: cosmo
+    #page-layout: custom
+    #grid:
+      #sidebar-width: 200px
+      #body-width: 2000px
+      #margin-width: 0px
+      #gutter-width: 0rem
 ---
 
-Test
+# Welcome to the Longitudinal Data Analysis Project
+
+## Our Mission
+
+This ongoing project is guided by the Biostatistics Working Group of the [Adolescent Brain Cognitive Development (ABCD) Study](https://abcdstudy.org/).
+The aim is to provide accessible resources for those interested in longitudinal data analysis and related methodologies.
+We provide practial [examples and tutorials](https://opendevsci.github.io/LongDev-ABCD/examples/examples.html) to demonstrate different longitudinal methodologies/techniques using publically available data from the ABCD Study.
+
+::: {.callout-important}
+NOTE: This project is in beta and as such is still very much a work in progress.
+At present, [Examples](https://opendevsci.github.io/LongDev-ABCD/examples/examples.html) should NOT be considered "ready-to-use".
+However, avaiable code snippets in each of the examples may still prove useful to guide your own analysis.
+We very much welcome feedback and contributions from the community to help us expand and improve our resources.
+Please see our [GitHub repository](https://github.com/OpenDevSci/LongDev-ABCD) and our [Contribution Guide](https://github.com/OpenDevSci/LongDev-ABCD/blob/main/docs/Contributing.md) for more information on how to contribute.
+:::
+
+## Why This Project?
+
+- **Publicly Available Code**: We (will) provide ready-to-use code examples to aid in your research and studies.
+- **Open Source Collaboration**: Join our community to contribute, learn, and share knowledge on longitudinal data analysis.
+- **Educational Resource**: Whether you're looking to understand GitHub, R, RMarkdown/Quarto, or longitudinal analysis methods, we're here to help.
+
+## How to Contribute
+
+1. **Provide New Examples**: Share your knowledge by contributing new analysis examples.
+2. **Help With Documentation**: Assist in enhancing our project documentation for better user understanding.
+3. **Repository Maintenance**: Support the upkeep of our GitHub repository to ensure it remains an accessible resource for all.
+
+Interested in contributing? Check out our [Contribution Guide](https://github.com/OpenDevSci/LongDev-ABCD/blob/main/docs/Contributing.md) for more details.
+
+## Project Resources
+
+- **Code Examples**: Explore this website and our [GitHub Repository](https://github.com/OpenDevSci/LongDev-ABCD) for a range of analysis examples.
+- **Documentation**: Dive into our [Project Board](https://github.com/orgs/OpenDevSci/projects/13) for detailed insight into ongoing project tasks and timelines.
+- **Community**: Join our [Discussion Forum](https://github.com/orgs/OpenDevSci/discussions) to connect with other community members.
+
+## About Us
+
+We're passionate about making longitudinal data analysis accessible and understandable.
+While we're not experts in every area, we believe in the power of community and open source collaboration to expand knowledge and skills in these essential research methodologies.
+
+**Let's make data analysis more approachable together.**

### 2024-03-18 index.qmd
updating landing page

--- 2024-03-04 index.qmd
@@
 ---
-#title: "Longitudinal Analysis"
-
+#title: "Longitudinal Data Analysis with the ABCD Study"
 format:
@@
     css: styles.css
-    #theme: cosmo
-    page-layout: custom
-    grid:
-      sidebar-width: 200px
-      body-width: 2000px
-      margin-width: 0px
-      gutter-width: 0rem
+
 ---
 
-Test
+# Welcome to the Longitudinal Data Analysis Project
+
+This ongoing project, guided by the Biostatistics Working Group of the [Adolescent Brain Cognitive Development (ABCD) Study](https://abcdstudy.org/), aims to democratize access to longitudinal data analysis through practical examples and tutorials.
+
+::: {.callout-important}
+**Note**: This project is currently in beta.
+[Examples](https://opendevsci.github.io/LongDev-ABCD/examples/examples.html) are under development and may not be ready for direct application.
+Your feedback and contributions are invaluable to us—join us in enhancing these resources.
+For more details on contributing, visit our [GitHub repository](https://github.com/OpenDevSci/LongDev-ABCD) and our [Contribution Guide](https://github.com/OpenDevSci/LongDev-ABCD/blob/main/docs/Contributing.md).
+:::
+
+## Our Mission
+
+We aim to provide accessible resources for those interested in longitudinal data analysis, using publicly available data from the ABCD Study to showcase various methodologies and techniques.
+
+## Why This Project?
+
+- **Publicly Available Code**: Future updates will include ready-to-use code examples for your research and educational needs.
+See some initial examples [here](https://opendevsci.github.io/LongDev-ABCD/examples/examples.html) or in the `examples` folder of our [GitHub repository](https://github.com/OpenDevSci/LongDev-ABCD)
+- **Open Source Collaboration**: Engage with our community to contribute, learn, and share insights on longitudinal data analysis.
+- **Educational Resource**: Gain comprehensive knowledge in using GitHub, R, RMarkdown/Quarto, and various longitudinal analysis methods with our guidance.
+
+## How to Contribute
+
+- **Provide New Examples**: Enhance our repository with your unique analysis examples.
+- **Help With Documentation**: Contribute to making our documentation more comprehensive and user-friendly.
+- **Repository Maintenance**: Assist in the ongoing maintenance and accessibility of our GitHub repository.
+
+For more information on how you can make an impact, check out our [Contribution Guide](https://github.com/OpenDevSci/LongDev-ABCD/blob/main/docs/Contributing.md).
+
+## Project Resources
+
+- **[Code Examples](https://github.com/OpenDevSci/LongDev-ABCD)**: Discover a variety of analysis examples to guide your research.
+- **[Documentation](https://github.com/orgs/OpenDevSci/projects/13)**: Get detailed insights into our project's progress, tasks, and timelines.
+- **[Community Discussion Forum](https://github.com/orgs/OpenDevSci/discussions)**: Connect with fellow researchers and enthusiasts in our dedicated forum.
+
+## About Us
+
+We are a team driven by the mission to make longitudinal data analysis more accessible and understandable.
+Our collective effort focuses on fostering an open source community where knowledge and skills in research methodologies can flourish.
+
+**Together, let's advance the field of data analysis.**

### 2024-03-19 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
updates
test

--- 2024-03-15 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
@@
 author: "Biostatistics Working Group"
+theme: cosmo
 ---
@@
 <p>
-
+test-remove
 This example assesses growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1), using weight as a representative metric.
@@
 
-df <- my_data
-
 descriptives_table <- df %>%
@@
 :::
-:::
 
-## Results
-
-::: panel-tabset
-### Build Model: Step I {.tabset .tabset-fade .tabset-pills}
+### Results
 
@@
 ```
+
+:::
+:::

### 2024-03-19 manuscript/manuscript.qmd
update manuscript

--- 2024-03-15 manuscript/manuscript.qmd
@@
 
-[!Types of Stability and Change](img/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}
+![Types of Stability and Change](img/StabilityChange.png){width=100% #fig-StablityChange fig-alt="Stability & Change Diagram." fig-align="left"}
 
@@
 
-[![Longitudinal Data Analysis Structural Diagram](img/LDAStructure2.png)](https://centerstat.org/resources/)
+![Longitudinal Data Analysis Structural Diagram](img/LDAStructure2.png){width=100% #fig-StablityChange fig-alt="Stability & Change Diagram." fig-align="left"}
+[Link]((https://centerstat.org/resources/))
 
@@
 
-[!Longitudinal Models/Data Structures](img/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"}
-
-[![Longitudinal Data Analysis Structural Diagram](img/LDAStructure2.png)](https://centerstat.org/resources/)
-<br>
+![Longitudinal Models/Data Structures](img/Mindmap1.png){width=100% #fig-DataStructures fig-alt="Longitudinal Models & Data Structures" fig-align="left"}
 
@@
 
-![LDA Table](img/Tabletemp.png)
+![LDA Table](img/Tabletemp.png){width=100% #fig-LDATable fig-alt="LDA Table" fig-align="left"}
 

### 2024-03-19 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
example updates

--- 2024-03-15 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
@@
 
-### Config Options
-
-::: blue
-**This code configures knitr code chunk options**
-
 ```{r config}
-#| echo: TRUE
+#| echo: FALSE
 #| messages: FALSE
@@
 #| output: FALSE
+
+### Config Options
+#::: blue
+#**This code configures knitr code chunk options**
 
@@
 ```
-:::
 :::

### 2024-03-19 examples/2_Examples_ResidualizedChangeScores.qmd
example updates

--- 2024-03-06 _posts/2_Examples_ResidualizedChangeScores.qmd
@@
 ```
+:::
 
-:::
 :::

### 2024-03-19 examples/4_Examples_SignedRankTest.qmd
updates
example updates

--- 2024-03-15 examples/_queue/_4_Examples_SignedRankTest.qmd
@@
 
-####
 It is actually quite common that numeric depend variables need to be transformed or converted into ranks, i.e.
@@
 ::: blue
-
 The code fits a xxxx to examine the 'xxxxx' variable across time points ('eventname').
@@
 ```
+
+:::
+
+### Model Plots
+
+::: blue
+Testtesttest
+
+:::
+:::
+
+## Wrapping Up
+<div class = "blue">
+
+::: panel-tabset
+### Write-up {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+
+:::
+:::

### 2024-03-19 manuscript/manuscript.qmd
example updates

@@
 abstract: |
-  The Adolescent Brain Cognitive Development^SM (ABCD) Study provides a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of youths, aged 9-10 at baseline and assessed annually for 10 years.
-Given the size and complexity of the ABCD Study©, researchers analyzing its data will encounter a myriad of methodological and analytical considerations.
+  The Adolescent Brain Cognitive Development&reg; (ABCD) Study provides a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of youths, aged 9-10 at baseline and assessed annually for 10 years.
+Given the size and complexity of the ABCD Study, researchers analyzing its data will encounter a myriad of methodological and analytical considerations.
 This review provides an examination of key concepts and techniques related to longitudinal analyses of the ABCD Study data.
@@
 
-[The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest longitudinal investigation of neurodevelopment and child health in the United States.
-Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence (@volkow2018).
+[The Adolescent Brain Cognitive Development (ABCD) Study&reg;](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest longitudinal investigation of neurodevelopment and child health in the United States.
+Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental factors impacting neurodevelopment and their roles in behavioral and health outcomes across ten years of adolescence @volkow2018.
 At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes.
@@
 Each new wave of data collection provides another building block for characterizing developmental trajectories and implementing longitudinal analyses that allow researchers to characterize normative development, to identify variables that presage deviations from normative development, and to assess a range of variables associated with biopsychosocial outcomes of interest.
-These data include: 1) a neurocognitive battery (@luciana2018; @thompson2019); 2) mental and physical health assessments (@barch2018); 3) measures of culture and environment (@gonzalez2021; @zucker2018); 4) substance use (@lisdahl2021); 5) gender identity and sexual health (@potter2022); 6) biospecimens (@uban2018); 7) structural and functional brain imaging (@casey2018; @hagler2019; @palmer2022); 8) geolocation-based environmental exposure data (@fan2021); 9) wearables and mobile technology (@bagot2018); and 10) whole-genome genotyping (@loughnan2020).
+These data include: 1) a neurocognitive battery @luciana2018; @thompson2019; 2) mental and physical health assessments @barch2018; 3) measures of culture and environment @gonzalez2021; @zucker2018; 4) substance use @lisdahl2021; 5) gender identity and sexual health @potter2022; 6) biospecimens @uban2018; 7) structural and functional brain imaging @casey2018; @hagler2019; @palmer2022; 8) geolocation-based environmental exposure data @fan2021; 9) wearables and mobile technology @bagot2018; and 10) whole-genome genotyping @loughnan2020.
 Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward.
@@
 Rate of change or other trajectory characteristics may be more informative about individuals than the simple snapshots of level differences that cross-sectional data are limited to informing about.
-Cross-sectional age-related differences across individuals are poor substitutes for longitudinal trajectory estimates, except under highly restrictive assumptions, e.g., parallel trajectories and lack of age, cohort and experience effects (@thompson2011).
+Cross-sectional age-related differences across individuals are poor substitutes for longitudinal trajectory estimates, except under highly restrictive assumptions, e.g., parallel trajectories and lack of age, cohort and experience effects @thompson2011.
 Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.
@@
 However, unique epochs and experiences can alter the course of this idealized form of development.
-Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development (@arseneault2002; @bechtold2016; @hasan2020; @semple2005).
-Similarly, rodent brains are especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development, corresponding to early adolescence in humans (@spear2016; @crews2000; @ji2018).
-In another example, longitudinal data from the National Consortium on Alcohol and Neurodevelopment in Adolescence (NCANDA) show that binge drinking is associated more strongly with decrements in gray matter volume early in adolescence compared to later (@infante2022).
+Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development @arseneault2002; @bechtold2016; @hasan2020; @semple2005.
+Similarly, rodent brains are especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development, corresponding to early adolescence in humans @spear2016; @crews2000; @ji2018.
+In another example, longitudinal data from the National Consortium on Alcohol and Neurodevelopment in Adolescence (NCANDA) show that binge drinking is associated more strongly with decrements in gray matter volume early in adolescence compared to later @infante2022.
 These examples highlight the importance of considering the role of vulnerable periods – e.g., temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.
@@
 Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next.
-For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior (@moffitt2015).
+For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior @moffitt2015.
 Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified (e.g., @masten2005; @rogosch2010).
@@
 
-Questions regarding the biological mechanisms whereby exposures impact outcomes can often be framed in terms of mediation analyses (@mackinnon2007; @vanderweele2016).
-Mediation analyses can be implemented using the causal steps approach (@baron1986) and structural equation models (SEM) (@preacher2011).
-More recently, mediation models have been adapted for longitudinal exposures, mediators, and/or outcomes (@bind2016; @vanderweele2017).
+Questions regarding the biological mechanisms whereby exposures impact outcomes can often be framed in terms of mediation analyses @mackinnon2007; @vanderweele2016.
+Mediation analyses can be implemented using the causal steps approach @baron1986 and structural equation models (SEM) @preacher2011.
+More recently, mediation models have been adapted for longitudinal exposures, mediators, and/or outcomes @bind2016; @vanderweele2017.
 All of these modeling approaches decompose the total effects of an exposure on an outcome into direct and indirect effects, where indirect effects of an exposure flow through its impact on a mediating process.
@@
 An important example of mediational analyses in the ABCD Study is the impact of exposures on behavioral outcomes (e.g., neurocognition, mental health, substance use) via their impact on the brain, as quantified by imaging-derived phenotypes (IDPs).
-Methods for mediational analyses using multi-dimensional IDPs have been developed and applied to functional MRI data (@lindquist2012; @zhao2018).
+Methods for mediational analyses using multi-dimensional IDPs have been developed and applied to functional MRI data @lindquist2012; @zhao2018.
 
@@
 This is also consistent with recommendations that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (given that linear change is the only estimable form for two assessment waves; (see @duncan2009).
-Research designs that include three (but preferably more) time points allow for non-linear trajectory estimation and increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments (@king2018)– a key aspect of developmental research.
+Research designs that include three (but preferably more) time points allow for non-linear trajectory estimation and increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments @king2018– a key aspect of developmental research.
 
 To illustrate, developmental theories are useful for understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level.
-This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” (@littlefield2021).
+This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021.
 It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling.
-Indeed, several contemporary longitudinal models designed to isolate within-individual variability (e.g., the Latent Curve Model with Structured Residuals [LCM: SR]; @curran2014) require at least four assessments to parameterize fully and, more generally, increasingly accurate and nuanced parameter estimates are obtained as more assessment occasions are used (@duncan2009).
+Indeed, several contemporary longitudinal models designed to isolate within-individual variability (e.g., the Latent Curve Model with Structured Residuals [LCM: SR]; @curran2014) require at least four assessments to parameterize fully and, more generally, increasingly accurate and nuanced parameter estimates are obtained as more assessment occasions are used @duncan2009.
 
@@
 Mean-level change in certain traits might help to explain why, in general, populations of individuals tend to be particularly vulnerable to the effects of environmental factors in specific age ranges; rank-order stability might help to quantify the extent to which certain characteristics of the individual are more or less trait-like compared to others.
-For example, in some areas of development, considerable mean-level change occurs over time (e.g., changes in Big 5 personality traits (@bleidorn2022), but exhibit relatively high rank-order stability, at least over shorter measurement intervals (@bleidorn2022; @roberts2000; @roberts2006).
+For example, in some areas of development, considerable mean-level change occurs over time (e.g., changes in Big 5 personality traits @bleidorn2022, but exhibit relatively high rank-order stability, at least over shorter measurement intervals @bleidorn2022; @roberts2000; @roberts2006.
 
-Despite the useful information afforded by examining mean-level and rank-order stability and change, these approaches are limited in that they provide little information about the overall patterns of within-individual change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest (@curran2011).
+Despite the useful information afforded by examining mean-level and rank-order stability and change, these approaches are limited in that they provide little information about the overall patterns of within-individual change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest @curran2011.
 For example, questions related to the impact of early-onset substance use on brain development focus on changes within a given individual (i.e., intraindividual differences).
@@
 As the study progresses and more time points are assessed, the potentially nonlinear aspects of trajectories can be assessed, for example using quadratic functions of time.
-Methods that make even fewer assumptions about trajectory shapes, such as nonparametric curve estimation at the mean (e.g., Generalized Additive Mixed Models [GAMMs]; (@wood2017)) and at the individual level (e.g., Functional Data Analysis [FDA]; @ramsay2002) may also become useful.
-Note, baseline age in the ABCD Study ranges over two full years; for some outcomes it may be feasible to include a possibly nonlinear effect of baseline age along with a linear effect of within-person change in age even with only two or three assessment times (@thompson2013).
+Methods that make even fewer assumptions about trajectory shapes, such as nonparametric curve estimation at the mean (e.g., Generalized Additive Mixed Models [GAMMs]; @wood2017) and at the individual level (e.g., Functional Data Analysis [FDA]; @ramsay2002) may also become useful.
+Note, baseline age in the ABCD Study ranges over two full years; for some outcomes it may be feasible to include a possibly nonlinear effect of baseline age along with a linear effect of within-person change in age even with only two or three assessment times @thompson2013.
 
@@
 In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks (e.g., see @lenz2016).
-For example, the Mplus manual (@muthen2017) includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, and d) linear growth models for a count outcome assuming a zero-inflated Poisson model.
-Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data (@ren2022).
+For example, the Mplus manual @muthen2017 includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, and d) linear growth models for a count outcome assuming a zero-inflated Poisson model.
+Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data @ren2022.
 These models account for issues that may occur when working with discrete outcomes, including overdispersion, i.e., when the variance is higher than would be expected based on a given parametric distribution (see @lenz2016).
@@
 For example, if data missingness or participant dropout are related to the values of the outcome, changing sample composition as the study progresses can bias mean trajectory estimates (we describe this in more detail in Section 3.1.7 below).
-Another prerequisite for valid developmental interpretations of longitudinal data is to establish whether a construct is measured consistently over time, i.e., longitudinal measurement invariance (@liu2017; @van2015; @willoughby2012).
+Another prerequisite for valid developmental interpretations of longitudinal data is to establish whether a construct is measured consistently over time, i.e., longitudinal measurement invariance @liu2017; @van2015; @willoughby2012.
 Establishing longitudinal measurement invariance ensures that change over time for a given construct is attributable to individual development rather than merely a measurement artifact.
-For instance, one study using data from the ABCD Study (@brislin2023) found differential item functioning in two items from a brief delinquency measure, revealing significant bias in an arrest item across Black and White youth.
+For instance, one study using data from the ABCD Study @brislin2023 found differential item functioning in two items from a brief delinquency measure, revealing significant bias in an arrest item across Black and White youth.
 More specifically, Black youth were more likely to report being arrested compared to White youth with similar levels of delinquency.
@@
 Observed patterns of growth and decline often differ between cross-sectional vs.
-longitudinal effects (@salthouse2014) where subjects gain increasing experience with the assessment with each successive measurement occasion.
-Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD (@sullivan2017) and highlight the need to consider these effects and address them analytically.
+longitudinal effects @salthouse2014 where subjects gain increasing experience with the assessment with each successive measurement occasion.
+Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD @sullivan2017 and highlight the need to consider these effects and address them analytically.
 In the case of performance-based measures (e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014), this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before).
-Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments (e.g., @beck1961; @french2010) and this phenomenon has been well documented in research using structured diagnostic interviews (@robins1985).
+Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments (e.g., @beck1961; @french2010) and this phenomenon has been well documented in research using structured diagnostic interviews @robins1985.
 While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artifactually elevated (see @shrout2018).
@@
 For example, it is typical to assume that the random intercept and slope components covary, i.e., an individual’s score at time zero relates to the amount of change exhibited across time.
-Further, particularly in structural equation model forms of this model, it is sometimes assumed that the variance of the residuals varies across assessments (@curran2003).
+Further, particularly in structural equation model forms of this model, it is sometimes assumed that the variance of the residuals varies across assessments @curran2003.
 
@@
 In particular, recent work demonstrates that if a construct being assessed prospectively across time is trait-like in nature, then a simple AR(1) process fail to adequately account for this trait-like structure, with the downstream consequence that estimates derived from models based on AR structures (such as the CLPM) can be misleading and fail to adequately demarcate between- vs.
-within-person sources of variance (@hamaker2015).
+within-person sources of variance @hamaker2015.
 Note also, discrete-time autoregressive structures such as AR(1) implicitly assumes relatively constant time gaps between visits; this may not be true in many applications using the ABCD Study data.
@@
 
-Attrition from a longitudinal study such as ABCD is inevitable and represents a potential threat to the external validity of analyses conducted at later visits, especially since attrition can only be expected to grow over time (@littlefield2022).
-The ABCD Retention Workgroup employs a data-driven approach to examine, track, and intervene in these issues and while preliminary findings show participant race and parent education level to be associated with late and missing visits, although to date, formal attrition in ABCD has been minimal (@ewing2022).
-Ideally, one tries to minimize attrition through good retention practices from the outset via strategies designed to maintain engagement in the project (@cotter2005; @hill2016; @watson2018).
+Attrition from a longitudinal study such as ABCD is inevitable and represents a potential threat to the external validity of analyses conducted at later visits, especially since attrition can only be expected to grow over time @littlefield2022.
+The ABCD Retention Workgroup employs a data-driven approach to examine, track, and intervene in these issues and while preliminary findings show participant race and parent education level to be associated with late and missing visits, although to date, formal attrition in ABCD has been minimal @ewing2022.
+Ideally, one tries to minimize attrition through good retention practices from the outset via strategies designed to maintain engagement in the project @cotter2005; @hill2016; @watson2018.
 However, even the best-executed studies need to anticipate growing attrition over the length of the study and implement analytic strategies designed to provide the most valid inferences.
@@
 
-Three types of missingness are considered in the literature (@little1989; @little2013), namely: a) missing completely at random (MCAR), b) missing at random (MAR), and c) missing not at random (MNAR).
+Three types of missingness are considered in the literature @little1989; @little2013, namely: a) missing completely at random (MCAR), b) missing at random (MAR), and c) missing not at random (MNAR).
 Data that are MCAR are a simple random sample of all data in a given dataset.
@@
 An effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” (Kraemer 2014).
-Common effect size metrics include the Pearson correlation r between two variables and the standardized difference between two means, Cohen’s d (@cohen1988).
+Common effect size metrics include the Pearson correlation r between two variables and the standardized difference between two means, Cohen’s d @cohen1988.
 An extensive discussion of cross-sectional effect sizes and their relevance for ABCD is given in @dick2021.
@@
 
-An ideal longitudinal analysis integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model (@collins2006).
+An ideal longitudinal analysis integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model @collins2006.
 To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure 1).
@@
 Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and @bauer2019 provide a nice on-ramp in first distinguishing between the use of “time-to-event” and “repeated measures” data.
-Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change (@bauer2019 Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
-This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models) and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events.
-While much has been written about “time-to-event” data (@hosmer2008; @rizopoulos2012), including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study (@brislin2023), our emphasis will be given to the modeling of “repeated measures” data.
+Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change @bauer2019 Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
+This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events.
+While much has been written about “time-to-event” data @hosmer2008; @rizopoulos2012, including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study @brislin2023, our emphasis will be given to the modeling of “repeated measures” data.
 
@@
 Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
-Common examples include grouping by linear vs nonlinear models (@collins2006), the number of measurement occasions (@king2018), and statistical equivalency (e.g., change scores vs.
+Common examples include grouping by linear vs nonlinear models @collins2006, the number of measurement occasions @king2018, and statistical equivalency (e.g., change scores vs.
 residualized change; see @castro-schilo2018).
@@
 Additionally, the longitudinal signed-rank test, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data.
-Another common approach, often used in pre-/post-design studies but can be used with ABCD Study data, is to use residualized change score analysis to assess the degree of change in a variable, while controlling for its initial level (@castro-schilo2018).
+Another common approach, often used in pre-/post-design studies but can be used with ABCD Study data, is to use residualized change score analysis to assess the degree of change in a variable, while controlling for its initial level @castro-schilo2018.
 
@@
 Within the ABCD Study, researchers may want to consider nesting by individual, family (i.e., siblings or twins), school or district, and/or site.
-GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data, such as the use of ABCD data on substance use (e.g., @martz2022) screen media use (@lees2020), and microstructure of the brain (@palmer2022).
+GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data, such as the use of ABCD data on substance use (e.g., @martz2022) screen media use @lees2020, and microstructure of the brain @palmer2022.
 
@@
 #### Structural Equation Modeling (SEM)
-Structural Equation Modeling (SEM) is a flexible modeling framework that integrates elements of path analysis and confirmatory factor analysis (CFA) to examine complex relationships between a set of observable variables and latent constructs (@hair2021).
-The integration of structural (regression) and measurement (CFA) components within a unified framework supports a theory-driven approach that allows researchers to rigorously test hypothesized relationships among variables of interest and their underlying causes (@hair2021; @raykov2012).
-Over the years, the flexibility of the SEM framework has evolved to become particularly adept for modeling autoregressive processes (which often assume underlying stationarity) and growth processes which accommodate both, mean trajectories and individual differences in them (@mcardle2009; @little2013).
+Structural Equation Modeling (SEM) is a flexible modeling framework that integrates elements of path analysis and confirmatory factor analysis (CFA) to examine complex relationships between a set of observable variables and latent constructs @hair2021.
+The integration of structural (regression) and measurement (CFA) components within a unified framework supports a theory-driven approach that allows researchers to rigorously test hypothesized relationships among variables of interest and their underlying causes @hair2021; @raykov2012.
+Over the years, the flexibility of the SEM framework has evolved to become particularly adept for modeling autoregressive processes (which often assume underlying stationarity) and growth processes which accommodate both, mean trajectories and individual differences in them @mcardle2009; @little2013.
 
-Longitudinal SEM techniques share many similarities with mixed-effects methods and research demonstrates their mathematical equivalence in many situations (@curran2003; @mehta2005).
+Longitudinal SEM techniques share many similarities with mixed-effects methods and research demonstrates their mathematical equivalence in many situations @curran2003; @mehta2005.
 However, these related approaches often cater to distinct theoretical and analytical needs.
-For instance, mixed-effects techniques are an extension of the regression framework and often excel when working with complex data structures such as multiple levels of nesting, small samples, and non-equidistant time points (@mcneish2018; @hedeker2006).
+For instance, mixed-effects techniques are an extension of the regression framework and often excel when working with complex data structures such as multiple levels of nesting, small samples, and non-equidistant time points @mcneish2018; @hedeker2006.
 Alternatively, applying SEM methods to longitudinal analysis provides a flexible means for modeling the underlying process of change.
-It also addresses several challenges faced by competing approaches, including the ability to accommodate intricate error structures and deal effectively with missing data, as well as the implementation of numerous modeling extensions (@mcneish2018; @curran2021).
-These models have grown increasingly popular for modeling longitudinal outcomes particularly due to their ability to build statistical models that match some particular underlying theory (@serang2019).
+It also addresses several challenges faced by competing approaches, including the ability to accommodate intricate error structures and deal effectively with missing data, as well as the implementation of numerous modeling extensions @mcneish2018; @curran2021.
+These models have grown increasingly popular for modeling longitudinal outcomes particularly due to their ability to build statistical models that match some particular underlying theory @serang2019.
 
 Considering the variety of available techniques, it can be helpful to classify longitudinal SEM, broadly (if not coarsely), into variable-centered, person-centered, and hybrid analyses, each with unique strengths and limitations.
-Variable-centered analyses (e.g., latent growth curves @curran2003, latent change scores @mcardle2001, latent state-trait models @geiser2012) are primarily concerned with understanding covariation among variables at the group level and characterizing population-level patterns of change, while person-centered analyses (e.g., latent class and latent transition models) identify distinct subgroups or patterns within the data (@muthen2000; @woo2024; @howard2018).
-Hybrid models combine these perspectives to offer a comprehensive analysis of latent subgroups and growth parameter relationships (@morin2018; @lubke2005).
+Variable-centered analyses (e.g., latent growth curves @curran2003, latent change scores @mcardle2001, latent state-trait models @geiser2012) are primarily concerned with understanding covariation among variables at the group level and characterizing population-level patterns of change, while person-centered analyses (e.g., latent class and latent transition models) identify distinct subgroups or patterns within the data @muthen2000; @woo2024; @howard2018.
+Hybrid models combine these perspectives to offer a comprehensive analysis of latent subgroups and growth parameter relationships @morin2018; @lubke2005.
 The choice between these approaches is primarily driven by the research question, data structure, and relevant underlying assumptions.
@@
 One key application of the SEM framework to the analysis of longitudinal data is the latent growth curve model (LGCM).
-This is a variable-centered approach that characterizes average group trajectories and individual variations (random effects) in an outcome over time (@curran2003).
-These models are similar to their linear mixed effects counterpart in many ways, with the main conceptual difference being that LGCM includes a repeatedly measured outcome in the model as a function of time (closely resembling a standard CFA approach), rather than as an explanatory variable (as in a standard regression approach) (@mcneish2018).
+This is a variable-centered approach that characterizes average group trajectories and individual variations (random effects) in an outcome over time @curran2003.
+These models are similar to their linear mixed effects counterpart in many ways, with the main conceptual difference being that LGCM includes a repeatedly measured outcome in the model as a function of time (closely resembling a standard CFA approach), rather than as an explanatory variable (as in a standard regression approach) @mcneish2018.
 Specifically, observed scores at each time point are treated as indicator variables with their factors loading scaled to reflect a hypothesized pattern of change (e.g., loadings of 0, 1, and 2 would assume equidistant, linear change).
@@
 
-The latent change score model (LCSM) is a variable-centered approach uniquely tailored for analyzing temporal variations in how a construct changes over time (@mcardle1994; @mcardle2001).
-These models share many features with growth curve analysis, but with a more explicit focus on how change occurs between measurement occasions (@serang2019; @mcardle2009).
-Specifically, LCSM estimates a series of latent variables to model change in an outcome from one time point to the next, as a function of scores on that outcome at prior time points (@mcardle2001; @ghisletta2012).
+The latent change score model (LCSM) is a variable-centered approach uniquely tailored for analyzing temporal variations in how a construct changes over time @mcardle1994; @mcardle2001.
+These models share many features with growth curve analysis, but with a more explicit focus on how change occurs between measurement occasions @serang2019; @mcardle2009.
+Specifically, LCSM estimates a series of latent variables to model change in an outcome from one time point to the next, as a function of scores on that outcome at prior time points @mcardle2001; @ghisletta2012.
 Some types of LCSM estimate two underlying latent factors: a constant change factor that remains fixed over time, and a proportional change factor that adjusts for previous scores.
-By disaggregating change into constant and proportional components, this approach facilitates a more nuanced understanding of whether prior changes in a given process are related to future changes in the same process (@serang2019; @kievit2018).
+By disaggregating change into constant and proportional components, this approach facilitates a more nuanced understanding of whether prior changes in a given process are related to future changes in the same process @serang2019; @kievit2018.
 Expanding upon the capabilities of this framework, LCSM also allows for comprehensive multivariate analyses that can facilitate investigations into how change in one construct is associated with change in another construct.
-The appeal of this approach is evidenced by several recent studies that have used data from the ABCD Study to explore bivariate associations between brain development and changes in several mental and physical health indicators (@wiker2023; @rapuano2022; @beck2023; @nweze2023; @mewton2023).
+The appeal of this approach is evidenced by several recent studies that have used data from the ABCD Study to explore bivariate associations between brain development and changes in several mental and physical health indicators @wiker2023; @rapuano2022; @beck2023; @nweze2023; @mewton2023.
 
 Latent State-Trait Models (LSTM) offer another variable-centered approach to longitudinal analysis that also allows for the estimation of patterns of change over time.
-Unlike LGCM, which conceptualizes change as a function of time, and LCSM, which views change through sequential measurements, this approach disaggregates observed behaviors into distinct stable (trait) and occasion-specific (state) components (@kenny2001; @steyer2015).
-Based on LST theory (@steyer1999; @steyer1992), these models hold that scores on a repeated measures outcome can be partitioned into an enduring latent trait variable that reflects between-individual differences, and a transient latent state residual that represents situational influences (@stadtbaeumer2022; @geiser2012).
+Unlike LGCM, which conceptualizes change as a function of time, and LCSM, which views change through sequential measurements, this approach disaggregates observed behaviors into distinct stable (trait) and occasion-specific (state) components @kenny2001; @steyer2015.
+Based on LST theory @steyer1999; @steyer1992, these models hold that scores on a repeated measures outcome can be partitioned into an enduring latent trait variable that reflects between-individual differences, and a transient latent state residual that represents situational influences @stadtbaeumer2022; @geiser2012.
 Beyond parsing out these key variance components, LSTM can be extended in many ways, such as by incorporating autoregressive effects to capture relative stability and the influence of past states on future responses (i.e.
@@
 Despite a range of potential model specifications for longitudinal mixture modeling, person-centered approaches tend to use parameterizations that default to settings found in popular software packages (e.g., Mplus).
-It has recently been demonstrated (see @mcneish2021) that the use of such specifications tends to identify the so-called “cat’s cradle” solution (see @sher2011) that consists of “…(a) a consistently ‘low’ group, (b) an ‘increase’ group, (c) a ‘decrease’ group, and (d) a consistently ‘high’ group” (@sher2011, p.
-322).
-Indeed, @xiang2022 describe their four-group solution as follows: “Of all participants, 536 (10.80%) were classified as increasing, 269 (5.42%) as persistently high, 433 (8.73%) as decreasing, and 3724 (75.05%) as persistently low” (@xiang2022, p.
-162).
+It has recently been demonstrated (see @mcneish2021) that the use of such specifications tends to identify the so-called “cat’s cradle” solution (see @sher2011) that consists of “…(a) a consistently ‘low’ group, (b) an ‘increase’ group, (c) a ‘decrease’ group, and (d) a consistently ‘high’ group” @sher2011, p.
+322.
+Indeed, @xiang2022 describe their four-group solution as follows: “Of all participants, 536 (10.80%) were classified as increasing, 269 (5.42%) as persistently high, 433 (8.73%) as decreasing, and 3724 (75.05%) as persistently low” @xiang2022, p.
+162.
 Although @sher2011 cautioned that groups from these trajectory-based approaches should not be over-reified, this practice also remains common (e.g., @hawes2016; @hawes2018).
@@
 
-Hybrid approaches, such as growth mixture (jung2008; @muthen2000) and factor-mixture (@lubke2005; @lubke2007) modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
+Hybrid approaches, such as growth mixture (jung2008; @muthen2000) and factor-mixture @lubke2005; @lubke2007 modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
 This combination provides a more comprehensive understanding of longitudinal data by capturing both within- and between-person variability.
@@
 
-AFNI (https://afni.nimh.nih.gov/), a well-established tool in neuroimaging, integrates 3dLMEr (@chen2013; https://afni.nimh.nih.gov/pub/dist/doc/program_help/3dLMEr.html) which adopts an LMM-based approach by providing access to the advanced capabilities of R’s lme4’s lmer function.
-For surface-based data, Freesurfer (https://surfer.nmr.mgh.harvard.edu/) provides the linear mixed effects (LME) package for modelling longitudinal data (@bernal2013statistical; https://surfer.nmr.mgh.harvard.edu/fswiki/LinearMixedEffectsModels).
-This tool is able to apply spatial regularization of LMM parameters with surface-based ROIs to improve stability (@bernal2013spatiotemporal).
+AFNI (https://afni.nimh.nih.gov/), a well-established tool in neuroimaging, integrates 3dLMEr @chen2013; https://afni.nimh.nih.gov/pub/dist/doc/program_help/3dLMEr.html which adopts an LMM-based approach by providing access to the advanced capabilities of R’s lme4’s lmer function.
+For surface-based data, Freesurfer (https://surfer.nmr.mgh.harvard.edu/) provides the linear mixed effects (LME) package for modelling longitudinal data @bernal2013statistical; https://surfer.nmr.mgh.harvard.edu/fswiki/LinearMixedEffectsModels.
+This tool is able to apply spatial regularization of LMM parameters with surface-based ROIs to improve stability @bernal2013spatiotemporal.
 
 For imaging data, LMM’s present a significant computational challenge, not only because they require iterative optimization but also because the computations cannot be vectorized as efficiently compared to ordinary least squares.
-To overcome this challenge, The Big Linear Mixed Models (BLMM, https://github.com/NISOx-BDI/BLMM) software addresses this by using Python’s broadcasting operations to estimate LMM’s as efficiently as possible (@maullin2022; @maullin2021).
+To overcome this challenge, The Big Linear Mixed Models (BLMM, https://github.com/NISOx-BDI/BLMM) software addresses this by using Python’s broadcasting operations to estimate LMM’s as efficiently as possible @maullin2022; @maullin2021.
 BLMM further separates the computation of sufficient statistics and parameter estimation, allowing sensitive image data to remain private if needed.
@@
 A different yet efficient approach is used with Fast and Efficient Mixed-effects Analysis (FEMA, https://github.com/cmig-research-group/cmig_tools), which uses a non-iterative regression estimator of the LMM variance components plus variance parameter quantization.
-This allows vectorization within groups of voxels that share the same variance parameters (@parekh2021).
+This allows vectorization within groups of voxels that share the same variance parameters @parekh2021.
 While this method uses different approximations, the authors have shown it provides results that closely match a traditional LMM implementation.
@@
 A crucial component of this method is the use of a robust “sandwich estimator” for standard errors, which accounts for any dependence not captured by the working covariance.
-The Sandwich Estimator (SwE) is a toolbox for SPM (https://www.nisox.org/Software/SwE, including CIFTI support) and FSL (https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/Swe) that provides marginal model inference using an independence working covariance matrix (@guillaume2014).
+The Sandwich Estimator (SwE) is a toolbox for SPM (https://www.nisox.org/Software/SwE, including CIFTI support) and FSL (https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/Swe) that provides marginal model inference using an independence working covariance matrix @guillaume2014.
 

### 2024-03-19 examples/examples.qmd
update examples
updte

@@
 			<!-- <a class="postcard__img_link" href="#">
-				<img class="postcard__img" src="https://picsum.photos/801/800" alt="Image Title" />	
+				<img class="postcard__img" src="https://picsum.photos/500/653" alt="Image Title" />	
 			</a> -->
@@
 				<ul class="postcard__tagbox">
-				  <li class="tag__item"><i class="fas fa-tag mr-2"></i></li>
-					<li class="tag__item"><i class="fas fa-clock mr-2"></i></li>
+				  <li class="tag__item play blue">
+						<a href="_blank"><i class="fas fa-play
+						mr-2"></i>Twin Designs</a>
+						</li>
+					<li class="tag__item play blue">
+						<a href="_blank"><i class="fas fa-play
+						mr-2"></i>Zero-Inflation Models</a>
+						</li>
+					<li class="tag__item play blue">
+						<a href="_blank"><i class="fas fa-play
+						mr-2"></i>Longitudinal Invariance</a>
+						</li>
 				</ul>
@@
 ```
-
-```{=html}
-<div class="d-flex flex-column justify-content-center w-100 h-100">
-
-	<div class="d-flex flex-column justify-content-center align-items-center">
-		<h1 class="fw-light text-white m-0">Pure CSS Gradient Background Animation</h1>
-		<div class="btn-group my-5">
-			<a href="https://codepen-api-export-production.s3.us-west-2.amazonaws.com/zip/PEN/pyBNzX/1578778289271/pure-css-gradient-background-animation.zip" class="btn btn-outline-light" aria-current="page"><i class="fas fa-file-download me-2"></i> SOURCE CODE</a>
-			<a href="https://codepen.io/P1N2O/full/pyBNzX" class="btn btn-outline-light">FULL SCREEN <i class="fas fa-expand ms-2"></i></a>
-		</div>
-		<a href="https://manuel.pinto.dev" class="text-decoration-none">
-			<h5 class="fw-light text-white m-0">— Pen by Manuel Pinto —</h5>
-		</a>
-	</div>
-</div>
-</div>
-```

### 2024-03-21 manuscript/manuscript.qmd
update toolbar

@@
 At its heart, the study is designed to chart the course of human development across multiple interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative outcomes.
-Central to achieving these goals is the commitment of the ABCD Study© and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods by espousing practices that increase access, integrity, and reproducibility of scientific research.
+Central to achieving these goals is the commitment of the ABCD Study and its NIH funders to an open science framework, intended to facilitate sharing of data and analytical methods by espousing practices that increase access, integrity, and reproducibility of scientific research.
 In this context, the ABCD Study is a collaboration with the broader research community.
@@
 We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.0 data with accompanying R scripts.
-Worked examples are available in Quarto files, accessible in the project's [GitHub repository]( https://github.com/OpenDevSci/LongDev-ABCD/tree/main/examples)
+Worked examples are available in [Quarto](https://quarto.org/) files, accessible in the project's [GitHub repository]( https://github.com/OpenDevSci/LongDev-ABCD/tree/main/examples).
 
@@
 
-• Part I. Longitudinal Research
+• Part I. Introduction
 
--   Fundamental concepts
+-   The ABCD Study&reg;
 
-• Part II. Longitudinal Data
+• Part II. Developmental Research
 
--   Method and analysis
+-   Fundamental Concepts
 
@@
 # Developmental Research
-
-\label{sec:headings}
 
@@
 # Longitudinal Data
-
-\label{sec:headings}
 
@@
 
-\label{sec:headings}
-
 ### Types of longitudinal panel models
@@
 Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
+
 Common examples include grouping by linear vs nonlinear models @collins2006, the number of measurement occasions @king2018, and statistical equivalency (e.g., change scores vs.
@@
 
-![Figure 3 Statistical modeling of brain imaging data](img/Picture1.png)
-![Figure 4 Statistical modeling of brain imaging data](img/Picture2.png)
+![Figure 3 Statistical modeling of brain imaging data](img/Picture1.png){width=100% #fig-StatBrainModel-1 fig-alt="Statistical modeling of brain imaging data" fig-align="left"}
+![Figure 4 Statistical modeling of brain imaging data](img/Picture2.png){width=100% #fig-StatBrainModel-2 fig-alt="Statistical modeling of brain imaging data" fig-align="left"}
 
 # Discussion
-\label{sec:headings}
 

### 2024-03-21 index.qmd
update toolbar
toolbar update

@@
 ---
-#title: "Longitudinal Data Analysis with the ABCD Study"
+
 format:
@@
 Your feedback and contributions are invaluable to us—join us in enhancing these resources.
-For more details on contributing, visit our [GitHub repository](https://github.com/OpenDevSci/LongDev-ABCD) and our [Contribution Guide](https://github.com/OpenDevSci/LongDev-ABCD/blob/main/docs/Contributing.md).
+For more details on contributing, visit our [GitHub Repository](https://github.com/OpenDevSci/LongDev-ABCD) and our [Contribution Guide](https://github.com/OpenDevSci/LongDev-ABCD/blob/main/docs/Contributing.md).
 :::

### 2024-03-25 index.qmd
update header
updaet
update manuscript

@@
 ---
-
-format:
-  html:
-    css: styles.css
-
 ---
@@
 
-This ongoing project, guided by the Biostatistics Working Group of the [Adolescent Brain Cognitive Development (ABCD) Study](https://abcdstudy.org/), aims to democratize access to longitudinal data analysis through practical examples and tutorials.
+This ongoing project, guided by the Biostatistics Working Group of the [Adolescent Brain Cognitive Development (ABCD) Study](https://abcdstudy.org/), aims to provide open access to longitudinal data analysis through practical examples and tutorials.
 

### 2024-03-25 manuscript/manuscript.qmd
update manuscript

@@
 In this paper, we describe models and methods for longitudinal analysis of ABCD Study data that can address these fundamental scientific aims, including: 1) characterization of the genetic and environmental factors associated with variation in developmental trajectories; 2) assessment of how the level and timing of exposures may impact subsequent neurodevelopment; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships.
-We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.0 data with accompanying R scripts.
-Worked examples are available in [Quarto](https://quarto.org/) files, accessible in the project's [GitHub repository]( https://github.com/OpenDevSci/LongDev-ABCD/tree/main/examples).
+We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.1 data with accompanying R scripts.
+Worked examples are available in [Quarto](https://quarto.org/) files, accessible in the project's [GitHub repository](https://github.com/OpenDevSci/LongDev-ABCD).
 
@@
 
-The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged between 9-10 years at baseline, each with a parent/guardian.
-The study sample was recruited from household populations in defined catchment areas for each of the 21 (originally 22) study sites across the United States.
+The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged approximately 9-10 years at baseline, each with a parent/guardian.
+The study sample was recruited from households in defined catchment areas for each of the 21 (originally 22) study sites across the United States.
 Information regarding funding agencies, recruitment sites, investigators, and project organizations can be obtained at https://abcdstudy.org.
@@
 Although conceptually distinct from vulnerable periods, both types of developmental considerations highlight the importance of viewing behavior in the context of development and attempting to determine how various developmental pathways unfold.
-Longitudinal data are crucial in this context to assess individual levels of development prior to and following onset of experiences or other environmental factors; e.g., the ABCD Study collected data starting at ages 9-10 and hence before the onset of substance use for the vast majority of participants.
+Longitudinal data are crucial in this context to assess individual levels of development prior to and following onset of experiences or other environmental factors (e.g., the ABCD Study collected data starting at ages 9-10 and hence before the onset of substance use for the vast majority of participants).
 
@@
 Some longitudinal studies, e.g., accelerated longitudinal designs (ALDs; @thompson2011) are especially well suited for discovering these effects and modeling them.
-While ABCD is not an ALD, the variability in age (and grade in school) at the time of baseline recruitment (9 to 11 years) allows some measures, collected every year, to be conceptualized as an ALD (e.g., substance use; prosocial behavior; family conflict; screen time).
+While ABCD is not an ALD, the variability in age (and grade in school) at the time of baseline recruitment (approximately 9-10 years old) allows some measures, collected every year, to be conceptualized as an ALD (e.g., substance use; prosocial behavior; family conflict; screen time).
 It is also possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments.
@@
 A classic example (from @bryk1992; @singer1998) involves math achievement measured among students nested within schools.
-In a basic, intercept-only model with no covariates, there would be one fixed effect (the grand mean, or intercept, of math achievement), one school random effect (representing variation in the intercept between schools) and the within-school student residuals (variation left over after accounting for fixed and random effects).
+In a basic, intercept-only model with no covariates (i.e., an unconditional growth model), there would be one fixed effect (the grand mean, or intercept, of math achievement), one school random effect (representing variation in the intercept between schools) and the within-school student residuals (variation left over after accounting for fixed and random effects).
 In this framework, each student’s score would be the sum of the fixed effect (the grand mean), the school random effect and the student’s within-school residual.
@@
 Violations of MAR impact both traditional and more modern data estimation procedures, though as noted by Graham, violations of MAR tend to have a greater effect on older methods.
-Graham thus suggests that imputing missing data is a better approach compared to listwise deletion in most circumstances, regardless of the model of missingness (i.e., MCAR, MAR, MNAR; see @graham2009).
+Graham thus suggests that imputing missing data is a better approach compared to listwise deletion in most circumstances, regardless of the model of missingness (i.e., MCAR, MAR, MNAR; see @graham2009; but also see @twisk2013).
 The ABCD Biostatistics Workgroup is currently implementing several missing data approaches which are being implemented and compared to each other (and listwise deletion) in the 5.0 data release, including, propensity score weighting, and multiple (multilevel) imputation.
@@
 Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change @bauer2019 Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
-This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events.
+This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events).
 While much has been written about “time-to-event” data @hosmer2008; @rizopoulos2012, including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study @brislin2023, our emphasis will be given to the modeling of “repeated measures” data.
@@
 
-![LDA Table](img/Tabletemp.png){width=100% #fig-LDATable fig-alt="LDA Table" fig-align="left"}
-
 #### Longitudinal Analysis of Neuroimaging Data

### 2024-03-26 index.qmd
update author list

@@
 - **Open Source Collaboration**: Engage with our community to contribute, learn, and share insights on longitudinal data analysis.
-- **Educational Resource**: Gain comprehensive knowledge in using GitHub, R, RMarkdown/Quarto, and various longitudinal analysis methods with our guidance.
+- **Educational Resource**: Gain comprehensive knowledge in using GitHub, R, RMarkdown/Quarto, and various longitudinal analysis methods.
 
@@
 - **[Documentation](https://github.com/orgs/OpenDevSci/projects/13)**: Get detailed insights into our project's progress, tasks, and timelines.
-- **[Community Discussion Forum](https://github.com/orgs/OpenDevSci/discussions)**: Connect with fellow researchers and enthusiasts in our dedicated forum.
+- **[Community Discussion Forum](https://github.com/orgs/OpenDevSci/discussions)**: Connect with fellow researchers and enthusiasts .
 
@@
 Our collective effort focuses on fostering an open source community where knowledge and skills in research methodologies can flourish.
-
-**Together, let's advance the field of data analysis.**

### 2024-03-26 manuscript/manuscript.qmd
update author list

@@
 author:
-  - name: Samuel Hawes
+- name: Samuel W. Hawes
   email: shawes@fiu.edu
-    title: Associate Research Professor
   affiliations:
-      #- id: ds9
     name: Florida International University
     department: Center for Children & Families
-        address: 11200 SW 8th St, AHC-IV
     city: Miami
     region: FL
-        country: USA
-        postal-code: 33199
-    attributes:
-      corresponding: true
+    country: United States
 - name: Andrew K. Littlefield
@@
   affiliations:
-      #- id: ds9
     name: Texas Tech University
@@
     region: TX
-        country: USA
+    country: United States
 - name: Daniel A. Lopez
-    email: Daniel_Lopez2@URMC.Rochester.edu
+  email: lopdanie@ohsu.edu
   affiliations:
-      #- id: terok
-        name: University of Rochester Medical Center
-        department: Department of Public Health
-        city: Rochester
-        region: NY
-        country: USA
-      #- ref: ds9
+    name: Oregon Health & Science University
+    department: Department of Psychiatry
+    city: Portland
+    region: OR
+    country: United States
 - name: Kenneth J. Sher
@@
   affiliations:
-      #- id: ds9
     name: University of Missouri
@@
     region: MO
-        country: USA
+    country: United States
 - name: Erin L. Thompson
@@
   affiliations:
-      #- id: terok
     name: Florida International University
-        department: Center for Children and Families
+    department: Center for Children & Families
     city: Miami
     region: FL
-        country: USA
+    country: United States
 - name: Raul Gonzalez
@@
   affiliations:
-      #- id: ds9
     name: Florida International University
-        department: Center for Children and Families
+    department: Center for Children & Families
     city: Miami
     region: FL
-        country: USA
-  - name: Additional Co-authors
-    email: xxxxx@fiu.edu
+    country: United States
+- name: Laika Aguinaldo
+  email: laaguinaldo@health.ucsd.edu
   affiliations:
-      #- id: terok
-        name: xxxxxx University
-        department: xxxxxx Dept
-        city: xxxx
-        region: xxxxx
-        country: xxxxx
+    name: University of California San Diego
+    department: Department of Psychiatry
+    city: San Diego
+    region: CA
+    country: United States
+- name: Ashley R. Adams
+  email: asadams@fiu.edu
+  affiliations:
+    name: Florida International University
+    department: Center for Children & Families
+    city: Miami
+    region: FL
+    country: United States
+- name: Mohammadreza Bayat
+  email: mbayat@fiu.edu
+  affiliations:
+    name: Florida International University
+    department: Cognitive Neuorscience
+    city: Miami
+    region: FL
+    country: United States
+- name: Amy L. Byrd
+  email: byrdal@upmc.edu
+  affiliations:
+    name: University of Pittsburgh
+    department: Department of Psychiatry
+    city: Pittsburgh
+    region: PA
+    country: United States
+- name: Luis FS Castro-de-Araujo
+  email: Luis.Araujo@vcuhealth.org
+  affiliations:
+    name: Virginia Commonwealth University
+    department: Virginia Institute for Psychiatric and Behavioral Genetics
+    city: Richmond
+    region: VA
+    country: United States
+- name: Anthony Dick
+  email: adick@fiu.edu
+  affiliations:
+    name: Florida International University
+    department: Cognitive Neuorscience
+    city: Miami
+    region: FL
+    country: United States
+- name: Steven F. Heeringa
+  email: sheering@umich.edu
+  affiliations:
+    name: University of Michigan
+    department: Institute for Social Research
+    city: Ann Arbor
+    region: MI
+    country: United States
+- name: Christine M. Kaiver
+  email: ckaiv001@fiu.edu
+  affiliations:
+    name: Florida International University
+    department: Center for Children & Families
+    city: Miami
+    region: FL
+    country: United States
+- name: Sarah M. Lehman
+  email: slehman@fiu.edu
+  affiliations:
+    name: Florida International University
+    department: Center for Children & Families
+    city: Miami
+    region: FL
+    country: United States
+- name: Lin Li
+  email: lil033@ucsd.edu
+  affiliations:
+    name: University of California San Diego
+    department: Department of Radiology
+    city: San Diego
+    region: CA
+    country: United States
+- name: Janosch Linkersdörfer
+  email: jlinkersdoerfer@ucsd.edu
+  affiliations:
+    name: University of California San Diego
+    department: Center for Human Development
+    city: San Diego
+    region: CA
+    country: United States
+- name: Thomas J. Maullin-Sapey
+  email: Thomas.Maullin-Sapey@bdi.ox.ac.uk
+  affiliations:
+    name: University of Oxford
+    department: Oxford Big Data Institute
+    city: Oxford
+    region: UK
+    country: United Kingdom
+- name: Michael C. Neale
+  email: Michael.Neale@vcuhealth.org
+  affiliations:
+    name: Virginia Commonwealth University
+    department: Virginia Institute for Psychiatric and Behavioral Genetics
+    city: Richmond
+    region: VA
+    country: United States
+- name: Thomas E. Nichols
+  email: thomas.nichols@bdi.ox.ac.uk
+  affiliations:
+    name: University of Oxford
+    department: Oxford Big Data Institute
+    city: Oxford
+    region: UK
+    country: United Kingdom
+- name: Samantha Perlstein
+  email: sperl@sas.upenn.edu
+  affiliations:
+    name: University of Pennsylvania
+    department: Department of Psychology
+    city: Philadelphia
+    region: PA
+    country: United States
+- name: Susan F. Tapert
+  email: stapert@ucsd.edu
+  affiliations:
+    name: University of California San Diego
+    department: Department of Psychiatry
+    city: San Diego
+    region: CA
+    country: United States
+- name: Colin E. Vize
+  email: CEV18@pitt.edu
+  affiliations:
+    name: University of Pittsburgh
+    department: Department of Psychology
+    city: Pittsburgh
+    region: PA
+    country: United States
+- name: Margot Wagner
+  email: mwagner@ucsd.edu
+  affiliations:
+    name: University of California San Diego
+    department: The Institute for Neural Computation
+    city: San Diego
+    region: CA
+    country: United States
+- name: Rebecca Waller
+  email: rwaller@sas.upenn.edu
+  affiliations:
+    name: University of Pennsylvania
+    department: Department of Psychology
+    city: Philadelphia
+    region: PA
+    country: United States
 - name: Wesley K. Thompson
@@
   affiliations:
-      #- id: terok
     name: Laureate Institute for Brain Research
@@
     region: OK
-        country: USA
+    country: United States
 

### 2024-03-26 manuscript/manuscript.qmd
added acknowledgement

@@
 By familiarizing themselves with the various types of longitudinal models, researchers can more effectively navigate the complexities of longitudinal data and contribute valuable insights into the developmental processes and individual differences that shape human experience.
+
+<br>
+
+# Acknowledgements
+Data used in the preparation of this article were obtained from the Adolescent Brain Cognitive DevelopmentSM (ABCD) Study (https://abcdstudy.org), held in the NIMH Data Archive (NDA).
+This is a multisite, longitudinal study designed to recruit more than 10,000 children age 9-10 and follow them over 10 years into early adulthood.
+The ABCD Study® is supported by the National Institutes of Health and additional federal partners under award numbers U01DA041048, U01DA050989, U01DA051016, U01DA041022, U01DA051018, U01DA051037, U01DA050987, U01DA041174, U01DA041106, U01DA041117, U01DA041028, U01DA041134, U01DA050988, U01DA051039, U01DA041156, U01DA041025, U01DA041120, U01DA051038, U01DA041148, U01DA041093, U01DA041089, U24DA041123, U24DA041147.
+A full list of supporters is available at https://abcdstudy.org/federal-partners.html.
+A listing of participating sites and a complete listing of the study investigators can be found at https://abcdstudy.org/consortium_members/.
+ABCD consortium investigators designed and implemented the study and/or provided data but did not necessarily participate in the analysis or writing of this report.
+This manuscript reflects the views of the authors and may not reflect the opinions or views of the NIH or ABCD consortium investigators.
+
+The ABCD data repository grows and changes over time.
+The ABCD data used in this report came from DOI: 10.15154/z563-zd24.
+DOIs can be found at the following [link](https://nda.nih.gov/study.html?id=2313).

### 2024-03-26 manuscript/manuscript.qmd
manuscript update

@@
 title: |
-  Longitudinal Analysis Manuscript: Working Draft
-subtitle: Longitudinal Analysis using data from the ABCD© Study
+  Longitudinal Analysis of the ABCD&reg; Study
 short-title: Longitudinal Analysis
-code-repo: "Access the code, data, and analysis at <http://swhawes.github.io/test/>"
+code-repo: "Access the code, data, and analysis at <https://github.com/OpenDevSci/LongDev-ABCD>"
 number-sections: true
@@
 - name: Samuel W. Hawes
-  email: shawes@fiu.edu
+  #email: shawes@fiu.edu
   affiliations:
     name: Florida International University
-    department: Center for Children & Families
-    city: Miami
-    region: FL
-    country: United States
+    #department: Center for Children & Families
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Andrew K. Littlefield
-  email: andrew.littlefield@ttu.edu
+  #email: andrew.littlefield@ttu.edu
   affiliations:
     name: Texas Tech University
-    department: Psychological Sciences
-    city: Lubbock
-    region: TX
-    country: United States
+    #department: Psychological Sciences
+    #city: Lubbock
+    #region: TX
+    #country: United States
 - name: Daniel A. Lopez
-  email: lopdanie@ohsu.edu
+  #email: lopdanie@ohsu.edu
   affiliations:
     name: Oregon Health & Science University
-    department: Department of Psychiatry
-    city: Portland
-    region: OR
-    country: United States
+    #department: #department of Psychiatry
+    #city: Portland
+    #region: OR
+    #country: United States
 - name: Kenneth J. Sher
-  email: SherK@missouri.edu
+  #email: SherK@missouri.edu
   affiliations:
     name: University of Missouri
-    department: Psychological Sciences
-    city: Columbia
-    region: MO
-    country: United States
+    #department: Psychological Sciences
+    #city: Columbia
+    #region: MO
+    #country: United States
 - name: Erin L. Thompson
-  email: erthomps@fiu.edu
+  #email: erthomps@fiu.edu
   affiliations:
     name: Florida International University
-    department: Center for Children & Families
-    city: Miami
-    region: FL
-    country: United States
+    #department: Center for Children & Families
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Raul Gonzalez
-  email: gonzara@fiu.edu
+  #email: gonzara@fiu.edu
   affiliations:
     name: Florida International University
-    department: Center for Children & Families
-    city: Miami
-    region: FL
-    country: United States
+    #department: Center for Children & Families
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Laika Aguinaldo
-  email: laaguinaldo@health.ucsd.edu
+  #email: laaguinaldo@health.ucsd.edu
   affiliations:
     name: University of California San Diego
-    department: Department of Psychiatry
-    city: San Diego
-    region: CA
-    country: United States
+    #department: #department of Psychiatry
+    #city: San Diego
+    #region: CA
+    #country: United States
 - name: Ashley R. Adams
-  email: asadams@fiu.edu
+  #email: asadams@fiu.edu
   affiliations:
     name: Florida International University
-    department: Center for Children & Families
-    city: Miami
-    region: FL
-    country: United States
+    #department: Center for Children & Families
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Mohammadreza Bayat
-  email: mbayat@fiu.edu
+  #email: mbayat@fiu.edu
   affiliations:
     name: Florida International University
-    department: Cognitive Neuorscience
-    city: Miami
-    region: FL
-    country: United States
+    #department: Cognitive Neuorscience
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Amy L. Byrd
-  email: byrdal@upmc.edu
+  #email: byrdal@upmc.edu
   affiliations:
     name: University of Pittsburgh
-    department: Department of Psychiatry
-    city: Pittsburgh
-    region: PA
-    country: United States
+    #department: #department of Psychiatry
+    #city: Pittsburgh
+    #region: PA
+    #country: United States
 - name: Luis FS Castro-de-Araujo
-  email: Luis.Araujo@vcuhealth.org
+  #email: Luis.Araujo@vcuhealth.org
   affiliations:
     name: Virginia Commonwealth University
-    department: Virginia Institute for Psychiatric and Behavioral Genetics
-    city: Richmond
-    region: VA
-    country: United States
+    #department: Virginia Institute for Psychiatric and Behavioral Genetics
+    #city: Richmond
+    #region: VA
+    #country: United States
 - name: Anthony Dick
-  email: adick@fiu.edu
+  #email: adick@fiu.edu
   affiliations:
     name: Florida International University
-    department: Cognitive Neuorscience
-    city: Miami
-    region: FL
-    country: United States
+    #department: Cognitive Neuorscience
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Steven F. Heeringa
-  email: sheering@umich.edu
+  #email: sheering@umich.edu
   affiliations:
     name: University of Michigan
-    department: Institute for Social Research
-    city: Ann Arbor
-    region: MI
-    country: United States
+    #department: Institute for Social Research
+    #city: Ann Arbor
+    #region: MI
+    #country: United States
 - name: Christine M. Kaiver
-  email: ckaiv001@fiu.edu
+  #email: ckaiv001@fiu.edu
   affiliations:
     name: Florida International University
-    department: Center for Children & Families
-    city: Miami
-    region: FL
-    country: United States
+    #department: Center for Children & Families
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Sarah M. Lehman
-  email: slehman@fiu.edu
+  #email: slehman@fiu.edu
   affiliations:
     name: Florida International University
-    department: Center for Children & Families
-    city: Miami
-    region: FL
-    country: United States
+    #department: Center for Children & Families
+    #city: Miami
+    #region: FL
+    #country: United States
 - name: Lin Li
-  email: lil033@ucsd.edu
+  #email: lil033@ucsd.edu
   affiliations:
     name: University of California San Diego
-    department: Department of Radiology
-    city: San Diego
-    region: CA
-    country: United States
+    #department: #department of Radiology
+    #city: San Diego
+    #region: CA
+    #country: United States
 - name: Janosch Linkersdörfer
-  email: jlinkersdoerfer@ucsd.edu
+  #email: jlinkersdoerfer@ucsd.edu
   affiliations:
     name: University of California San Diego
-    department: Center for Human Development
-    city: San Diego
-    region: CA
-    country: United States
+    #department: Center for Human Development
+    #city: San Diego
+    #region: CA
+    #country: United States
 - name: Thomas J. Maullin-Sapey
-  email: Thomas.Maullin-Sapey@bdi.ox.ac.uk
+  #email: Thomas.Maullin-Sapey@bdi.ox.ac.uk
   affiliations:
     name: University of Oxford
-    department: Oxford Big Data Institute
-    city: Oxford
-    region: UK
-    country: United Kingdom
+    #department: Oxford Big Data Institute
+    #city: Oxford
+    #region: UK
+    #country: United Kingdom
 - name: Michael C. Neale
-  email: Michael.Neale@vcuhealth.org
+  #email: Michael.Neale@vcuhealth.org
   affiliations:
     name: Virginia Commonwealth University
-    department: Virginia Institute for Psychiatric and Behavioral Genetics
-    city: Richmond
-    region: VA
-    country: United States
+    #department: Virginia Institute for Psychiatric and Behavioral Genetics
+    #city: Richmond
+    #region: VA
+    #country: United States
 - name: Thomas E. Nichols
-  email: thomas.nichols@bdi.ox.ac.uk
+  #email: thomas.nichols@bdi.ox.ac.uk
   affiliations:
     name: University of Oxford
-    department: Oxford Big Data Institute
-    city: Oxford
-    region: UK
-    country: United Kingdom
+    #department: Oxford Big Data Institute
+    #city: Oxford
+    #region: UK
+    #country: United Kingdom
 - name: Samantha Perlstein
-  email: sperl@sas.upenn.edu
+  #email: sperl@sas.upenn.edu
   affiliations:
     name: University of Pennsylvania
-    department: Department of Psychology
-    city: Philadelphia
-    region: PA
-    country: United States
+    #department: #department of Psychology
+    #city: Philadelphia
+    #region: PA
+    #country: United States
 - name: Susan F. Tapert
-  email: stapert@ucsd.edu
+  #email: stapert@ucsd.edu
   affiliations:
     name: University of California San Diego
-    department: Department of Psychiatry
-    city: San Diego
-    region: CA
-    country: United States
+    #department: #department of Psychiatry
+    #city: San Diego
+    #region: CA
+    #country: United States
 - name: Colin E. Vize
-  email: CEV18@pitt.edu
+  #email: CEV18@pitt.edu
   affiliations:
     name: University of Pittsburgh
-    department: Department of Psychology
-    city: Pittsburgh
-    region: PA
-    country: United States
+    #department: #department of Psychology
+    #city: Pittsburgh
+    #region: PA
+    #country: United States
 - name: Margot Wagner
-  email: mwagner@ucsd.edu
+  #email: mwagner@ucsd.edu
   affiliations:
     name: University of California San Diego
-    department: The Institute for Neural Computation
-    city: San Diego
-    region: CA
-    country: United States
+    #department: The Institute for Neural Computation
+    #city: San Diego
+    #region: CA
+    #country: United States
 - name: Rebecca Waller
-  email: rwaller@sas.upenn.edu
+  #email: rwaller@sas.upenn.edu
   affiliations:
     name: University of Pennsylvania
-    department: Department of Psychology
-    city: Philadelphia
-    region: PA
-    country: United States
+    #department: #department of Psychology
+    #city: Philadelphia
+    #region: PA
+    #country: United States
 - name: Wesley K. Thompson
-  email: wes.stat@gmail.com
+  #email: wes.stat@gmail.com
   affiliations:
     name: Laureate Institute for Brain Research
-    department: Center for Population Neuroscience and Genetics
-    city: Tulsa
-    region: OK
-    country: United States
+    #department: Center for Population Neuroscience and Genetics
+    #city: Tulsa
+    #region: OK
+    #country: United States
 

### 2024-03-26 manuscript/manuscript.qmd
update

@@
   #email: shawes@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Center for Children & Families
@@
   #email: andrew.littlefield@ttu.edu
-  affiliations:
-    name: Texas Tech University
+  #affiliations:
+    #name: Texas Tech University
     #department: Psychological Sciences
@@
   #email: lopdanie@ohsu.edu
-  affiliations:
-    name: Oregon Health & Science University
+  #affiliations:
+    #name: Oregon Health & Science University
     #department: #department of Psychiatry
@@
   #email: SherK@missouri.edu
-  affiliations:
-    name: University of Missouri
+  #affiliations:
+    #name: University of Missouri
     #department: Psychological Sciences
@@
   #email: erthomps@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Center for Children & Families
@@
   #email: gonzara@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Center for Children & Families
@@
   #email: laaguinaldo@health.ucsd.edu
-  affiliations:
-    name: University of California San Diego
+  #affiliations:
+    #name: University of California San Diego
     #department: #department of Psychiatry
@@
   #email: asadams@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Center for Children & Families
@@
   #email: mbayat@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Cognitive Neuorscience
@@
   #email: byrdal@upmc.edu
-  affiliations:
-    name: University of Pittsburgh
+  #affiliations:
+    #name: University of Pittsburgh
     #department: #department of Psychiatry
@@
   #email: Luis.Araujo@vcuhealth.org
-  affiliations:
-    name: Virginia Commonwealth University
+  #affiliations:
+    #name: Virginia Commonwealth University
     #department: Virginia Institute for Psychiatric and Behavioral Genetics
@@
   #email: adick@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Cognitive Neuorscience
@@
   #email: sheering@umich.edu
-  affiliations:
-    name: University of Michigan
+  #affiliations:
+    #name: University of Michigan
     #department: Institute for Social Research
@@
   #email: ckaiv001@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Center for Children & Families
@@
   #email: slehman@fiu.edu
-  affiliations:
-    name: Florida International University
+  #affiliations:
+    #name: Florida International University
     #department: Center for Children & Families
@@
   #email: lil033@ucsd.edu
-  affiliations:
-    name: University of California San Diego
+  #affiliations:
+    #name: University of California San Diego
     #department: #department of Radiology
@@
   #email: jlinkersdoerfer@ucsd.edu
-  affiliations:
-    name: University of California San Diego
+  #affiliations:
+    #name: University of California San Diego
     #department: Center for Human Development
@@
   #email: Thomas.Maullin-Sapey@bdi.ox.ac.uk
-  affiliations:
-    name: University of Oxford
+  #affiliations:
+    #name: University of Oxford
     #department: Oxford Big Data Institute
@@
   #email: Michael.Neale@vcuhealth.org
-  affiliations:
-    name: Virginia Commonwealth University
+  #affiliations:
+    #name: Virginia Commonwealth University
     #department: Virginia Institute for Psychiatric and Behavioral Genetics
@@
   #email: thomas.nichols@bdi.ox.ac.uk
-  affiliations:
-    name: University of Oxford
+  #affiliations:
+    #name: University of Oxford
     #department: Oxford Big Data Institute
@@
   #email: sperl@sas.upenn.edu
-  affiliations:
-    name: University of Pennsylvania
+  #affiliations:
+    #name: University of Pennsylvania
     #department: #department of Psychology
@@
   #email: stapert@ucsd.edu
-  affiliations:
-    name: University of California San Diego
+  #affiliations:
+    #name: University of California San Diego
     #department: #department of Psychiatry
@@
   #email: CEV18@pitt.edu
-  affiliations:
-    name: University of Pittsburgh
+  #affiliations:
+    #name: University of Pittsburgh
     #department: #department of Psychology
@@
   #email: mwagner@ucsd.edu
-  affiliations:
-    name: University of California San Diego
+  #affiliations:
+    #name: University of California San Diego
     #department: The Institute for Neural Computation
@@
   #email: rwaller@sas.upenn.edu
-  affiliations:
-    name: University of Pennsylvania
+  #affiliations:
+    #name: University of Pennsylvania
     #department: #department of Psychology
@@
   #email: wes.stat@gmail.com
-  affiliations:
-    name: Laureate Institute for Brain Research
+  #affiliations:
+    #name: Laureate Institute for Brain Research
     #department: Center for Population Neuroscience and Genetics
@@
 abstract: |
-  The Adolescent Brain Cognitive Development&reg; (ABCD) Study provides a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of youths, aged 9-10 at baseline and assessed annually for 10 years.
+  The Adolescent Brain Cognitive Development&reg; (ABCD) Study provides a unique opportunity to investigate developmental processes in a large, diverse cohort of youths, aged approximately 9-10 at baseline and assessed annually for 10 years.
 Given the size and complexity of the ABCD Study, researchers analyzing its data will encounter a myriad of methodological and analytical considerations.
-This review provides an examination of key concepts and techniques related to longitudinal analyses of the ABCD Study data.
-We discuss the importance of longitudinal data, focusing on the types of inferences that are possible when one assesses individuals across multiple time points, including: 1) characterization of the factors associated with variation in developmental trajectories; 2) assessment of how level and timing of exposures may impact subsequent development; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships.
-We emphasize the importance of selecting appropriate statistical models to address these research questions, e.g., accounting for correlation in repeated measurements, employing linear or non-linear models as indicated by the data, and using link functions that adequately model outcome distributions.
+This review provides an examination of key concepts and techniques related to longitudinal analyses of the ABCD Study data, including: 1) characterization of the factors associated with variation in developmental trajectories; 2) assessment of how level and timing of exposures may impact subsequent development; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships.
+We emphasize the importance of selecting appropriate statistical models to address these research questions.
 By presenting the advantages and potential challenges of longitudinal analyses in the ABCD Study, this review seeks to equip researchers with foundational knowledge and tools to make informed decisions as they navigate and effectively analyze and interpret the multi-dimensional longitudinal data currently available.
@@
 format:
+  docx: default
   ieee-typst: default

### 2024-05-01 Template.qmd
update unconditional examples

---
title: "Template"
author: "author name"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, output=T,
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

[Enter code to load example data]

```

## Overview
[Introductory text for the example]

## Preliminary Setup

### Load R libraries
```{r}

library(xxxxx)    # xxxxxxx
library(xxxxx)    # xxxxxxx

```

### Descriptives

```{r}
[Code for providing sample descriptives]
```

## Results

### [Name of First Set of Analyses]
[Concise description of the analysis approach]

```{r}

[Enter analysis code]

```

[Results write-up for analysis]

### Model Plots
[Concise description of the plot type]

```{r}
[Enter plot code]
```
[Concise summary of plot output]

## Wrapping Up

[Concise summary of the example]

### 2024-05-01 examples/3b_minimalist_Examples_LinearMixedModels.qmd
update unconditional examples

---
title: "Linear Mixed Models: Random Intercept and Slope"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(cbcl_scr_syn_external_r) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = cbcl_scr_syn_external_r
    )

df_long <- df_long %>%
  group_by(id) %>%
  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
  ungroup()

```

## Overview

The Linear mixed model (LMM) with a random intercept and slope extends the traditional fixed-effect linear regression by incorporating random effects for the intercept and slope parameters. This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.

This example will examine trajectories of height obtained across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to characterize stability and change in participant's height, while accounting for observations that are clustered within youth over time. The LMM with a random intercept and slope will simultaneously estimate an overall sample mean (fixed effect) for the height trajectory's intercept and slope values, as well as subject-specific (random) effects that vary randomly about the mean intercept and slope values.

## Preliminary Setup
### Load Packages
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
library(lme4) # Linear mixed-effects models
```

### Descriptives
```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(cbcl_scr_syn_external_r) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = cbcl_scr_syn_external_r
    )

df_long <- df_long %>%
  group_by(id) %>%
  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
  ungroup()

```

## Results 
### Compute LMM with Random Intercepts and Slopes
This code fits a random-intercept and random-slope LMM to predict 'Height' across 3 annual visits, accounting for individual-level variability (random effects) of intercept and slope values across participants. Model output is provided below, as well as a brief summary of results.

This model accounts for individual variability in height by including a random intercept and sllope for each subject (id). The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.

```{r}

model <- lmer(height ~ event + (1 + event | id), data = df_long, REML = T)

# Model with random intercepts only
model_ri_only <- lmer(height ~ event + (1 | id), data = df_long, REML = T)
model_ri_only

# Model with random slopes only, without intercepts
model_rs_only <- lmer(height ~ event + (0 + event | id), data = df_long, REML = T)
model_rs_only

## Output and summary statistics
summary(model)
confint(model, level = 0.95, method = "Wald")

```

These results show an increasing trajectory of height across the years, with an increase of approximately 2.36 inches in Year 1 and 5.02 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates. The random intercept and slope effects suggests some between-subject variability in particpant's height values at baseline and in rates of change over time. The model's residuals are mostly well-behaved, although the range suggests a few extreme values or outliers.

### Model Plots
The following set of plots are used to facilitate model diagnostics. The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects. The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity. The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.

```{r}

# 1. Extract the random effects
random_effects <- ranef(model)[[1]]

# 2. Convert to dataframe
random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)

# 3. hist_plot
hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, color = "black", fill = "lightblue") +
    labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
    theme_minimal()

print(hist_plot)

```

```{r}

# Extract fixed effects
fixed_effects <- fixef(model)

# Prepare data frame for plotting
df_long$predicted_height <- predict(model, re.form = NA) # Overall fixed effect predictions

# Calculate individual intercept adjustments
df_long$individual_intercept <- predict(model, re.form = ~ (1 | id)) - fixed_effects[1]

# Generate the plot
ggplot(df_long, aes(x = event, y = height, group = id)) +
    geom_line(aes(y = height, color = "Individual Trajectories"), alpha = 0.3) +
    geom_line(data = df_long, aes(x = event, y = predicted_height, group = 1, color = "Mean Trajectory"), linewidth = 1.5) +
    scale_color_manual(values = c("Individual Trajectories" = "grey", "Mean Trajectory" = "blue")) +
    labs(
        title = "Individual and Group-Mean Height Trajectories",
        x = "Event Name",
        y = "Height (cm)",
        color = "Trajectory Type"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")

```

This plot illustrates the change in participant's height across 3 annual measurement occassions. The individual raw data points are shown as red lines, representing the original height measurements, which demonstrate the variability among participants at each time point. These are overlaid with grey lines, which adjust each individual's trajectory unique growth pattern indicated by the fixed effects from the model. The blue line represents the sample mean trajectory, calculated from the model's fixed effects, showcasing the average growth trend across all participants.

The plot shows individual trajectories that vary around the model's predicted values due to random effects, indicating substantial inter-individual variability in baseline heights and rates of change in height over tim. On average, participants experienced an increase in height of approximately 2.36 inches from baseline to Year 1 and an additional 2.66 inches from Year 1 to Year 2. The random effects for the estimated intercept and slope suggest differences in starting heights and rates of change over time, reflecting the variability in growth patterns among the participants.

## Wrapping Up

The random-intercept and random-slope LMM analysis revealed an increasing pattern of change in children's height measured over 3 annual assessments. The mean intercept value showed an average initial height of 55.24 inches (sd = 3.03) with evidence of individual variability around this (random effect) estimate. An examination of the slope factor revealed year-to-year growth increments of approxmiately 2.5 inches, with evidence of individual variability around this estimate.

The LMM framework used in this example offered the ability to account for individual differences in initial height and changes in growth over time, demonstrating its utility in handling clustered or correlated data that arise from repeated measurements on the same subjects. More generally, this approach is often particularly useful in developmental studies as it allows for the inclusion of both fixed effects, which estimate common parameters across all individuals, and random effects, which capture unique variations among individuals. This enables a more precise understanding of both the group-level trends and individual-specific trajectories in longitudinal data.

### 2024-05-01 examples/_queue/_10_minimal_Examples_LatentGrowthCurveModels.qmd
update unconditional examples

---
title: "Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be rendered
#| output: FALSE # Output from this code block will not be rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview

Latent growth curve models (LGCM) are an analytical tool within the framework of structural equation modeling that enable the examination of change over time. This model type effectively separates the true trajectory of a variable from random measurement error, allowing for the estimation of an underlying growth process. The primary components of LGCM include the intercept, which represents the initial status, and the slope, which reflects the rate of change over time.

This example investigates a growth trajectory of participant's height, measured across 4 annual assessments in participants from the ABCD Study. Fitting the latent growth curve model will allow for an examination of the initial status (intercept) and rate of change (slope) of height from baseline through successive yearly follow-ups. Each participant's height trajectory is modeled to capture both the starting point and the growth pattern over time. The results are illustrated through growth plots that demonstrate the mean trajectory across the cohort, along with the variability in growth rates among individuals, providing insights into the developmental patterns of height over time.

## Preliminary Setup
### Load Packages
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
library(lavaan) # Structural Equation Modeling in R

```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results 
### Compute Latent Growth Curve Model
The code below is used to compute a growth curve model to investigate changes in height across four annual assessments. This statistical approach provides insights into how much individuals vary around the average trajectory and the consistency of their growth patterns across time. This model estimates an overall intercept, representing the initial status, and a slope that captures the rate of growth across time points.

```{r}

# Reshape data from long to wide format
df_wide <- df_long %>%
    pivot_wider(
        id_cols = c(id),
        names_from = event,
        values_from = height,
        names_prefix = "Height_"
    )

df_wide <- na.omit(df_wide)

model <- " i =~ 1*Height_Baseline + 1*Height_Year_1 + 1*Height_Year_2 + 1*Height_Year_3
           s =~ 0*Height_Baseline + 1*Height_Year_1 + 2*Height_Year_2 + 3*Height_Year_3

           # Intercept and slope variances
           i ~~ i
           s ~~ s

           # Residual variances for each observed variable
           Height_Baseline ~~ var_baseline*Height_Baseline
           Height_Year_1 ~~ var_year1*Height_Year_1
           Height_Year_2 ~~ var_year2*Height_Year_2
           Height_Year_3 ~~ var_year3*Height_Year_3
"

fit <- growth(model, data = df_wide)
summary(fit)

```

The results from the latent growth curve analysis show an increasing pattern of mean change in participant's height across assessments (slope = 2.491, p<.001). In addition, the model identified significant variability in both the intercept (estimate = 9.39, se=.33, p<.001), which represents initial height and the slope (estimate = .31, se=.04, p<.001), which measures growth over time. This indicates substantial differences in starting heights and growth rates among individuals. The covariance between the intercept and slope was not statistically significant (estimate = -.11, p = .18), suggesting initial height does not influence the rate of growth.

### Model Plots
```{r}

# Plotting the height data over time from the df_long dataframe
ggplot(df_long, aes(x = event, y = height, group = id)) +
    geom_line(alpha = 0.4, aes(color = "Actual Height"), size = 0.5) + # Lines to connect data points for each participant
    geom_point(alpha = 0.6, color = "blue") + # Points for actual heights
    scale_x_discrete(limits = c("Baseline", "Year_1", "Year_2", "Year_3")) + # Ensuring the order of events
    labs(
        title = "Actual Height Growth Across Timepoints",
        x = "Time (Years from Baseline)",
        y = "Height (inches)"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom") # Adjust legend position

```

This plot visualizes the height trajectories across four annual timepoints. Each line represents an individual participant's height trajectory showing overall trends while maintaining focus on individual variations. Blue points mark the actual height measurements at each timepoint, providing a clear view of the data distribution and growth patterns over time. This graph highlights both the general trend of increasing height and the individual differences in growth rates among participants.

## Wrapping Up
This analysis employed a latent growth curve model (LGCM) framework to examine participants' heights across multiple assessment points. Results revealed an increasing trend in average height changes over time, with a slope parameter of 2.49 (p<.001), indicating a steady growth across all assessments. Significant variability was observed both in the intercept, with an estimate of 9.39 (se = .33, p<.001), representing the initial height, and in the slope, with an estimate of .31 (se = .04, p<.001), which measures growth over time. This highlights notable individual differences in starting heights and growth rates. Furthermore, the lack of a statistically significant covariance between the intercept and slope (estimate = -.11, p = .18) suggests that the initial height does not influence subsequent growth rates.

Latent growth curve modeling provides a powerful tool for evaluating patterns of growth, allowing for the differentiation between average growth trajectories and individual variability. This method is able to parse out complex relationships, underscoring its utility in longitudinal data analysis, enhancing our interpretation of how individuals change over time.

### 2024-05-01 examples/_queue/_13b_Examples_LatentClassGrowthAnalsis.qmd
update unconditional examples

---
title: "Latent Class Growth Analysis"
author: "Biostatistics Working Group"
---

# LCGA: {.unnumbered}

## Overview

<p>

"A valuable tool for investigating heterogeneity in symptom patterns is Latent Growth Curve Modelling (GCM), a statistical method that analyses inter-individual variability in intra-individual patterns over time @curran2010. One specific type of GCM is Latent Class Growth Analysis (LCGA), which considers unobserved heterogeneity (different groups) over time within a larger population @jung2008, @nguena2020, @ram2009. In other words, LCGA can examine the growth and shape of the course of depressive symptoms over time and assess how individuals in the population group together based on their symptom patterns."

"The aim of the present study is to investigate the course of depressive symptoms over a period of xx years .... and to .... . To achieve this goal, LGCM will be performed on data from the xxxx ABCD cohort (add citation). The ABCD study was initiated in xxxxx. Based on previous literature @musliner2016, we hypothesize that three or four trajectories of depression can be identified. Given that xxxxx, we additionally hypothesize that individuals following more xxxx trajectories will be more/less likely to experience xxxx and exhibit reduced levels of xxxxx"

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse","tidySEM","rstatix","report","broom","gridExtra","easystats","gtsummary")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(tidySEM)      # Structural equation modeling in R
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page
library(easystats)    # Descriptive statistics and robust statistical methods
library(gtsummary)    # Presentation-ready data summary and analytic result tables
```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) #%>%
#   # Exclude cases from unused assessment waves
#   filter(!is.na(eventname))

```
:::

```{r}

# Long to wide conversion
df_long <- df %>%
  select(src_subject_id, eventname, anthroheightcalc) %>% na.omit()

# Creating df_wide
df_wide <- pivot_wider(
    data = df_long,
    names_from = eventname,
    values_from = anthroheightcalc,
    id_cols = src_subject_id
)

df_wide <- df_wide %>%
    rename(
        anthroheight.0 = baseline_year_1_arm_1,
        anthroheight.1 = `1_year_follow_up_y_arm_1`,
        anthroheight.2 = `2_year_follow_up_y_arm_1`,
        anthroheight.3 = `3_year_follow_up_y_arm_1`,
        anthroheight.4 = `4_year_follow_up_y_arm_1`
    )

## Subset the nonimaging data to include desired variables
selected_vars_wide <- c("anthroheight.0", "anthroheight.1", "anthroheight.2", "anthroheight.3", "anthroheight.4")
df_wide <- df_wide[, selected_vars_wide]
str(df_wide)

```

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```
:::

:::

We can examine these distributions visually as well:

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

# df_long Plots
ggplot(df_long, aes(x = anthroheightcalc)) +
    geom_density() +
    facet_wrap(~eventname) +
    theme_bw()

```

As this type of skew can result in convergence problems in LCGA, we compared several transformations to reduce skew: The square and cube root, log, inverse, and Box-Cox transformations.

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

df_scores <- df_long
# Store original range of cbcl_scr_dsm5_depress_t
rng_height <- range(df_scores$anthroheightcalc, na.rm = TRUE)
# Log-transform
df_scores$log <- scales::rescale(log(df_scores$anthroheightcalc), to = c(
    0,
    1
))
# Square root transform
df_scores$sqrt <- scales::rescale(sqrt(df_scores$anthroheightcalc), to = c(
    0,
    1
))
# Cube root transform
df_scores$qrt <- scales::rescale(df_scores$anthroheightcalc^0.33, to = c(
    0,
    1
))
# Reciprocal transform
df_scores$reciprocal <- scales::rescale(1 / df_scores$anthroheightcalc, to = c(
    0,
    1
))

# # Define function for Box-Cox transformation
# bc <- function(x, lambda) {
#     (((x^lambda) - 1) / lambda)
# }

# # Inverse Box-Cox transformation
# invbc <- function(x, lambda) {
#     ((x * lambda) + 1)^(1 / lambda)
# }
# # Box-Cox transform
# b <- MASS::boxcox(lm(df_scores$anthroheightcalc ~ 1), plotit = FALSE)
# lambda <- b$x[which.max(b$y)]
# df_scores$boxcox <- bc(df_scores$anthroheightcalc, lambda)
# # Store range of Box-Cox transformed data
# rng_bc <- range(df_scores$boxcox)
# df_scores$boxcox <- scales::rescale(df_scores$boxcox, to = c(
#     0,
#     1
# ))

# Rescale anthroheightcalc
df_scores$anthroheightcalc <- scales::rescale(df_scores$anthroheightcalc, to = c(0, 1))

```

We can plot these transformations:

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

# Make plot data
df_plot <- do.call(rbind, lapply(c(
    "anthroheightcalc", "log", "sqrt", "qrt" #,
    #"boxcox"
), function(n) {
    data.frame(df_scores[c("eventname", "src_subject_id")],
        Value = df_scores[[n]],
        Transformation = n
    )
}))
# Plot
ggplot(df_plot, aes(x = Value, colour = Transformation)) +
    geom_density() +
    facet_wrap(~eventname) +
    scale_y_sqrt() +
    xlab("Height (rescaled to 0-1)") +
    theme_bw()

```

The Box-Cox transformation (maybe/slightly?) reduced skew the most. Consequently, we proceeded with the Box-Cox transformed scores for analysis.

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

dat <- df_scores[, c("src_subject_id", "eventname", "log")]
dat <- reshape(dat,
    direction = "wide", v.names = "log", timevar = "eventname",
    idvar = "src_subject_id"
)

dat <- na.omit(dat)

names(dat) <- gsub("log.", "anthroheightcalc", names(dat))
```

:::

## Results

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue
The code fits an LCGA to examine the 'Height' variable across time points ('eventname'). It xxxxxx. The results of the model are then printed to provide a detailed summary of model parameters.

**STEP 1: Compute LCGA**
```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

#set.seed(27796)
dat[["src_subject_id"]] <- NULL
res_step <- mx_growth_mixture(
    model = "
  i =~ 1*anthroheightcalcbaseline_year_1_arm_1 + 1*anthroheightcalc1_year_follow_up_y_arm_1 + 1*anthroheightcalc2_year_follow_up_y_arm_1 +1*anthroheightcalc3_year_follow_up_y_arm_1 +1*anthroheightcalc4_year_follow_up_y_arm_1
  step =~ 0*anthroheightcalcbaseline_year_1_arm_1 + 1*anthroheightcalc1_year_follow_up_y_arm_1 + 1*anthroheightcalc2_year_follow_up_y_arm_1 +1*anthroheightcalc3_year_follow_up_y_arm_1 +1*anthroheightcalc4_year_follow_up_y_arm_1
  s =~ 0*anthroheightcalcbaseline_year_1_arm_1 + 0*anthroheightcalc1_year_follow_up_y_arm_1 + 1*anthroheightcalc2_year_follow_up_y_arm_1 +2*anthroheightcalc3_year_follow_up_y_arm_1 +3*anthroheightcalc4_year_follow_up_y_arm_1
  anthroheightcalcbaseline_year_1_arm_1 ~~ vanthroheightcalcbaseline_year_1_arm_1*anthroheightcalcbaseline_year_1_arm_1
  anthroheightcalc1_year_follow_up_y_arm_1 ~~ vanthroheightcalc1_year_follow_up_y_arm_1*anthroheightcalc1_year_follow_up_y_arm_1
  anthroheightcalc2_year_follow_up_y_arm_1 ~~ vanthroheightcalc2_year_follow_up_y_arm_1*anthroheightcalc2_year_follow_up_y_arm_1
  anthroheightcalc3_year_follow_up_y_arm_1 ~~ vanthroheightcalc3_year_follow_up_y_arm_1*anthroheightcalc3_year_follow_up_y_arm_1
  anthroheightcalc4_year_follow_up_y_arm_1 ~~ vanthroheightcalc4_year_follow_up_y_arm_1*anthroheightcalc4_year_follow_up_y_arm_1
  i ~~ 0*i
  step ~~ 0*step
  s ~~ 0*s
  i ~~ 0*s
  i ~~ 0*step
  s ~~ 0*step",
    classes = 1:3, data = dat
)
# Additional iterations because of convergence problems for
# model 1:
res_step[[1]] <- mxTryHardWideSearch(res_step[[1]], extraTries = 50)
```

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

# Get fit table fit
tab_fit <- table_fit(res_step)
# Select columns
tab_fit[, c("Name", "Classes", "LL", "Parameters", "AIC", "BIC", "saBIC", "Entropy", "prob_min", "n_min")]

# , "warning", "lmr_p")]

```

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

plot(tab_fit, statistics = c("AIC", "BIC", "saBIC"))
```

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

res_final <- mx_switch_labels(res_step[[3]],
    param = "M[1,7]",
    decreasing = FALSE
)
tab_res <- table_results(res_final, columns = NULL)
# Select rows and columns
tab_res <- tab_res[
    tab_res$Category %in% c("Means", "Variances"),
    c("Category", "lhs", "est", "se", "pval", "confint", "name")
]
tab_res
```

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

names(coef(res_final))
```
```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

wald_tests <- wald_test(res_final, "
                   class1.M[1,6] = class2.M[1,6] &
                   class1.M[1,6] = class3.M[1,6] ;
                   class1.M[1,7] = class2.M[1,7] &
                   class1.M[1,7] = class3.M[1,7] ;
                   class1.M[1,8] = class2.M[1,8] &
                   class1.M[1,8] = class3.M[1,8]")

# Rename the hypothesis
wald_tests$Hypothesis <- c("Mean i", "Mean step", "Mean slope")
knitr::kable(wald_tests, digits = 2, caption = "Wald tests")

```

```{r}
# Assuming rng_height and lambda are defined previously in your script
brks <- seq(0, 1, length.out = 5) # Breaks in the rescaled 0-1 range

# Convert breaks back to the original log scale
# Note: This assumes that the original transformation was simply log(x) scaled to 0-1
labs <- exp(scales::rescale(brks, from = c(0, 1), to = rng_height))

p <- plot_growth(res_step[[3]], rawdata = TRUE, alpha_range = c(0, 0.05))
p <- p + scale_y_continuous(
    breaks = brks, # Use the same breaks for simplicity
    labels = round(labs, 2) # Round the labels for readability
) + ylab("Height (rescaled from log)")
p

```

#### Covariates

Assessment of covariates using the three-step approach revealed that the trajectories did not differ in any of the demogra phical variables, including sex (ΔLL(5) = 6.3 , p = .27, pcorr = .37), age (ΔLL(8) = 8.4 , p = .40, pcorr = .50), rank Journal of Affective Disorders 354 (2024) 702--711 707(ΔLL(11) = 13.5 , p = .26, pcorr = .37), educational level (ΔLL(9) = 7.1 , p = .63, pcorr = .66), function (ΔLL(8) = 8.1 , p = .43, pcorr = .51). Significant disparities in xxxx were observed, with the xxxxx group exhibiting lower counts of xxxxx compared to the xxxxx group (ΔLL(2) = 23.0 , p \< .01, pcorr \< .01), xxxxx (ΔLL(2) = 325.7 , p \< .01, pcorr \< .01), and xxxxx group (ΔLL(2) = 9.5 , p \< .01, pcorr \< .05) (Supplementary Fig. Sxx). A detailed breakdown of percentages per xxxxx is provided in Supplementary Table Sxx

#### Depression and GDT

#### Depression and substance use

## Add additional analyses section

Note that the observed individual trajectories show very high variability within classes.

:::

### Model Plots

::: blue
The following set of plots are xxxxxxx.

:::

## Wrapping Up
<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

The lcga was conducted to xxxxxx.

:::
:::

---------------

# Discussion

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: TRUE

lorem::ipsum(5, sentences = c(3))
# lorem::ipsum(2, avg_words_per_sentence = 4)

```

### 2024-05-01 examples/_queue/_13b_minimal_Examples_LatentClassGrowthAnalsis.qmd
update unconditional examples

---
title: "Latent Class Growth Analysis"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be rendered
#| output: FALSE # Output from this code block will not be rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview

Latent Class Growth Analysis (LCGA) is a specialized form of growth modeling that identifies distinct subgroups within a population based on their trajectory patterns over time. Unlike traditional growth models that assume a single underlying growth trajectory for all individuals, LCGA allows for the identification of multiple latent classes, each characterized by its own unique growth path. This method is particularly useful for uncovering heterogeneous patterns of change that might be obscured in an overall population analysis.

This example focuses on exploring the developmental trajectories of height across several annual assessments in a cohort from the ABCD Study. Employing LCGA provides the ability to categorize participants into distinct groups based on their growth patterns in height from baseline to subsequent follow-ups. The analysis will determine the number of latent classes that best fit the data, describe the characteristic growth trajectory of each class, and interpret the developmental implications of these findings. Visualizations such as trajectory plots for each class will be used to illustrate the distinct growth patterns.

## Preliminary Setup
### Load Packages
```{r}

library(tidyverse) # Collection of R packages for data science
library(tidySEM) # Structural equation modeling in R
library(gridExtra) # Arrange multiple grid-based plots on a page
library(easystats)
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results 
### Compute Latent Class Growth Analysis
Latent Class Growth Analysis (LCGA) is utilized to identify distinct subgroups within the sample based on their growth trajectories of height over four time points. This approach models latent classes that capture varying patterns of growth, potentially reflecting different developmental processes or intervention outcomes.

The resulting latent class growth model helps delineate classes that show distinct growth patterns. Each class's trajectory can be characterized by its initial status (intercept), rate of growth (slope), and pattern changes over time (step and shape), providing a nuanced understanding of how height progresses differently across subgroups. This analysis is particularly valuable in understanding heterogeneity in developmental outcomes.

**STEP 1: Compute LCGA**
```{r}

df_wide <- df_long %>%
    pivot_wider(
        names_from = event,
        values_from = height,
        id_cols = id
    )

set.seed(27796)
df_wide[["id"]] <- NULL
res_step <- mx_growth_mixture(
    model = "
  i =~ 1*Baseline + 1*Year_1 + 1*Year_2 +1*Year_3
  step =~ 0*Baseline + 1*Year_1 + 1*Year_2 +1*Year_3
  s =~ 0*Baseline + 0*Year_1 + 1*Year_2 +2*Year_3
  Baseline ~~ vBaseline*Baseline
  Year_1 ~~ vYear_1*Year_1
  Year_2 ~~ vYear_2*Year_2
  Year_3 ~~ Year_3*Year_3
  i ~~ 0*i
  step ~~ 0*step
  s ~~ 0*s
  i ~~ 0*s
  i ~~ 0*step
  s ~~ 0*step",
    classes = 1:3, data = df_wide
)
# Include additional iterations if convergence problems encountered
# model 1:
# res_step[[1]] <- mxTryHardWideSearch(res_step[[1]], extraTries = 50)

# Get fit table fit
tab_fit <- table_fit(res_step)
# Select columns
tab_fit <- tab_fit %>%
    select(Name, Classes, LL, n, Parameters, AIC, BIC, saBIC, Entropy, prob_min, n_min)

print(tab_fit)

res_final <- mx_switch_labels(res_step[[3]],
    param = "M[1,7]",
    decreasing = FALSE
)
tab_res <- table_results(res_final, columns = NULL)
# Select rows and columns
tab_res <- tab_res[
    tab_res$Category %in% c("Means", "Variances"),
    c("Category", "lhs", "est", "se", "pval", "confint", "name")
]
tab_res

names(coef(res_final))

# Wald test for model comparison
wald_tests <- wald_test(res_final, "
    class1.M[1,6] = class2.M[1,6] &
    class1.M[1,6] = class3.M[1,6] ;
    class1.M[1,7] = class2.M[1,7] &
    class1.M[1,7] = class3.M[1,7]
")

# Rename the hypothesis
wald_tests$Hypothesis <- c("Mean Year_1", "Mean Year_2")
knitr::kable(wald_tests, digits = 2, caption = "Wald tests")

```

The results from the latent class growth analysis indicate a substantial improvement in model fit as the number of classes increases from one to three, as evidenced by lower Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) values. Specifically, the three-class model shows a log-likelihood (LL) of -28081.19, with an entropy of 0.80, suggesting a good classification accuracy. Notably, each class exhibits distinct growth parameters: the first class shows a mean intercept (i) of 59.45 and slope (s) of 2.29, indicating steady growth; while the third class has a notably lower mean intercept of 51.76 but a comparable slope of 2.36, reflecting similar growth rates from different starting points. The variances across time points also differ significantly between classes, highlighting variability in growth patterns across the sample.

### Model Plots
```{r}
# Assuming rng_height and lambda are defined previously in your script
brks <- seq(0, 1, length.out = 5) # Breaks in the rescaled 0-1 range

# Convert breaks back to the original log scale
# Note: This assumes that the original transformation was simply log(x) scaled to 0-1
labs <- exp(scales::rescale(brks, from = c(0, 1)))

p <- plot_growth(res_step[[3]], rawdata = TRUE, alpha_range = c(0, 0.05))
p <- p + scale_y_continuous(
    breaks = brks, # Use the same breaks for simplicity
    labels = round(labs, 2) # Round the labels for readability
) + ylab("Height (rescaled from log)")
p

```

The provided plot visualizes the growth trajectories from the latent class growth analysis, showcasing how height develops across different classes over time. The plot uses transformed height data, rescaled from log values to a normalized 0-1 range for easier interpretation and comparison. By setting specific breaks on the y-axis, corresponding to the original height values before transformation, the plot facilitates an intuitive understanding of growth patterns across the four time points. The plot highlights distinct growth curves for each class, with the shaded areas representing confidence intervals that help gauge the variability within each class. The plot shows classes overall appear to following similar patterns of growth.

## Wrapping Up

This analysis assessed how participants' heights change across multiple time points, revealing improved model fit with an increase in the number of classes from one to three. This is supported by reductions in the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) values, suggesting more precise class specifications enhance model accuracy. The best-fitting three-class model, which attained a log-likelihood of -28081.19 and an entropy of .80, indicates good classification accuracy. Distinct growth patterns emerged across classes and there was a notable variance in intercepts and consistent slopes across classes, emphasizing the variability of developmental patterns within the population.

Latent Class Growth Analysis (LCGA) is particularly advantageous when exploring heterogeneous population subgroups that follow distinct developmental trajectories over time. This method can effectively identify and categorize latent classes within the data based on differing patterns of change, without the assumption of normally distributed growth within each class. By classifying individuals into discrete groups that share similar growth patterns, LCGA allows for a clearer interpretation of how subgroups differ in their developmental trajectories. This approach is especially valuable in contexts where understanding the diverse patterns of change trajectories is important. It offers robust insights that are not merely averaged across the population but reflect subgroup-specific trends, making it ideal for studies where baseline characteristics and their influence on subsequent outcomes vary significantly across individuals.

### 2024-05-01 examples/_queue/_1a_minimal_Examples_DifferenceScores_PairedTtests.qmd
update unconditional examples

---
title: "Difference Scores: Paired Samples T-test"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

```

## Overview
Difference scores can be used to assess change in a variable across two timepoints. The value of the variable at the first timepoint is substracted from the value at the second timepoint. A one-sample t-test can then be applied to determine whether the mean change is significantly different from zero.

This example will examine height values obtained across 2 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to determine whether height values in the sample, on average, increase across the two timempoints. A difference score will be calculated between participant's height values collected at baseline and again at the 1-year follow-up. Next, a one-sample t-test will be used to test whether the difference score is statistically different than a null hypothesis of zero change. Finally, a visual inspection is further conducted via a scatterplot to graphically represent the relationship between participant's height across assessments.

## Preliminary Setup
### Load R libraries
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format

```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results
### Compute Difference Score
This code computes a difference score by subtracting participant's height at baseline from height at the 1-year follow-up. Next, a one-sample t-test is conducted to determine whether the mean difference score is different from zero. Model output is provided below, as well as a brief summary of results.

```{r}

# Reshaping data from long to wide
df_wide <- df_long %>%
  pivot_wider(
    id_cols = id,
    names_from = event,
    values_from = height,
    names_prefix = "Height_"
  )

# Compute the height difference score
df_wide <- df_wide %>%
    mutate(
        height_diff = Height_Year_1 - Height_Baseline
    )

# Calculate summary statistics for the difference score
summary(df_wide$height_diff)

# Compute summary statistics for Height by eventname
summary <- df_long %>%
    group_by(event) %>%
    get_summary_stats(height, type = "mean_sd")

data.frame(summary)

```

The summary statistics provide insights into participant's height across two annual assessments. At baseline, the average height is 55.24 inches (sd = 3.33). Over the year, there is an increase of 2.35 inches in average height to 57.59 inches at the time of the 1-year follow-up.

### Conduct t-test on Difference Score
A one-sample t-test is computed on the difference scores to examine whether there is evidence of change in participant's height between the two timepoints.
```{r}

# Perform a one-sample t-test on the difference scores for height
model <- t.test(df_wide$height_diff, mu = 0)
model

```

The results from the one-sample t-test conducted on the height difference scores reveals an increase in height from Baseline to Year 1. The t-test yields a t-value of 98.86 with a p-value less than <.001, indicating that the average height increase of 2.36 inches is significantly different from zero. Additionally, the 95% confidence interval
indicates that the true mean difference score in the population lies between approximately 2.31 to 2.41 inches.

### Model Plots
```{r}

# Scatterplot to compare height differences across two events
ggplot(df_wide, aes(x = Height_Baseline, y = Height_Year_1)) +
    geom_point(aes(color = Height_Baseline), alpha = 0.6) + # Color points by event type, adjust for your data
    labs(
        x = "Height at Baseline",
        y = "Height at Year 1",
        title = "Scatterplot of Heights at Baseline and Year 1",
        subtitle = "Each point represents a participant"
    ) +
    theme_minimal() +
    geom_smooth(method = "lm", se = FALSE) + # Add a regression line without confidence interval
    theme(legend.position = "bottom")

```

The scatterplot visually depicts the relationship between individuals' heights at baseline and their heights at year 1. Each point on the plot represents an individual, with their baseline height plotted on the x-axis and their year 1 height on the y-axis. A noticeable positive linear trend can be observed, as highlighted by the blue regression line, indicating that those who were taller at baseline generally remained taller at year 1. The strength and direction of this relationship suggests a strong positive association between baseline and year 1 height values.

## Wrapping Up
This analysis examined difference scores of participant's height using a one-sample t-test. Findings showed a significant increase in height values over the one-year interval of 2.36 inches, with confidence intervals ranging from 2.31 to 2.41 (t = 98.864, df = 11135, p-value < 2.2e-16). Further, a scatterplot visualizing the relationship between baseline and Year_1 weights showed a strong positive linear trend. This suggests that participants who were taller at baseline generally remained taller at Year_1, reaffirming the consistent growth trend observed in the data.

This difference score approach used in this example quantifies the change in height values over time within a single group and uses a one-sample t-test to test whether this value differs significanctly from 0. More generally, this approach is often implemented when data used to evalaute change is only available at two timepoints, as more flexible approaches are typically implemented when data from additional (>2) time points is available.

### 2024-05-01 examples/_queue/_1a_minimaltest_Examples_DifferenceScores_PairedTtests.qmd
update unconditional examples

---
title: "Difference Scores: Paired Samples T-test"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE
#| output: FALSE

# This code configures knitr code chunk options

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, output=T,
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(dplyr)  # Ensure dplyr is loaded

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
  select(src_subject_id, eventname, anthroheightcalc) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = as.factor(eventname),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
  na.omit()

```

## Overview

This example assesses growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1), using height as a representative metric. The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and year 1 height measurements for each participant; 2) a one-sample t-test is used to test whether the difference score is statistically different than a null hypothesis of zero change. Finally, a visual inspection is further conducted via a boxplot to graphically represent the relationship between participant's height at Baseline and the Year 1 follow-up.

## Preliminary Setup

### Load R libraries
```{r}

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(gridExtra)    # Arrange multiple grid-based plots on a page
library(rstatix)      # 
```

### Descriptives

```{r}

#Mean response over time
group_by(df_long, eventname) %>% 
  get_summary_stats(anthroheightcalc)

```

## Results

### Compute Difference Score
The code snippet below tells R to compute a difference score by subtracting each participant's height at baseline and the one-year follow-up. Relevant summary statistics are also provided.

```{r}

# Reshape data from long to wide format
df_wide <- df_long %>%
  pivot_wider(
    names_from = eventname,
    values_from = anthroheightcalc,
    names_prefix = "height_"
  )

# Compute the height difference
# Replace 'baseline_year_1_arm_1' and '1_year_follow_up_y_arm_1' with the exact names used in your dataset
df_wide <- df_wide %>%
  mutate(
    height_diff = height_baseline_year_1_arm_1 - height_1_year_follow_up_y_arm_1
  )

# Calculate summary statistics for the difference score
summary_stats <- summary(df_wide$height_diff)
print(summary_stats)

```

This R code effectively calculates the difference in height for each participant across two time points. The summary statistics, such as the mean, median, and range of these differences, provide a quantitative measure of change.

This summary of the difference score variable indicates that the differences range from a minimum xx units to an increase of xx units. The median difference is xx units, and the average difference is approximately xx units.

The summary statistics provide insights into the height variable across different xxx categories. At baseline, the average height is approximately xxx units with a standard deviation of xxxxx. Over the years, there’s a noticeable increase in average height: by the first year, it’s about xxxx units, and it continues to rise, reaching approximately xxxxx units by the fourth year. The standard deviation remains relatively consistent over the years, suggesting similar variability in height measurements across these time points.

### One-Sample T-test on Difference Score
After computing the difference scores, a one-sample t-test was conducted to examine if the average change in height (i.e., the difference score) is significantly different from zero (indicating change). This is key to determining whether there is a statistically significant growth on average across all participants.

```{r}

# Assuming df_wide is your dataframe and height_diff is your column of interest
# Perform a one-sample t-test
t_test_result <- t.test(df_wide$height_diff, mu = 0)

# Print the results
print(t_test_result)
```
This one-sample t-test assesses the null hypothesis that there is no change in height on average (mean difference = 0). The results of this test will indicate whether the observed changes are likely to be due to random variation or represent a true change in height over the period. A significant p-value (< 0.05) would reject the null hypothesis, suggesting that the average height change is statistically significant, either as an increase or a decrease.

### Model Plots
The boxplot visualization is a powerful tool for showing the distribution of height at both time points and visually assessing changes.

```{r}

# Assuming df_long is your dataframe
df_long <- df_long %>%
  mutate(eventname = factor(eventname,
                            levels = rev(c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")),
                            labels = c("Baseline", "1 Year Follow-up")))

# Create a boxplot with jitter for the height data
ggplot(df_long, aes(x = eventname, y = anthroheightcalc, fill = eventname)) +
  geom_boxplot() +  # Adds the boxplot
  geom_jitter(width = 0.2, alpha = 0.5, color = "black") +  # Adds jitter to show individual data points
  guides(fill = "none") +  # Removes the legend for fill, since it's redundant
  labs(x = "", y = "Height (inches)") +  # Sets the labels for axes
  theme_minimal() +  # Applies a minimal theme for a cleaner look
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels for better readability

```
The boxplot clearly delineates the range and distribution of heights at baseline and one-year follow-up. The central box represents the interquartile range (IQR) where the middle 50% of scores lie, and the whiskers extend to show the range of the data. This visualization helps highlight outliers and potential anomalies in data collection or entry. Observing how the IQR shifts from Baseline to 1 Year Follow-up can inform us about overall growth trends and variances.

## Wrapping Up

In the study sample, the average weight at baseline was approximately xxxx units with a standard deviation of xx, which showed a noticeable increase to an average of xx units by Year_1. A one-sample t-test was conducted to determine if the mean difference in weight from baseline to Year_1 significantly deviated from zero. The results indicated a statistically significant increase with a mean difference of approximately xx units (xxx to xxx), t(xx) = xxx, p < xx. Further, a scatterplot visualizing the relationship between baseline and Year_1 weights showed a strong positive linear trend. This suggests that participants who were taller at baseline generally remained taller at Year_1, reaffirming the consistent growth trend observed in the data.

### 2024-05-01 examples/1b_Examples_DifferenceScores_SimpleRegression.qmd
update toolbar

@@
 author: "Biostatistics Working Group"
-theme: cosmo
 ---
@@
 
-Compared to boys (the reference group), girls have an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units.
-This effect was marginally significant with a p-value of `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`.
-Overall, the model explained a very small portion of the variance in height difference, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 6)`, and the overall model significance was not statistically significant with a p-value of `r pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)`.
+Compared to girls (the reference group), boys have an average height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units.
+This effect was statistically significant with a p-value of `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`.
+Overall, the model explained a very small portion of the variance in height difference, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 6)`, and the overall model significance was  statistically significant with a p-value of `r pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)`.
 

### 2024-05-01 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
update toolbar
example 1a update

--- 2024-03-15 examples/_1a_Examples_DifferenceScores_PairedTtests.qmd
@@
 
-## Overview
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
 
-<p>
-
-This example assesses growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1), using weight as a representative metric.
-The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 weight measurements for each participant; 2) a one-sample t-test is used to test whether the difference score is statistically different than a null hypothesis of zero change.
-Finally, a visual inspection is further conducted via a scatterplot to graphically represent the relationship between participant's weight at Baseline and Year_1.
-The ensuing analysis and interpretations are detailed in the subsequent sections.
-
-</p>
-
-## Preliminary Setup
-
-::: panel-tabset
-### Install Packages {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
-
-```{r pckg-install}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-# Create a list of required packages
-packages_required <- c("tidyverse", "easystats", "lme4", "gtsummary", "report", "broom", "gridExtra")
-
-# Check which packages are not installed and install them
-packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
-if (length(packages_to_install) > 0) {
-    install.packages(packages_to_install)
-}
-
-# Load the required packages
-lapply(packages_required, library, character.only = TRUE)
-
-```
-:::
-
-### Load Packages
-
-::: blue
-**This code loads the r libraries necessary for this example**
-
-```{r lib-load}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-library(tidyverse) # Collection of R packages for data science
-library(easystats) # Collection of R packages statistical modeling, visualization, and reporting
-library(gtsummary) # Publication-ready tables
-library(report) # Easy reporting of regression analyses
-library(broom) # Tidy and augment statistical models output
-library(gridExtra) # Arrange multiple grid-based plots on a page
-
-```
-:::
-
-### Config Options
-
-::: blue
-**This code configures knitr code chunk options**
-
-```{r config}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
+# This code configures knitr code chunk options
 
 knitr::opts_chunk$set(
-    echo = T, message = F, warning = F, error = F,
+    echo = T, message = F, warning = F, error = F, output = T,
     comment = NA, cache = T, code_folding = T,
@@
 ```
-:::
-:::
 
-## Descriptives Overview
-
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-**This code reads in and shows the data to be used in the current example**
-
-```{r read-data}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
 #| cache: FALSE
 
-# Set the data paths
-data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
-data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
+# This code loads the data to be used for the example
+
+library(tidyverse)
+
+# Set the data path
+data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
 
 # Read the data
-data_demographics <- readRDS(data_path_1)
-data_nonimaging <- readRDS(data_path_2)
+df <- readRDS(data_path)
 
-# Subset the nonimaging data to include desired variables
-selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroweightcalc")
-subset_data <- data_nonimaging[, selected_vars]
-
-library(dplyr)
-# # Merge the datasets on 'src_subject_id' and 'eventname'
-merged_data <- data_demographics %>%
-    full_join(subset_data, by = c("src_subject_id", "eventname"))
-
-# Inspect the merged data structure
-str(merged_data)
-
-# Define event names to be retained in the analysis and convert variables to appropriate data types
-eventnames_to_include <- c(
-    "baseline_year_1_arm_1",
-    "1_year_follow_up_y_arm_1",
-    "2_year_follow_up_y_arm_1",
-    "3_year_follow_up_y_arm_1",
-    "4_year_follow_up_y_arm_1"
+# Create and mutate the dataframe
+df_long <- df %>%
+    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
+    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
+    drop_na(anthroheightcalc) %>%
+    arrange(src_subject_id, eventname) %>%
+    mutate(
+        src_subject_id = factor(src_subject_id),
+        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
+        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
+    ) %>%
+    rename(
+        id = src_subject_id,
+        event = eventname,
+        height = anthroheightcalc
     )
 
-df <- merged_data %>%
-    filter(eventname %in% eventnames_to_include) %>%
-    mutate(
-        src_subject_id = as.factor(src_subject_id),
-        eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
-        age = as.numeric(age),
-        sex = as.factor(sex),
-        race.4level = as.factor(race.4level),
-        hisp = as.factor(hisp),
-        high.educ.bl = as.factor(high.educ.bl),
-        household.income.bl = as.factor(household.income.bl),
-        acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
-        rel_family_id.bl = as.factor(rel_family_id.bl),
-        site_id_l = as.factor(site_id_l),
-        nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
-        anthroweightcalc = as.numeric(anthroweightcalc),
-        anthroweightcalc = as.numeric(anthroweightcalc)
-    ) %>%
-    # Exclude cases from unused assessment waves
-    filter(!is.na(eventname))
+```
+
+## Overview
+Difference scores can be used to assess change in a variable across two timepoints.
+The value of the variable at the first timepoint is substracted from the value at the second timepoint.
+A one-sample t-test can then be applied to determine whether the mean change is significantly different from zero.
+
+This example will examine height values obtained across 2 annual measurement occasions in a sample of youth taking part in the ABCD Study.
+The primary aim is to determine whether height values in the sample, on average, increase across the two timempoints.
+A difference score will be calculated between participant's height values collected at baseline and again at the 1-year follow-up.
+Next, a one-sample t-test will be used to test whether the difference score is statistically different than a null hypothesis of zero change.
+Finally, a visual inspection is further conducted via a scatterplot to graphically represent the relationship between participant's height across assessments.
+
+## Preliminary Setup
+### Load R libraries
+```{r}
+
+library(tidyverse) # Collection of R packages for data science
+library(gtsummary) # Publication-ready tables
+library(rstatix) # Statistical Tests in tidy format
 
 ```
-:::
 
 ### Descriptives
+```{r}
 
-::: blue
-**This code creates a descriptives table**
-
-```{r descriptives}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-df <- my_data
-
-descriptives_table <- df %>%
-    select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
-    mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1", "Year 2", "Year 3", "Year 4"))) %>%
-    mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
+descriptives_table <- df_long %>%
+    select(event, height) %>%
+    mutate(event = factor(event)) %>%
     tbl_summary(
-        by = eventname,
+        by = event,
         missing = "no",
-        label = list(
-            sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic",
-            anthroweightcalc ~ "Weight"
-        ),
-        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
+        label = list(height ~ "Height"),
+        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
     ) %>%
@@
 descriptives_table
+
 ```
 
-:::
-:::
+## Results
+### Compute Difference Score
+This code computes a difference score by subtracting participant's height at baseline from height at the 1-year follow-up.
+Next, a one-sample t-test is conducted to determine whether the mean difference score is different from zero.
+Model output is provided below, as well as a brief summary of results.
 
-## Results
+```{r}
 
-::: panel-tabset
-### Build Model: Step I {.tabset .tabset-fade .tabset-pills}
+# Reshaping data from long to wide
+df_wide <- df_long %>%
+    pivot_wider(
+        id_cols = id,
+        names_from = event,
+        values_from = height,
+        names_prefix = "Height_"
+    )
 
-::: blue
-The code snippet below tells R to compute a difference score by subtracting each participant's weight at T1 from their weight at T0. Relevant summary statistics are also provided.
+# Compute the height difference score
+df_wide <- df_wide %>%
+    mutate(
+        height_diff = Height_Year_1 - Height_Baseline
+    )
 
-**STEP 1: Compute Difference Score**
+# Calculate summary statistics for the difference score
+summary(df_wide$height_diff)
 
-```{r build-model-ttest-1}
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-#| class.source: 'fold-hide'
+# Compute summary statistics for Height by eventname
+summary <- df_long %>%
+    group_by(event) %>%
+    get_summary_stats(height, type = "mean_sd")
 
-# Function to compute difference scores
-compute_difference <- function(df, variable_name) {
-    # Define the event names of interest
-    baseline_event <- "baseline_year_1_arm_1"
-    followup_event <- "1_year_follow_up_y_arm_1"
-
-    # Compute the difference between Baseline and Year 1 data for the given variable
-    diff_data <- df %>%
-        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
-        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
-        spread(eventname, variable_name) %>% # Convert data from long to wide format
-        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
-        mutate(group_resid = (diff) - mean(diff)) %>% # compute residuals (group_resid).
-        drop_na(diff) # Exclude rows with NA in the computed difference
-
-    # Summarize the computed difference scores
-    diff_summary <- summary(diff_data$diff)
-
-    # Return the difference data and its summary
-    list(data = diff_data, summary = diff_summary)
-}
-
-# List of variables for which difference scores are to be computed
-variables_of_interest <- c("anthroweightcalc")
-
-# Compute and store difference scores and summaries for each variable in a list
-difference_and_summary_list <- lapply(variables_of_interest, function(var) {
-    compute_difference(df, var)
-})
-
-# Extract the difference data for the 'anthroweightcalc' variable
-weight_diff_data <- difference_and_summary_list[[1]]$data
-
-# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
-df <- left_join(df, weight_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")
+data.frame(summary)
 
 ```
+
+The summary statistics provide insights into participant's height across two annual assessments.
+At baseline, the average height is 55.24 inches (sd = 3.33).
+Over the year, there is an increase of 2.35 inches in average height to 57.59 inches at the time of the 1-year follow-up.
+
+### Conduct t-test on Difference Score
+A one-sample t-test is computed on the difference scores to examine whether there is evidence of change in participant's height between the two timepoints.
+```{r}
+
+# Perform a one-sample t-test on the difference scores for height
+model <- t.test(df_wide$height_diff, mu = 0)
+model
+
+```
+
+The results from the one-sample t-test conducted on the height difference scores reveals an increase in height from Baseline to Year 1.
+The t-test yields a t-value of 98.86 with a p-value less than <.001, indicating that the average height increase of 2.36 inches is significantly different from zero.
+Additionally, the 95% confidence interval indicates that the true mean difference score in the population lies between approximately 2.31 to 2.41 inches.
+
+### Model Plots
+```{r}
+
+# Scatterplot to compare height differences across two events
+ggplot(df_wide, aes(x = Height_Baseline, y = Height_Year_1)) +
+    geom_point(aes(color = Height_Baseline), alpha = 0.6) + # Color points by event type, adjust for your data
+    labs(
+        x = "Height at Baseline",
+        y = "Height at Year 1",
+        title = "Scatterplot of Heights at Baseline and Year 1",
+        subtitle = "Each point represents a participant"
+    ) +
+    theme_minimal() +
+    geom_smooth(method = "lm", se = FALSE) + # Add a regression line without confidence interval
+    theme(legend.position = "bottom")
+
+```
+
+The scatterplot visually depicts the relationship between individuals' heights at baseline and their heights at year 1.
+Each point on the plot represents an individual, with their baseline height plotted on the x-axis and their year 1 height on the y-axis.
+A noticeable positive linear trend can be observed, as highlighted by the blue regression line, indicating that those who were taller at baseline generally remained taller at year 1.
+The strength and direction of this relationship suggests a strong positive association between baseline and year 1 height values.
+
+## Wrapping Up
+This analysis examined difference scores of participant's height using a one-sample t-test.
+Findings showed a significant increase in height values over the one-year interval of 2.36 inches, with confidence intervals ranging from 2.31 to 2.41 (t = 98.864, df = 11135, p-value < 2.2e-16).
+Further, a scatterplot visualizing the relationship between baseline and Year_1 weights showed a strong positive linear trend.
+This suggests that participants who were taller at baseline generally remained taller at Year_1, reaffirming the consistent growth trend observed in the data.
+
+This difference score approach used in this example quantifies the change in height values over time within a single group and uses a one-sample t-test to test whether this value differs significanctly from 0.
+More generally, this approach is often implemented when data used to evalaute change is only available at two timepoints, as more flexible approaches are typically implemented when data from additional (>2) time points is available.

### 2024-05-01 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
update example 1b

@@
     echo = T, message = F, warning = F, error = F, output = T,
-    comment = NA, cache = T, code_folding = T,
+    comment = NA, cache = T, code_folding = show,
     R.options = list(width = 220, digits = 3),

### 2024-05-01 examples/1b_Examples_DifferenceScores_SimpleRegression.qmd
update example 1b

@@
 
-# Simple Regression {.unnumbered}
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+
+# This code configures knitr code chunk options
+
+knitr::opts_chunk$set(
+    echo = T, message = F, warning = F, error = F, output = T,
+    comment = NA, cache = T, code_folding = show,
+    R.options = list(width = 220, digits = 3),
+    fig.align = "center",
+    out.width = "75%", fig.asp = .75
+)
+```
+
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+#| cache: FALSE
+
+# This code loads the data to be used for the example
+
+library(tidyverse)
+
+# Set the data path
+data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
+
+# Read the data
+df <- readRDS(data_path)
+
+# Create and mutate the dataframe
+df_long <- df %>%
+    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
+    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
+    drop_na(anthroheightcalc) %>%
+    arrange(src_subject_id, eventname) %>%
+    group_by(src_subject_id) %>%
+    fill(ehi_y_ss_scoreb, .direction = "downup") %>%
+    ungroup() %>%
+    filter(ehi_y_ss_scoreb %in% c(1, 2)) %>%
+    mutate(
+        src_subject_id = factor(src_subject_id),
+        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
+        ehi_y_ss_scoreb = factor(ehi_y_ss_scoreb, levels = c(1, 2), labels = c("Right", "Left")),
+        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
+    ) %>%
+    rename(
+        id = src_subject_id,
+        event = eventname,
+        handedness = ehi_y_ss_scoreb,
+        height = anthroheightcalc
+    )
+
+```
 
 ## Overview
+Difference scores can be used to assess change in a variable across two timepoints.
+The value of the variable at the first timepoint is substracted from the value at the second timepoint.
+A simple regression can then be applied to determine whether a variable of interest predicts the computed difference score.
 
-<p>
-
-This example assesses whether growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1) differs significantly based on handedness, using height ("anthroheightcalc") as a representative metric of growth.
-The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 height measurements for each participant; 2) a simple regression analysis is used to test whether sex (boy, girl) predicts the average difference value in participants height from baseline to the 1-Year follow-up.
-Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between difference scores and sex.
-The ensuing analysis and interpretations are detailed in the subsequent sections.
-
-</p>
+This example will examine height values obtained across 2 annual measurement occasions in a sample of youth taking part in the ABCD Study.
+The primary aim is to determine whether a variable of interest (handedness) predicts the difference in height values between the two timepoints.
+A difference score will be calculated between participant's height values collected at baseline and again at the 1-year follow-up.
+Next, a simple regression analysis will be conducted to test whether handedness (left, right) predicts the average difference value in participants' height from baseline to the 1-Year follow-up.
+Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between the difference score and handedness.
 
 ## Preliminary Setup
+### Load R libraries
+```{r}
 
-::: panel-tabset
-### Install Packages {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
-
-```{r pckg-install}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-# rm(list = ls())
-
-# Create a list of required packages
-packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")
-
-# Check which packages are not installed and install them
-packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
-if (length(packages_to_install) > 0) {
-    install.packages(packages_to_install)
-}
-
-# Load the required packages
-lapply(packages_required, library, character.only = TRUE)
+library(tidyverse) # Collection of R packages for data science
+library(gtsummary) # Publication-ready tables
+library(rstatix) # Statistical Tests in tidy format
 
 ```
-:::
-
-### Load Packages
-
-::: blue
-**This code loads the r libraries necessary for this example**
-
-```{r lib-load}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-library(tidyverse)    # Collection of R packages for data science
-library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
-library(rstatix)      # Pipe-friendly framework for performing common stats
-library(gtsummary)    # Publication-ready tables
-library(report)       # Easy reporting of regression analyses
-library(broom)        # Tidy and augment statistical models output
-library(gridExtra)    # Arrange multiple grid-based plots on a page
-
-```
-:::
-
-```{r config}
-#| echo: FALSE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-### Config Options
-#::: blue
-#**This code configures knitr code chunk options**
-
-knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
-                      comment=NA, cache=T, code_folding=T,
-                      R.options=list(width=220, digits = 3),
-                      fig.align='center', 
-                      out.width='75%', fig.asp=.75)
-```
-:::
-
-## Descriptives Overview
-
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-**This code reads in and shows the data to be used in the current
-example**
-
-```{r read-data}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-#| cache: FALSE
-
-# Set the data paths
-data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
-data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
-
-# Read the data
-data_demographics <- readRDS(data_path_1)
-data_nonimaging <- readRDS(data_path_2)
-
-# Subset the nonimaging data to include desired variables
-selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
-subset_data <- data_nonimaging[, selected_vars]
-
-library(dplyr)
-# # Merge the datasets on 'src_subject_id' and 'eventname'
-merged_data <- data_demographics %>%
-  full_join(subset_data, by = c("src_subject_id", "eventname"))
-
-# Inspect the merged data structure
-str(merged_data)
-
-# Define event names to be retained in the analysis and convert variables to appropriate data types
-eventnames_to_include <- c("baseline_year_1_arm_1",
-                           "1_year_follow_up_y_arm_1",
-                           "2_year_follow_up_y_arm_1",
-                           "3_year_follow_up_y_arm_1",
-                           "4_year_follow_up_y_arm_1")
-
-df <- merged_data %>%
-  filter(eventname %in% eventnames_to_include) %>%
-  mutate(
-    src_subject_id = as.factor(src_subject_id),
-    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
-    age = as.numeric(age),
-    sex = as.factor(sex),
-    race.4level = as.factor(race.4level),
-    hisp = as.factor(hisp),
-    high.educ.bl = as.factor(high.educ.bl),
-    household.income.bl = as.factor(household.income.bl),
-    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
-    rel_family_id.bl = as.factor(rel_family_id.bl),
-    site_id_l = as.factor(site_id_l),
-    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
-    anthroweightcalc = as.numeric(anthroweightcalc),
-    anthroheightcalc = as.numeric(anthroheightcalc)
-  ) %>%
-  # Exclude cases from unused assessment waves
-  filter(!is.na(eventname))
-
-```
-:::
 
 ### Descriptives
+```{r}
 
-::: blue
-**This code creates a descriptives table**
-
-```{r descriptives}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-descriptives_table <- df %>%
-  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
-  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
-  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
+descriptives_table <- df_long %>%
+    select(event, handedness, height) %>%
+    mutate(event = factor(event)) %>%
+    mutate(handedness = factor(handedness)) %>%
     tbl_summary(
-    by = eventname,
+        by = event,
         missing = "no",
-    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
-                 anthroweightcalc ~ "Weight"),
+        label = list(handedness ~ "Handedness", height ~ "Height"),
         statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
@@
 descriptives_table
-```
-
-:::
-:::
-
-## Results
-
-::: panel-tabset
-### Build Model: Step 1 {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-The code snippet below tells R to compute a difference score by subtracting each participant’s height at T1 from their height at T0. Relevant summary statistics are also provided.
-
-**STEP 1: Compute Difference Score**
-```{r build-model-regression-1}
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-#| class.source: 'fold-hide'
-
-# Define the function to compute difference scores for a given variable and provide a summary
-# Function to compute difference scores for a given variable and provide a summary
-compute_difference_and_summary <- function(df, variable_name) {
-    # Define the event names of interest
-    baseline_event <- "baseline_year_1_arm_1"
-    followup_event <- "1_year_follow_up_y_arm_1"
-
-    # Compute the difference between Baseline and Year 1 data for the given variable
-    diff_data <- df %>%
-        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
-        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
-        spread(eventname, variable_name) %>% # Convert data from long to wide format
-        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
-        drop_na(diff) # Exclude rows with NA in the computed difference
-
-    # Summarize the computed difference scores
-    diff_summary <- summary(diff_data$diff)
-
-    # Return the difference data and its summary
-    list(data = diff_data, summary = diff_summary)
-}
-
-# List of variables for which difference scores are to be computed
-variables_of_interest <- c("anthroheightcalc")
-
-# Compute and store difference scores and summaries for each variable in a list
-difference_and_summary_list <- lapply(variables_of_interest, function(var) {
-    compute_difference_and_summary(df, var)
-})
-
-# Extract the difference data for the 'anthroheightcalc' variable
-height_diff_data <- difference_and_summary_list[[1]]$data
-
-# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
-df <- left_join(df, height_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")
 
@@
 
-**Descriptive Statistics for the Difference Score**
-```{r model-checking}
-# Compute statistical summaries for the difference score variable
-lapply(difference_and_summary_list, function(item) {
-    print(item$summary)
-})
-```
+## Results
+### Compute Difference Score
+This code computes a difference score by subtracting participant's height at baseline from height at the 1-year follow-up.
+Next, a simple regression analysis is conducted to determine whether handedness predicts the difference in height values between the two timepoints.
 
-This summary of the difference score variable indicates that the differences range from a decrease of `r min(height_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(height_diff_data$diff, na.rm = TRUE)` units.
-The median difference is `r median(height_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(height_diff_data$diff, digits = 2, na.rm = TRUE)` units.
-There are `r sum(is.na(height_diff_data$diff))` missing values in this difference score dataset.
+```{r}
 
-**Descriptive Statistics for the Height Variable**
-```{r}
-## Summary statistics
+# Reshape data from long to wide format
+df_wide <- df_long %>%
+  pivot_wider(
+    id_cols = c(id, handedness),
+    names_from = event,
+    values_from = height,
+    names_prefix = "Height_"
+  )
+
+# Compute the height difference score
+df_wide <- df_wide %>%
+    mutate(
+        height_diff = Height_Year_1 - Height_Baseline
+    )
+
+# Calculate summary statistics for the difference score
+summary(df_wide$height_diff)
+
 # Compute summary statistics for Height by eventname
-summary <- df %>%
-    group_by(eventname) %>%
-    get_summary_stats(anthroheightcalc, type = "mean_sd")
+summary <- df_long %>%
+    group_by(event) %>%
+    get_summary_stats(height, type = "mean_sd")
 
 data.frame(summary)
+
 ```
+The summary statistics provide insights into participant's height across two annual assessments.
+At baseline, the average height is 55.24 inches (sd = 3.33).
+Over the year, there is an increase of 2.35 inches in average height to 57.59 inches at the time of the 1-year follow-up.
 
-The summary statistics provide insights into the Height variable across different `eventname` categories.
-At baseline, the average height is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`.
-Over the years, there’s a noticeable increase in average height: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year.
-The standard deviation remains relatively consistent over the years, suggesting similar variability in height measurements across these time points.
-
-:::
-
-### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-A simple regression analyses is conducted to examine whether a grouping variable (sex) significantly predicts the difference score value (indicating significant group differences in the average difference score).
-
-**STEP 2: Conduct regression on Difference Score**
-```{r build-model-regression-2}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
-merged_data <- height_diff_data %>%
-  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")
-
-# Ensure 'sex' is a factor
-merged_data$sex <- as.factor(merged_data$sex)
+### Conduct regression on Difference Score
+A simple regression analyses is conducted to examine whether a grouping variable (handedness) predicts the difference score value.
+```{r}
 
 # Run the regression
-model <- lm(diff ~ sex, data = merged_data)
+model <- lm(height_diff ~ handedness, data = df_wide)
 
 # Get the summary of the regression model
-model_summary <- summary(model)
-
-model_summary
+summary(model)
 
@@
 
-This regression analysis evaluates whether sex (boy, girl) predicts the average difference in participants' height from baseline to the 1-Year follow-up. The output from our model provides:
-
-i. an F-statistic of `r model_summary$fstatistic[1]`; 
-ii. degrees of freedom of `r model_summary$df[2]`; 
-iii. a parameter estimate of `r model_summary$coefficients["sexMale or other", "Estimate"]`; 
-iv. standard error of `r model_summary$coefficients["sexMale or other", "Std. Error"]`; 
-v. p-value of `r model_summary$coefficients["sexMale or other", "Pr(>|t|)"]`.
-
-Compared to girls (the reference group), boys have an average height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units.
-This effect was statistically significant with a p-value of `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`.
-Overall, the model explained a very small portion of the variance in height difference, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 6)`, and the overall model significance was  statistically significant with a p-value of `r pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)`.
-
-:::
+The results from the simple regression analysis suggest handedness does not significantly predict change in height.
+The intercept, estimated at 2.35 with a standard error of .02, indicates the average height difference for right-handed individuals is approximately 2.35 units, with this effect differing by only .03 inches (p=.69) for left-handed individuals
 
 ### Model Plots
+```{r}
 
-::: blue
-```{r violin-plots}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
-
-# Visualize the difference scores across different levels of sex
+# Visualize the difference scores across different levels of handedness
 
@@
 # Load necessary libraries
-library(dplyr)
-library(ggplot2)
 library(RColorBrewer)
 
-# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
-merged_data <- height_diff_data %>%
-  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")
-
 # Plotting using ggplot2
-ggplot(merged_data, aes(x = sex, y = diff, fill = sex)) +
+ggplot(df_long, aes(x = handedness, y = height, fill = handedness)) +
     geom_violin() +
@@
     labs(
-    title = "Difference Scores by Sex",
-    x = "Sex",
-    y = "Difference Score"
+        title = "Difference Scores by Handedness",
+        x = "Handedness",
+        y = "Height Difference Score"
     ) +
@@
 
-The violin plot visualizes the distribution of difference scores in children's heights by their sex: Right, Left, and Mixed.
+The violin plot visualizes the distribution of difference scores in children's heights by their handedness (Right, Left).
 Each violin shape provides insight into the density of the data at different height differences, with wider sections representing higher densities of data points.
 Superimposed jittered points offer a granular view of individual observations.
-From the plot, it appears that the distributions of height differences across the three handedness categories are somewhat similar, though there might be subtle variations in median and spread.
-
-:::
-:::
+From the plot, it appears that the distributions of height differences across the two handedness categories are markedly similar.
 
 ## Wrapping Up
+This analysis examined difference scores of participant's height across a one-year period and used a simple regression analysis to explore the impact of handedness on the year-to-year change in children's height.
+Findings showed an increase in height values from baseline (55.23 inches) to the year 1 follow-up (57.57 inches).
+The regression model's output shows that handedness, specifically being left-handed as compared to right-handed, is not associated with any change in height (r=.03, se=.08, p=.69).
+Further, a violin plot visualizing the distribution of difference scores in children's heights by their handedness showed no substantive effect.
 
-::: panel-tabset
-### Write-up {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-Differences in children's heights between baseline and a subsequent 1-year follow-up, we sought to understand the potential influence of sex on this difference.
-Descriptive statistics revealed that height differences ranged from a decrease of `r min(merged_data$diff, na.rm = TRUE)` units to an increase of `r max(merged_data$diff, na.rm = TRUE)` units, with a median difference of `r median(merged_data$diff, na.rm = TRUE)` units and an average difference of approximately `r mean(merged_data$diff, na.rm = TRUE)` units.
-Subsequent regression analysis indicated that, when compared to boys, girls had an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units, though this effect was marginally significant (p = `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`).
-Complementary to these findings, violin plots visually underscored the subtle variations in height differences across sex categories, suggesting broadly similar distributions but with nuanced variations in central tendency and spread.
-In conclusion, while sex exhibited a potential influence on height difference over the year, the observed effect was notably minor.
-
-:::
-:::
+This difference score approach used in this example quantifies the change in height values over time and uses a simple regression to test whether participant handedness predicts this value.
+More generally, this approach is often implemented when data used to evalaute change is only available at two timepoints, as more flexible approaches are typically implemented when data from additional (>2) time points is available.

### 2024-05-01 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
update examples 9, 10, 13

--- 2024-05-01 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
@@
     echo = T, message = F, warning = F, error = F, output = T,
-    comment = NA, cache = T, code_folding = T,
+    comment = NA, cache = T,
     R.options = list(width = 220, digits = 3),

### 2024-05-01 examples/1b_Examples_DifferenceScores_SimpleRegression.qmd
update examples 9, 10, 13

@@
     echo = T, message = F, warning = F, error = F, output = T,
-    comment = NA, cache = T, code_folding = show,
+    comment = NA, cache = T,
     R.options = list(width = 220, digits = 3),

### 2024-05-01 examples/2_Examples_ResidualizedChangeScores.qmd
update examples 9, 10, 13

--- 2024-03-06 _posts/2_Examples_ResidualizedChangeScores.qmd
@@
 
-# Mulitple Regression {.unnumbered}
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+
+# This code configures knitr code chunk options
+
+knitr::opts_chunk$set(
+    echo = T, message = F, warning = F, error = F, output = T,
+    comment = NA, cache = T, code_folding = T,
+    R.options = list(width = 220, digits = 3),
+    fig.align = "center",
+    out.width = "75%", fig.asp = .75
+)
+```
+
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+#| cache: FALSE
+
+# This code loads the data to be used for the example
+
+library(tidyverse)
+
+# Set the data path
+data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
+
+# Read the data
+df <- readRDS(data_path)
+
+# Create and mutate the dataframe
+df_long <- df %>%
+    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
+    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
+    drop_na(anthroheightcalc) %>%
+    arrange(src_subject_id, eventname) %>%
+    group_by(src_subject_id) %>%
+    fill(ehi_y_ss_scoreb, .direction = "downup") %>%
+    ungroup() %>%
+    filter(ehi_y_ss_scoreb %in% c(1, 2)) %>%
+    mutate(
+        src_subject_id = factor(src_subject_id),
+        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
+        ehi_y_ss_scoreb = factor(ehi_y_ss_scoreb, levels = c(1, 2), labels = c("Right", "Left")),
+        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
+    ) %>%
+    rename(
+        id = src_subject_id,
+        event = eventname,
+        handedness = ehi_y_ss_scoreb,
+        height = anthroheightcalc
+    )
+
+```
 
 ## Overview
+A residualized change score is a statistical approach used to account for the baseline (initial) values of a variable when measuring across two time points.
+First, a linear regression is conducted with the follow-up score predicted by the baseline score.
+The residual values from this model can be interpreted as change scores that have been adjusted for the baseline values.
+In a next step, regression analysis can be applied to determine whether a variable of interest predicts the residualized change score.
 
-<p>
-
-This example assesses whether height, measured at the 1-Year follow-up (T1) in a subsample of ABCD participants, is predicted by weight and height measured at baseline (T0).
-A visual inspection is further conducted via a scatterplot to graphically represent the relationship between height at Year_1 and weight at baseline, controlling for height at baseline.
-The ensuing analysis and interpretations are detailed in the subsequent sections.
-
-</p>
+This example will examine whether handedness predicts changes in height from baseline to the 1 year follow-up, controlling for initial height measurements, in a sample of participants taking part in the ABCD Study.
+Initially, a linear regression analysis predicts height measurements at the 1 year follow-up from their baseline values to calculate residualized change scores.
+Subsequently, a second regression assesses whether handedness (left vs.
+right) predicts these residualized changes in height measurements.
+The relationship is also visually represented through a violin plot, illustrating differences in the residualized height values between left-handed and right-handed individuals.
 
 ## Preliminary Setup
-
-::: panel-tabset
-### Install Packages {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
-
-```{r pckg-install}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-# Create a list of required packages
-packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")
-
-# Check which packages are not installed and install them
-packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
-if (length(packages_to_install) > 0) {
-    install.packages(packages_to_install)
-}
-
-# Load the required packages
-lapply(packages_required, library, character.only = TRUE)
-
-```
-:::
-
 ### Load Packages
-
-::: blue
-**This code loads the r libraries necessary for this example**
-
-```{r lib-load}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
+```{r}
 
 library(tidyverse) # Collection of R packages for data science
-library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
 library(gtsummary) # Publication-ready tables
-library(report)       # Easy reporting of regression analyses
-library(broom)        # Tidy and augment statistical models output
-library(gridExtra)    # Arrange multiple grid-based plots on a page
-
+library(rstatix) # Statistical Tests in tidy format
+library(lme4) # Linear mixed-effects models
 ```
-:::
-
-### Config Options
-
-::: blue
-**This code configures knitr code chunk options**
-
-```{r config}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
-                      comment=NA, cache=T, code_folding=T,
-                      R.options=list(width=220, digits = 3),
-                      fig.align='center', 
-                      out.width='75%', fig.asp=.75)
-```
-:::
-:::
-
-## Descriptives Overview
-
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-**This code reads in and shows the data to be used in the current
-example**
-
-```{r read-data}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-#| cache: FALSE
-
-# Set the data paths
-data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
-data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
-
-# Read the data
-data_demographics <- readRDS(data_path_1)
-data_nonimaging <- readRDS(data_path_2)
-
-# Subset the nonimaging data to include desired variables
-selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
-subset_data <- data_nonimaging[, selected_vars]
-
-library(dplyr)
-# # Merge the datasets on 'src_subject_id' and 'eventname'
-merged_data <- data_demographics %>%
-  full_join(subset_data, by = c("src_subject_id", "eventname"))
-
-# Inspect the merged data structure
-str(merged_data)
-
-# Define event names to be retained in the analysis and convert variables to appropriate data types
-eventnames_to_include <- c("baseline_year_1_arm_1",
-                           "1_year_follow_up_y_arm_1",
-                           "2_year_follow_up_y_arm_1",
-                           "3_year_follow_up_y_arm_1",
-                           "4_year_follow_up_y_arm_1")
-
-df <- merged_data %>%
-  filter(eventname %in% eventnames_to_include) %>%
-  mutate(
-    src_subject_id = as.factor(src_subject_id),
-    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
-    age = as.numeric(age),
-    sex = as.factor(sex),
-    race.4level = as.factor(race.4level),
-    hisp = as.factor(hisp),
-    high.educ.bl = as.factor(high.educ.bl),
-    household.income.bl = as.factor(household.income.bl),
-    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
-    rel_family_id.bl = as.factor(rel_family_id.bl),
-    site_id_l = as.factor(site_id_l),
-    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
-    anthroweightcalc = as.numeric(anthroweightcalc),
-    anthroheightcalc = as.numeric(anthroheightcalc)
-  ) %>%
-  # Exclude cases from unused assessment waves
-  filter(!is.na(eventname))
-
-```
-:::
 
 ### Descriptives
+```{r}
 
-::: blue
-**This code creates a descriptives table**
-
-```{r descriptives}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-descriptives_table <- df %>%
-  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
-  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
-  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
+descriptives_table <- df_long %>%
+    select(event, handedness, height) %>%
+    mutate(event = factor(event)) %>%
+    mutate(handedness = factor(handedness)) %>%
     tbl_summary(
-    by = eventname,
+        by = event,
         missing = "no",
-    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
-                 anthroweightcalc ~ "Weight"),
+        label = list(handedness ~ "Handedness", height ~ "Height"),
         statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
@@
 descriptives_table
-```
-
-:::
-:::
-
-## Results
-
-::: panel-tabset
-### Build Model {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-The code snippet below tells R to conduct a multiple regression analysis by subtracting each participant's height at T1 from their height at T0. Relevant summary statistics are also provided.
-
-**Compute Multiple Regression Model**
-```{r build-model-resid-regression-1}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# Split data using the correct eventname values
-baseline_data <- df %>% 
-  filter(eventname == "baseline_year_1_arm_1") %>% 
-  select(src_subject_id, Height_baseline = anthroheightcalc, Weight_baseline = anthroweightcalc)
-
-followup_data <- df %>% 
-  filter(eventname == "1_year_follow_up_y_arm_1") %>% 
-  select(src_subject_id, Height_followup = anthroheightcalc)
-
-# Merge baseline and follow-up data
-merged_data <- baseline_data %>%
-  left_join(followup_data, by = "src_subject_id")
-
-# Check the first few rows of the merged data
-#head(merged_data)
-
-# Fit the regression model excluding NA values
-model <- lm(Height_followup ~ Height_baseline + Weight_baseline, data = merged_data, na.action = na.exclude)
 
@@
 
-:::
+## Results
+### Compute Multiple Regression Model
+The code below is used to compute a residualized change score by first adjusting participants' height at the 1-year follow-up for their baseline height through a regression analysis.
+The residuals from this model represent the adjusted height changes that are not explained by their initial measurements.
+Subsequently, these residualized change scores are used in a simple regression analysis to explore whether handedness predicts the changes in height between the two timepoints.
+This approach helps to isolate any effect of handedness on height growth, independent of the baseline height values.
 
-### Model Summary {.tabset .tabset-fade .tabset-pills}
+```{r}
 
-::: blue
+# Reshape data from long to wide format
+df_wide <- df_long %>%
+    pivot_wider(
+        names_from = event,
+        values_from = height,
+        names_prefix = "Height_"
+    )
 
-A plot to show xxxxx.
-```{r model-checking}
-model_summary <- summary(model)
-print(model_summary)
+# Step 1: Fit the linear model to get residuals
+model_baseline <- lm(Height_Year_1 ~ Height_Baseline, data = df_wide, na.action = na.exclude)
+df_wide <- df_wide %>%
+    mutate(residualized_change = residuals(model_baseline))
 
-par(mfrow = c(2, 2))  # Set up a 2x2 plotting area
-plot(model)  # Generate diagnostic plots
+# Step 2: Model residualized change scores predicted by handedness
+model_residualized_change <- lm(residualized_change ~ handedness, data = df_wide, na.action = na.exclude)
+
+# View the summary of the model to see the effect of handedness
+summary(model_residualized_change)
 
@@
 
-This regression analysis evaluates how baseline height and weight predict children's height at a 1-year follow-up. The output from our model provides several key statistics:
+The results from the residualized change score analysis suggest that while individual differences in growth occurs, handedness does not appear to be a determining factor.
+The model intercept, close to zero (-0.005), is consistent with the expectation that if there is no effect for handedness, there would be no change when adjusting for baseline heights.
+Further, examination of the model residuals (range = -53.90 to 65.53, se = 2.32) and the negligible r-squared value indicates that handedness does not explain a substantive portion of the variance in residualized height changes.
 
-i. an F-statistic of `r model_summary$fstatistic[1]`;
-ii. degrees of freedom of `r model_summary$df[2]`;
-iii. parameter estimates for baseline height and weight;
-iv. standard errors for these estimates; and
-v. p-values for the significance of these estimates.
+### Model Plots
+```{r}
 
-For every one unit increase in baseline height, there's an associated increase of approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units in the follow-up height, and this effect was statistically significant with a p-value of `r model_summary$coefficients["Height_baseline", "Pr(>|t|)"]`.
-Similarly, for every one unit increase in baseline weight, there's an associated increase of approximately `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units in the follow-up height, which is also statistically significant with a p-value of `r model_summary$coefficients["Weight_baseline", "Pr(>|t|)"]`.
-Overall, this model explains a substantial portion of the variance in follow-up height, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 4)`.
-The overall model is highly significant with a p-value less than `r model_summary$coefficients[1,4]`, indicating that the predictors, collectively, have a significant relationship with the dependent variable.
+par(mfrow = c(2, 2)) # Set up a 2x2 plotting area
+plot(model_baseline) # Generate diagnostic plots
 
-:::
+# Visualize the residualized change scores across different levels of handedness
 
-### Model Plots  {.tabset .tabset-fade .tabset-pills}
+# Create a violin plot to show the distribution of residualized change scores by handedness
+# Jittered points are added to provide a more granular view of individual observations
+# Load necessary libraries
+library(RColorBrewer)
 
-::: blue
-```{r scatterplot}
-
-# Scatterplot with regression line for Weight_baseline
-ggplot(merged_data, aes(x = Weight_baseline, y = Height_followup)) +
-  geom_point(alpha = 0.5) +
-  geom_smooth(method = "lm", color = "black") + 
-  labs(title = "Relationship between Baseline Weight and Follow-up Height",
-       x = "Baseline Weight",
-       y = "Follow-up Height") +
-  theme_minimal()
+# Plotting using ggplot2
+ggplot(df_long, aes(x = handedness, y = height, fill = handedness)) +
+    geom_violin() +
+    geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
+    scale_fill_brewer(palette = "Set2") +
+    labs(
+        title = "Residualized Change Score by Handedness",
+        x = "Handedness",
+        y = "Height Change Score"
+    ) +
+    theme_minimal() +
+    theme(
+        axis.text.x = element_text(angle = 45, hjust = 1),
+        legend.position = "none"
+    )
 
@@
 
-The scatterplot visually depicts the relationship between children's weights at baseline (`r names(merged_data)[3]`) and their heights at Year_1 (`r names(merged_data)[4]`).
-Each point on the plot represents a child, with their baseline weight plotted on the x-axis and their Year_1 height on the y-axis.
-The clear positive linear trend, as illustrated by the black regression line, indicates that children who had higher weights at baseline generally had higher heights at Year_1.
-While further statistical analyses can quantify the strength and direction of this relationship, visually, the data suggests a pronounced positive association between baseline weight and follow-up height.
-
-:::
-:::
+The violin plot visualizes the distribution of residualized change scores in children's heights by their handedness (Right, Left).
+Each violin shape provides insight into the density of the data at different (residualized) height values, with wider sections representing higher densities of data points.
+Superimposed jittered points offer a granular view of individual observations.
+From the plot, it appears that the overall distributions of the residualized change scores for the height outcome across the two handedness categories are markedly similar.
 
 ## Wrapping Up
-<div class = "blue">
+This analysis evaluated residualized change scores of participants' height over a one-year period, controlling for initial height measurements at baseline.
+It then utilized a simple regression analysis to investigate handedness as a predictor of these adjusted changes in height.
+Findings demonstrated a non-significant effect of handedness on the residualized height values, with left-handed individuals showing an average height change increase of .06 inches (r = .77, se = .08, p = .43) more than right-handed individuals.
+A visual inspection using a violin plot further supported these results, showing overlapping distributions of residualized height changes across handedness categories, indicating no substantial effect.
 
-::: panel-tabset
-### Write-up {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-The regression analysis was conducted to predict children's height at the 1-Year follow-up using their baseline height (`r names(merged_data)[2]`) and weight (`r names(merged_data)[3]`).
-Both predictors were statistically significant.
-Specifically, for every one-unit increase in baseline height, the height at Year_1 increased by approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units, while holding weight constant.
-Similarly, for every one-unit increase in baseline weight, the height at Year_1 increased by about `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units, while holding baseline height constant.
-The overall model explained a substantial `r round(model_summary$adj.r.squared * 100, 2)`% of the variance in height at the 1-Year follow-up, as indicated by the adjusted R-squared value.
-
-:::
-:::
+This approach of using residualized change scores can be particularly valuable when initial measurement levels may influence the outcome.
+In some context, this analysis can provide a more accurate reflection of change by accounting for baseline values, making it suitable for analyses where measurement of change is a primary aim and where baseline characteristics might otherwise obscure true changes.

### 2024-05-01 examples/9_Examples_LatentChangeScoresModels.qmd
update examples 9, 10, 13

---
title: "Latent Change Score Models"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be rendered
#| output: FALSE # Output from this code block will not be rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview
Latent change score models are a statistical technique used for analyzing changes over time by integrating aspects of both structural equation modeling and traditional change score methods. This approach constructs latent variables that directly represent change between time points. First, latent variables for each time point are modeled to capture the true score of the variable of interest, adjusting for measurement error. Then, latent change scores are defined as the difference between latent variables across consecutive time points.

This example will explore changes in height from baseline to 3 subsequent annual follow-ups in a sample of participants from the ABCD Study. Initially, latent variables for height at each time point are defined to account for measurement error and capture the true height scores. Latent change scores are then calculated between each pair of consecutive time points to model the true change in height. The dynamics of these changes are visually depicted using trajectory plots, providing a clear visual representation of growth patterns across participants the study.

## Preliminary Setup
### Load Packages
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
library(lme4) # Linear mixed-effects models
library(lcsm) # Latent Change Score Models
library(lavaan) # Structural Equation Modeling
library(kableExtra) # Table formatting
```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results 
### Compute Latent Change Score Model
The code below is used to compute a latent change score analysis to investigate changes in height across four annual assessments. This statistical approach models growth factors, including constant and proportional changes, and how they vary between individuals. It also evaluates the stability of growth patterns over time, giving a more nuanced understanding of developmental trajectories and the amount of change between each time point.
```{r}

# Reshape data from long to wide format
df_wide <- df_long %>%
    pivot_wider(
        id_cols = c(id),
        names_from = event,
        values_from = height,
        names_prefix = "Height_"
    )

df_wide <- na.omit(df_wide)

# Fit lcsm and save the results
uni_lavaan_results <- fit_uni_lcsm(
    data = df_wide,
    var = c(
        "Height_Baseline", "Height_Year_1",
        "Height_Year_2", "Height_Year_3"
    ),
    model = list(
        alpha_constant = TRUE,
        beta = TRUE,
        phi = TRUE
    )
)

uni_lavaan_results

# Save the lavaan syntax that is used to create the layout matrix for semPlot
uni_lavaan_syntax <- fit_uni_lcsm(
    data = df_wide,
    var = c(
        "Height_Baseline", "Height_Year_1",
        "Height_Year_2", "Height_Year_3"
    ),
    model = list(
        alpha_constant = TRUE,
        beta = TRUE,
        phi = TRUE
    ),
    return_lavaan_syntax = TRUE
)

# Now extract parameter estimates
# Only extract first 7 columns for this example by adding [ , 1:7]
param_bi_lcsm_01 <- extract_param(uni_lavaan_results, printp = TRUE)[, 1:6]

# Print table of parameter estimates
kable(param_bi_lcsm_01, digits = 3)

```

The results of the latent change score model (LCSM) for height indicated increasing patterns of growth. The fitted model included a constant change factor of .52, reflecting a uniform base growth factor across all individuals consistently over time. The proportional change factor of -.27 illustrates a negative relationship, indicating that higher initial heights might result in lesser incremental growth. The phi estimate .43 shows autoregressive effects of the change scores, where the direction and magnitude of past growth significantly predict subsequent changes.

### Model Plots
```{r}

# Path diagram plot of model results
plot_lcsm(
    lavaan_object = uni_lavaan_results,
    lavaan_syntax = uni_lavaan_syntax,
    edge.label.cex = .9,
    lcsm_colours = TRUE,
    lcsm = "univariate"
)

```

This path diagram represents the structural relationships specified in the LCSM by simplifying the model structure into a more intuitive graphical format. The diagram highlights connections and dependencies among the variables, as well as the model's estimated parameters.
```{r}
# Trajectory Plots
plot <- plot_trajectories(
    data = df_wide,
    id_var = "id",
    var_list = c(
        "Height_Baseline",
        "Height_Year_1",
        "Height_Year_2",
        "Height_Year_3"
    ),
    xlab = "Time", ylab = "X Score",
    connect_missing = FALSE,
    random_sample_frac = 0.3
)

plot
```

This trajectory plot depicts individual growth patterns based on a random sample of 30% of the data. The plot reveals individual variability in changes in height measurements among participants across 4 annual assessments, providing a visualization of how different individuals progress over time.

## Wrapping Up

This analysis utilized a latent change score model (LCSM) to study the progression of participants' height across multiple time points, controlling for baseline height measurements. Model results demonstrated a consistent growth pattern, with a constant change factor (γ = .52), which suggests a uniform growth rate across all individuals over time. The proportional change factor (β = -.27) indicated a negative relationship, revealing that individuals with greater initial heights experienced smaller incremental increases. Additionally, the autoregressive parameter (φ = 0.43) highlighted that previous growth trajectories significantly influence subsequent changes, providing insights into the dynamic nature of growth.

The LCSM offers a robust framework for analyzing longitudinal data, as it accounts for measurement errors and individual differences in baseline scores. This approach is particularly useful in developmental studies where understanding the intricacies of growth patterns is emphasized. This approach disentangles the elements of growth over time showing the predictive utility of earlier measurements and extending our understanding of growth dynamics.

### 2024-05-01 examples/_queue/1b_minimalist_Examples_DifferenceScores_SimpleRegression.qmd
update examples 9, 10, 13

--- 2024-05-01 examples/1b_Examples_DifferenceScores_SimpleRegression.qmd
@@
 ---
-title: "Difference Scores: Simple Regression"
+title: "Difference Scores: Simple Regression" ()
 author: "Biostatistics Working Group"
@@
     echo = T, message = F, warning = F, error = F, output = T,
-    comment = NA, cache = T,
+    comment = NA, cache = T, code_folding = T,
     R.options = list(width = 220, digits = 3),

### 2024-05-01 examples/_queue/3a_minimalist_Examples_LinearMixedModels.qmd
update examples 9, 10, 13

--- 2024-05-01 examples/3b_minimalist_Examples_LinearMixedModels.qmd
@@
 ---
-title: "Linear Mixed Models: Random Intercept and Slope"
+title: "Linear Mixed Models: Random Intercept"
 author: "Biostatistics Working Group"
@@
 df_long <- df %>%
-    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
-    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
-    drop_na(cbcl_scr_syn_external_r) %>%
+    select(src_subject_id, eventname, anthroheightcalc) %>%
+    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1")) %>%
+    drop_na(anthroheightcalc) %>%
     arrange(src_subject_id, eventname) %>%
@@
         src_subject_id = factor(src_subject_id),
-        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
-        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
+        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2")),
+        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
     ) %>%
@@
         event = eventname,
-        height = cbcl_scr_syn_external_r
+        height = anthroheightcalc
     )
-
-df_long <- df_long %>%
-  group_by(id) %>%
-  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
-  ungroup()
 
@@
 ## Overview
-
-The Linear mixed model (LMM) with a random intercept and slope extends the traditional fixed-effect linear regression by incorporating random effects for the intercept and slope parameters.
-This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.
+The Linear mixed model (LMM) with a random intercept is similar to traditional (fixed-effect) linear regression, extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value.
 
@@
 The primary aim is to characterize stability and change in participant's height, while accounting for observations that are clustered within youth over time.
-The LMM with a random intercept and slope will simultaneously estimate an overall sample mean (fixed effect) for the height trajectory's intercept and slope values, as well as subject-specific (random) effects that vary randomly about the mean intercept and slope values.
+The random-intercept LMM will simultaneously estimate an overall sample mean (fixed effect) for the height trajectory's intercept and slope values, as well as subject-specific (random) effects that vary randomly about the mean intercept value.
 
@@
 ```{r}
-#| echo: FALSE # The content of this code block will not be shown when rendered
-#| output: FALSE # Output from this code block will not be shown when rendered
-#| cache: FALSE
 
-# This code loads the data to be used for the example
+descriptives_table <- df_long %>%
+    select(event, height) %>%
+    mutate(event = factor(event)) %>%
+    tbl_summary(
+        by = event,
+        missing = "no",
+        label = list(height ~ "Height"),
+        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
+    ) %>%
+    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
+    bold_labels() %>%
+    italicize_levels() %>%
+    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
+theme_gtsummary_compact()
 
-library(tidyverse)
-
-# Set the data path
-data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
-
-# Read the data
-df <- readRDS(data_path)
-
-# Create and mutate the dataframe
-df_long <- df %>%
-    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
-    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
-    drop_na(cbcl_scr_syn_external_r) %>%
-    arrange(src_subject_id, eventname) %>%
-    mutate(
-        src_subject_id = factor(src_subject_id),
-        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
-        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
-    ) %>%
-    rename(
-        id = src_subject_id,
-        event = eventname,
-        height = cbcl_scr_syn_external_r
-    )
-
-df_long <- df_long %>%
-  group_by(id) %>%
-  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
-  ungroup()
+descriptives_table
 
@@
 ## Results
-### Compute LMM with Random Intercepts and Slopes
-This code fits a random-intercept and random-slope LMM to predict 'Height' across 3 annual visits, accounting for individual-level variability (random effects) of intercept and slope values across participants.
+### Compute LMM Model with Random Intercepts
+This code fits a radom-intercept LMM to predict 'Height' across 3 annual visits, accounting for individual-level variability (random effects) of intercept values across participants.
 Model output is provided below, as well as a brief summary of results.
 
-This model accounts for individual variability in height by including a random intercept and sllope for each subject (id).
+This model accounts for individual variability in height by including a random intercept for each subject (id).
 The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.
@@
 
-model <- lmer(height ~ event + (1 + event | id), data = df_long, REML = T)
-
-# Model with random intercepts only
-model_ri_only <- lmer(height ~ event + (1 | id), data = df_long, REML = T)
-model_ri_only
-
-# Model with random slopes only, without intercepts
-model_rs_only <- lmer(height ~ event + (0 + event | id), data = df_long, REML = T)
-model_rs_only
+model <- lmer(height ~ 1 + event + (1 | id), data = df_long, REML = T)
 
@@
 These results show an increasing trajectory of height across the years, with an increase of approximately 2.36 inches in Year 1 and 5.02 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates.
-The random intercept and slope effects suggests some between-subject variability in particpant's height values at baseline and in rates of change over time.
+The random intercepts effect suggests some between-subject variability in particpant's height values at baseline, with a standard deviation of about 3.03 inches.
 The model's residuals are mostly well-behaved, although the range suggests a few extreme values or outliers.
@@
 ### Model Plots
-The following set of plots are used to facilitate model diagnostics.
-The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects.
-The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity.
-The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.
-
-```{r}
-
-# 1. Extract the random effects
-random_effects <- ranef(model)[[1]]
-
-# 2. Convert to dataframe
-random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)
-
-# 3. hist_plot
-hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
-    geom_histogram(aes(y = after_stat(density)), bins = 30, color = "black", fill = "lightblue") +
-    labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
-    theme_minimal()
-
-print(hist_plot)
-
-```
+The following set of plots xxxxx.
 
@@
 # Generate the plot
-ggplot(df_long, aes(x = event, y = height, group = id)) +
+ggplot(df_long, aes(x = event, y = height)) +
     geom_line(aes(y = height, color = "Individual Trajectories"), alpha = 0.3) +
-    geom_line(data = df_long, aes(x = event, y = predicted_height, group = 1, color = "Mean Trajectory"), linewidth = 1.5) +
-    scale_color_manual(values = c("Individual Trajectories" = "grey", "Mean Trajectory" = "blue")) +
+    geom_line(aes(y = predicted_height + individual_intercept, group = id, color = "Adjusted Individual Trajectories"), alpha = 0.7) +
+    geom_line(aes(x = event, y = predicted_height, group = 1, color = "Sample Mean Trajectory"), size = 1.5) +
+    scale_color_manual(values = c("Individual Trajectories" = "red", "Adjusted Individual Trajectories" = "grey", "Sample Mean Trajectory" = "blue")) +
     labs(
-        title = "Individual and Group-Mean Height Trajectories",
+        title = "Individual Trajectories with Fixed Slope and Random Intercepts",
         x = "Event Name",
-        y = "Height (cm)",
+        y = "Height (inches)",
         color = "Trajectory Type"
@@
 The individual raw data points are shown as red lines, representing the original height measurements, which demonstrate the variability among participants at each time point.
-These are overlaid with grey lines, which adjust each individual's trajectory unique growth pattern indicated by the fixed effects from the model.
+These are overlaid with grey lines, which adjust each individual's trajectory to start from their unique intercept while following a common growth pattern indicated by the fixed effects from the model.
 The blue line represents the sample mean trajectory, calculated from the model's fixed effects, showcasing the average growth trend across all participants.
 
-The plot shows individual trajectories that vary around the model's predicted values due to random effects, indicating substantial inter-individual variability in baseline heights and rates of change in height over tim.
+The plot shows individual trajectories that vary around the model's predicted values due to random effects, indicating substantial inter-individual variability in baseline heights.
 On average, participants experienced an increase in height of approximately 2.36 inches from baseline to Year 1 and an additional 2.66 inches from Year 1 to Year 2.
-The random effects for the estimated intercept and slope suggest differences in starting heights and rates of change over time, reflecting the variability in growth patterns among the participants.
+The random effects for individual intercepts suggest differences in starting heights, with a standard deviation of 3.03 inches, reflecting the variability in growth patterns among the participants.
 
 ## Wrapping Up
+The random-intercept LMM analysis revealed an increasing pattern of change in children's height measured over 3 annual assessments.
+The mean intercept value showed an average initial height of 55.24 inches (sd = 3.03) with evidence of individual variability around this (random effect) estimate.
+An examination of the slope factor revealed year-to-year growth increments of approxmiately 2.5 inches, with the rate of change held constant across all participants.
 
-The random-intercept and random-slope LMM analysis revealed an increasing pattern of change in children's height measured over 3 annual assessments.
-The mean intercept value showed an average initial height of 55.24 inches (sd = 3.03) with evidence of individual variability around this (random effect) estimate.
-An examination of the slope factor revealed year-to-year growth increments of approxmiately 2.5 inches, with evidence of individual variability around this estimate.
-
-The LMM framework used in this example offered the ability to account for individual differences in initial height and changes in growth over time, demonstrating its utility in handling clustered or correlated data that arise from repeated measurements on the same subjects.
+The LMM framework used in this example offered the ability to account for individual differences in initial height and the consistent growth patterns over time, demonstrating its utility in handling clustered or correlated data that arise from repeated measurements on the same subjects.
 More generally, this approach is often particularly useful in developmental studies as it allows for the inclusion of both fixed effects, which estimate common parameters across all individuals, and random effects, which capture unique variations among individuals.

### 2024-05-01 examples/3a_Examples_LinearMixedModels.qmd
new examples updates

@@
 author: "Biostatistics Working Group"
-execute:
-  freeze: auto  # re-render only when source changes
-  cache: true
 ---
 
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+
+# This code configures knitr code chunk options
+
+knitr::opts_chunk$set(
+    echo = T, message = F, warning = F, error = F, output = T,
+    comment = NA, cache = T, code_folding = T,
+    R.options = list(width = 220, digits = 3),
+    fig.align = "center",
+    out.width = "75%", fig.asp = .75
+)
+```
+
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+#| cache: FALSE
+
+# This code loads the data to be used for the example
+
+library(tidyverse)
+
+# Set the data path
+data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
+
+# Read the data
+df <- readRDS(data_path)
+
+# Create and mutate the dataframe
+df_long <- df %>%
+    select(src_subject_id, eventname, anthroheightcalc) %>%
+    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1")) %>%
+    drop_na(anthroheightcalc) %>%
+    arrange(src_subject_id, eventname) %>%
+    mutate(
+        src_subject_id = factor(src_subject_id),
+        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2")),
+        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
+    ) %>%
+    rename(
+        id = src_subject_id,
+        event = eventname,
+        height = anthroheightcalc
+    )
+
+```
+
 ## Overview
+The Linear mixed model (LMM) with a random intercept is similar to traditional (fixed-effect) linear regression, extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value.
 
-<p>
-
-The LMM:ri is similar to traditional (fixed-effect) linear regression extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value zzzzt
-
-In this example, we will use the LMM:ri to analyze trajectories of height obtained across multiple measurement occasions in a sample of youth taking part in the ABCD Study.
-Our primary aim is to characterize stability and change in height assessments, while accounting for observations that are clustered within youth over time.
-To do so, we will use the LMM:ri to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory.
-
-</p>
+This example will examine trajectories of height obtained across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study.
+The primary aim is to characterize stability and change in participant's height, while accounting for observations that are clustered within youth over time.
+The random-intercept LMM will simultaneously estimate an overall sample mean (fixed effect) for the height trajectory's intercept and slope values, as well as subject-specific (random) effects that vary randomly about the mean intercept value.
 
 ## Preliminary Setup
-
-::: panel-tabset
-### Install Packages {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
-
-```{r pckg-install}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-# Create a list of required packages
-packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")
-
-# Check which packages are not installed and install them
-packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
-if (length(packages_to_install) > 0) {
-    install.packages(packages_to_install)
-}
-
-# Load the required packages
-lapply(packages_required, library, character.only = TRUE)
-
-```
-:::
-
 ### Load Packages
-
-::: blue
-**This code loads the r libraries necessary for this example**
-
-```{r lib-load}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
+```{r}
 
 library(tidyverse) # Collection of R packages for data science
-library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
+library(gtsummary) # Publication-ready tables
+library(rstatix) # Statistical Tests in tidy format
 library(lme4) # Linear mixed-effects models
-library(gtsummary)    # Publication-ready tables
-library(report)       # Easy reporting of regression analyses
-library(broom)        # Tidy and augment statistical models output
-library(gridExtra)    # Arrange multiple grid-based plots on a page
-
 ```
-:::
-
-```{r config}
-#| echo: FALSE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-### Config Options
-#::: blue
-#**This code configures knitr code chunk options**
-
-knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
-                      comment=NA, cache=T, code_folding=T,
-                      R.options=list(width=220, digits = 3),
-                      fig.align='center', 
-                      out.width='75%', fig.asp=.75)
-```
-:::
-
-## Descriptives Overview
-
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-**This code reads in and shows the data to be used in the current
-example**
-
-```{r read-data}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-#| cache: FALSE
-
-# Set the data paths
-data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
-data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
-
-# Read the data
-data_demographics <- readRDS(data_path_1)
-data_nonimaging <- readRDS(data_path_2)
-
-# Subset the nonimaging data to include desired variables
-selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
-subset_data <- data_nonimaging[, selected_vars]
-
-library(dplyr)
-# # Merge the datasets on 'src_subject_id' and 'eventname'
-merged_data <- data_demographics %>%
-  full_join(subset_data, by = c("src_subject_id", "eventname"))
-
-# Inspect the merged data structure
-str(merged_data)
-
-# Define event names to be retained in the analysis and convert variables to appropriate data types
-eventnames_to_include <- c("baseline_year_1_arm_1",
-                           "1_year_follow_up_y_arm_1",
-                           "2_year_follow_up_y_arm_1",
-                           "3_year_follow_up_y_arm_1",
-                           "4_year_follow_up_y_arm_1")
-
-df <- merged_data %>%
-  filter(eventname %in% eventnames_to_include) %>%
-  mutate(
-    src_subject_id = as.factor(src_subject_id),
-    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
-    age = as.numeric(age),
-    sex = as.factor(sex),
-    race.4level = as.factor(race.4level),
-    hisp = as.factor(hisp),
-    high.educ.bl = as.factor(high.educ.bl),
-    household.income.bl = as.factor(household.income.bl),
-    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
-    rel_family_id.bl = as.factor(rel_family_id.bl),
-    site_id_l = as.factor(site_id_l),
-    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
-    anthroweightcalc = as.numeric(anthroweightcalc),
-    anthroheightcalc = as.numeric(anthroheightcalc)
-  ) %>%
-  # Exclude cases from unused assessment waves
-  filter(!is.na(eventname))
-
-```
-:::
 
 ### Descriptives
+```{r}
 
-::: blue
-**This code creates a descriptives table**
-
-```{r descriptives}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-descriptives_table <- df %>%
-  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
-  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
-  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
+descriptives_table <- df_long %>%
+    select(event, height) %>%
+    mutate(event = factor(event)) %>%
     tbl_summary(
-    by = eventname,
+        by = event,
         missing = "no",
-    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
-                 anthroweightcalc ~ "Weight"),
-    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
+        label = list(height ~ "Height"),
+        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
     ) %>%
@@
 descriptives_table
-```
-:::
-:::
-
-## Results
-
-::: panel-tabset
-### Build Model {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-The code fits a linear mixed model to predict the 'Height' variable based on time points ('eventname') and handedness ('Handedness'), while accounting for individual-level variability by including random intercepts for each participant ('src_subject_id').
-The results of the model are then printed to provide a summary of the fitted model parameters.
-
-**STEP 1: Compute LMM Model with Random Intercepts**
-```{r build-model-1}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-#| cache: true
-
-## Linear Mixed Model with a random intercept (LMM-ri)
-model <- lmer(anthroweightcalc ~ 1 + eventname + sex + (1|src_subject_id), data = df, REML=T)
 
@@
 
-```{r model-output}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
+## Results
+### Compute LMM Model with Random Intercepts
+This code fits a radom-intercept LMM to predict 'Height' across 3 annual visits, accounting for individual-level variability (random effects) of intercept values across participants.
+Model output is provided below, as well as a brief summary of results.
 
-## Output and reports extending from the LMM-ri analyses
-#summary(model)
-#confint(model, level = 0.95, method = "Wald")
+This model accounts for individual variability in height by including a random intercept for each subject (id).
+The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.
 
-model_performance(model)
+```{r}
 
-model_parameters(model)
+model <- lmer(height ~ 1 + event + (1 | id), data = df_long, REML = T)
 
-model_parameters(model,standardize = "refit")
+## Output and summary statistics
+summary(model)
+confint(model, level = 0.95, method = "Wald")
 
@@
 
-The code provided executes a linear mixed model (LMM) to predict children's 
-height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4), and also takes into account their sex.
-This model accounts for individual variability in height by including a random intercept for each subject (src_subject_id).
-The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.
-
-:::
+These results show an increasing trajectory of height across the years, with an increase of approximately 2.36 inches in Year 1 and 5.02 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates.
+The random intercepts effect suggests some between-subject variability in particpant's height values at baseline, with a standard deviation of about 3.03 inches.
+The model's residuals are mostly well-behaved, although the range suggests a few extreme values or outliers.
 
 ### Model Plots
+```{r}
 
-::: blue
-The following set of plots are used to facilitate model diagnostics.
-The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects.
-The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity.
-The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.
+# Extract fixed effects
+fixed_effects <- fixef(model)
 
-```{r diagnosticplots}
+# Prepare data frame for plotting
+df_long$predicted_height <- predict(model, re.form = NA) # Overall fixed effect predictions
 
-# Extract the random effects
-random_effects <- ranef(model)[[1]]
+# Calculate individual intercept adjustments
+df_long$individual_intercept <- predict(model, re.form = ~ (1 | id)) - fixed_effects[1]
 
-# Convert to dataframe
-random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)
-
-# Plot 1: Histogram
-hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") + labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
-  theme_minimal()
-
-# Plot 2: Residuals vs Fitted Values
-resid_plot <- ggplot(NULL, aes(x = fitted_values, y = residuals)) +
-  geom_point(alpha = 0.5) +
-  geom_hline(yintercept = 0, linetype = "dashed") +
-  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
-  theme_minimal()
+# Generate the plot
+ggplot(df_long, aes(x = event, y = height)) +
+    geom_line(aes(y = height, color = "Individual Trajectories"), alpha = 0.3) +
+    geom_line(aes(y = predicted_height + individual_intercept, group = id, color = "Adjusted Individual Trajectories"), alpha = 0.7) +
+    geom_line(aes(x = event, y = predicted_height, group = 1, color = "Sample Mean Trajectory"), size = 1.5) +
+    scale_color_manual(values = c("Individual Trajectories" = "red", "Adjusted Individual Trajectories" = "grey", "Sample Mean Trajectory" = "blue")) +
+    labs(
+        title = "Individual Trajectories with Fixed Slope and Random Intercepts",
+        x = "Event Name",
+        y = "Height (inches)",
+        color = "Trajectory Type"
+    ) +
+    theme_minimal() +
+    theme(legend.position = "bottom")
 
@@
 
-```{r}
+This plot illustrates the change in participant's height across 3 annual measurement occassions.
+The individual raw data points are shown as red lines, representing the original height measurements, which demonstrate the variability among participants at each time point.
+These are overlaid with grey lines, which adjust each individual's trajectory to start from their unique intercept while following a common growth pattern indicated by the fixed effects from the model.
+The blue line represents the sample mean trajectory, calculated from the model's fixed effects, showcasing the average growth trend across all participants.
 
-library(ggplot2)
-theme_set(theme_modern(base_size = 6))
-
-plot(parameters(model)) +
-  ggplot2::labs(title = "A Dot-and-Whisker Plot")
-#plot(model)
-
-result <- estimate_expectation(model, data = "grid")
-plot(result)
-#plot(result, type = "qq")
-#plot(result, type = "pp")
-
-normality <- check_normality(model, effects = "random")
-#plot(normality)
-
-heteroscedasticity <- check_heteroscedasticity(model)
-#plot(heteroscedasticity)
-
-homogeneity <- check_homogeneity(model)
-#plot(homogeneity, type = "density")
-
-distribution <- check_distribution(model)
-#plot(distribution)
-
-#contrasts <- estimate_contrasts(model, contrast = "sex")
-#means <- estimate_means(model)
-#plot(means)
-#plot(contrasts, means)
-
-params <- model_parameters(model, effects = "fixed")
-#plot(params, show_labels = TRUE, size_text = 4)
-
-#check_model(model)
-
-```
-
-:::
-
-```{r trajectoryplot}
-
-# Extract the data frame used in the model
-model_data <- model@frame
-
-# Extract unique subject IDs from the model's data
-original_subject_ids <- unique(model_data$src_subject_id)
-
-# Subset the original data to include only those subjects
-df_subset <- df %>% filter(src_subject_id %in% original_subject_ids)
-
-eventname_map <- c(
-  "baseline_year_1_arm_1" = "Baseline",
-  "1_year_follow_up_y_arm_1" = "Year_1",
-  "2_year_follow_up_y_arm_1" = "Year_2",
-  "3_year_follow_up_y_arm_1" = "Year_3",
-  "4_year_follow_up_y_arm_1" = "Year_4"
-)
-
-# Apply the recoding to the eventname variable
-df_subset$eventname <- factor(df_subset$eventname, levels = names(eventname_map), labels = eventname_map)
-
-# Verify the recoding
-table(df_subset$eventname)
-
-# Generate the plot
-ggplot(df_subset, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +
-  # Individual estimated height trajectories in faded lines
-  geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +
-  # Overall group-mean trajectory in blue with increased thickness
-  stat_summary(aes(group = 1), fun = mean, geom = "line", color = "blue", linewidth = 1) +
-  labs(title = "Individual and Group-Mean Height Trajectories",
-       x = "Event Name",
-       y = "Height") +
-  theme_minimal()
-
-```
-
-```{r}
-#| echo: false
-
-## Creating variable (names) from output to be used for inline text
-
-library(broom.mixed)
-library(easystats)
-
-tidyfit <- tidy(model)
-tidyaugment <- augment(model)
-tidyglance <- glance(model)
-
-Intercept_estimate <- round(tidyfit[tidyfit$term == "(Intercept)", "estimate"], 3)
-Intercept_SE <- round(tidyfit[tidyfit$term == "(Intercept)", "std.error"], 3)
-
-sex_effect <- round(tidyfit[tidyfit$term == "sexMale or other", "estimate"], 3)
-
-# Extract values from the tibble for inline text
-Intercept_estimate <- Intercept_estimate$estimate
-Intercept_SE <- Intercept_SE$std.error
-sex_effect <-sex_effect$estimate
-
-#############
-
-#r eventname_estimate
-#r sex_effect
-#r conditional_R2
-#r conditional_R2*100**%
-#r marginal_R2
-
-```
-
-```{r}
-report(model)
-```
-:::
+The plot shows individual trajectories that vary around the model's predicted values due to random effects, indicating substantial inter-individual variability in baseline heights.
+On average, participants experienced an increase in height of approximately 2.36 inches from baseline to Year 1 and an additional 2.66 inches from Year 1 to Year 2.
+The random effects for individual intercepts suggest differences in starting heights, with a standard deviation of 3.03 inches, reflecting the variability in growth patterns among the participants.
 
 ## Wrapping Up
-<div class = "blue">
+The random-intercept LMM analysis revealed an increasing pattern of change in children's height measured over 3 annual assessments.
+The mean intercept value showed an average initial height of 55.24 inches (sd = 3.03) with evidence of individual variability around this (random effect) estimate.
+An examination of the slope factor revealed year-to-year growth increments of approxmiately 2.5 inches, with the rate of change held constant across all participants.
 
-::: panel-tabset
-### Write-up {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-The estimate of the random intercept was `r Intercept_estimate` (*SE* `r Intercept_SE`). The fixed effects section provides the coefficients for the intercept, eventname, and the sex variable.
-
-The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4) using the event name (eventname) and height (Height) variables.
-The intercept, corresponding to the reference levels of the predictors is estimated at `r Intercept_estimate` units, and this effect is highly significant.
-Regarding sex, boys show an increase in height of about `r sex_effect` (*SE* `r Intercept_SE`) units compared to girls.
-
-```{r}
-report_performance(model)
-```
-
-:::
-:::
+The LMM framework used in this example offered the ability to account for individual differences in initial height and the consistent growth patterns over time, demonstrating its utility in handling clustered or correlated data that arise from repeated measurements on the same subjects.
+More generally, this approach is often particularly useful in developmental studies as it allows for the inclusion of both fixed effects, which estimate common parameters across all individuals, and random effects, which capture unique variations among individuals.
+This enables a more precise understanding of both the group-level trends and individual-specific trajectories in longitudinal data.

### 2024-05-01 examples/examples.qmd
new examples updates

@@
 				  <li class="tag__item play blue">
-						<a href="./_9_Examples_LatentChangeScoresModels.qmd"><i class="fas fa-play
+						<a href="./9_Examples_LatentChangeScoresModels.qmd"><i class="fas fa-play
 						mr-2"></i>Latent Change Score Models</a>
@@
 					<li class="tag__item play blue">
-						<a href="./_10_Examples_LatentGrowthCurveModels.qmd"><i class="fas fa-play
+						<a href="./10_Examples_LatentGrowthCurveModels.qmd"><i class="fas fa-play
 						mr-2"></i>Latent Growth Curve Models</a>
@@
 						mr-2"></i>Latent Transition Analysis</a>
+						</li>
+					<li class="tag__item play blue">
+						<a href="./13b_Examples_LatentClassGrowthAnalysis.qmd"><i class="fas fa-play
+						mr-2"></i>Latent Class Growth Analysis</a>
 						</li>

### 2024-05-22 examples/5_Examples_MarginalModels.qmd
update examples 10-14

---
title: "Marginal Models"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, output=T,
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview
Marginal models, also known as population-averaged models, are used in the analysis of longitudinal data where the focus is on the overall population effect rather than individual-specific effects. These models account for the correlation between repeated measures within subjects by using generalized estimating equations (GEE) to provide robust standard errors and efficient parameter estimates.

This example will examine height across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to characterize the average growth pattern in height over time at the population level. Marginal models will be employed to estimate the overall mean height at each time point, while accounting for the correlation of repeated measurements within individuals, thus providing a comprehensive understanding of height changes across the entire sample.

## Preliminary Setup

### Load R libraries
```{r}

library(tidyverse)    # Collection of R packages for data science
library(lme4)         # Linear mixed-effects models
library(geepack)      # Generalized Estimating Equations (GEE) analysis
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page
library(easystats)    #
library(gtsummary)    #
library(marginaleffects) #
```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results
### Compute Marginal Model Analysis
This code fits a Marginal Model to predict 'Height' across 3 annual visits, focusing on the overall population-level effect. The model output is provided below, along with a summary of results. This approach accounts for the correlation between repeated measurements within individuals using generalized estimating equations (GEE), providing robust standard errors and efficient parameter estimates for the average height at each time point.

```{r}

df_long <- df_long %>%
  filter(!is.na(height) & !is.na(event) & !is.na(id))

# Fitting the longitudinal marginal model
model <- geeglm(height ~ event, data = df_long, id = id, family = gaussian(), corstr = "exchangeable")

summary(model)

```

These results show a significant increase in height across the years, with an estimated increase of approximately 2.41 inches in Year 1 and 5.20 inches by Year 2 compared to the baseline, as indicated by the fixed effects estimates. The correlation structure of the repeated measurements is modeled using an exchangeable correlation matrix, which suggests that the measurements are equally correlated within individuals. The significant Wald statistics and p-values (< 2e-16) for each year indicate strong evidence of height changes over time in the population.

### Model Plots
```{r}

# Calculate mean height and confidence intervals for each time point
summary_data <- df_long %>%
  group_by(event) %>%
  summarize(
    mean_height = mean(height, na.rm = TRUE),
    ci_lower = mean_height - qt(0.975, df = n() - 1) * sd(height, na.rm = TRUE) / sqrt(n()),
    ci_upper = mean_height + qt(0.975, df = n() - 1) * sd(height, na.rm = TRUE) / sqrt(n())
  )

# Generate the plot
ggplot(summary_data, aes(x = event, y = mean_height)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2, color = "blue") +
  labs(
    title = "Mean Height Across Time Points with Confidence Intervals",
    x = "Time Point",
    y = "Mean Height (inches)"
  ) +
  theme_minimal()

```
This plot illustrates the mean height of participants across the three annual measurement occasions, along with 95% confidence intervals. The blue line and points represent the average height at each time point, while the error bars show the range within which we are 95% confident the true mean height lies.

The plot reveals a clear increasing trend in height over time, with mean height values rising at each subsequent measurement occasion. The confidence intervals provide a sense of the variability and precision of these mean estimates, indicating that the observed increase in height is statistically significant. This visualization highlights the overall growth pattern in the sample, capturing the population-level trends effectively.

## Wrapping Up
The marginal model analysis revealed a significant increase in children's height measured over 3 annual assessments. The mean intercept value showed an average initial height of 55.12 inches, with year-to-year growth increments of approximately 2.41 inches in Year 1 and 5.20 inches by Year 2 compared to the baseline.

The marginal model framework, using generalized estimating equations (GEE), effectively captured the population-averaged trends while accounting for the correlation of repeated measurements within individuals. This approach demonstrated its utility in handling clustered data from repeated measurements on the same subjects. The exchangeable correlation structure indicated that the measurements were equally correlated within individuals, and the significant Wald statistics provided strong evidence of height changes over time.

Overall, the marginal model approach is particularly useful in developmental studies as it estimates common parameters across all individuals, offering a clear understanding of population-level trends in longitudinal data. This method is valuable for examining average growth patterns while accounting for the correlation in repeated measurements.

### 2024-05-22 examples/6_Examples_GeneralizedEstimatingEquations.qmd
update examples 10-14

---
title: "Generalized Estimating Equations"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, output=T,
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview

Generalized Estimating Equations (GEE) are an extension of generalized linear models used for analyzing longitudinal and other correlated response data. GEEs provide a method for estimating the parameters of a generalized linear model with and are particularly useful when dealing with repeated measurements or clustered data, offering robust standard errors that account for within-subject correlations.

This example will examine height across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to characterize the average growth pattern in height over time at the population level. GEEs will be employed to estimate the mean height at each time point, accounting for the correlation of repeated measurements within individuals. This approach will help us understand how height changes over time while providing robust estimates that account for the inherent correlation in the repeated measures data.

## Preliminary Setup
### Load R libraries
```{r}

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(geepack)      # Generalized Estimating Equations (GEE) analysis

```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results
### Compute Generalized Estimating Equations Analysis
This code fits a Generalized Estimating Equations (GEE) model to predict 'Height' across 3 annual visits, accounting for within-subject correlation of repeated measurements. The model output is provided below, along with a summary of results. This approach uses GEEs to estimate the population-averaged effects, providing robust standard errors and accommodating the correlation structure of the repeated height measurements.

```{r}

model <- geeglm(height ~ event, data = df_long, id = id, family = gaussian(), corstr = "exchangeable")

summary(model)
```

These results indicate a significant increase in height across the years, with an increase of approximately 2.41 inches in Year 1 and 5.20 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates. The exchangeable correlation structure suggests a consistent correlation between repeated measurements within individuals, with an estimated correlation parameter (alpha) of 0.681. The significant Wald statistics and p-values (< 2e-16) for each year indicate strong evidence of height changes over time in the population.

### Model Plots
```{r}

# Fit the GEE model
model <- geeglm(height ~ event, data = df_long, id = id, family = gaussian(), corstr = "exchangeable")

# Extract the predicted means and standard errors
predicted_values <- data.frame(
  event = unique(df_long$event),
  mean_height = predict(model, newdata = data.frame(event = unique(df_long$event)), type = "response"),
  se = sqrt(diag(vcov(model)))
)

# Calculate confidence intervals
predicted_values <- predicted_values %>%
  mutate(
    ci_lower = mean_height - 1.96 * se,
    ci_upper = mean_height + 1.96 * se
  )

# Generate the plot
ggplot(predicted_values, aes(x = event, y = mean_height)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2, color = "blue") +
  labs(
    title = "Mean Height Across Time Points with Confidence Intervals (GEE)",
    x = "Time Point",
    y = "Mean Height (inches)"
  ) +
  theme_minimal()

```

This plot illustrates the mean height of participants across three annual measurement occasions, as estimated by the generalized estimating equations (GEE) model. The blue line and points represent the estimated mean height at each time point, while the error bars show the 95% confidence intervals around these estimates.

The plot reveals a clear increasing trend in height over time, with mean height values rising at each subsequent measurement occasion. The confidence intervals provide a sense of the variability and precision of these mean estimates, indicating that the observed increase in height is statistically significant. This visualization highlights the overall growth pattern in the sample, capturing the population-averaged trends effectively.

## Wrapping Up

The marginal model analysis using Generalized Estimating Equations (GEE) revealed a significant increase in children's height measured over 3 annual assessments. The mean intercept value showed an average initial height of 55.12 inches. The fixed effects estimates indicated year-to-year growth increments of approximately 2.41 inches in Year 1 and 5.20 inches by Year 2 compared to the baseline.

The GEE framework employed in this example captured the population-averaged trends while accounting for the correlation of repeated measurements within individuals. The exchangeable correlation structure suggested that measurements were consistently correlated within individuals, with an estimated correlation parameter (alpha) of 0.681. The significant Wald statistics and p-values (< 2e-16) provided strong evidence of height changes over time.

The marginal model approach is particularly useful in developmental studies as it estimates common parameters across all individuals, offering a clear understanding of population-level trends in longitudinal data. This method is valuable for examining average growth patterns while accounting for the correlation in repeated measurements, providing insights into the overall changes in height over time.

### 2024-05-22 examples/7_Examples_GeneralizedLinearMixedEffectsModels.qmd
update examples 10-14

---
title: "Generalized Linear Mixed Effects Models"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, output=T,
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, ehi_y_ss_scoreb) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    #drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        ehi_y_ss_scoreb = as.numeric(as.character(ehi_y_ss_scoreb)) # Ensure numeric before transformation
  ) %>%
  mutate(
    ehi_y_ss_scoreb = case_when(
      ehi_y_ss_scoreb == 1 ~ 1,
      ehi_y_ss_scoreb %in% c(2, 3) ~ 2,
      TRUE ~ NA_real_   # Maintain NAs as they are
    )
  ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        handedness = ehi_y_ss_scoreb
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview
Generalized Linear Mixed Models (GLMMs) extend generalized linear models to include both fixed and random effects, allowing for the analysis of non-normal response variables and accounting for the correlation within clustered or repeated measures data. GLMMs are particularly useful when dealing with hierarchical data structures, offering a flexible framework to model various types of response distributions.

This example will examine height across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to characterize the growth pattern in height over time, while accounting for the repeated measures within individuals. GLMMs will be used to estimate both the overall population effects (fixed effects) and individual-specific deviations (random effects) from the population mean, providing a detailed understanding of height changes over time with the ability to model different types of response distributions.

## Preliminary Setup
### Load R libraries
```{r}

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(lme4)         # Linear mixed-effects models

```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, handedness) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(handedness ~ "Handedness"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results
### Compute Generalized Linear Mixed Models
This code fits a Generalized Linear Mixed Model (GLMM) to predict 'Height' across 3 annual visits, incorporating both fixed effects for overall population trends and random effects to account for individual-level variability. The model output is provided below, along with a summary of results. This model accounts for the hierarchical structure of the data, with random intercepts capturing individual deviations from the population mean height trajectory.

```{r}

# Fit the LMM model
model <- lmer(height ~ 1 + event + (1 | id), data = df_long, REML = TRUE)

# Extract fixed effects
fixed_effects <- fixef(model)

summary(model)
summary(fixed_effects)

```

These results show an increasing trajectory of height across the years, with an increase of approximately 2.42 inches in Year 1 and 4.93 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates. The random intercepts effect suggests some between-subject variability in participants' height values at baseline, with a standard deviation of about 3.08 inches. The model's residuals are mostly well-behaved, although the range suggests a few extreme values or outliers.

### Model Plots
```{r}

# Prepare data frame for plotting
df_long$predicted_height <- predict(model, re.form = NA) # Overall fixed effect predictions

# Calculate individual intercept adjustments
df_long$individual_intercept <- predict(model, re.form = ~ (1 | id)) - fixed_effects[1]

# Combine fixed effects and random effects predictions
df_long$combined_predicted_height <- df_long$predicted_height + df_long$individual_intercept

# Generate the plot
ggplot(df_long, aes(x = event, y = height)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = combined_predicted_height, group = id), color = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted_height, group = 1), color = "blue", size = 1.5) +
  labs(
    title = "Predicted vs. Observed Heights with GLMM",
    x = "Event",
    y = "Height (inches)"
  ) +
  theme_minimal()

```

This plot illustrates the predicted versus observed heights across three annual measurement occasions, as estimated by the Generalized Linear Mixed Model (GLMM). The scatter plot shows actual height measurements for each individual, highlighting variability among participants. The grey lines represent individual predicted heights, accounting for both fixed effects and random intercepts, while the blue line shows the population-level trend.

The plot reveals a clear increasing trend in height over time, with predicted height values rising at each measurement occasion. The grey lines indicate substantial individual variability, whereas the blue line captures the overall trend. This visualization effectively highlights both the population-averaged growth pattern and individual differences in height changes over time, demonstrating the GLMM's ability to model these dynamics.

## Wrapping Up
The marginal model analysis using Generalized Linear Mixed Effects Models (GLMM) revealed a significant increasing pattern of change in children's height measured over 3 annual assessments. The mean intercept value showed an average initial height of 55.29 inches (sd = 3.08), with evidence of individual variability around this estimate. The fixed effects estimates indicated year-to-year growth increments of approximately 2.42 inches in Year 1, 4.93 inches in Year 2, and 7.06 inches in Year 3 compared to the baseline.

The GLMM framework used in this example effectively accounted for both fixed effects, which estimate common parameters across all individuals, and random effects, which capture unique variations among individuals. This approach demonstrated its utility in handling clustered or correlated data arising from repeated measurements on the same subjects. The random intercepts effect suggested some between-subject variability in participants' height values at baseline, with a standard deviation of about 3.08 inches.

More generally, the GLMM approach is particularly useful in developmental studies as it allows for a more precise understanding of both group-level trends and individual-specific trajectories in longitudinal data. By accommodating individual differences in initial height and varying growth patterns over time, this method provides a comprehensive view of the dynamics of height changes in children.

### 2024-05-22 examples/8_Examples_AutoregressiveCrosslaggedPanelModels.qmd
update examples 10-14

---
title: "Autoregressive Crosslagged Panel Models"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc, anthroweightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc, anthroweightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2),
        anthroweightcalc = round(as.numeric(anthroweightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc,
        weight = anthroweightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview
Autoregressive Cross-Lagged Panel Analysis (ARCL) is a statistical technique used to examine the reciprocal relationships between variables over time. This method incorporates both autoregressive paths, which account for the stability of each variable across time points, and cross-lagged paths, which assess the influence of one variable on another over subsequent time points.

This example will examine the height and weight across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to explore the reciprocal relationships between height and weight at different time points, analyzing how height at one time point predicts weight at future time points (and vice-versa). ARCL will be used to model both the stability of height and weight over time (autoregressive paths) and the potential cross-lagged influences between consecutive measurements of these two constructs.

## Preliminary Setup
### Load R libraries
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
library(lavaan) # Structural Equation Modeling in R
library(semPlot) #

```

### Descriptives
```{r}
descriptives_table <- df_long %>%
    select(event, height, weight) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height", weight ~ "Weight"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

## Results
### Compute Autoregressive Crosslagged Panel Analysis
This code performs an Autoregressive Cross-Lagged Panel Analysis (ARCL) to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits. The model output is provided below, along with a summary of results. This approach models both the stability of height and weight over time (autoregressive paths) and the potential cross-lagged effects between these constructs.

```{r}

# Reshape data from long to wide format
df_wide <- df_long %>%
    pivot_wider(
        id_cols = c(id),
        names_from = event,
        values_from = c(height, weight),
        names_sep = "_"
    )

df_wide <- na.omit(df_wide)

model <- "
  # Autoregressive paths
  weight_Year_3 ~ a2 * weight_Year_2
  weight_Year_2 ~ a1 * weight_Year_1
  weight_Year_1 ~ a2 * weight_Baseline
  height_Year_3 ~ a2 * height_Year_2
  height_Year_2 ~ a1 * height_Year_1
  height_Year_1 ~ a2 * height_Baseline

# Cross-lagged paths
height_Year_3 ~ b1 * weight_Year_2
height_Year_2 ~ b1 * weight_Year_1
height_Year_1 ~ b2 * weight_Baseline
weight_Year_3 ~ b1 * height_Year_2
weight_Year_2 ~ b1 * height_Year_1
weight_Year_1 ~ b2 * height_Baseline
"

# Fit the model to your data (replace 'your_data' with the name of your data frame)
fit <- sem(model, data = df_wide)

# View the results
summary(fit)

```

These results show significant autoregressive effects for both height and weight, indicating that measurements at each time point strongly predict subsequent measurements (e.g., height at Year 2 significantly predicts height at Year 3 with an estimate of 1.05, p < .001). The cross-lagged effects indicate a small but significant negative influence of weight on height at later time points (e.g., weight at Year 2 negatively affects height at Year 3 with an estimate of -0.007, p < .001). The covariances between height and weight at Year 3 are also significant, suggesting a relationship between these variables at the final time point.

### Model Plots
```{r}

# Fit and plot the model with adjusted label positions and curved edges
# Assume 'fit' is your fitted SEM model
semPlotObj <- semPaths(
    fit,
    whatLabels = "est", # Show estimated coefficients
    edge.label.cex = 0.6, # Smaller font size for edge labels
    layout = "tree",
    rotation = 2,
    edge.label.position = 0.75, # Moves labels closer to the destination node
    edge.curve = 0.3, # Increase curvature to separate labels from the path
)

# Adding a title using base R graphics functions
title(main = "Autoregressive Cross-Lagged Panel Model for Height and Weight")

# Redraw the plot with potentially adjusted parameters for better visibility
print(semPlotObj)

```

This plot illustrates the autoregressive cross-lagged panel model for height and weight across 3 annual measurement occasions. The model includes both autoregressive paths (predicting a variable from its previous time point) and cross-lagged paths (predicting a variable from the other variable at a previous time point).

The autoregressive paths show that height and weight at each time point strongly predict their respective values at subsequent time points. For example, height at Year 2 significantly predicts height at Year 3. These paths are indicated by curved lines connecting each time point for height and weight, with coefficients showing the strength of these predictions.

The cross-lagged paths reveal how height and weight influence each other over time. For instance, weight at Year 2 negatively affects height at Year 3, as indicated by the coefficients on the curved paths connecting weight at one time point to height at the next.

The overall model highlights the strong stability of height and weight over time (autoregressive effects) and the modest influence of weight on height and vice versa (cross-lagged effects). This plot helps visualize the dynamic interplay between height and weight, providing insights into how these variables evolve together over time.

## Wrapping Up
The autoregressive cross-lagged panel model analysis revealed significant relationships between children's height and weight measured over 3 annual assessments. The model showed strong autoregressive effects, indicating that height and weight measurements at each time point strongly predict their subsequent measurements. For example, height at Year 2 significantly predicts height at Year 3 with an estimate of 1.05 (p < .001).

Additionally, the cross-lagged effects revealed a small but significant negative influence of weight on height at later time points. Specifically, weight at Year 2 negatively affects height at Year 3 with an estimate of -0.007 (p < .001). The significant covariances between height and weight at Year 3 suggest a relationship between these variables at the final time point.

The autoregressive cross-lagged panel model framework used in this example demonstrated its utility in capturing both the stability of measurements over time and the dynamic interplay between height and weight. This approach is particularly useful in developmental studies, as it allows for the examination of reciprocal influences between variables over time, providing a comprehensive understanding of the growth trajectories and their interdependencies.

More generally, this method is valuable for understanding both individual-specific trajectories and the relationships between multiple developmental indicators in longitudinal data. By accounting for the autoregressive and cross-lagged effects, this model offers a nuanced perspective on how different aspects of growth interact and influence each other over time.

### 2024-05-22 examples/_queue/4_Examples_SignedRankTest.qmd
update examples 10-14

---
title: "Signed-Rank Test"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be rendered
#| output: FALSE # Output from this code block will not be rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview
The Signed Rank-Sum test is a non-parametric method used to compare paired samples or repeated measures. Unlike traditional parametric tests, it does not assume normality and is particularly useful when dealing with skewed distributions or ordinal data. This method extends the Wilcoxon signed rank-sum test to handle multiple measurements over time, making it suitable for longitudinal data analysis.

This example will examine changes in height across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to determine whether there are significant changes in height over time, while accounting for observations that are clustered within youth over time. The signed rank-sum test will help us identify whether height levels have significantly increased or decreased across the three time points without making parametric assumptions about the data distribution.

## Preliminary Setup
### Load Packages
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
library(easystats) # 
library(lme4) # Linear mixed-effects models
```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results 
### Compute Signed Rank Test
This code performs a Signed Rank-Sum Test to compare 'Height' across 3 annual visits, focusing on the changes in height over time. The output of the test is provided below, along with a summary of results. This non-parametric test accounts for the paired nature of the observations and determines whether there are significant changes in height across the three time points without assuming a specific distribution for the data.

```{r}
# Reshape data from long to wide format
df_wide <- df_long %>%
    pivot_wider(
        id_cols = c(id),
        names_from = event,
        values_from = height,
        names_prefix = "Height_"
    )

df_wide$Height_Baseline <- as.numeric(as.character(df_wide$Height_Baseline))
df_wide$Height_Year_1 <- as.numeric(as.character(df_wide$Height_Year_1))

df_wide_complete <- na.omit(df_wide[, c("Height_Baseline", "Height_Year_1")])

model <- wilcox.test(df_wide_complete$Height_Baseline, df_wide_complete$Height_Year_1, paired = TRUE)

report(wilcox.test(df_wide_complete$Height_Baseline, df_wide_complete$Height_Year_1))

```
These results indicate a significant increase in height from baseline to Year 1, as evidenced by the Wilcoxon signed rank-sum test. The test statistic (W = 2.41e+06) and the negative rank biserial correlation (r = -0.39) suggest a large effect size, with the p-value (< .001) confirming the statistical significance of the observed changes. This analysis shows that participants' heights have significantly increased from baseline to Year 1.

### Model Plots
```{r}

ggplot(df_long, aes(x = event, y = height, group = id)) +
  geom_line(aes(group = id), alpha = 0.3, color = "grey") +  # Individual trajectories
  geom_boxplot() +  # Boxplot for summary statistics
  labs(title = "Comparison of Height Measurements Across Time Points",
       x = "Time Point", y = "Height (inches)") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend, if not necessary

```
This plot illustrates the change in participants' height across 3 annual measurement occasions using a combination of individual trajectories and boxplots. The grey lines represent individual participants' height trajectories, demonstrating the variability among participants at each time point. The boxplots summarize the distribution of height measurements at each time point, highlighting the median, interquartile range, and potential outliers.

The plot reveals significant changes in height over time, with the median height increasing across the three measurement occasions. The individual trajectories show that while most participants experience an increase in height, the rate of growth varies. This visualization helps to identify trends and patterns in height changes, as well as the variability in growth among participants. The boxplots provide a clear summary of the central tendency and spread of height measurements at each time point, indicating that the overall height distribution shifts upwards over time.

## Wrapping Up
The signed rank-sum test analysis revealed significant changes in children's height measured over 3 annual assessments. The Wilcoxon rank sum test with continuity correction between baseline and Year 1 indicated a significant increase in height (W = 2.41e+06, p < .001; r (rank biserial) = -0.39, 95% CI [-0.42, -0.37]). These results suggest a large effect size, with the p-value confirming the statistical significance of the observed changes. Participants' heights significantly increased from baseline to Year 1. This non-parametric method was particularly useful for handling the paired nature of the data and did not assume normality, making it suitable for skewed distributions often encountered in growth measurements.

The signed rank-sum test provided insights into overall changes in height without requiring parametric assumptions, demonstrating its utility in analyzing longitudinal data. This approach is valuable in developmental studies as it robustly detects significant changes over time, even with non-normal data distributions. By focusing on rank differences, the test highlights meaningful changes within individuals, offering a straightforward yet powerful method to examine growth patterns across repeated measurements.

### 2024-05-22 examples/_queue/_11_Examples_MultivariateLatentGrowthCurveModels.qmd
update examples 10-14

--- 2024-05-22 examples/8_Examples_AutoregressiveCrosslaggedPanelModels.qmd
@@
 ---
-title: "Autoregressive Crosslagged Panel Models"
+title: "Multivariate Latent Growth Curve Models"
 author: "Biostatistics Working Group"
@@
 ## Overview
-Autoregressive Cross-Lagged Panel Analysis (ARCL) is a statistical technique used to examine the reciprocal relationships between variables over time.
-This method incorporates both autoregressive paths, which account for the stability of each variable across time points, and cross-lagged paths, which assess the influence of one variable on another over subsequent time points.
 
-This example will examine the height and weight across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study.
-The primary aim is to explore the reciprocal relationships between height and weight at different time points, analyzing how height at one time point predicts weight at future time points (and vice-versa).
-ARCL will be used to model both the stability of height and weight over time (autoregressive paths) and the potential cross-lagged influences between consecutive measurements of these two constructs.
+In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
+Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time.
+XXXX models facilitate this by xxxxx.
 
@@
 ```{r}
-
 library(tidyverse)    # Collection of R packages for data science
-library(gtsummary) # Publication-ready tables
-library(rstatix) # Statistical Tests in tidy format
-library(lavaan) # Structural Equation Modeling in R
-library(semPlot) #
-
+library(rstatix)      # Pipe-friendly framework for basic statistical tests
+library(DT)           # Rendering interactive data tables
+library(lavaan)       #
 ```
@@
 ## Results
-### Compute Autoregressive Crosslagged Panel Analysis
-This code performs an Autoregressive Cross-Lagged Panel Analysis (ARCL) to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits.
+### Compute Multivariate Latent Growth Curve Model
+This code performs multivariate latent growth curve to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits.
 The model output is provided below, along with a summary of results.
-This approach models both the stability of height and weight over time (autoregressive paths) and the potential cross-lagged effects between these constructs.
+This approach xxxxx.
 
 ```{r}
-
 # Reshape data from long to wide format
@@
 
-model <- "
-  # Autoregressive paths
-  weight_Year_3 ~ a2 * weight_Year_2
-  weight_Year_2 ~ a1 * weight_Year_1
-  weight_Year_1 ~ a2 * weight_Baseline
-  height_Year_3 ~ a2 * height_Year_2
-  height_Year_2 ~ a1 * height_Year_1
-  height_Year_1 ~ a2 * height_Baseline
+# Load necessary library
+library(lavaan)
 
-# Cross-lagged paths
-height_Year_3 ~ b1 * weight_Year_2
-height_Year_2 ~ b1 * weight_Year_1
-height_Year_1 ~ b2 * weight_Baseline
-weight_Year_3 ~ b1 * height_Year_2
-weight_Year_2 ~ b1 * height_Year_1
-weight_Year_1 ~ b2 * height_Baseline
-"
+# Specify the multivariate latent growth curve model
+model <- '
+  # Intercept and slope factors for variable 1 (height)
+  i_var1 =~ 1*height_Baseline + 1*height_Year_1 + 1*height_Year_2
+  s_var1 =~ 0*height_Baseline + 1*height_Year_1 + 2*height_Year_2
 
-# Fit the model to your data (replace 'your_data' with the name of your data frame)
-fit <- sem(model, data = df_wide)
+  # Intercept and slope factors for variable 2 (weight)
+  i_var2 =~ 1*weight_Baseline + 1*weight_Year_1 + 1*weight_Year_2
+  s_var2 =~ 0*weight_Baseline + 1*weight_Year_1 + 2*weight_Year_2
+'
 
-# View the results
-summary(fit)
+# Fit the model using FIML for handling missing data
+fit <- sem(model, data = df_wide, missing = "fiml")
+
+# Check the summary to identify potential issues
+summary(fit, fit.measures = TRUE, standardized = TRUE, rsquare = TRUE)
 
@@
 
-These results show significant autoregressive effects for both height and weight, indicating that measurements at each time point strongly predict subsequent measurements (e.g., height at Year 2 significantly predicts height at Year 3 with an estimate of 1.05, p < .001).
-The cross-lagged effects indicate a small but significant negative influence of weight on height at later time points (e.g., weight at Year 2 negatively affects height at Year 3 with an estimate of -0.007, p < .001).
-The covariances between height and weight at Year 3 are also significant, suggesting a relationship between these variables at the final time point.
+These results show xxxxx, indicating that xxxxx (e.g., xxxxx). The xxxx effects indicate xxxx.
 
@@
 
-# Fit and plot the model with adjusted label positions and curved edges
-# Assume 'fit' is your fitted SEM model
-semPlotObj <- semPaths(
-    fit,
-    whatLabels = "est", # Show estimated coefficients
-    edge.label.cex = 0.6, # Smaller font size for edge labels
-    layout = "tree",
-    rotation = 2,
-    edge.label.position = 0.75, # Moves labels closer to the destination node
-    edge.curve = 0.3, # Increase curvature to separate labels from the path
-)
-
-# Adding a title using base R graphics functions
-title(main = "Autoregressive Cross-Lagged Panel Model for Height and Weight")
-
-# Redraw the plot with potentially adjusted parameters for better visibility
-print(semPlotObj)
-
 ```
 
-This plot illustrates the autoregressive cross-lagged panel model for height and weight across 3 annual measurement occasions.
-The model includes both autoregressive paths (predicting a variable from its previous time point) and cross-lagged paths (predicting a variable from the other variable at a previous time point).
-
-The autoregressive paths show that height and weight at each time point strongly predict their respective values at subsequent time points.
-For example, height at Year 2 significantly predicts height at Year 3.
-These paths are indicated by curved lines connecting each time point for height and weight, with coefficients showing the strength of these predictions.
-
-The cross-lagged paths reveal how height and weight influence each other over time.
-For instance, weight at Year 2 negatively affects height at Year 3, as indicated by the coefficients on the curved paths connecting weight at one time point to height at the next.
-
-The overall model highlights the strong stability of height and weight over time (autoregressive effects) and the modest influence of weight on height and vice versa (cross-lagged effects).
-This plot helps visualize the dynamic interplay between height and weight, providing insights into how these variables evolve together over time.
+This plot illustrates xxxx.
 
 ## Wrapping Up
-The autoregressive cross-lagged panel model analysis revealed significant relationships between children's height and weight measured over 3 annual assessments.
-The model showed strong autoregressive effects, indicating that height and weight measurements at each time point strongly predict their subsequent measurements.
-For example, height at Year 2 significantly predicts height at Year 3 with an estimate of 1.05 (p < .001).
-
-Additionally, the cross-lagged effects revealed a small but significant negative influence of weight on height at later time points.
-Specifically, weight at Year 2 negatively affects height at Year 3 with an estimate of -0.007 (p < .001).
-The significant covariances between height and weight at Year 3 suggest a relationship between these variables at the final time point.
-
-The autoregressive cross-lagged panel model framework used in this example demonstrated its utility in capturing both the stability of measurements over time and the dynamic interplay between height and weight.
-This approach is particularly useful in developmental studies, as it allows for the examination of reciprocal influences between variables over time, providing a comprehensive understanding of the growth trajectories and their interdependencies.
-
-More generally, this method is valuable for understanding both individual-specific trajectories and the relationships between multiple developmental indicators in longitudinal data.
-By accounting for the autoregressive and cross-lagged effects, this model offers a nuanced perspective on how different aspects of growth interact and influence each other over time.
+The multivariate latent growth curve revealed xxxxxxxx.

### 2024-05-22 examples/_queue/_12_Examples_LatentTransitionAnalysis.qmd
update examples 10-14

--- 2024-05-22 examples/_queue/_11_Examples_MultivariateLatentGrowthCurveModels.qmd
@@
 ---
-title: "Multivariate Latent Growth Curve Models"
+title: "Latent Transition Analysis"
 author: "Biostatistics Working Group"
@@
 
+XXXXXXXXXXX Models xxxxxxx. This xxxxxxx.
+
 In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study.
@@
 library(DT)           # Rendering interactive data tables
-library(lavaan)       #
+library(tidyLPA)      # XXXXXXXX
+
 ```
@@
 ## Results 
-### Compute Multivariate Latent Growth Curve Model
-This code performs multivariate latent growth curve to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits.
+### Compute Latent Transition Analysis
+This code performs latent transition analysis to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits.
 The model output is provided below, along with a summary of results.
@@
 ```{r}
-# Reshape data from long to wide format
-df_wide <- df_long %>%
-    pivot_wider(
-        id_cols = c(id),
-        names_from = event,
-        values_from = c(height, weight),
-        names_sep = "_"
-    )
+# Estimate the LTA model
+model <- tidyLPA::estimate_profiles(df_wide, 2, variances = "equal", covariances = "zero")
 
-df_wide <- na.omit(df_wide)
+summary(model, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)
 
-# Load necessary library
-library(lavaan)
-
-# Specify the multivariate latent growth curve model
-model <- '
-  # Intercept and slope factors for variable 1 (height)
-  i_var1 =~ 1*height_Baseline + 1*height_Year_1 + 1*height_Year_2
-  s_var1 =~ 0*height_Baseline + 1*height_Year_1 + 2*height_Year_2
-
-  # Intercept and slope factors for variable 2 (weight)
-  i_var2 =~ 1*weight_Baseline + 1*weight_Year_1 + 1*weight_Year_2
-  s_var2 =~ 0*weight_Baseline + 1*weight_Year_1 + 2*weight_Year_2
-'
-
-# Fit the model using FIML for handling missing data
-fit <- sem(model, data = df_wide, missing = "fiml")
-
-# Check the summary to identify potential issues
-summary(fit, fit.measures = TRUE, standardized = TRUE, rsquare = TRUE)
+plot(model)
 

### 2024-05-22 examples/_queue/_13_Examples_GrowthMixtureModels.qmd
update examples 10-14

@@
 
-# XXXXXXXXXXX {.unnumbered}
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+
+# This code configures knitr code chunk options
+
+knitr::opts_chunk$set(
+    echo = T, message = F, warning = F, error = F, output = T,
+    comment = NA, cache = T, code_folding = T,
+    R.options = list(width = 220, digits = 3),
+    fig.align = "center",
+    out.width = "75%", fig.asp = .75
+)
+```
+
+```{r}
+#| echo: FALSE # The content of this code block will not be shown when rendered
+#| output: FALSE # Output from this code block will not be shown when rendered
+#| cache: FALSE
+
+# This code loads the data to be used for the example
+
+library(tidyverse)
+
+# Set the data path
+data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
+
+# Read the data
+df <- readRDS(data_path)
+
+# Create and mutate the dataframe
+df_long <- df %>%
+    select(src_subject_id, eventname, anthroheightcalc, anthroweightcalc) %>%
+    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
+    drop_na(anthroheightcalc, anthroweightcalc) %>%
+    arrange(src_subject_id, eventname) %>%
+    mutate(
+        src_subject_id = factor(src_subject_id),
+        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
+        anthroheightcalc = round(as.numeric(anthroheightcalc), 2),
+        anthroweightcalc = round(as.numeric(anthroweightcalc), 2)
+    ) %>%
+    rename(
+        id = src_subject_id,
+        event = eventname,
+        height = anthroheightcalc,
+        weight = anthroweightcalc
+    )
+
+df_long <- df_long %>%
+    group_by(id) %>%
+    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
+    ungroup()
+
+```
 
 ## Overview
-
-<p>
 
@@
 
-</p>
-
 ## Preliminary Setup
-
-::: panel-tabset
-### Install Packages {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
-
-```{r pckg-install}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-# Create a list of required packages
-packages_required <- c("tidyverse","rstatix","DT")
-
-# Check which packages are not installed and install them
-packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
-if (length(packages_to_install) > 0) {
-    install.packages(packages_to_install)
-}
-
-# Load the required packages
-lapply(packages_required, library, character.only = TRUE)
-
-```
-:::
-
-### Load Packages
-
-::: blue
-**This code loads the r libraries necessary for this example**
-
-```{r lib-load}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
+### Load R libraries
+```{r}
 library(tidyverse)    # Collection of R packages for data science
+library(lavaan)       #
 library(rstatix)      # Pipe-friendly framework for basic statistical tests
@@
 ```
-:::
-
-### Config Options
-
-::: blue
-**This code configures knitr code chunk options**
-
-```{r config}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-
-knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
-                      comment=NA, cache=T, code_folding=T,
-                      R.options=list(width=220, digits = 3),
-                      fig.align='center', 
-                      out.width='75%', fig.asp=.75)
-```
-:::
-:::
-
-## Descriptives Overview
-
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-**This code reads in and shows the data to be used in the current
-example**
-
-```{r read-data}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| output: FALSE
-#| cache: FALSE
-
-# Set the data paths
-data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
-data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
-
-# Read the data
-data_demographics <- readRDS(data_path_1)
-data_nonimaging <- readRDS(data_path_2)
-
-# Subset the nonimaging data to include desired variables
-selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
-subset_data <- data_nonimaging[, selected_vars]
-
-library(dplyr)
-# # Merge the datasets on 'src_subject_id' and 'eventname'
-merged_data <- data_demographics %>%
-  full_join(subset_data, by = c("src_subject_id", "eventname"))
-
-# Inspect the merged data structure
-str(merged_data)
-
-# Define event names to be retained in the analysis and convert variables to appropriate data types
-eventnames_to_include <- c("baseline_year_1_arm_1",
-                           "1_year_follow_up_y_arm_1",
-                           "2_year_follow_up_y_arm_1",
-                           "3_year_follow_up_y_arm_1",
-                           "4_year_follow_up_y_arm_1")
-
-df <- merged_data %>%
-  filter(eventname %in% eventnames_to_include) %>%
-  mutate(
-    src_subject_id = as.factor(src_subject_id),
-    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
-    age = as.numeric(age),
-    sex = as.factor(sex),
-    race.4level = as.factor(race.4level),
-    hisp = as.factor(hisp),
-    high.educ.bl = as.factor(high.educ.bl),
-    household.income.bl = as.factor(household.income.bl),
-    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
-    rel_family_id.bl = as.factor(rel_family_id.bl),
-    site_id_l = as.factor(site_id_l),
-    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
-    anthroweightcalc = as.numeric(anthroweightcalc),
-    anthroheightcalc = as.numeric(anthroheightcalc)
-  ) %>%
-  # Exclude cases from unused assessment waves
-  filter(!is.na(eventname))
-
-```
-:::
 
 ### Descriptives
+```{r}
+descriptives_table <- df_long %>%
+    select(event, height, weight) %>%
+    mutate(event = factor(event)) %>%
+    tbl_summary(
+        by = event,
+        missing = "no",
+        label = list(height ~ "Height", weight ~ "Weight"),
+        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
+    ) %>%
+    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
+    bold_labels() %>%
+    italicize_levels() %>%
+    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
+theme_gtsummary_compact()
 
-::: blue
-**This code creates a descriptives table**
-
-```{r descriptives}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-# Define a function to compute descriptives
-compute_descriptives <- function(data, event_name) {
-  # For factor variables
-  sex_desc <- paste0(table(data$sex), " (", round(100 * prop.table(table(data$sex)), 1), "%)")
-  race_desc <- paste0(table(data$race.4level), " (", round(100 * prop.table(table(data$race.4level)), 1), "%)")
-
-  # For numeric variables
-  age_desc <- paste0(round(mean(data$age, na.rm = TRUE), 2), " (", round(sd(data$age, na.rm = TRUE), 2), ")")
-  weight_desc <- paste0(round(mean(data$anthroweightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroweightcalc, na.rm = TRUE), 2), ")")
-  height_desc <- paste0(round(mean(data$anthroheightcalc, na.rm = TRUE), 2), " (", round(sd(data$anthroheightcalc, na.rm = TRUE), 2), ")")
-
-  # Combine into a data frame
-  desc_df <- data.frame(
-    Variable = c("Sex - Female", "Sex - Male or other", "Race - Asian", "Race - Black", "Race - Other/Mixed", "Race - White", "Age", "Weight", "Height"),
-    Value = c(sex_desc, race_desc, age_desc, weight_desc, height_desc)
-  )
-
-  # Rename the Value column based on event name
-  colnames(desc_df)[2] <- event_name
-  return(desc_df)
-}
-
-# Compute descriptives for each event
-baseline_desc <- compute_descriptives(subset(df, eventname == "baseline_year_1_arm_1"), "baseline_year_1_arm_1")
-one_year_desc <- compute_descriptives(subset(df, eventname == "1_year_follow_up_y_arm_1"), "1_year_follow_up_y_arm_1")
-two_year_desc <- compute_descriptives(subset(df, eventname == "2_year_follow_up_y_arm_1"), "2_year_follow_up_y_arm_1")
-three_year_desc <- compute_descriptives(subset(df, eventname == "3_year_follow_up_y_arm_1"), "3_year_follow_up_y_arm_1")
-
-# Join all data frames side-by-side
-final_table <- baseline_desc %>%
-  left_join(one_year_desc, by = "Variable") %>%
-  left_join(two_year_desc, by = "Variable") %>%
-  left_join(three_year_desc, by = "Variable")
-
-# Adjust for the required format
-final_table[1:6, 3:5] <- NA
-
-# Round numeric values to two decimal places
-numeric_cols <- sapply(final_table, is.numeric)
-final_table[numeric_cols] <- lapply(final_table[numeric_cols], round, 2)
-
-# Create heading rows with the same columns as final_table
-heading_rows <- data.frame(
-  Variable = c("Sex", "Race"),
-  baseline_year_1_arm_1 = NA_real_,
-  `1_year_follow_up_y_arm_1` = NA_real_,
-  `2_year_follow_up_y_arm_1` = NA_real_,
-  `3_year_follow_up_y_arm_1` = NA_real_
-)
-
-# Set column names of heading_rows to match final_table
-colnames(heading_rows) <- colnames(final_table)
-
-# Introduce group labels and adjust the "Variable" column
-final_table <- rbind(
-                    heading_rows[1,], 
-                    final_table[1:2,],
-                    heading_rows[2,], 
-                    final_table[3:6,], 
-                    final_table[7:9,]
-                    )
-
-# Update the Variable column to remove redundant factor variable name
-final_table$Variable <- gsub("Sex - ", "", final_table$Variable)
-final_table$Variable <- gsub("Race - ", "", final_table$Variable)
-final_table$Variable[final_table$Variable == "Male or other"] <- "Male"
-
-# Add non-breaking spaces for increased indentation
-final_table$Variable[final_table$Variable %in% c("Female", "Male", "Asian", "Black", "Other/Mixed", "White")] <- 
-  paste0(rep("&nbsp;", 6), final_table$Variable[final_table$Variable %in% c("Female", "Male", "Asian", "Black", "Other/Mixed", "White")])
-
-# Update column headers
-colnames(final_table)[2:5] <- c("Baseline", "Year 1", "Year 2", "Year 3")
-
-# Display the table interactively without row names, with updated column headers, and with HTML entities rendered
-datatable(final_table, 
-          colnames = c("", "Baseline", "Year 1", "Year 2", "Year 3"),
-          options = list(pageLength = nrow(final_table), autoWidth = TRUE),
-          rownames = FALSE, escape = FALSE,
-          caption = "Descriptives Table")  # Add table title
-
+descriptives_table
 ```
 
-:::
-:::
-
 ## Results 
-<div class = "blue">
-
-::: panel-tabset
-### Build Model {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-**STEP 1: XXXXXXXXX **
-
-----
+### Compute Growth Mixture Model
+This code performs growth mixture modeling to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits.
+The model output is provided below, along with a summary of results.
+This approach xxxxx.
 
@@
 # Combine the data from all classes
-data <- rbind(class1, class2, class3)
-```
+#data <- rbind(class1, class2, class3)
 
-### Model Specification and Estimation
-Now that we have our simulated data, we can fit a Growth Mixture Model.
-In this example, we will assume that there are three latent classes.
-However, in practice, the number of classes is usually unknown and must be determined through model comparisons or other methods.
-```{r, eval=FALSE}
 # Fit the Growth Mixture Model
@@
             B = list(value = c(1, 0, 0, 0)
-```
 
-```{r}
 # Estimate the model
@@
 ```
+
+These results show xxxxx, indicating that xxxxx (e.g., xxxxx). The xxxx effects indicate xxxx.
+
+### Model Plots
+```{r}
+
+```
+
+This plot illustrates xxxx.
+
+## Wrapping Up
+The multivariate latent growth curve revealed xxxxxxxx.

### 2024-05-22 examples/_queue/_14_Examples_StateTraitModels.qmd
update examples 10-14

--- 2024-05-22 examples/_queue/_13_Examples_GrowthMixtureModels.qmd
@@
 ---
-title: "Growth Mixture Models"
+title: "State-Trait Models"
 author: "Biostatistics Working Group"
@@
 ## Results 
-### Compute Growth Mixture Model
-This code performs growth mixture modeling to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits.
-The model output is provided below, along with a summary of results.
-This approach xxxxx.
+### Compute Latent State-Trait Model
+This code performs latent state-trait analysis to examine xxxxx. The model output is provided below, along with a summary of results. This approach xxxxx.
 
 ```{r}
-# Combine the data from all classes
-#data <- rbind(class1, class2, class3)
 
-# Fit the Growth Mixture Model
-gmm <- hlme(value ~ time,
-            random = ~ time,
-            subject = "id",
-            mixture = ~ time,
-            data = data,
-            ng = 3,
-            B = list(value = c(1, 0, 0, 0)
+# Reshape data from long to wide format
+df_wide <- df_long %>%
+    pivot_wider(
+        id_cols = c(id),
+        names_from = event,
+        values_from = c(height, weight),
+        names_sep = "_"
+    )
 
-# Estimate the model
-fit <- lavaan::sem(model, data = df_wide, missing = "FIML")
+df_wide <- na.omit(df_wide)
+
+# Define the model
+model <- '
+  # Trait component
+  trait =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3
+
+  # State component
+  state =~ anthroheightcalc_0 + anthroheightcalc_1 + anthroheightcalc_2 + anthroheightcalc_3
+
+  # Residual variances
+  anthroheightcalc_0 ~~ anthroheightcalc_0
+  anthroheightcalc_1 ~~ anthroheightcalc_1
+  anthroheightcalc_2 ~~ anthroheightcalc_2
+  anthroheightcalc_3 ~~ anthroheightcalc_3
+'
+
+# Estimate the State-Trait Model
+fit <- sem(model, data = df_wide, missing = "FIML")
 
@@
 ## Wrapping Up
-The multivariate latent growth curve revealed xxxxxxxx.
+The latent state-trait analysis revealed xxxxxxxx.

### 2024-05-23 examples/6_Examples_GeneralizedEstimatingEquations.qmd
updates ex 5-10

@@
 library(geepack)      # Generalized Estimating Equations (GEE) analysis
+library(gtsummary) # Publication-ready tables
 
@@
 ggplot(predicted_values, aes(x = event, y = mean_height)) +
-  geom_line(color = "blue", size = 1) +
+  geom_line(color = "blue", linewidth = 1) +
   geom_point(color = "blue", size = 3) +

### 2024-05-23 examples/7_Examples_GeneralizedLinearMixedEffectsModels.qmd
updates ex 5-10

@@
 library(lme4)         # Linear mixed-effects models
+library(gtsummary) # Publication-ready tables
 
@@
 
-# Fit the LMM model
-model <- lmer(height ~ 1 + event + (1 | id), data = df_long, REML = TRUE)
-
-# Extract fixed effects
-fixed_effects <- fixef(model)
-
-summary(model)
-summary(fixed_effects)
+# # Fit the GLMM model
+# model <- lmer(handedness ~ 1 + event + (1 | id), data = df_long, REML = TRUE)
+# 
+# # Extract fixed effects
+# fixed_effects <- fixef(model)
+# 
+# summary(model)
+# summary(fixed_effects)
 
@@
 
-# Prepare data frame for plotting
-df_long$predicted_height <- predict(model, re.form = NA) # Overall fixed effect predictions
-
-# Calculate individual intercept adjustments
-df_long$individual_intercept <- predict(model, re.form = ~ (1 | id)) - fixed_effects[1]
-
-# Combine fixed effects and random effects predictions
-df_long$combined_predicted_height <- df_long$predicted_height + df_long$individual_intercept
-
-# Generate the plot
-ggplot(df_long, aes(x = event, y = height)) +
-  geom_point(alpha = 0.3) +
-  geom_line(aes(y = combined_predicted_height, group = id), color = "grey", alpha = 0.5) +
-  geom_line(aes(y = predicted_height, group = 1), color = "blue", size = 1.5) +
-  labs(
-    title = "Predicted vs. Observed Heights with GLMM",
-    x = "Event",
-    y = "Height (inches)"
-  ) +
-  theme_minimal()
+# # Prepare data frame for plotting
+# df_long$predicted_height <- predict(model, re.form = NA) # Overall fixed effect predictions
+# 
+# # Calculate individual intercept adjustments
+# df_long$individual_intercept <- predict(model, re.form = ~ (1 | id)) - fixed_effects[1]
+# 
+# # Combine fixed effects and random effects predictions
+# df_long$combined_predicted_height <- df_long$predicted_height + df_long$individual_intercept
+# 
+# # Generate the plot
+# ggplot(df_long, aes(x = event, y = height)) +
+#   geom_point(alpha = 0.3) +
+#   geom_line(aes(y = combined_predicted_height, group = id), color = "grey", alpha = 0.5) +
+#   geom_line(aes(y = predicted_height, group = 1), color = "blue", size = 1.5) +
+#   labs(
+#     title = "Predicted vs. Observed Heights with GLMM",
+#     x = "Event",
+#     y = "Height (inches)"
+#   ) +
+#   theme_minimal()
 

### 2024-06-06 examples/_3b_Examples_LinearMixedModels.qmd
updated output directory to the docs folder

--- 2024-05-01 examples/3b_minimalist_Examples_LinearMixedModels.qmd
@@
 
-```{r}
+```{r- BLOCK 1}
 #| echo: FALSE # The content of this code block will not be shown when rendered
@@
 df_long <- df %>%
-    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
+    select(src_subject_id, eventname, interview_age, anthroheightcalc) %>%
     filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
-    drop_na(cbcl_scr_syn_external_r) %>%
+    drop_na(anthroheightcalc) %>%
     arrange(src_subject_id, eventname) %>%
@@
         eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
-        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
+        interview_age = as.numeric(interview_age),
+        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
     ) %>%
@@
         event = eventname,
-        height = cbcl_scr_syn_external_r
+        age = interview_age,
+        height = anthroheightcalc
     )
-
-df_long <- df_long %>%
-  group_by(id) %>%
-  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
-  ungroup()
 
@@
 ```{r}
-#| echo: FALSE # The content of this code block will not be shown when rendered
-#| output: FALSE # Output from this code block will not be shown when rendered
-#| cache: FALSE
 
-# This code loads the data to be used for the example
+descriptives_table <- df_long %>%
+    select(event, age, height) %>%
+    mutate(event = factor(event)) %>%
+    tbl_summary(
+        by = event,
+        missing = "no",
+        label = list(height ~ "Height"),
+        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
+    ) %>%
+    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
+    bold_labels() %>%
+    italicize_levels() %>%
+    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
+theme_gtsummary_compact()
 
-library(tidyverse)
-
-# Set the data path
-data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"
-
-# Read the data
-df <- readRDS(data_path)
-
-# Create and mutate the dataframe
-df_long <- df %>%
-    select(src_subject_id, eventname, cbcl_scr_syn_external_r) %>%
-    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
-    drop_na(cbcl_scr_syn_external_r) %>%
-    arrange(src_subject_id, eventname) %>%
-    mutate(
-        src_subject_id = factor(src_subject_id),
-        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
-        cbcl_scr_syn_external_r = round(as.numeric(cbcl_scr_syn_external_r), 2)
-    ) %>%
-    rename(
-        id = src_subject_id,
-        event = eventname,
-        height = cbcl_scr_syn_external_r
-    )
-
-df_long <- df_long %>%
-  group_by(id) %>%
-  filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
-  ungroup()
+descriptives_table
 
@@
 
-model <- lmer(height ~ event + (1 + event | id), data = df_long, REML = T)
+df_long <- na.omit(df_long)
+
+model <- lmer(height ~ age + (1 + age | id), data = df_long, REML = T)
 
 # Model with random intercepts only
-model_ri_only <- lmer(height ~ event + (1 | id), data = df_long, REML = T)
-model_ri_only
+#model_ri_only <- lmer(height ~ age + (1 | id), data = df_long, REML = T)
+#model_ri_only
 
 # Model with random slopes only, without intercepts
-model_rs_only <- lmer(height ~ event + (0 + event | id), data = df_long, REML = T)
-model_rs_only
+#model_rs_only <- lmer(height ~ age + (0 + age | id), data = df_long, REML = T)
+#model_rs_only
 

### 2024-06-06 examples/_queue/_11_Examples_MultivariateLatentGrowthCurveModels.qmd
updated output directory to the docs folder

@@
 library(lavaan)       #
+library(patchwork)
 ```
@@
   # Intercept and slope factors for variable 1 (height)
-  i_var1 =~ 1*height_Baseline + 1*height_Year_1 + 1*height_Year_2
-  s_var1 =~ 0*height_Baseline + 1*height_Year_1 + 2*height_Year_2
+  i_var1 =~ 1*height_Baseline + 1*height_Year_1 + 1*height_Year_2 + 1*height_Year_3
+  s_var1 =~ 0*height_Baseline + 1*height_Year_1 + 2*height_Year_2 + 3*height_Year_3
 
   # Intercept and slope factors for variable 2 (weight)
-  i_var2 =~ 1*weight_Baseline + 1*weight_Year_1 + 1*weight_Year_2
-  s_var2 =~ 0*weight_Baseline + 1*weight_Year_1 + 2*weight_Year_2
+  i_var2 =~ 1*weight_Baseline + 1*weight_Year_1 + 1*weight_Year_2 + 1*weight_Year_3
+  s_var2 =~ 0*weight_Baseline + 1*weight_Year_1 + 2*weight_Year_2 + 3*weight_Year_3
 '
@@
 
+# Load necessary libraries
+library(tidyverse)
+library(patchwork)
+
+# Calculate means for each time point for height and weight
+mean_height <- df_wide %>%
+  summarise(across(starts_with("height"), mean, na.rm = TRUE))
+
+mean_weight <- df_wide %>%
+  summarise(across(starts_with("weight"), mean, na.rm = TRUE))
+
+# Reshape the data for plotting
+mean_height_long <- pivot_longer(mean_height, cols = everything(), names_to = "Time", values_to = "Mean_Height")
+mean_weight_long <- pivot_longer(mean_weight, cols = everything(), names_to = "Time", values_to = "Mean_Weight")
+
+# Plot the mean trajectories for height
+height_plot <- ggplot(mean_height_long, aes(x = Time, y = Mean_Height, group = 1)) +
+  geom_line(color = "blue", size = 1.2) +
+  geom_point(color = "blue") +
+  labs(title = "Mean Growth Trajectory for Height", y = "Mean Height", x = "Time Point") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# Plot the mean trajectories for weight
+weight_plot <- ggplot(mean_weight_long, aes(x = Time, y = Mean_Weight, group = 1)) +
+  geom_line(color = "red", size = 1.2) +
+  geom_point(color = "red") +
+  labs(title = "Mean Growth Trajectory for Weight", y = "Mean Weight", x = "Time Point") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# Combine the plots side by side using patchwork
+#combined_plot <- height_plot + weight_plot
+
+# Display plots
+print(height_plot)
+print(weight_plot)
+
+####
+
+# Extract factor scores
+factor_scores <- lavPredict(fit, type = "lv")
+
+# Convert to a data frame
+factor_scores_df <- as.data.frame(factor_scores)
+
+# Rename the columns for clarity
+colnames(factor_scores_df) <- c("Intercept_Height", "Slope_Height", "Intercept_Weight", "Slope_Weight")
+
+# Summary of the model
+summary(fit, fit.measures = TRUE, standardized = TRUE, rsquare = TRUE)
+
+# Plot histograms of intercepts and slopes
+ggplot(factor_scores_df, aes(x = Intercept_Height)) +
+  geom_histogram(binwidth = 0.5, fill = "blue", alpha = 0.7) +
+  labs(title = "Distribution of Intercepts for Height", x = "Intercept", y = "Frequency") +
+  theme_minimal()
+
+ggplot(factor_scores_df, aes(x = Slope_Height)) +
+  geom_histogram(binwidth = 0.05, fill = "blue", alpha = 0.7) +
+  labs(title = "Distribution of Slopes for Height", x = "Slope", y = "Frequency") +
+  theme_minimal()
+
+ggplot(factor_scores_df, aes(x = Intercept_Weight)) +
+  geom_histogram(binwidth = 0.5, fill = "red", alpha = 0.7) +
+  labs(title = "Distribution of Intercepts for Weight", x = "Intercept", y = "Frequency") +
+  theme_minimal()
+
+ggplot(factor_scores_df, aes(x = Slope_Weight)) +
+  geom_histogram(binwidth = 0.05, fill = "red", alpha = 0.7) +
+  labs(title = "Distribution of Slopes for Weight", x = "Slope", y = "Frequency") +
+  theme_minimal()
+
+#####
+
 ```
 
-This plot illustrates xxxx.
+These plots illustrate xxxx.
 

### 2024-06-06 examples/_queue/_13_Examples_GrowthMixtureModels.qmd
updated output directory to the docs folder

@@
     mutate(
-        src_subject_id = factor(src_subject_id),
+        src_subject_id = as.numeric(src_subject_id),
         eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
@@
 library(tidyverse)    # Collection of R packages for data science
+library(lcmm)
 library(lavaan)       #
@@
 ```{r}
+set.seed(2002)
+gmm1 <- hlme(height ~ event, subject = "id", random=~1, ng = 1, data =
+df_long)
+
+gmm2 <- gridsearch(rep = 100, maxiter = 10, minit = gmm1,
+hlme(height ~ event, subject = "id_numeric", random=~1,
+ng = 2, data = df_long, mixture = ~ event,
+nwg=T))
+
+gmm3 <- gridsearch(rep = 100, maxiter = 10, minit = gmm1,
+hlme(height ~ event, subject = "id_numeric", random=~1,
+ng = 3, data = df_long, mixture = ~ event,
+nwg=T))
+
+# make table with results for the 4 models:
+summarytable(gmm1, gmm2, gmm3)
+```
+
+```{r}
 # Combine the data from all classes

### 2024-06-06 examples/_queue/_14_Examples_StateTraitModels.qmd
updated output directory to the docs folder

@@
   # Trait component
-  trait =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3
+  trait =~ 1 * height_Baseline + 1 * height_Year_1 + 1 * height_Year_2 + 1 * height_Year_3
 
   # State component
-  state =~ anthroheightcalc_0 + anthroheightcalc_1 + anthroheightcalc_2 + anthroheightcalc_3
+  state =~ height_Baseline + height_Year_1 + height_Year_2 + height_Year_3
 
   # Residual variances
-  anthroheightcalc_0 ~~ anthroheightcalc_0
-  anthroheightcalc_1 ~~ anthroheightcalc_1
-  anthroheightcalc_2 ~~ anthroheightcalc_2
-  anthroheightcalc_3 ~~ anthroheightcalc_3
+  height_Baseline ~~ height_Baseline
+  height_Year_1 ~~ height_Year_1
+  height_Year_2 ~~ height_Year_2
+  height_Year_3 ~~ height_Year_3
 '
